### 2021-04-28 article/article.tex
article init

\documentclass{llncs}

\title{Synthesis of Linear Functional Programs}
\author{Rodrigo Mesquita}
\date{April 2021}
\institute{NOVA School of Science and Technology}

\begin{document}

\maketitle

\section{Introduction}

This work explores the process of synthetising linear functional programs from types based in linear logic, leveraging the Curry-Howard correspondence.
- forma resumida a correspondência: "obs que proposições lógicas correspondem a tipos e programas bem tipificados correspondem a provas das proposições"
Having a type be a proposition in linear logic, and from the Curry-Howard correspondence, we can relate a proof of that proposition directly to a linear functional program - finding a proof is finding a program with that type, so we formulate the synthesis as proof search in this logic.

\section{Background}

\section{Related Work}

\bibliographystyle{splncs04}
\bibliography{references}
\end{document}

### 2021-04-29 article/article.tex
some notes

@@
 \documentclass{llncs}
+
+\usepackage[inline]{todo}
+\usepackage{cmll}
+\usepackage{mathpartir}
+
+\newcommand{\lolli}{\multimap}
+\newcommand{\tensor}{\otimes}
+\newcommand{\one}{\mathbf{1}}
+\newcommand{\bang}{{!}}
+
+\newcommand{\llet}[2]{\mathsf{let}\,#1\,\mathsf{in}\,#2}
 
@@
 
-This work explores the process of synthetising linear functional programs from types based in linear logic, leveraging the Curry-Howard correspondence.
-- forma resumida a correspondência: "obs que proposições lógicas correspondem a tipos e programas bem tipificados correspondem a provas das proposições"
-Having a type be a proposition in linear logic, and from the Curry-Howard correspondence, we can relate a proof of that proposition directly to a linear functional program - finding a proof is finding a program with that type, so we formulate the synthesis as proof search in this logic.
+This work explores the process of synthetising linear functional
+programs from types based in linear logic, leveraging the Curry-Howard
+correspondence.  - forma resumida a correspondência: "obs que
+proposições lógicas correspondem a tipos e programas bem tipificados
+correspondem a provas das proposições" Having a type be a proposition
+in linear logic, and from the Curry-Howard correspondence, we can
+relate a proof of that proposition directly to a linear functional
+program - finding a proof is finding a program with that type, so we
+formulate the synthesis as proof search in this logic.
+
+\begin{enumerate}
+\item General motivation
+  \begin{itemize}
+\item Program synthesis is an automated or semi-automated process of
+  deriving a program from a specification of its behavior.
+\item Specifications appear in many forms\dots
+\item Program synthesis as a means of improving program
+  correctness/programmer productivity.
+\item Challenges: space of valid programs, space of valid
+  specifications,  etc.
+\end{itemize}
+\item \emph{Type-based} program synthesis
+  \begin{itemize}
+ \item Rich types as specifications.
+\item Languages with richer type systems allow for more precise
+  types. This precision can statically eliminate various kinds of
+  logical errors by making certain invalid program states ill-typed
+  (e.g. a ``null aware'' type system, will ensure you cannot
+  dereference a null-pointer).
+\item But can also be a burden -- the whole point of these type
+  systems is to ensure that ``less'' things are well-typed, and
+  sometimes its hard to convince the checker\dots
+\item Type-based synthesis leverages rich types as a way of pruning
+  the search space and by using types as a more ``familiar''
+  specification.
+\end{itemize}
+\item ``My'' problem: Synthesis based on linear types
+  \begin{itemize}
+  \item Linear types constrain resource usage in programs by \emph{statically} limiting
+    the number of times certain resources can be used during its
+    lifetime.
+  \item Applications in resource-aware programming such as concurrent
+    programming (e.g. session types) and memory-management
+    (e.g. Rust).
+  \item This work will study program synthesis in a functional,
+    linearly-typed setting, by leveraging the propositions-as-types
+    correspondence between linear logic and the (linear)
+    $\lambda$-calculus.
+  \item What is props-as-types? What is linear logic? Linear $\lambda$
+    is ``just'' a small functional core language.
+  \item Through props-as-types, synthesis can be formulated as proof
+    search and so we can leverage a lot of related technology from the
+    literature.
+  \end{itemize}
+\item Goals:
+  \begin{itemize}
+\item Full and partial synthesis of well-typed (in the above sense)
+  programs. Full synthesis means producing a function (or set of)
+  satisfying the spec. Partial synthesis means being given a partial
+  program (i.e. a function with a hole in it) and ``completing it''.
+\item Start from a small system, build up to recursive
+  types/functions.
+\item Potential to go further: type refinements; interactive
+  synthesis?
+ \item Evaluation through expressiveness benchmarks (can we synthesize
+   ``X''?); time measurements (how fast can we synthesize ``X''?');
+   etc.
+  \end{itemize}
+
+\end{enumerate}
 
@@
 
+More about the concepts than citing concrete works (but citations are
+needed!):
+\begin{enumerate}
+\item PL/Type systems in a ``formal'' sense (inference rules, etc)
+\item Propositions as types: Proofs as programs, proof search as
+  synthesis, etc.
+\item Linear logic: a ``resource aware logic''
+\item Proof search in LL: challenges, focusing as a ``solution''
+\item Linear $\lambda$-calculus: what does it look like.
+\item Type-based synthesis: ``inverting'' a type system.
+\end{enumerate}
+
 \section{Related Work}
+
+Concrete works:
+\begin{itemize}
+\item Synthesis for graded types (no recursion   -- in that work, no
+  obvious connection of grading with
+  concurrency)~\cite{DBLP:conf/lopstr/HughesO20}
+
+\item Type-and-example-directed
+  synthesis~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
+  or how to turn your type system upside down.
+
+\item Synthesis from (polymorphic) refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16}
+
+Refinement types provide very precise specifications to assist the
+synthesis process.
+
+\item Synthesis of heap-manipulating
+  programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}
+
+Less related, but more evidence of this idea of turning ``checking''
+systems into synthesis frameworks.
+
+\item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}
+
+This resource-guided means something a bit different. Programs satisfy
+a functional specification and a symbolic resource bound in the sense
+of amortized analysis, but can provide technical insights.
+
+\end{itemize}
+
+\section{Goals and Work Plan}
+
+Expand on earlier points. Start from a functional language with linear
+types ($\tensor$, $\lolli$, $\oplus$, $\bang$); build on it with more
+``stuff''. General techniques, drawn from proof theory via props as
+types: explore focusing to tame non-determinism/search space. Partial
+synthesis is still proof search! Add other techniques as we go along
+(e.g. for recursion we need to constrain recursive calls).
+
+Mention that you've already implemented a type-checker for this
+(useful as a prelim. exercise but also later, for \emph{validation}).
+
+Validation and evaluation: validation is as simple as ``does it typecheck''? can you synth? how fast?
+
+\section{Bonus}
+
+\[
+  \begin{array}{lcl}
+    A, B & ::= & A \tensor B \mid A \lolli B \mid A \with B \mid A
+                 \oplus B \mid \bang A \mid \one \mid \dots\\[1ex]
+    M,N & ::= & \lambda x{:}A.M\\
+         & \mid & M\,N\\
+         & \mid & (M \tensor N)\\
+         & \mid & \llet{x\tensor y = M}{N} \\
+         & \mid & x \\
+         & \mid & \dots\\
+    \end{array}
+\]
+
+\[
+  \infer[$(\lolli\! I)$]
+  {\Delta , x{:}A \vdash M : B }
+  {\Delta \vdash \lambda x {:} A . M : A \lolli B}
+  \quad
+  \infer[$(\lolli\! E)$]
+  {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
+  {\Delta_1, \Delta_2 \vdash M\,N : B}
+\]
+\[
+  \infer*[left=($\tensor I$)]
+  {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
+  {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
+  \quad
+  \infer*[right=($\tensor E$)]
+  {\Delta_1 \vdash M : A \tensor B \and \Delta_2 , x{:}A, y{:}B\vdash
+    N : C }
+  {\Delta_1 , \Delta_2\vdash \llet{x\tensor y = M}{N} : C }
+\]
 

### 2021-04-30 article/article.tex
more notes
write one more line :)

--- 2021-04-28 article/article.tex
@@
 This work explores the process of synthetising linear functional programs from types based in linear logic, leveraging the Curry-Howard correspondence.
-- forma resumida a correspondência: "obs que proposições lógicas correspondem a tipos e programas bem tipificados correspondem a provas das proposições"
-Having a type be a proposition in linear logic, and from the Curry-Howard correspondence, we can relate a proof of that proposition directly to a linear functional program - finding a proof is finding a program with that type, so we formulate the synthesis as proof search in this logic.
+Said correspondence states that propositions in a logic have a direct mapping to types, and well typed programs also have a one to one correspendence to proofs of those propositions.
+As such, having a type be a proposition in linear logic, we can relate a proof of that proposition directly to a linear functional program - finding a proof is finding a program with that type, and so we formulate the synthesis as proof search in linear logic.
 

### 2021-04-30 article/article.tex

--- 2021-04-29 article/article.tex
@@
 
-This work explores the process of synthetising linear functional
-programs from types based in linear logic, leveraging the Curry-Howard
-correspondence.  - forma resumida a correspondência: "obs que
-proposições lógicas correspondem a tipos e programas bem tipificados
-correspondem a provas das proposições" Having a type be a proposition
-in linear logic, and from the Curry-Howard correspondence, we can
-relate a proof of that proposition directly to a linear functional
-program - finding a proof is finding a program with that type, so we
-formulate the synthesis as proof search in this logic.
+This work explores the process of synthetising linear functional programs from types based in linear logic, leveraging the Curry-Howard correspondence.
+Said correspondence states that propositions in a logic have a direct mapping to types, and well typed programs also have a one to one correspendence to proofs of those propositions.
+As such, having a type be a proposition in linear logic, we can relate a proof of that proposition directly to a linear functional program - finding a proof is finding a program with that type, and so we formulate the synthesis as proof search in linear logic.
 

### 2021-05-05 article/article.tex
minor change
more introduction

@@
 
-\usepackage[inline]{todo}
 \usepackage{cmll}
@@
 \title{Synthesis of Linear Functional Programs}
-\author{Rodrigo Mesquita}
+\author{Rodrigo Mesquita \and Bernardo Toninho}
 \date{April 2021}
@@
 
-This work explores the process of synthetising linear functional programs from types based in linear logic, leveraging the Curry-Howard correspondence.
-Said correspondence states that propositions in a logic have a direct mapping to types, and well typed programs also have a one to one correspendence to proofs of those propositions.
-As such, having a type be a proposition in linear logic, we can relate a proof of that proposition directly to a linear functional program - finding a proof is finding a program with that type, and so we formulate the synthesis as proof search in linear logic.
+% General motivation?
+Program synthesis is an automated or semi-automated process of
+ deriving a program (generating code) from a high-level specification.
+ Specifications appear in many forms (i.e.
+ polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
+ examples~\cite{DBLP:conf/popl/FrankleOWZ16},
+ graded linear types~\cite{DBLP:conf/lopstr/HughesO20}, ...),
+ and in this work the synthesis specification will be a linear type, ?possibly with refinements.?
+ The chosen specification form dictates the synthesis process, and different ones might
+ be approached in completely different ways.
+ In spite of the specification, most program synthetizer's have to deal with two main
+ inherent complexities: searching over a vast space of valid programs, and interpreting user intent.
+ Synthesis has many motivations, it can be seen as a means to improve a programmer's productivity and/or
+ program correctness (i.e. through suggestion and/or autocompletion), or as a tool to automate certain parts
+ of the programming process (i.e. in the same way AI might generate some boilerplate text to auxiliate a journalist, rather
+ than to replace), just to name a few examples.
+
+% Type based synthesis
+Type-based synthesis takes a given type and produces a program with that same type.
+
+This work explores the process of synthetising linear functional
+programs from types based in linear logic, leveraging the Curry-Howard correspondence.
+Said correspondence states that propositions in a logic have a direct mapping to types,
+ and well typed programs also have a one to one correspendence to proofs of those propositions.
+As such, having a type be a proposition in linear logic, we can relate a proof of that
+proposition directly to a linear functional program
+ — finding a proof is finding a program with that type, and so we
+ formulate the synthesis as proof search in linear logic.
 
@@
   \begin{itemize}
-\item Program synthesis is an automated or semi-automated process of
-  deriving a program from a specification of its behavior.
+\item 
 \item Specifications appear in many forms\dots
@@
   systems is to ensure that ``less'' things are well-typed, and
-  sometimes its hard to convince the checker\dots
+  sometimes it's hard to convince the checker\dots
 \item Type-based synthesis leverages rich types as a way of pruning
@@
 \]
+
 \[

### 2021-05-06 article/article.tex
First introduction iteration

@@
  deriving a program (generating code) from a high-level specification.
- Specifications appear in many forms (i.e.
+ Specifications appear in many forms (e.g.
  polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
  examples~\cite{DBLP:conf/popl/FrankleOWZ16},
- graded linear types~\cite{DBLP:conf/lopstr/HughesO20}, ...),
- and in this work the synthesis specification will be a linear type, ?possibly with refinements.?
+ graded linear types~\cite{DBLP:conf/lopstr/HughesO20}, ...).
+ In this work the synthesis specification will be a linear type, ?possibly with refinements.?
  The chosen specification form dictates the synthesis process, and different ones might
@@
  program correctness (i.e. through suggestion and/or autocompletion), or as a tool to automate certain parts
- of the programming process (i.e. in the same way AI might generate some boilerplate text to auxiliate a journalist, rather
+ of the programming process (e.g. in the same way AI might generate some boilerplate text to auxiliate a journalist, rather
  than to replace), just to name a few examples.
@@
 % Type based synthesis
-Type-based synthesis takes a given type and produces a program with that same type.
+Type-based synthesis takes a type and produces a program with that given type. 
+Types can be further explored to give the programmer more expressiveness (e.g. by adding constraints??), and are called rich types.
+-- Gostava de ver com o prof aqui. [Estive a procurar e fiquei um pouco confuso. polimorfismo é rich types?
+dependent types are rich types? aqueles constraints são os dependent types pelo que percebi... mas n me
+parece correto. arithmetic types? temos de rever. Por agora vou por apenas os tópicos que o professor escreveu]
+Languages with richer type systems allow for more precise
+  types. This precision can statically eliminate various kinds of
+  logical errors by making certain invalid program states ill-typed
+  (e.g. a ``null aware'' type system, will ensure you cannot
+  dereference a null-pointer).
+However, they can also be a burden -- the whole point of these type
+  systems is to ensure that ``less'' things are well-typed, and
+  sometimes it's hard to convince the checker.
+Regarding the challenges, type-based synthesis leverages rich types as a way of pruning
+  the search space, and by using types gives the user as a more ``familiar'' specification.
 
 This work explores the process of synthetising linear functional
-programs from types based in linear logic, leveraging the Curry-Howard correspondence.
+ programs from types based in linear logic (linear types), leveraging the Curry-Howard correspondence.
 Said correspondence states that propositions in a logic have a direct mapping to types,
@@
  proposition directly to a linear functional program
- — finding a proof is finding a program with that type, and so we
- formulate the synthesis as proof search in linear logic.
+ — finding a proof is finding a program with that type, so we
+ formulate the synthesis as proof search in linear logic, which makes to us available
+ a lot of related technology from the proof search literature.
+Linear types differ from usual types for they constrain resource usage in programs by
+ \emph{statically?} limiting the number of times certain resources can be used during their lifetime.
+ They can be applied to resource-aware programming such as concurrent programming (e.g. session types??), and to memory-management (e.g. Rust)
+ [Quero reescrever a frase acima, não gosto de como falamos de session types e rust em dois exemplos paralelos visto que não são coisas bem comparáveis. Revemos na reunião.]
+
+ % Goals
+In the end, we intend to be able to do full and partial synthesis of well-typed programs --
+ Full synthesis being the production of a function (or set of) satisfying the specification;
+ Partial synthesis being the ``completion'' of a partial program (i.e. a function with a \emph{hole} in it) --
+ Starting from a small core linear -- i.e. \emph{resource-aware} functional language, and building up to recursive types/functions.
+ Eventually, possibly, enhancing the synthesis process with type refinements, and making it interactive -- i.e. the synthetiser offers the user
+ ``choices'' mid-process.
+[Devia estar a escrever isto eu ahah?]The evaluation will be done through expressiveness
+benchmarks (i.e. can we synthesize ``X''?); time measurements (i.e. how fast can we synthesize ``X''?'); and so on.
 

### 2021-05-06 article/article.tex
Add background first lines

@@
 
+A type system can be formally described through natural deduction -- as a set of rules that determines the concept of deduction for some language~\cite{prawitznd65}.
+The language -- the propositions of a type system are the types themselves, the rules define how to construct types, and being able to deduct a type (i.e.
+constructing a derivation) proves the type is valid.
+[Preciso de ajuda nesta última frase, não é isto que quero dizer mas não sei descrever de forma correta e satisfatória :)]
+The natural deduction judgment is extended to include a \emph{term} (i.e. to make proofs explicit with a formal definition), and modified to include a list of hypothesis.
+[Também preciso de ajuda a reescrever a última, eu bem que procuro muito mas precisava de alcançar uma mestria de natural deduction - será mais rápido o professor ajudar-me a dar nomes às coisas :)]
+This extended judgment, called a typing judgment has the form:
+\[
+    \Gamma \vdash M : A
+\]
+Where the natural deduction list of hypothesis, $\Gamma$, is now called \emph{typing environment} (i.e.
+set of variables and their types), $M$ is still a \emph{term} but also known as ``program'', and $A$, the natural deduction proposition, is now called \emph{type}.
+This judgment can be read -- $M$ has type $A$ in $\Gamma$.
+% As a concrete example, let us describe the typing rule for the tensor type $A \tensor B$
+
+\section{Separate}
+
 More about the concepts than citing concrete works (but citations are

### 2021-05-06 article/article.tex
tweaks

@@
 \usepackage{mathpartir}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{hyperref}
+\usepackage{todonotes}
 
@@
 Program synthesis is an automated or semi-automated process of
- deriving a program (generating code) from a high-level specification.
- Specifications appear in many forms (e.g.
- polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
- examples~\cite{DBLP:conf/popl/FrankleOWZ16},
- graded linear types~\cite{DBLP:conf/lopstr/HughesO20}, ...).
- In this work the synthesis specification will be a linear type, ?possibly with refinements.?
- The chosen specification form dictates the synthesis process, and different ones might
- be approached in completely different ways.
- In spite of the specification, most program synthetizer's have to deal with two main
- inherent complexities: searching over a vast space of valid programs, and interpreting user intent.
- Synthesis has many motivations, it can be seen as a means to improve a programmer's productivity and/or
- program correctness (i.e. through suggestion and/or autocompletion), or as a tool to automate certain parts
- of the programming process (e.g. in the same way AI might generate some boilerplate text to auxiliate a journalist, rather
- than to replace), just to name a few examples.
+deriving a program (i.e.~generating code) from a high-level specification.
+Specifications can take many forms (e.g.  polymorphic refinement
+types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
+examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
+types~\cite{DBLP:conf/lopstr/HughesO20}, ...).  In this work we
+explore type-based synthesis where specifications take the form of
+linear types.\todo{O que é um linear type?}
+%
+The chosen specification form dictates the synthesis process, and
+different ones might be approached in completely different
+ways.\todo{Esta frase não parece ter grande utilidade. O que queres
+  dizer aqui?}
+%
+Regardless of the kind of specification, program synthesizers have to
+deal with two main inherent sources of complexity: searching over a
+vast space of (potentially) valid programs, and interpreting user
+intent.
+%
+Synthesis has many motivations, it can be seen as a means to
+improve a programmer's productivity and/or program correctness
+(i.e. through suggestion and/or autocompletion), or as a tool to
+automate certain parts of the programming process (e.g. in the same
+way AI might generate some boilerplate text to auxiliate a journalist,
+rather than to replace), just to name a few examples.
+\todo[inline]{Acho que estes 3 ``paragrafos'' deviam estar por ordens diferentes.}
 
 % Type based synthesis
-Type-based synthesis takes a type and produces a program with that given type. 
-Types can be further explored to give the programmer more expressiveness (e.g. by adding constraints??), and are called rich types.
--- Gostava de ver com o prof aqui. [Estive a procurar e fiquei um pouco confuso. polimorfismo é rich types?
-dependent types are rich types? aqueles constraints são os dependent types pelo que percebi... mas n me
-parece correto. arithmetic types? temos de rever. Por agora vou por apenas os tópicos que o professor escreveu]
-Languages with richer type systems allow for more precise
-  types. This precision can statically eliminate various kinds of
-  logical errors by making certain invalid program states ill-typed
-  (e.g. a ``null aware'' type system, will ensure you cannot
-  dereference a null-pointer).
+Type-based synthesis takes a type and produces a program with that
+given type.  Types can be further explored to give the programmer more
+expressiveness (e.g. by adding constraints??), and are called rich
+types.  -- Gostava de ver com o prof aqui. [Estive a procurar e fiquei
+um pouco confuso. polimorfismo é rich types?  dependent types are rich
+types? aqueles constraints são os dependent types pelo que
+percebi... mas n me parece correto. arithmetic types? temos de
+rever. Por agora vou por apenas os tópicos que o professor escreveu]
+Languages with richer type systems allow for more precise types. This
+precision can statically eliminate various kinds of logical errors by
+making certain invalid program states ill-typed (e.g. a ``null aware''
+type system, will ensure you cannot dereference a null-pointer).
 However, they can also be a burden -- the whole point of these type
 systems is to ensure that ``less'' things are well-typed, and
-  sometimes it's hard to convince the checker.
-Regarding the challenges, type-based synthesis leverages rich types as a way of pruning
-  the search space, and by using types gives the user as a more ``familiar'' specification.
+sometimes it's hard to convince the checker.  Regarding the
+challenges, type-based synthesis leverages rich types as a way of
+pruning the search space, and by using types gives the user as a more
+``familiar'' specification.
 
 This work explores the process of synthetising linear functional
- programs from types based in linear logic (linear types), leveraging the Curry-Howard correspondence.
-Said correspondence states that propositions in a logic have a direct mapping to types,
- and well typed programs also have a one to one correspendence to proofs of those propositions.
-As such, having a type be a proposition in linear logic, we can relate a proof of that
- proposition directly to a linear functional program
- — finding a proof is finding a program with that type, so we
- formulate the synthesis as proof search in linear logic, which makes to us available
- a lot of related technology from the proof search literature.
-Linear types differ from usual types for they constrain resource usage in programs by
- \emph{statically?} limiting the number of times certain resources can be used during their lifetime.
- They can be applied to resource-aware programming such as concurrent programming (e.g. session types??), and to memory-management (e.g. Rust)
- [Quero reescrever a frase acima, não gosto de como falamos de session types e rust em dois exemplos paralelos visto que não são coisas bem comparáveis. Revemos na reunião.]
+programs from types based in linear logic (linear types), leveraging
+the Curry-Howard correspondence.  Said correspondence states that
+propositions in a logic have a direct mapping to types, and well typed
+programs also have a one to one correspendence to proofs of those
+propositions.  As such, having a type be a proposition in linear
+logic, we can relate a proof of that proposition directly to a linear
+functional program — finding a proof is finding a program with that
+type, so we formulate the synthesis as proof search in linear logic,
+which makes to us available a lot of related technology from the proof
+search literature.  Linear types differ from usual types for they
+constrain resource usage in programs by \emph{statically?} limiting
+the number of times certain resources can be used during their
+lifetime.  They can be applied to resource-aware programming such as
+concurrent programming (e.g. session types??), and to
+memory-management (e.g. Rust) [Quero reescrever a frase acima, não
+gosto de como falamos de session types e rust em dois exemplos
+paralelos visto que não são coisas bem comparáveis. Revemos na
+reunião.]
 
  % Goals
-In the end, we intend to be able to do full and partial synthesis of well-typed programs --
- Full synthesis being the production of a function (or set of) satisfying the specification;
- Partial synthesis being the ``completion'' of a partial program (i.e. a function with a \emph{hole} in it) --
- Starting from a small core linear -- i.e. \emph{resource-aware} functional language, and building up to recursive types/functions.
- Eventually, possibly, enhancing the synthesis process with type refinements, and making it interactive -- i.e. the synthetiser offers the user
- ``choices'' mid-process.
-[Devia estar a escrever isto eu ahah?]The evaluation will be done through expressiveness
-benchmarks (i.e. can we synthesize ``X''?); time measurements (i.e. how fast can we synthesize ``X''?'); and so on.
+In the end, we intend to be able to do full and partial synthesis of
+well-typed programs -- Full synthesis being the production of a
+function (or set of) satisfying the specification; Partial synthesis
+being the ``completion'' of a partial program (i.e. a function with a
+\emph{hole} in it) -- Starting from a small core linear --
+i.e. \emph{resource-aware} functional language, and building up to
+recursive types/functions.  Eventually, possibly, enhancing the
+synthesis process with type refinements, and making it interactive --
+i.e. the synthetiser offers the user ``choices'' mid-process.  [Devia
+estar a escrever isto eu ahah?]The evaluation will be done through
+expressiveness benchmarks (i.e. can we synthesize ``X''?); time
+measurements (i.e. how fast can we synthesize ``X''?'); and so on.
 
@@
 
-A type system can be formally described through natural deduction -- as a set of rules that determines the concept of deduction for some language~\cite{prawitznd65}.
+A type system can be formally described through a set of inference
+rules that inductively define a judgment of the form $\Gamma \vdash M
+: A$, stating that program expression $M$ has type $A$ according to
+the \emph{typing assumptions} for variables tracked in $\Gamma$.
+\dots
+
+\todo[inline]{Daqui, para tipificar uma função, para Curry-Howard,
+  para lógica linear, para proof search, etc\dots Só faz sentido falar de dedução natural no
+  contexto de ``lógica'' / Curry-Howard.}
+
+natural deduction -- as a set of rules that determines the concept of deduction for some language~\cite{prawitznd65}.
 The language -- the propositions of a type system are the types themselves, the rules define how to construct types, and being able to deduct a type (i.e.

### 2021-05-07 article/article.tex
more on background

--- 2021-05-06 article/article.tex
@@
 
-\begin{enumerate}
-\item General motivation
-  \begin{itemize}
-\item 
-\item Specifications appear in many forms\dots
-\item Program synthesis as a means of improving program
-  correctness/programmer productivity.
-\item Challenges: space of valid programs, space of valid
-  specifications,  etc.
-\end{itemize}
-\item \emph{Type-based} program synthesis
-  \begin{itemize}
- \item Rich types as specifications.
-\item Languages with richer type systems allow for more precise
-  types. This precision can statically eliminate various kinds of
-  logical errors by making certain invalid program states ill-typed
-  (e.g. a ``null aware'' type system, will ensure you cannot
-  dereference a null-pointer).
-\item But can also be a burden -- the whole point of these type
-  systems is to ensure that ``less'' things are well-typed, and
-  sometimes it's hard to convince the checker\dots
-\item Type-based synthesis leverages rich types as a way of pruning
-  the search space and by using types as a more ``familiar''
-  specification.
-\end{itemize}
-\item ``My'' problem: Synthesis based on linear types
-  \begin{itemize}
-  \item Linear types constrain resource usage in programs by \emph{statically} limiting
-    the number of times certain resources can be used during its
-    lifetime.
-  \item Applications in resource-aware programming such as concurrent
-    programming (e.g. session types) and memory-management
-    (e.g. Rust).
-  \item This work will study program synthesis in a functional,
-    linearly-typed setting, by leveraging the propositions-as-types
-    correspondence between linear logic and the (linear)
-    $\lambda$-calculus.
-  \item What is props-as-types? What is linear logic? Linear $\lambda$
-    is ``just'' a small functional core language.
-  \item Through props-as-types, synthesis can be formulated as proof
-    search and so we can leverage a lot of related technology from the
-    literature.
-  \end{itemize}
-\item Goals:
-  \begin{itemize}
-\item Full and partial synthesis of well-typed (in the above sense)
-  programs. Full synthesis means producing a function (or set of)
-  satisfying the spec. Partial synthesis means being given a partial
-  program (i.e. a function with a hole in it) and ``completing it''.
-\item Start from a small system, build up to recursive
-  types/functions.
-\item Potential to go further: type refinements; interactive
-  synthesis?
- \item Evaluation through expressiveness benchmarks (can we synthesize
-   ``X''?); time measurements (how fast can we synthesize ``X''?');
-   etc.
-  \end{itemize}
-
-\end{enumerate}
+% \begin{enumerate}
+% \item General motivation
+%   \begin{itemize}
+% \item 
+% \item Specifications appear in many forms\dots
+% \item Program synthesis as a means of improving program
+%   correctness/programmer productivity.
+% \item Challenges: space of valid programs, space of valid
+%   specifications,  etc.
+% \end{itemize}
+% \item \emph{Type-based} program synthesis
+%   \begin{itemize}
+%  \item Rich types as specifications.
+% \item Languages with richer type systems allow for more precise
+%   types. This precision can statically eliminate various kinds of
+%   logical errors by making certain invalid program states ill-typed
+%   (e.g. a ``null aware'' type system, will ensure you cannot
+%   dereference a null-pointer).
+% \item But can also be a burden -- the whole point of these type
+%   systems is to ensure that ``less'' things are well-typed, and
+%   sometimes it's hard to convince the checker\dots
+% \item Type-based synthesis leverages rich types as a way of pruning
+%   the search space and by using types as a more ``familiar''
+%   specification.
+% \end{itemize}
+% \item ``My'' problem: Synthesis based on linear types
+%   \begin{itemize}
+%   \item Linear types constrain resource usage in programs by \emph{statically} limiting
+%     the number of times certain resources can be used during its
+%     lifetime.
+%   \item Applications in resource-aware programming such as concurrent
+%     programming (e.g. session types) and memory-management
+%     (e.g. Rust).
+%   \item This work will study program synthesis in a functional,
+%     linearly-typed setting, by leveraging the propositions-as-types
+%     correspondence between linear logic and the (linear)
+%     $\lambda$-calculus.
+%   \item What is props-as-types? What is linear logic? Linear $\lambda$
+%     is ``just'' a small functional core language.
+%   \item Through props-as-types, synthesis can be formulated as proof
+%     search and so we can leverage a lot of related technology from the
+%     literature.
+%   \end{itemize}
+% \item Goals:
+%   \begin{itemize}
+% \item Full and partial synthesis of well-typed (in the above sense)
+%   programs. Full synthesis means producing a function (or set of)
+%   satisfying the spec. Partial synthesis means being given a partial
+%   program (i.e. a function with a hole in it) and ``completing it''.
+% \item Start from a small system, build up to recursive
+%   types/functions.
+% \item Potential to go further: type refinements; interactive
+%   synthesis?
+%  \item Evaluation through expressiveness benchmarks (can we synthesize
+%    ``X''?); time measurements (how fast can we synthesize ``X''?');
+%    etc.
+%   \end{itemize}
+% \end{enumerate}
 
 \section{Background}
+
+A type system can be formally described through natural deduction -- as a set of rules that determines the concept of deduction for some language~\cite{prawitznd65}.
+This comes from the close relation between logic and programming languages, called the Curry-Howard isomorphism or correspondence. A logical proof describes a program, and a program describes a proof.
+The main observations made are ``propositions as types'', ``proofs as programs'', ``proof checking as type checking'' \cite{https://www.cs.cmu.edu/~fp/courses/15312-f04/handouts/23-curryhoward.pdf}.
+Formalizing our type system, the language i.e.
+the propositions of the logic are the types themselves, the inference rules are used to construct a proof which corresponds directly to a program, and the correctness of the proof assures the program is well-typed.
+The typical natural deduction judgment
+\[
+    A \textrm{ is a proposition}
+\]
+is extended to include a \emph{term}
+(i.e. to make proofs explicit with a formal definition -- these proofs map directly to a $\lambda$-calculus program as stated by the Curry-Howard correspondence????),
+and modified to include a list of hypothesis. 
+This extended judgment, called?? a typing judgment, has the form:
+\[
+    \Gamma \vdash M : A
+\]
+In the type system context, the list of hypothesis, $\Gamma$, is called \emph{typing environment} (i.e. set of variables and their types),
+ the term $M$ is the \emph{program}, and $A$, the natural deduction proposition, is the \emph{type}.
+ This judgment can be read -- $M$ has type $A$ in $\Gamma$.
+
+Type rules assert the validity of certain judgments on the basis of other judgments already known to be valid \cite{http://homepage.cs.uiowa.edu/~tinelli/classes/185/Fall06/notes/cardelli-95.pdf}.
+We introduce a few typing i.e. inference rules to display how a type system would be formalized.
+In natural deduction, there are both \emph{introduction} and \emph{elimination} rules for almost all logical connectives.
+The following rule is a hypothesis rule ??, não sei muito bem como sair daqui:)
+\[
+    \infer*[right=(var)]
+    {  }
+    {\Gamma, u: A \vdash u : A}
+\]
+It reads -- the program $u$ has type $A$ if our typing environment holds a variable labeled $u$ of type $A$.
+[também não sei muito bem onde quero chegar com isto :)]
+As another concrete example, let us describe the inference rule for the function type ($A \supset B$):
+\[
+    \infer*[right=($\supset I$)]
+    {\Gamma, u: A \vdash M : B}
+    {\Gamma \vdash \lambda u. M : A \supset B }
+\]
+It reads -- the program $\lambda u. M$ has type $A \supset B$ ($A$ implies $B$) as long as $M$ is a program of type $B$ knowning that $u$ is a variable of type $A$.
+The rule shown above is an \emph{introduction} rule, that introduces the ``implication'' connective,
+
+\section{Separate}
 

### 2021-05-07 article/article.tex
more

@@
 
+\todo{Merge! These 2 paragraphs}
+A type system can be formally described through a set of inference
+rules that inductively define a judgment of the form $\Gamma \vdash M
+: A$, stating that program expression $M$ has type $A$ according to
+the \emph{typing assumptions} for variables tracked in $\Gamma$.
+\dots
+
+\todo[inline]{Daqui, para tipificar uma função, para Curry-Howard,
+  para lógica linear, para proof search, etc\dots Só faz sentido falar de dedução natural no
+  contexto de ``lógica'' / Curry-Howard.}
+
 A type system can be formally described through natural deduction -- as a set of rules that determines the concept of deduction for some language~\cite{prawitznd65}.
-This comes from the close relation between logic and programming languages, called the Curry-Howard isomorphism or correspondence. A logical proof describes a program, and a program describes a proof.
+This comes from the close relation between logic and programming languages, the Curry-Howard isomorphism or correspondence. A logical proof describes a program, and a program describes a proof.
 The main observations made are ``propositions as types'', ``proofs as programs'', ``proof checking as type checking'' \cite{https://www.cs.cmu.edu/~fp/courses/15312-f04/handouts/23-curryhoward.pdf}.
@@
 The rule shown above is an \emph{introduction} rule, that introduces the ``implication'' connective,
+
+% Linear Logic
+Linear logic \cite{DBLP:journals/tcs/Girard87} can be thought of as a logic of resources.
+Where we were previously able to utilize a variable i.e. a hypothesis as many times as we wanted,
+now we'll keep track of the amount available (the ``resources''), and guarantee that every resource is used exactly once, hence the name ``Linear Logic''.
+It comes that connectives must be decomposed to get more primitive ones that reflect our resource aware system.
+So what would be the $\wedge$ (and) in classical logic, decomposes into its multiplicative
 

### 2021-05-07 article/article.tex
tweaks

@@
 \newcommand{\bang}{{!}}
+\newcommand{\mypara}[1]{\paragraph{\textbf{#1}.}}
 
@@
 Program synthesis is an automated or semi-automated process of
-deriving a program (i.e.~generating code) from a high-level specification.
+deriving a program (i.e.~generating code) from a high-level
+specification.  
+%
+Synthesis can be seen as a means to improve a programmer's
+productivity and/or program correctness (i.e. through suggestion
+and/or autocompletion), or as a tool to automate certain parts of the
+programming process (e.g. in the same way AI might generate some
+boilerplate text to aid a journalist, rather than to replace the
+journalist outright), just to name a few examples.
+
 Specifications can take many forms (e.g.  polymorphic refinement
@@
 examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
-types~\cite{DBLP:conf/lopstr/HughesO20}, ...).  In this work we
-explore type-based synthesis where specifications take the form of
-linear types.\todo{O que é um linear type?}
+types~\cite{DBLP:conf/lopstr/HughesO20}, ...).  Regardless of the kind
+of specification, program synthesizers have to deal with two main
+inherent sources of complexity: searching over a vast space of
+(potentially) valid programs, and interpreting user intent.
 %
-The chosen specification form dictates the synthesis process, and
-different ones might be approached in completely different
-ways.\todo{Esta frase não parece ter grande utilidade. O que queres
-  dizer aqui?}
+In this work we will explore type-based synthesis where specifications
+take the form of linear types, which are types that limit resource
+usage in programs.  Concretely, we will explore the proof-theoretic
+foundations of linear types via the Curry-Howard correspondence with
+linear logic, viewing program synthesis under the lens of (linear
+logic) bottom-up proof search.
 %
-Regardless of the kind of specification, program synthesizers have to
-deal with two main inherent sources of complexity: searching over a
-vast space of (potentially) valid programs, and interpreting user
-intent.
+
 %
-Synthesis has many motivations, it can be seen as a means to
-improve a programmer's productivity and/or program correctness
-(i.e. through suggestion and/or autocompletion), or as a tool to
-automate certain parts of the programming process (e.g. in the same
-way AI might generate some boilerplate text to auxiliate a journalist,
-rather than to replace), just to name a few examples.
-\todo[inline]{Acho que estes 3 ``paragrafos'' deviam estar por ordens diferentes.}
 
+\mypara{Type-Based Synthesis}
 % Type based synthesis
-Type-based synthesis takes a type and produces a program with that
-given type.  Types can be further explored to give the programmer more
-expressiveness (e.g. by adding constraints??), and are called rich
-types.  -- Gostava de ver com o prof aqui. [Estive a procurar e fiquei
+Type-based synthesis uses types as a form of program specification,
+automatically or semi-automatically producing a program expression
+with the given type and so matching the specification.
+%
+In order to make type-based specifications more expressive, type-based
+synthesis frameworks use rich type systems such as refinement
+types~\cite{} and polymorphic types~\cite{}.
+\todo[inline]{[Estive a procurar e fiquei
 um pouco confuso. polimorfismo é rich types?  dependent types are rich
@@
 rever. Por agora vou por apenas os tópicos que o professor escreveu]
-Languages with richer type systems allow for more precise types. This
-precision can statically eliminate various kinds of logical errors by
-making certain invalid program states ill-typed (e.g. a ``null aware''
-type system, will ensure you cannot dereference a null-pointer).
-However, they can also be a burden -- the whole point of these type
-systems is to ensure that ``less'' things are well-typed, and
-sometimes it's hard to convince the checker.  Regarding the
-challenges, type-based synthesis leverages rich types as a way of
-pruning the search space, and by using types gives the user as a more
-``familiar'' specification.
+-- Não há propriamente uma definição rigorosa de ``rich types'',
+geralmente tudo o que vá além do tipico pode ser qualificado como
+rico.}
 
-This work explores the process of synthetising linear functional
-programs from types based in linear logic (linear types), leveraging
-the Curry-Howard correspondence.  Said correspondence states that
-propositions in a logic have a direct mapping to types, and well typed
-programs also have a one to one correspendence to proofs of those
-propositions.  As such, having a type be a proposition in linear
+Richer type systems allow for more precise types, which can
+statically eliminate various kinds of logical errors by making certain
+invalid program states ill-typed (e.g.,~a ``null aware'' type system,
+will ensure at compile-time that you cannot dereference a
+null-pointer). However, they can also be a burden -- the whole point
+of these type systems is to ensure that \emph{fewer} programs are
+deemed well-typed, which can pose additional challenges to the
+development process. Type-based synthesis leverages rich types as a
+way of pruning the search space, and by using types gives the user as
+a more ``familiar'' specification. For instance, the type
+$\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
+specifies a (curried) function that takes two integers and produces an
+integer. Viewed as a specification, it is extremely imprecise (there
+are an infinite number of functions that satisfy this specification).
+However, the richer type $(x{:}\mathsf{Int}) \rightarrow
+(y{:}\mathsf{Int}) \rightarrow \{z{:}\mathsf{Int} \mid z = x+y\}$
+very precisely specifies a function that takes integer arguments $x$
+and $y$ and returns an integer $z$ that is the sum of $x$ and $y$.
+
+%
+This work explores the process of synthesizing linear functional
+programs from types based on linear logic propositions (i.e.~linear
+types) by leveraging the Curry-Howard correspondence.  The
+correspondence states that propositions in a logic have a direct
+mapping to types, and well-typed programs correspond to to proofs of
+those propositions.  As such, having a type be a proposition in linear
 logic, we can relate a proof of that proposition directly to a linear
 functional program — finding a proof is finding a program with that
-type, so we formulate the synthesis as proof search in linear logic,
-which makes to us available a lot of related technology from the proof
-search literature.  Linear types differ from usual types for they
-constrain resource usage in programs by \emph{statically?} limiting
-the number of times certain resources can be used during their
-lifetime.  They can be applied to resource-aware programming such as
-concurrent programming (e.g. session types??), and to
-memory-management (e.g. Rust) [Quero reescrever a frase acima, não
-gosto de como falamos de session types e rust em dois exemplos
-paralelos visto que não são coisas bem comparáveis. Revemos na
-reunião.]
+type. Thus, we formulate synthesis as proof search in linear logic,
+which allows us to inform our work with approaches from the proof
+search literature.  Linear types differ from more traditional types in
+that they constrain resource usage in programs by \emph{statically}
+limiting the number of times certain resources can be used during
+their lifetime.  They can be applied to resource-aware programming
+such as concurrent programming (e.g. session types for message passing
+concurrency~\cite{}), and to memory-management (e.g.~Rust's ownership
+types).
 
+\mypara{Goals.}
  % Goals
 In the end, we intend to be able to do full and partial synthesis of
-well-typed programs -- Full synthesis being the production of a
+well-typed programs. Full synthesis consists of the production of a
 function (or set of) satisfying the specification; Partial synthesis
-being the ``completion'' of a partial program (i.e. a function with a
-\emph{hole} in it) -- Starting from a small core linear --
-i.e. \emph{resource-aware} functional language, and building up to
-recursive types/functions.  Eventually, possibly, enhancing the
-synthesis process with type refinements, and making it interactive --
-i.e. the synthetiser offers the user ``choices'' mid-process.  [Devia
-estar a escrever isto eu ahah?]The evaluation will be done through
-expressiveness benchmarks (i.e. can we synthesize ``X''?); time
-measurements (i.e. how fast can we synthesize ``X''?'); and so on.
+is the ``completion'' of a partial program (i.e. a function with a
+\emph{hole} in it). The work will start from a small core linear --
+i.e. \emph{resource-aware} functional language, building up to
+recursive types/functions, with potential other avenues of further
+extension. 
+%
+We will evaluate the work through
+expressiveness benchmarks (i.e. can we synthesize ``X''?) and time
+measurements (i.e. how fast can we synthesize ``X''?').
 
@@
 This comes from the close relation between logic and programming languages, the Curry-Howard isomorphism or correspondence. A logical proof describes a program, and a program describes a proof.
-The main observations made are ``propositions as types'', ``proofs as programs'', ``proof checking as type checking'' \cite{https://www.cs.cmu.edu/~fp/courses/15312-f04/handouts/23-curryhoward.pdf}.
+The main observations made are ``propositions as types'', ``proofs as
+programs'', ``proof checking as type checking''
+\cite{https://www.cs.cmu.edu/~fp/courses/15312-f04/handouts/23-curryhoward.pdf}.
 Formalizing our type system, the language i.e.
@@
 % Linear Logic
-Linear logic \cite{DBLP:journals/tcs/Girard87} can be thought of as a logic of resources.
-Where we were previously able to utilize a variable i.e. a hypothesis as many times as we wanted,
-now we'll keep track of the amount available (the ``resources''), and guarantee that every resource is used exactly once, hence the name ``Linear Logic''.
-It comes that connectives must be decomposed to get more primitive ones that reflect our resource aware system.
-So what would be the $\wedge$ (and) in classical logic, decomposes into its multiplicative
+\subsection{Linear Logic}
+
+Linear logic \cite{DBLP:journals/tcs/Girard87} can be seen as a
+resource-aware logic, where propositions are interpreted as resources
+that are consumed during the inference process.  Where in standard
+propositional logic we are able to use an assumption as many times as
+we want, in linear logic every resource (i.e., every assumption) must
+be used \emph{exactly once}, or \emph{linearly}. This usage
+restriction gives rise to new logical connectives, based on the way
+the ambient resources are used. For instance, conjunction, usually
+written as $A\wedge B$, appears in two forms in linear logic:
+multiplicative or simultaneous conjunction (written $A\tensor B$); and
+additive or alternative conjunction (written $A\with
+B$). Multiplicative conjunction denotes the simultaneous availability
+of resources $A$ and $B$, requiring both of them to be
+used. Alternative conjunction denotes the availability of $A$ and $B$,
+but where only one of the two resources may be used. Similarly,
+implication becomes into linear implication, written $A\lolli B$,
+denoting a resource that will consume (exactly one) resource $A$ to
+produce a resource $B$.
 

### 2021-05-07 article/article.tex
on linear

--- 2021-05-07 article/article.tex
@@
 In the end, we intend to be able to do full and partial synthesis of
-well-typed programs -- Full synthesis being the production of a
-function (or set of) satisfying the specification; Partial synthesis
+well-typed programs -- full synthesis being the production of a
+function (or set of) satisfying the specification; partial synthesis
 being the ``completion'' of a partial program (i.e. a function with a
@@
 Where we were previously able to utilize a variable i.e. a hypothesis as many times as we wanted,
-now we'll keep track of the amount available (the ``resources''), and guarantee that every resource is used exactly once, hence the name ``Linear Logic''.
+now we'll keep track of the amount available i.e. the ``resources'', and guarantee that every resource is used exactly once, hence the name ``Linear Logic''.
 It comes that connectives must be decomposed to get more primitive ones that reflect our resource aware system.
-So what would be the $\wedge$ (and) in classical logic, decomposes into its multiplicative
+So what would be the $\wedge$ (conjunction) in classical logic, decomposes into its respective multiplicative and additive connectives, $\tensor$ and $\with$.
+To present the formalization of this logic, besides the new connectives, we need to introduce the \emph{resource-aware context} $\Delta$.
+ In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types, but in which each and every variable must be used exactly once.
+ So, to introduce the connective $\tensor$ which defines a multiplicative pair of propositions, we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
+ to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
+\[
+    \infer*[right=($\tensor I$)]
+    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
+    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
+\]
+Out of the logical connectives, we need to mention one more, since it'll alter the judgment form, and it's the one 
+that ensures logical strength -- i.e. we're able to translate usual logic into linear logic.
+The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
+it useable more than once.
+To distinguish the ``infinite'' variables, a separate context is used -- $\Gamma$.
+So $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.
 

### 2021-05-07 article/article.tex
tweaks

--- 2021-05-07 article/article.tex
@@
 
-\mypara{Goals.}
+\mypara{Goals}
  % Goals
@@
 
-\todo{Merge! These 2 paragraphs}
-A type system can be formally described through a set of inference
-rules that inductively define a judgment of the form $\Gamma \vdash M
-: A$, stating that program expression $M$ has type $A$ according to
-the \emph{typing assumptions} for variables tracked in $\Gamma$.
-\dots
+\mypara{Type Systems} A type system can be formally described through
+a set of inference rules that inductively define a judgment of the
+form $\Gamma \vdash M : A$, stating that program expression $M$ has
+type $A$ according to the \emph{typing assumptions} for variables
+tracked in $\Gamma$. For instance,
+$x{:}\mathsf{Int}, y{:}\mathsf{Int} \vdash x+y : \mathsf{Int}$ states
+that $x+y$ has type $\mathsf{Int}$ under the assumption that $x$ and
+$y$ have type $\mathsf{Int}$.  An expression $M$ is deemed well-typed
+with a given type $A$ if one can construct a typing derivation with $M :
+A$ as its conclusion, by repeated application of the inference rules.
+
+The simply-typed $\lambda$-calculus is a typed core functional
+language~\cite{} that captures \dots. Its syntax consists of
+functional abstraction, written $\lambda x{:}A.M$, denoting the
+an (anonymous) function that takes an argument of type $A$, bound to
+$x$ in $M$; and application $M\,N$, with the standard meaning, and
+variables $x$. For instance, the \dots \todo{Identity function!}
+
+\mypara{Propositions as Types}
+%
+It turns out that the inference rules of the simply-typed
+$\lambda$-calculus are closely related to those of a system of natural
+deduction for intuitionistic logic~\cite{prawitznd65}. This
+relationship, known as the Curry-Howard correspondence~\cite{},
+identifies that the propositions of intuitionistic logic can be read
+as types for the simply-typed $\lambda$-calculus, and their proofs are
+exactly the \dots.
+The main observations made are ``propositions as types'', ``proofs as
+programs'', ``proof checking as type checking''~\cite{}.
+
+Inference in ND are categorized as introduction or elimination, these
+correspond to constructors and destructors in programming languages
+(e.g. function abstraction vs application, construction of a pair vs
+projection).
+
+We introduce a few typing rules to display how a type system can be
+formalized.  In natural deduction, there are both \emph{introduction}
+and \emph{elimination} rules for almost all logical
+connectives.\todo{se mencionares intro e elim antes nao precisas de
+  repetir aqui.}  The following rule captures the nature of a
+hypothetical judgment, allowing for reasoning from assumptions:
+\[
+    \infer*[right=(var)]
+    {  }
+    {\Gamma, u: A \vdash u : A}
+  \]
+  When seen as a typing rule for the $\lambda$-calculus, it
+  corresponds to the rule for typing variables -- variable $u$ has
+  type $A$ if the typing environment contains a variable $u$ of
+  type $A$.
+
+  As another concrete example, let us consider the rule for
+  implication ($A \supset B$):
+\[
+    \infer*[right=($\supset I$)]
+    {\Gamma, u: A \vdash M : B}
+    {\Gamma \vdash \lambda u. M : A \supset B }
+  \]
+  Logically, the rule states that to prove $A\supset B$, we assume $A$
+  and prove $B$. Through the Curry-Howard correspondence, implication
+  corresponds to the function type -- the program $\lambda u. M$ has
+  type $A \supset B$, provided $M$ has type
+  type $B$ under the assumption that $u$ is a variable of type $A$.  The rule
+  shown above is an \emph{introduction} rule, that introduces the
+  ``implication'' connective. 
+%$\rightarrow$ $\Rightarrow$, $\Longrightarrow$, $\longrightarrow$
+
+  \todo[inline]{Mesma historia para a aplicacao/modus ponens.
+Um bocadinho das ``consequencias'' do Curry-Howard. }
+
+The Curry-Howard correspondence generalizes beyond the simply-typed
+$\lambda$-calculus and propositional intuitionistic logic.
+\todo{Mencionar polimorfismo corresponde a logica de ordem superior,
+  etc. E tambem para logica linear \dots}
 
@@
 
-A type system can be formally described through natural deduction -- as a set of rules that determines the concept of deduction for some language~\cite{prawitznd65}.
-This comes from the close relation between logic and programming languages, the Curry-Howard isomorphism or correspondence. A logical proof describes a program, and a program describes a proof.
-The main observations made are ``propositions as types'', ``proofs as
-programs'', ``proof checking as type checking''
-\cite{https://www.cs.cmu.edu/~fp/courses/15312-f04/handouts/23-curryhoward.pdf}.
-Formalizing our type system, the language i.e.
-the propositions of the logic are the types themselves, the inference rules are used to construct a proof which corresponds directly to a program, and the correctness of the proof assures the program is well-typed.
-The typical natural deduction judgment
-\[
-    A \textrm{ is a proposition}
-\]
-is extended to include a \emph{term}
-(i.e. to make proofs explicit with a formal definition -- these proofs map directly to a $\lambda$-calculus program as stated by the Curry-Howard correspondence????),
-and modified to include a list of hypothesis. 
-This extended judgment, called?? a typing judgment, has the form:
-\[
-    \Gamma \vdash M : A
-\]
-In the type system context, the list of hypothesis, $\Gamma$, is called \emph{typing environment} (i.e. set of variables and their types),
- the term $M$ is the \emph{program}, and $A$, the natural deduction proposition, is the \emph{type}.
- This judgment can be read -- $M$ has type $A$ in $\Gamma$.
-
-Type rules assert the validity of certain judgments on the basis of other judgments already known to be valid \cite{http://homepage.cs.uiowa.edu/~tinelli/classes/185/Fall06/notes/cardelli-95.pdf}.
-We introduce a few typing i.e. inference rules to display how a type system would be formalized.
-In natural deduction, there are both \emph{introduction} and \emph{elimination} rules for almost all logical connectives.
-The following rule is a hypothesis rule ??, não sei muito bem como sair daqui:)
-\[
-    \infer*[right=(var)]
-    {  }
-    {\Gamma, u: A \vdash u : A}
-\]
-It reads -- the program $u$ has type $A$ if our typing environment holds a variable labeled $u$ of type $A$.
-[também não sei muito bem onde quero chegar com isto :)]
-As another concrete example, let us describe the inference rule for the function type ($A \supset B$):
-\[
-    \infer*[right=($\supset I$)]
-    {\Gamma, u: A \vdash M : B}
-    {\Gamma \vdash \lambda u. M : A \supset B }
-\]
-It reads -- the program $\lambda u. M$ has type $A \supset B$ ($A$ implies $B$) as long as $M$ is a program of type $B$ knowning that $u$ is a variable of type $A$.
-The rule shown above is an \emph{introduction} rule, that introduces the ``implication'' connective,
-
 % Linear Logic
-\subsection{Linear Logic}
+\mypara{Linear Logic}
 
@@
 
+\todo{Through the Curry-Howard correspondence, we can view the process
+  of finding a proof of a proposition as the process of synthesizing a
+program of a given type.}
+
 \section{Separate}
@@
 \begin{enumerate}
-\item PL/Type systems in a ``formal'' sense (inference rules, etc)
+\item PL/Type systems in a ``formal'' sense (inference rules, etc) [DONE]
 \item Propositions as types: Proofs as programs, proof search as
-  synthesis, etc.
-\item Linear logic: a ``resource aware logic''
-\item Proof search in LL: challenges, focusing as a ``solution''
+  synthesis, etc. [MOSTLY DONE]
+\item Linear logic: a ``resource aware logic'' [ONGOING]
+\item Proof search in LL: challenges, focusing as a ``solution'' [ONGOING]
 \item Linear $\lambda$-calculus: what does it look like.
@@
 
-Concrete works:
+\todo[inline]{Os~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
+  como ``estrategia geral'' de ``ler'' um sistema de tipos +
+  engenharia como um sistema de sintese de programas tipificados;
+  Aqui~\cite{DBLP:conf/pldi/PolikarpovaKS16} estamos num contexto
+  ``mais avançado'' (polimorfismo e tipos refinados -- ``logica de
+  primeira ordem'') -- tipos ricos -- trabalho aspiracional,
+  tratamento de funcao recursivas, tratamento de tipos de dados
+  indutivos/algebricos; Aqui~\cite{DBLP:conf/lopstr/HughesO20},
+  sintese para graded modal types, tem a ver com ``linearidade'' kind
+  of, abordagem baseado em focusing (portanto Curry-Howard++), no caso
+  deles acaba por ser uma lógica ``modal'' vs linear -- não permite
+  futuramente explorar relações com e.g. concorrência, análise de
+  recursos.}
+
+Concrete works
 \begin{itemize}
-\item Synthesis for graded types (no recursion   -- in that work, no
-  obvious connection of grading with
-  concurrency)~\cite{DBLP:conf/lopstr/HughesO20}
 
@@
   or how to turn your type system upside down.
+
+  \item Synthesis for graded types (no recursion   -- in that work, no
+  obvious connection of grading with
+  concurrency)~\cite{DBLP:conf/lopstr/HughesO20}
 
@@
 
-\item Synthesis of heap-manipulating
-  programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}
+% \item Synthesis of heap-manipulating
+%   programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}
 
-Less related, but more evidence of this idea of turning ``checking''
-systems into synthesis frameworks.
+% Less related, but more evidence of this idea of turning ``checking''
+% systems into synthesis frameworks.
 
-\item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}
+% \item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}
 
-This resource-guided means something a bit different. Programs satisfy
-a functional specification and a symbolic resource bound in the sense
-of amortized analysis, but can provide technical insights.
+% This resource-guided means something a bit different. Programs satisfy
+% a functional specification and a symbolic resource bound in the sense
+% of amortized analysis, but can provide technical insights.
 

### 2021-05-07 article/article.tex
merge linear logic background

@@
 well-typed programs. Full synthesis consists of the production of a
-function (or set of) satisfying the specification; Partial synthesis
+function (or set of) satisfying the specification; partial synthesis
 is the ``completion'' of a partial program (i.e. a function with a
@@
 measurements (i.e. how fast can we synthesize ``X''?').
-
-% \begin{enumerate}
-% \item General motivation
-%   \begin{itemize}
-% \item 
-% \item Specifications appear in many forms\dots
-% \item Program synthesis as a means of improving program
-%   correctness/programmer productivity.
-% \item Challenges: space of valid programs, space of valid
-%   specifications,  etc.
-% \end{itemize}
-% \item \emph{Type-based} program synthesis
-%   \begin{itemize}
-%  \item Rich types as specifications.
-% \item Languages with richer type systems allow for more precise
-%   types. This precision can statically eliminate various kinds of
-%   logical errors by making certain invalid program states ill-typed
-%   (e.g. a ``null aware'' type system, will ensure you cannot
-%   dereference a null-pointer).
-% \item But can also be a burden -- the whole point of these type
-%   systems is to ensure that ``less'' things are well-typed, and
-%   sometimes it's hard to convince the checker\dots
-% \item Type-based synthesis leverages rich types as a way of pruning
-%   the search space and by using types as a more ``familiar''
-%   specification.
-% \end{itemize}
-% \item ``My'' problem: Synthesis based on linear types
-%   \begin{itemize}
-%   \item Linear types constrain resource usage in programs by \emph{statically} limiting
-%     the number of times certain resources can be used during its
-%     lifetime.
-%   \item Applications in resource-aware programming such as concurrent
-%     programming (e.g. session types) and memory-management
-%     (e.g. Rust).
-%   \item This work will study program synthesis in a functional,
-%     linearly-typed setting, by leveraging the propositions-as-types
-%     correspondence between linear logic and the (linear)
-%     $\lambda$-calculus.
-%   \item What is props-as-types? What is linear logic? Linear $\lambda$
-%     is ``just'' a small functional core language.
-%   \item Through props-as-types, synthesis can be formulated as proof
-%     search and so we can leverage a lot of related technology from the
-%     literature.
-%   \end{itemize}
-% \item Goals:
-%   \begin{itemize}
-% \item Full and partial synthesis of well-typed (in the above sense)
-%   programs. Full synthesis means producing a function (or set of)
-%   satisfying the spec. Partial synthesis means being given a partial
-%   program (i.e. a function with a hole in it) and ``completing it''.
-% \item Start from a small system, build up to recursive
-%   types/functions.
-% \item Potential to go further: type refinements; interactive
-%   synthesis?
-%  \item Evaluation through expressiveness benchmarks (can we synthesize
-%    ``X''?); time measurements (how fast can we synthesize ``X''?');
-%    etc.
-%   \end{itemize}
-% \end{enumerate}
 
@@
 
-% Linear Logic
 \mypara{Linear Logic}
@@
 produce a resource $B$.
+To present the formalization of this logic, besides the new connectives, we need to introduce the \emph{resource-aware context} $\Delta$.
+ In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types, but in which each and every variable must be used exactly once.
+ So, to introduce the connective $\tensor$ which defines a multiplicative pair of propositions, we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
+ to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
+\[
+    \infer*[right=($\tensor I$)]
+    {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
+    {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
+\]
+Out of the logical connectives, we need to mention one more, since it'll alter the judgment form, and it's the one 
+that ensures logical strength -- i.e. we're able to translate usual logic into linear logic.
+The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
+it useable more than once.
+To distinguish the ``infinite'' variables, a separate context is used -- $\Gamma$.
+So $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.
 

### 2021-05-08 article/article.tex
uns tweaks

@@
 We will evaluate the work through
-expressiveness benchmarks (i.e. can we synthesize ``X''?) and time
-measurements (i.e. how fast can we synthesize ``X''?').
+expressiveness benchmarks (i.e. can we synthesize a term of type $T$?) and time
+measurements (i.e. how fast can we synthesize a term of type $T$?').
 
@@
 $x$ in $M$; and application $M\,N$, with the standard meaning, and
-variables $x$. For instance, the \dots \todo{Identity function!}
+variables $x$. For instance, the term $\lambda x{:}A.x$ denotes the
+identity function.
 
@@
 as types for the simply-typed $\lambda$-calculus, and their proofs are
-exactly the \dots.
+exactly the programs with the given type.
 The main observations made are ``propositions as types'', ``proofs as
@@
 
-Inference in ND are categorized as introduction or elimination, these
-correspond to constructors and destructors in programming languages
-(e.g. function abstraction vs application, construction of a pair vs
-projection).
+Inference rules in natural deduction are generally categorized as
+introduction or elimination rules, which correspond, respectively, to
+constructors and destructors in programming languages (e.g.~function
+abstraction vs application, construction of a pair vs projections).
 
-We introduce a few typing rules to display how a type system can be
-formalized.  In natural deduction, there are both \emph{introduction}
-and \emph{elimination} rules for almost all logical
-connectives.\todo{se mencionares intro e elim antes nao precisas de
-  repetir aqui.}  The following rule captures the nature of a
-hypothetical judgment, allowing for reasoning from assumptions:
+We introduce a few rules to both show how a type system can be
+formalized and to show the relationship with (intuitionistic)
+propositional logic.  In natural deduction, there are both
+\emph{introduction} and \emph{elimination} rules for almost all
+logical connectives.\todo{se mencionares intro e elim antes nao
+  precisas de repetir aqui.}  The following rule captures the nature
+of a hypothetical judgment, allowing for reasoning from assumptions:
 \[
@@
 produce a resource $B$.
-To present the formalization of this logic, besides the new connectives, we need to introduce the \emph{resource-aware context} $\Delta$.
- In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types, but in which each and every variable must be used exactly once.
- So, to introduce the connective $\tensor$ which defines a multiplicative pair of propositions, we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
- to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
+
+To present the formalization of this logic, besides the new
+connectives, we need to introduce the \emph{resource-aware context}
+$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
+also a list of variables and their types, but where each and every
+variable must be used exactly once during inference.  So, to introduce
+the connective $\tensor$ which defines a multiplicative pair of
+propositions, we must use exactly all the resources
+($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$) 
+proposition $A$, and ($\Delta_2$) proposition $B$:
 \[
@@
 \]
-Out of the logical connectives, we need to mention one more, since it'll alter the judgment form, and it's the one 
-that ensures logical strength -- i.e. we're able to translate usual logic into linear logic.
-The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
-it useable more than once.
-To distinguish the ``infinite'' variables, a separate context is used -- $\Gamma$.
-So $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.
+Out of the logical connectives, we need to mention one more, since
+it'll alter the judgment form, and it's the one that ensures logical
+strength -- i.e. we're able to translate intuitionistic logic into
+linear logic.  The proposition $\bang A$ (read \emph{of course} $A$)
+is used (under certain conditions) to make a resource ``infinite''
+i.e. to make it useable an arbitrary number of times. To distinguish the
+``infinite'' variables, a separate, unrestricted, context is used -- $\Gamma$. So
+$\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources
+that can only be used once.
 

### 2021-05-08 article/article.tex
slow work

--- 2021-05-07 article/article.tex
@@
 logic) bottom-up proof search.
-%
 
@@
 types~\cite{} and polymorphic types~\cite{}.
-\todo[inline]{[Estive a procurar e fiquei
-um pouco confuso. polimorfismo é rich types?  dependent types are rich
-types? aqueles constraints são os dependent types pelo que
-percebi... mas n me parece correto. arithmetic types? temos de
-rever. Por agora vou por apenas os tópicos que o professor escreveu]
--- Não há propriamente uma definição rigorosa de ``rich types'',
-geralmente tudo o que vá além do tipico pode ser qualificado como
-rico.}
 
@@
 correspondence states that propositions in a logic have a direct
-mapping to types, and well-typed programs correspond to to proofs of
+mapping to types, and well-typed programs correspond to proofs of
 those propositions.  As such, having a type be a proposition in linear
@@
 The simply-typed $\lambda$-calculus is a typed core functional
-language~\cite{} that captures \dots. Its syntax consists of
-functional abstraction, written $\lambda x{:}A.M$, denoting the
+language~\cite{} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
+functional abstraction, written $\lambda x{:}A.M$, denoting
 an (anonymous) function that takes an argument of type $A$, bound to
 $x$ in $M$; and application $M\,N$, with the standard meaning, and
-variables $x$. For instance, the \dots \todo{Identity function!}
+variables $x$. For instance, $\lambda x{:}A.x$, the identity function, is a functional abstraction
+taking an argument of type $A$ and returning it back.
 
@@
 identifies that the propositions of intuitionistic logic can be read
-as types for the simply-typed $\lambda$-calculus, and their proofs are
-exactly the \dots.
-The main observations made are ``propositions as types'', ``proofs as
-programs'', ``proof checking as type checking''~\cite{}.
+as types for the simply-typed $\lambda$-calculus (``propositions as types'' observation), their proofs are
+exactly the program with that type (``proofs as programs'' observation), and checking a proof is type checking a program (``proof checking as type checking'' observation)~\cite{}.
 
-Inference in ND are categorized as introduction or elimination, these
+Inference rules in natural deduction are categorized as introduction or elimination, these
 correspond to constructors and destructors in programming languages
@@
 We introduce a few typing rules to display how a type system can be
-formalized.  In natural deduction, there are both \emph{introduction}
-and \emph{elimination} rules for almost all logical
-connectives.\todo{se mencionares intro e elim antes nao precisas de
-  repetir aqui.}  The following rule captures the nature of a
+formalized. The following rule captures the nature of a
 hypothetical judgment, allowing for reasoning from assumptions:
@@
 As another concrete example, let us consider the rule for
-  implication ($A \supset B$):
+implication ($A\rightarrow~B$):
 \[
-    \infer*[right=($\supset I$)]
+    \infer*[right=($\rightarrow I$)]
     {\Gamma, u: A \vdash M : B}
-    {\Gamma \vdash \lambda u. M : A \supset B }
+    {\Gamma \vdash \lambda u. M : A \rightarrow B}
 \]
-  Logically, the rule states that to prove $A\supset B$, we assume $A$
+Logically, the rule states that to prove $A\rightarrow B$, we assume $A$
 and prove $B$. Through the Curry-Howard correspondence, implication
 corresponds to the function type -- the program $\lambda u. M$ has
-  type $A \supset B$, provided $M$ has type
-  type $B$ under the assumption that $u$ is a variable of type $A$.  The rule
+type $A \rightarrow B$, provided $M$ has type
+$B$ under the assumption that $u$ is a variable of type $A$. The rule
 shown above is an \emph{introduction} rule, that introduces the
 ``implication'' connective.
-%$\rightarrow$ $\Rightarrow$, $\Longrightarrow$, $\longrightarrow$
 
-  \todo[inline]{Mesma historia para a aplicacao/modus ponens.
-Um bocadinho das ``consequencias'' do Curry-Howard. }
+Finally, let us consider an \emph{elimination} rule, also for the implication connective:
+\[
+    \infer*[right=($\rightarrow E$)]
+    {\Gamma \vdash M : A \rightarrow B \and \Gamma \vdash N : A}
+    {\Gamma \vdash M\,N : B}
+\]
+Elimination rules are easier to think of in a top-down manner. Logically, this rule states
+that if we prove $A \rightarrow B$ and $A$ we can prove $B$. Through the Curry-Howard correspondence,
+implication elimination corresponds to function application, and its type is the function return type
+-- the program $M N$ has type $B$, provided $M$ has type $A \rightarrow B$ and $N$ has type $A$.
 
 The Curry-Howard correspondence generalizes beyond the simply-typed
-$\lambda$-calculus and propositional intuitionistic logic.
-\todo{Mencionar polimorfismo corresponde a logica de ordem superior,
-  etc. E tambem para logica linear \dots}
+$\lambda$-calculus and propositional intuitionistic logic. It extends to
+the realms of polymorphism (higher-order logic \todo{review - {https://en.wikipedia.org/wiki/Lambda_cube}} second-order propositional calculus),
+dependent types (first-order predicate logic), and every other extension
+of natural deduction and simply-typed $\lambda$-calculus -- which also means
+linear logic and linear types.
 
@@
 produce a resource $B$.
-To present the formalization of this logic, besides the new connectives, we need to introduce the \emph{resource-aware context} $\Delta$.
- In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types, but in which each and every variable must be used exactly once.
- So, to introduce the connective $\tensor$ which defines a multiplicative pair of propositions, we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
+To present the formalization of this logic, besides the new connectives, we 
+need to introduce the \emph{resource-aware context} $\Delta$.
+ In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types,
+ but in which each and every variable must be used exactly once.
+To introduce the connective $\tensor$ which defines a multiplicative pair of propositions,
+ we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
  to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
@@
 The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
-it useable more than once.
-To distinguish the ``infinite'' variables, a separate context is used -- $\Gamma$.
-So $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.
+it useable more than once. To distinguish the ``infinite'' variables, a separate context is used --
+$\Gamma$. $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.
+The linear typing judgment for the introduction of the exponential $\bang$ takes the form:
+\todo{A formula está um pouco estranha (faltam espaços)?}
+\[
+    \infer*[right=($\bang I$)]
+    {\Gamma ; \emptyset \vdash M : A}
+    {\Gamma ; \emptyset \vdash !M : !A}
+\]
+\todo{Describe exponent introduction and new gamma context}
+\todo{Add h spaces in formula}
+\[
+    \infer*[right=($\bang E$)]
+    {\Gamma ; \Delta_1 \vdash M : !A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
+    {\Gamma ; \Delta_1, \Delta_2 \vdash let !u = M \mathsf{in} N : C}
+\]
+\todo{Describe exponent elimination and new gamma context}
 
@@
 Concrete works
-\begin{itemize}
+
+\mypara{Type-and-Example-Directed Program Synthesis}~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
+In this paper, type synthesis is explored
 
 \item Type-and-example-directed
-  synthesis~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
+  synthesis~
   or how to turn your type system upside down.
@@
 
-\end{itemize}
-
 \section{Goals and Work Plan}
@@
 Mention that you've already implemented a type-checker for this
-(useful as a prelim. exercise but also later, for \emph{validation}).
+(useful as a prelim. Exercise but also later, for \emph{validation}).
 
@@
   \infer[$(\lolli\! I)$]
-  {\Delta , x{:}A \vdash M : B }
+  {\Delta , x{:}A \vdash M : B}
   {\Delta \vdash \lambda x {:} A . M : A \lolli B}

### 2021-05-08 article/article.tex
first related work small

@@
 $x$ in $M$; and application $M\,N$, with the standard meaning, and
-variables $x$. For instance, $\lambda x{:}A.x$, the identity function, is a functional abstraction
+variables $x$. For instance, the term $\lambda x{:}A.x$, denoting the identity function, is a functional abstraction
 taking an argument of type $A$ and returning it back.
@@
 identifies that the propositions of intuitionistic logic can be read
-as types for the simply-typed $\lambda$-calculus (``propositions as types'' observation), their proofs are
-exactly the program with that type (``proofs as programs'' observation), and checking a proof is type checking a program (``proof checking as type checking'' observation)~\cite{}.
+as types for the simply-typed $\lambda$-calculus (``propositions as types'' observation),
+their proofs are exactly the program with the given type (``proofs as programs'' observation),
+and checking a proof is type checking a program (``proof checking as type checking'' observation)~\cite{}.
 
@@
 
-We introduce a few typing rules to display how a type system can be
-formalized. The following rule captures the nature of a
+We introduce a few typing rules to both show how a type system can be
+formalized and to show the relationship with (intuitionistic)
+propositional logic. The following rule captures the nature of a
 hypothetical judgment, allowing for reasoning from assumptions:
@@
 produce a resource $B$.
-To present the formalization of this logic, besides the new connectives, we 
-need to introduce the \emph{resource-aware context} $\Delta$.
- In contrast to the previously seen $\Gamma$, $\Delta$ is also a list of variables and their types,
- but in which each and every variable must be used exactly once.
-To introduce the connective $\tensor$ which defines a multiplicative pair of propositions,
- we must use exactly all the resources ($\Delta_1, \Delta_2$) needed
- to make the ($\Delta_1$) first proposition, and ($\Delta_2$) second proposition:
+
+To present the formalization of this logic, besides the new
+connectives, we need to introduce the \emph{resource-aware context}
+$\Delta$.  In contrast to the previously seen $\Gamma$, $\Delta$ is
+also a list of variables and their types, but where each and every
+variable must be used exactly once during inference.  So, to introduce
+the connective $\tensor$ which defines a multiplicative pair of
+propositions, we must use exactly all the resources
+($\Delta_1, \Delta_2$) needed to realize the ($\Delta_1$) 
+proposition $A$, and ($\Delta_2$) proposition $B$:
 \[
@@
 \]
-Out of the logical connectives, we need to mention one more, since it'll alter the judgment form, and it's the one 
-that ensures logical strength -- i.e. we're able to translate usual logic into linear logic.
-The $\bang$ is used (under certain conditions) to make a variable ``infinite'' i.e. to make
-it useable more than once. To distinguish the ``infinite'' variables, a separate context is used --
-$\Gamma$. $\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources that can only be used once.
+Out of the logical connectives, we need to mention one more, since
+it'll alter the judgment form, and it's the one that ensures logical
+strength -- i.e. we're able to translate intuitionistic logic into
+linear logic.
+  The proposition $\bang A$ (read \emph{of course} $A$)
+is used (under certain conditions) to make a resource ``infinite''
+i.e. to make it useable an arbitrary number of times. To distinguish the
+``infinite'' variables, a separate, unrestricted, context is used -- $\Gamma$. So
+$\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources
+that can only be used once.
 The linear typing judgment for the introduction of the exponential $\bang$ takes the form:
@@
 \mypara{Type-and-Example-Directed Program Synthesis}~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
-In this paper, type synthesis is explored
+This paper explores a purely functional, recursive program synthesis technique based on types/proof search (as we intend to do),
+with examples as an auxiliar technique to trim down the program synthesis vast search space.
+\todo[inline]{more introduction to the paper}.
+A good use is made of the general strategy of using a type system
+to generate programs (by ``inverting'' it), rather than to type-check one.
+Moreover, the paper uses a data structure called ``refinement tree'' in conjunction
+with the extra information provided by the examples to allow for efficient synthesis
+of non-trivial programs.
 

### 2021-05-08 article/article.tex
related work introduction?

@@
 
-\todo[inline]{Os~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
-  como ``estrategia geral'' de ``ler'' um sistema de tipos +
-  engenharia como um sistema de sintese de programas tipificados;
-  Aqui~\cite{DBLP:conf/pldi/PolikarpovaKS16} estamos num contexto
-  ``mais avançado'' (polimorfismo e tipos refinados -- ``logica de
-  primeira ordem'') -- tipos ricos -- trabalho aspiracional,
-  tratamento de funcao recursivas, tratamento de tipos de dados
-  indutivos/algebricos; Aqui~\cite{DBLP:conf/lopstr/HughesO20},
-  sintese para graded modal types, tem a ver com ``linearidade'' kind
-  of, abordagem baseado em focusing (portanto Curry-Howard++), no caso
-  deles acaba por ser uma lógica ``modal'' vs linear -- não permite
-  futuramente explorar relações com e.g. concorrência, análise de
-  recursos.}
+% \todo[inline]{Os~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
+%   como ``estrategia geral'' de ``ler'' um sistema de tipos +
+%   engenharia como um sistema de sintese de programas tipificados;
+%   Aqui~\cite{DBLP:conf/pldi/PolikarpovaKS16} estamos num contexto
+%   ``mais avançado'' (polimorfismo e tipos refinados -- ``logica de
+%   primeira ordem'') -- tipos ricos -- trabalho aspiracional,
+%   tratamento de funcao recursivas, tratamento de tipos de dados
+%   indutivos/algebricos; Aqui~\cite{DBLP:conf/lopstr/HughesO20},
+%   sintese para graded modal types, tem a ver com ``linearidade'' kind
+%   of, abordagem baseado em focusing (portanto Curry-Howard++), no caso
+%   deles acaba por ser uma lógica ``modal'' vs linear -- não permite
+%   futuramente explorar relações com e.g. concorrência, análise de
+%   recursos.}
 
-Concrete works
+Type-based program synthesis is a studied field\todo{ dizer de uma forma melhor? }, and it follows
+that a lot of literature is available to inspire and complement our work.
+Most? follow the synthesis-as-proof-search pattern, however, the process is
+novel for each, due to a variety of different rich types explored, and their respective logics; or nuances done to the process, such
+as considering examples; or even the programming paradigm of the output produced (e.g. generating heap manipulating programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).
+Some other common patterns we might want to follow are the type refinements and the support for recursive functions.
+These works share some common ground, and base themselves of the same works we do -- most famously the Curry-Howard correspondence, but also,
+for example, works on the method of Focusing~\cite{focusing}.
 
@@
 A good use is made of the general strategy of using a type system
-to generate programs (by ``inverting'' it), rather than to type-check one.
+to generate programs (by ``inverting'' the type system), rather than using it to type-check one.
 Moreover, the paper uses a data structure called ``refinement tree'' in conjunction
 with the extra information provided by the examples to allow for efficient synthesis
-of non-trivial programs.
+of non-trivial programs. Overall it employs good engineering to achieve type-based synthesis
+of functional typed programs, with at least two additions that we might want to follow too
+(type refinements and recursive functions).
 
-\item Type-and-example-directed
-  synthesis~
-  or how to turn your type system upside down.
+% or how to turn your type system upside down ~ não apagar a Kubrick reference? :)
 
-  \item Synthesis for graded types (no recursion   -- in that work, no
+\mypara{Synthesis from (polymorphic) refinement types}~\cite{DBLP:conf/pldi/PolikarpovaKS16}
+
+\mypara{Synthesis for graded types} (no recursion   -- in that work, no
   obvious connection of grading with
   concurrency)~\cite{DBLP:conf/lopstr/HughesO20}
-
-\item Synthesis from (polymorphic) refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16}
 

### 2021-05-09 article/article.tex
.

@@
 identifies that the propositions of intuitionistic logic can be read
-as types for the simply-typed $\lambda$-calculus (``propositions as types'' observation),
-their proofs are exactly the program with the given type (``proofs as programs'' observation),
-and checking a proof is type checking a program (``proof checking as type checking'' observation)~\cite{}.
+as types for the simply-typed $\lambda$-calculus (``propositions as
+types''), their proofs are exactly the program with the given type
+(``proofs as programs''), and checking a proof is type checking a
+program (``proof checking as type checking'')~\cite{}.
 
-Inference rules in natural deduction are categorized as introduction or elimination, these
-correspond to constructors and destructors in programming languages
-(e.g. function abstraction vs application, construction of a pair vs
-projection).
+Inference rules in natural deduction are categorized as introduction
+or elimination rules, these correspond, respectively, to rules for
+constructors and destructors in programming languages (e.g. function
+abstraction vs application, construction of a pair vs projection).
 
-We introduce a few typing rules to both show how a type system can be
+We introduce select typing rules to both show how a type system can be
 formalized and to show the relationship with (intuitionistic)
@@
 
-Finally, let us consider an \emph{elimination} rule, also for the implication connective:
+Finally, let us consider an \emph{elimination} rule, also for implication:
 \[
@@
 \]
-Elimination rules are easier to think of in a top-down manner. Logically, this rule states
-that if we prove $A \rightarrow B$ and $A$ we can prove $B$. Through the Curry-Howard correspondence,
-implication elimination corresponds to function application, and its type is the function return type
--- the program $M N$ has type $B$, provided $M$ has type $A \rightarrow B$ and $N$ has type $A$.
+Elimination rules are easier to think of in a top-down
+manner. Logically, this rule states that if we prove $A \rightarrow B$
+and $A$ we can prove $B$. Through the Curry-Howard correspondence,
+implication elimination corresponds to function application, and its
+type is the function return type -- the application $M N$ has type
+$B$, provided $M$ has type $A \rightarrow B$ and $N$ has type $A$.
 
 The Curry-Howard correspondence generalizes beyond the simply-typed
-$\lambda$-calculus and propositional intuitionistic logic. It extends to
-the realms of polymorphism (higher-order logic \todo{review - {https://en.wikipedia.org/wiki/Lambda_cube}} second-order propositional calculus),
-dependent types (first-order predicate logic), and every other extension
-of natural deduction and simply-typed $\lambda$-calculus -- which also means
-linear logic and linear types.
-
-\todo[inline]{Daqui, para tipificar uma função, para Curry-Howard,
-  para lógica linear, para proof search, etc\dots Só faz sentido falar de dedução natural no
-  contexto de ``lógica'' / Curry-Howard.}
+$\lambda$-calculus and propositional intuitionistic logic. It extends
+to the realms of polymorphism (higher-order logic \todo{review -
+  \url{https://en.wikipedia.org/wiki/Lambda_cube}} second-order
+propositional calculus), dependent types (first-order predicate
+logic), and various other extensions of natural deduction and
+simply-typed $\lambda$-calculus -- which include linear logic and
+linear types.
 
@@
 \]
-Out of the logical connectives, we need to mention one more, since
-it'll alter the judgment form, and it's the one that ensures logical
-strength -- i.e. we're able to translate intuitionistic logic into
-linear logic.
-  The proposition $\bang A$ (read \emph{of course} $A$)
-is used (under certain conditions) to make a resource ``infinite''
-i.e. to make it useable an arbitrary number of times. To distinguish the
-``infinite'' variables, a separate, unrestricted, context is used -- $\Gamma$. So
-$\Gamma$ holds the ``infinite'' resources, and $\Delta$ the resources
-that can only be used once.
-The linear typing judgment for the introduction of the exponential $\bang$ takes the form:
-\todo{A formula está um pouco estranha (faltam espaços)?}
+Out of the logical connectives, we need to mention one more, since it
+augments the form of the judgment and it's the one that ensures
+logical strength -- i.e. we're able to translate intuitionistic logic
+into linear logic.  The proposition $\bang A$ (read \emph{of course}
+$A$) is used (under certain conditions) to make a resource
+``infinite'' i.e. to make it useable an arbitrary number of times. To
+distinguish the ``infinite'' variables, a separate, unrestricted,
+context is used -- $\Gamma$. So $\Gamma$ holds the ``infinite''
+resources, and $\Delta$ the resources that can only be used once.  The
+linear typing judgment for the introduction of the exponential $\bang A$
+takes the form: 
 \[
@@
     {\Gamma ; \emptyset \vdash M : A}
-    {\Gamma ; \emptyset \vdash !M : !A}
+    {\Gamma ; \emptyset \vdash \bang M : \bang A}
   \]
-\todo{Describe exponent introduction and new gamma context}
-\todo{Add h spaces in formula}
+  Logically, a proof of $\bang A$ cannot use linear resources since
+  $\bang A$ denotes an unbounded (potentially $0$) number of copies of
+  $A$. Proofs of $\bang A$ may use other unrestricted or exponential
+  resources, tracked by context $\Gamma$.
+  From a computational perspective, the type $\bang A$
+  internalizes the simply-typed $\lambda$-calculus in the linear
+  $\lambda$-calculus.
+
+  The elimination form for the exponential, written $\vdash \llet{ !u
+    = M}{ N}$, warrants the use of the resource $A$ an unbounded
+  number of times in $N$ via the variable $u$:  
 \[
     \infer*[right=($\bang E$)]
-    {\Gamma ; \Delta_1 \vdash M : !A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
-    {\Gamma ; \Delta_1, \Delta_2 \vdash let !u = M \mathsf{in} N : C}
+    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
+    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
 \]
-\todo{Describe exponent elimination and new gamma context}
 
@@
 
-Type-based program synthesis is a studied field\todo{ dizer de uma forma melhor? }, and it follows
-that a lot of literature is available to inspire and complement our work.
-Most? follow the synthesis-as-proof-search pattern, however, the process is
-novel for each, due to a variety of different rich types explored, and their respective logics; or nuances done to the process, such
-as considering examples; or even the programming paradigm of the output produced (e.g. generating heap manipulating programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).
-Some other common patterns we might want to follow are the type refinements and the support for recursive functions.
-These works share some common ground, and base themselves of the same works we do -- most famously the Curry-Howard correspondence, but also,
-for example, works on the method of Focusing~\cite{focusing}.
+Type-based program synthesis is a vast field of study and so it
+follows that a lot of literature is available to inspire and
+complement our work. Most works~\cite{} follow some variation of the
+synthesis-as-proof-search approach, however, the process is novel for
+each, due to a variety of different rich types explored, and their
+respective corresponding logics and programming languages; or nuances
+of the synthesis process itself, such as complementing types with
+program examples; or even the programming paradigm of the output
+produced (e.g. generating heap manipulating
+programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).  Some other
+common patterns we might want to follow are the use of type
+refinements~\cite{} and the support for synthesis of recursive
+functions~\cite{}.  These works share some common ground, and base
+themselves of the same works we do -- most famously the Curry-Howard
+correspondence, but also, for example, works on focusing-based
+proof-search~\cite{focusing}.
 
-\mypara{Type-and-Example-Directed Program Synthesis}~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
-This paper explores a purely functional, recursive program synthesis technique based on types/proof search (as we intend to do),
-with examples as an auxiliar technique to trim down the program synthesis vast search space.
-\todo[inline]{more introduction to the paper}.
-A good use is made of the general strategy of using a type system
-to generate programs (by ``inverting'' the type system), rather than using it to type-check one.
-Moreover, the paper uses a data structure called ``refinement tree'' in conjunction
-with the extra information provided by the examples to allow for efficient synthesis
-of non-trivial programs. Overall it employs good engineering to achieve type-based synthesis
-of functional typed programs, with at least two additions that we might want to follow too
-(type refinements and recursive functions).
+\mypara{Type-and-Example-Directed Program Synthesis} This
+work~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
+explores a purely functional, recursive program synthesis technique
+based on types/proof search (as we intend to do), with examples as an
+auxiliar technique to trim down the program synthesis vast search
+space.  \todo[inline]{more introduction to the paper}.  A good use is
+made of the general strategy of using a type system to generate
+programs (by ``inverting'' the type system), rather than using it to
+type-check one.  Moreover, the paper uses a data structure called
+``refinement tree'' in conjunction with the extra information provided
+by the examples to allow for efficient synthesis of non-trivial
+programs. Overall it employs good engineering to achieve type-based
+synthesis of functional typed programs, with at least two additions
+that we might want to follow too (type refinements and recursive
+functions).
 

### 2021-05-09 article/article.tex
Polikarpova work

--- 2021-05-08 article/article.tex
@@
 $\lambda$-calculus and propositional intuitionistic logic. It extends to
-the realms of polymorphism (higher-order logic \todo{review - {https://en.wikipedia.org/wiki/Lambda_cube}} second-order propositional calculus),
+the realms of polymorphism (higher-order logic $\todo{review - {https://en.wikipedia.org/wiki/Lambda_cube}}$ second-order propositional calculus),
 dependent types (first-order predicate logic), and every other extension
@@
 as considering examples; or even the programming paradigm of the output produced (e.g. generating heap manipulating programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).
-Some other common patterns we might want to follow are the type refinements and the support for recursive functions.
+Some other common aspects we might want to follow are the type refinements and the support for recursive functions.
 These works share some common ground, and base themselves of the same works we do -- most famously the Curry-Howard correspondence, but also,
@@
 \mypara{Synthesis from (polymorphic) refinement types}~\cite{DBLP:conf/pldi/PolikarpovaKS16}
+This paper also attempts to synthetize recursive functional programs, in a more ``advanced'' setting. The specification
+is a combination of two rich types, polymorfic and refinement types (which correspond to a first-order logic through the Curry-Howard isomorphism) -- this is an interesting
+combo that offers ``expressive power and decidability, which enables automatic verification--and hence synthesis--of nontrivial programs''.
+Their approach to the refinement types is a new algorithm that supports decomposition of the refinement specification,
+and they separation between the language of specification and programs.
 

### 2021-05-09 article/article.tex

--- 2021-05-09 article/article.tex
@@
 $\lambda$-calculus and propositional intuitionistic logic. It extends
-to the realms of polymorphism (higher-order logic \todo{review -
-  \url{https://en.wikipedia.org/wiki/Lambda_cube}} second-order
+to the realms of polymorphism (higher-order logic $\todo{review -
+  \url{https://en.wikipedia.org/wiki/Lambda_cube}}$ second-order
 propositional calculus), dependent types (first-order predicate

### 2021-05-09 article/article.tex
fix compile error

@@
 $\lambda$-calculus and propositional intuitionistic logic. It extends
-to the realms of polymorphism (higher-order logic $\todo{review -
-  \url{https://en.wikipedia.org/wiki/Lambda_cube}}$ second-order
+to the realms of polymorphism (higher-order logic \todo{review -
+  \url{https://en.wikipedia.org/wiki/Lambda_cube}} second-order
 propositional calculus), dependent types (first-order predicate

### 2021-05-09 article/article.tex
related work graded model types

@@
 
-\mypara{Synthesis from (polymorphic) refinement types}~\cite{DBLP:conf/pldi/PolikarpovaKS16}
+\mypara{Program Synthesis from Polymorfic Refinement Types}~\cite{DBLP:conf/pldi/PolikarpovaKS16}
 This paper also attempts to synthetize recursive functional programs, in a more ``advanced'' setting. The specification
@@
 
-\mypara{Synthesis for graded types} (no recursion   -- in that work, no
-  obvious connection of grading with
-  concurrency)~\cite{DBLP:conf/lopstr/HughesO20}
+\mypara{Resourceful Program Synthesis from Graded Linear Types}~\cite{DBLP:conf/lopstr/HughesO20}
+This paper attempts to synthetize programs in a manner very similar to ours.
+It uses graded modal types, which is an ``augment'' to linear types -- a more \emph{fine-grained} version, since it allows for
+quantitative specification of resource usage, in contrast to ours either \emph{linear} or \emph{unrestricted} (``banged'' values) use of assumptions.
+Their resource management is more complex, and so they provide solutions which adapt Hodas and Miller's input-output model~\cite{} -- the model we'll be using.
+They also use focusing as a solution to trim down search space, and our solution will use the same literature as a base.
+However, since the logic behind the their types is \emph{modal}, instead of \emph{linear}, it doesn't allow to, in the future, explore
+the tight relation between linear logic and concurrency based (message passing) programs with session types (e.g. synthetize concurrent programs with message passing), and other fields such as
+resource analysis? -- which ours will.
+
+% (no recursion   -- in that work, no
+%   obvious connection of grading with
+%   concurrency)
 

### 2021-05-09 article/article.tex
tweaks

--- 2021-05-09 article/article.tex
@@
 deduction for intuitionistic logic~\cite{prawitznd65}. This
-relationship, known as the Curry-Howard correspondence~\cite{},
+relationship, known as the Curry-Howard
+correspondence~(see~\cite{DBLP:journals/cacm/Wadler15} for a
+historical survey),
 identifies that the propositions of intuitionistic logic can be read
@@
 (``proofs as programs''), and checking a proof is type checking a
-program (``proof checking as type checking'')~\cite{}.
+program (``proof checking as type checking'').
 
@@
 
-  The elimination form for the exponential, written $\vdash \llet{ !u
-    = M}{ N}$, warrants the use of the resource $A$ an unbounded
+  The elimination form for the exponential, written $\llet{ !u
+    = M}{ N}$, warrants the use of resource $A$ an unbounded
   number of times in $N$ via the variable $u$:  
@@
 
-\mypara{Synthesis from (polymorphic) refinement types}~\cite{DBLP:conf/pldi/PolikarpovaKS16}
-This paper also attempts to synthetize recursive functional programs, in a more ``advanced'' setting. The specification
-is a combination of two rich types, polymorfic and refinement types (which correspond to a first-order logic through the Curry-Howard isomorphism) -- this is an interesting
-combo that offers ``expressive power and decidability, which enables automatic verification--and hence synthesis--of nontrivial programs''.
-Their approach to the refinement types is a new algorithm that supports decomposition of the refinement specification,
-and they separation between the language of specification and programs.
+\mypara{Program Synthesis from Polymorfic Refinement Types} The
+work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis
+recursive functional programs but in a more ``advanced'' setting. The
+specification is a combination of two rich forms of types: polymorphic and
+refinement types (which correspond to a first-order logic through the
+Curry-Howard isomorphism) -- this is an interesting combination that offers
+``expressive power and decidability, which enables automatic
+verification--and hence synthesis--of nontrivial programs''.  Their
+approach to refinement types consists a new algorithm that supports
+decomposition of the refinement specification, allowing for separation
+between the language of specification and programs and making the
+approach amenable to compositional synthesis techniques.
 
-\mypara{Synthesis for graded types} (no recursion   -- in that work, no
-  obvious connection of grading with
-  concurrency)~\cite{DBLP:conf/lopstr/HughesO20}
+\mypara{Resourceful Program Synthesis from Graded Linear Types} The
+work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
+approach very similar to our own.  The work employs so-called graded
+modal types, which is an refinement of pure linear types -- a more
+\emph{fine-grained} version, since it allows for quantitative
+specification of resource usage, in contrast to ours either
+\emph{linear} or \emph{unrestricted} (via the linear logic
+exponential) use of assumptions.  Their resource management is more
+complex, and so they provide solutions which adapt Hodas and Miller's
+resource management approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09} -- the model we'll be using.
+
+They also use focusing as a solution to trim down search space and
+to ensure that synthesis only produces well-typed programs, and
+our solution will use the same literature as a base.  However, since
+their underlying logic is \emph{modal} rather than purely
+\emph{linear}, it lacks a clear correspondence with concurrent
+session-typed programs~\cite{}, which is a crucial avenue of future
+work. Moreover, their use grading effectively requires constraint solving to
+be integrated with the synthesis procedure, which can limit the
+effectiveness of the overall approach as one scales to more
+sophisticated settings (e.g.~refinement types).
+
+% (no recursion   -- in that work, no
+%   obvious connection of grading with
+%   concurrency)
 

### 2021-05-09 article/article.tex
init goals and work plan

--- 2021-05-09 article/article.tex
@@
 
-\section{Separate}
+% \section{Separate}
 
-More about the concepts than citing concrete works (but citations are
-needed!):
-\begin{enumerate}
-\item PL/Type systems in a ``formal'' sense (inference rules, etc) [DONE]
-\item Propositions as types: Proofs as programs, proof search as
-  synthesis, etc. [MOSTLY DONE]
-\item Linear logic: a ``resource aware logic'' [ONGOING]
-\item Proof search in LL: challenges, focusing as a ``solution'' [ONGOING]
-\item Linear $\lambda$-calculus: what does it look like.
-\item Type-based synthesis: ``inverting'' a type system.
-\end{enumerate}
+% More about the concepts than citing concrete works (but citations are
+% needed!):
+% \begin{enumerate}
+% \item PL/Type systems in a ``formal'' sense (inference rules, etc) [DONE]
+% \item Propositions as types: Proofs as programs, proof search as
+%   synthesis, etc. [MOSTLY DONE]
+% \item Linear logic: a ``resource aware logic'' [ONGOING]
+% \item Proof search in LL: challenges, focusing as a ``solution'' [ONGOING]
+% \item Linear $\lambda$-calculus: what does it look like.
+% \item Type-based synthesis: ``inverting'' a type system.
+% \end{enumerate}
 
@@
 
-Refinement types provide very precise specifications to assist the
-synthesis process.
+% Refinement types provide very precise specifications to assist the
+% synthesis process.
 
@@
 
-Expand on earlier points. Start from a functional language with linear
-types ($\tensor$, $\lolli$, $\oplus$, $\bang$); build on it with more
-``stuff''. General techniques, drawn from proof theory via props as
-types: explore focusing to tame non-determinism/search space. Partial
-synthesis is still proof search! Add other techniques as we go along
-(e.g. for recursion we need to constrain recursive calls).
+We start with the definition of our type system and simply-typed functional language
+with linear types ($\lolli$, $\tensor$, $1$, $\with$, $\oplus$, $\bang$, $Bool?$).
+Through the Curry-Howard correspondence, we'll formulate the synthesis of a
+linear type in our functional lanugage as proof search in linear logic.
+Since \emph{natural deduction} is ``natural'' in practice, and thus, harder to streamline and reason about,
+we'll use the equivalent \emph{sequent calculus} to formulate our inference rules.
+In sequent calculus, all our rules can be read in a \emph{bottom-up} manner, simplifying the reasoning for the synthetizer.
+Therafter, a way to reason about our resource contexts is needed, and the Hodas and Miller's? input-output model~\cite{} comes into play.
+According to it, instead of a single context that we must split into parts when proving the judgments
+above the inference rule line, we'll utilize an \emph{input} and an \emph{output} context -- to prove $A\tensor B$ with $\Delta$, prove $A$ with $\Delta$, consuming the needed resources, and output a $\Delta'$, prove $B$ with $\Delta'$ and the output the remaining resources in $\Delta''$, and finally, after having proven $A\tensor B$, output $\Delta''$, which might be used by another judgment.
+With the input/output contexts, and the sequent calculus formulation of the inference rules, we're on a good path to synthetize programs.
+However, even though proof-search here might seem direct, it still encompasses a lot of non-determinism, e.g., which rule to apply at a certain point?
+To this we add focusing~\cite{} -- a method created to trim down the search space of valid proofs in a linear logic, by conducting the
+proof-search first with invertible rules, and then by focusing where no more direct (invertible) rules can be applied.
+With focusing we tame some of non-determinism from the synthesis aspect. \todo{talvez algo mais sobre focusing aqui.}
+We might, later on, add refinement types also as a way to constrain the valid programs space.
+Additionally, we might need to discuss strategies to synthetize recursive functions as well.
 
-Mention that you've already implemented a type-checker for this
-(useful as a prelim. Exercise but also later, for \emph{validation}).
+Putting together everything above, and the more detailed information in the background section -- linear types, sequent calculus, resource management, and focusing,
+we have a set of inference rules (written out in the appendix) that define how the synthesis process will progress.
 
-Validation and evaluation: validation is as simple as ``does it typecheck''? can you synth? how fast?
+% Expand on earlier points. Start from a functional language with linear
+% types ($\tensor$, $\lolli$, $\oplus$, $\bang$); build on it with more
+% ``stuff''. General techniques, drawn from proof theory via props as
+% types: explore focusing to tame non-determinism/search space. Partial
+% synthesis is still proof search! Add other techniques as we go along
+% (e.g. for recursion we need to constrain recursive calls).
 
-\section{Bonus}
+% Mention that you've already implemented a type-checker for this
+% (useful as a prelim. Exercise but also later, for \emph{validation}).
 
-\[
-  \begin{array}{lcl}
-    A, B & ::= & A \tensor B \mid A \lolli B \mid A \with B \mid A
-                 \oplus B \mid \bang A \mid \one \mid \dots\\[1ex]
-    M,N & ::= & \lambda x{:}A.M\\
-         & \mid & M\,N\\
-         & \mid & (M \tensor N)\\
-         & \mid & \llet{x\tensor y = M}{N} \\
-         & \mid & x \\
-         & \mid & \dots\\
-    \end{array}
-\]
+% Validation and evaluation: validation is as simple as ``does it typecheck''? can you synth? how fast?
 
-\[
-  \infer[$(\lolli\! I)$]
-  {\Delta , x{:}A \vdash M : B}
-  {\Delta \vdash \lambda x {:} A . M : A \lolli B}
-  \quad
-  \infer[$(\lolli\! E)$]
-  {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
-  {\Delta_1, \Delta_2 \vdash M\,N : B}
-\]
+% \section{Bonus}
 
-\[
-  \infer*[left=($\tensor I$)]
-  {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
-  {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
-  \quad
-  \infer*[right=($\tensor E$)]
-  {\Delta_1 \vdash M : A \tensor B \and \Delta_2 , x{:}A, y{:}B\vdash
-    N : C }
-  {\Delta_1 , \Delta_2\vdash \llet{x\tensor y = M}{N} : C }
-\]
+% \[
+%   \begin{array}{lcl}
+%     A, B & ::= & A \tensor B \mid A \lolli B \mid A \with B \mid A
+%                  \oplus B \mid \bang A \mid \one \mid \dots\\[1ex]
+%     M,N & ::= & \lambda x{:}A.M\\
+%          & \mid & M\,N\\
+%          & \mid & (M \tensor N)\\
+%          & \mid & \llet{x\tensor y = M}{N} \\
+%          & \mid & x \\
+%          & \mid & \dots\\
+%     \end{array}
+% \]
+
+% \[
+%   \infer[$(\lolli\! I)$]
+%   {\Delta , x{:}A \vdash M : B}
+%   {\Delta \vdash \lambda x {:} A . M : A \lolli B}
+%   \quad
+%   \infer[$(\lolli\! E)$]
+%   {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
+%   {\Delta_1, \Delta_2 \vdash M\,N : B}
+% \]
+
+% \[
+%   \infer*[left=($\tensor I$)]
+%   {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
+%   {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
+%   \quad
+%   \infer*[right=($\tensor E$)]
+%   {\Delta_1 \vdash M : A \tensor B \and \Delta_2 , x{:}A, y{:}B\vdash
+%     N : C }
+%   {\Delta_1 , \Delta_2\vdash \llet{x\tensor y = M}{N} : C }
+% \]
 

### 2021-05-09 article/article.tex

@@
 
-\mypara{Program Synthesis from Polymorfic Refinement Types}~\cite{DBLP:conf/pldi/PolikarpovaKS16}
-This paper also attempts to synthetize recursive functional programs, in a more ``advanced'' setting. The specification
-is a combination of two rich types, polymorfic and refinement types (which correspond to a first-order logic through the Curry-Howard isomorphism) -- this is an interesting
-combo that offers ``expressive power and decidability, which enables automatic verification--and hence synthesis--of nontrivial programs''.
-Their approach to the refinement types is a new algorithm that supports decomposition of the refinement specification,
-and they separation between the language of specification and programs.
+\mypara{Program Synthesis from Polymorfic Refinement Types} The
+work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis
+recursive functional programs but in a more ``advanced'' setting. The
+specification is a combination of two rich forms of types: polymorphic and
+refinement types (which correspond to a first-order logic through the
+Curry-Howard isomorphism) -- this is an interesting combination that offers
+``expressive power and decidability, which enables automatic
+verification--and hence synthesis--of nontrivial programs''.  Their
+approach to refinement types consists a new algorithm that supports
+decomposition of the refinement specification, allowing for separation
+between the language of specification and programs and making the
+approach amenable to compositional synthesis techniques.
 
-\mypara{Resourceful Program Synthesis from Graded Linear Types}~\cite{DBLP:conf/lopstr/HughesO20}
-This paper attempts to synthetize programs in a manner very similar to ours.
-It uses graded modal types, which is an ``augment'' to linear types -- a more \emph{fine-grained} version, since it allows for
-quantitative specification of resource usage, in contrast to ours either \emph{linear} or \emph{unrestricted} (``banged'' values) use of assumptions.
-Their resource management is more complex, and so they provide solutions which adapt Hodas and Miller's input-output model~\cite{} -- the model we'll be using.
-They also use focusing as a solution to trim down search space, and our solution will use the same literature as a base.
-However, since the logic behind the their types is \emph{modal}, instead of \emph{linear}, it doesn't allow to, in the future, explore
-the tight relation between linear logic and concurrency based (message passing) programs with session types (e.g. synthetize concurrent programs with message passing), and other fields such as
-resource analysis? -- which ours will.
+\mypara{Resourceful Program Synthesis from Graded Linear Types} The
+work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
+approach very similar to our own.  The work employs so-called graded
+modal types, which is an refinement of pure linear types -- a more
+\emph{fine-grained} version, since it allows for quantitative
+specification of resource usage, in contrast to ours either
+\emph{linear} or \emph{unrestricted} (via the linear logic
+exponential) use of assumptions.  Their resource management is more
+complex, and so they provide solutions which adapt Hodas and Miller's
+resource management approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09} -- the model we'll be using.
+
+They also use focusing as a solution to trim down search space and
+to ensure that synthesis only produces well-typed programs, and
+our solution will use the same literature as a base.  However, since
+their underlying logic is \emph{modal} rather than purely
+\emph{linear}, it lacks a clear correspondence with concurrent
+session-typed programs~\cite{}, which is a crucial avenue of future
+work. Moreover, their use grading effectively requires constraint solving to
+be integrated with the synthesis procedure, which can limit the
+effectiveness of the overall approach as one scales to more
+sophisticated settings (e.g.~refinement types).
 

### 2021-05-09 article/article.tex
more goals

@@
 
+The full and partial synthesis mentioned are closely related, in which the latter is also a complete synthesis, but inserted inside another, bigger, program that makes use the synthetized function.
+
+\mypara{Work Plan}
+The end goal is to have a synthetizer capable of taking a type (possibly refined) as input,
+and to produce (possibly interactively) a function or more (possibly recursive) with that same type.
+Besides the synth, we've already created a parser and a type-checker -- more than good preparation for the
+development of the synthetizer, they will be useful to \emph{validate} the functions synthetized, and to implement the
+synthesis of partial programs. The implementations are in Haskell, a functional language, since it's easier to reason about inference rules that way.
+
 % Expand on earlier points. Start from a functional language with linear

### 2021-05-09 article/article.tex
some evaluation

@@
 
-The full and partial synthesis mentioned are closely related, in which the latter is also a complete synthesis, but inserted inside another, bigger, program that makes use the synthetized function.
+The full and partial synthesis mentioned are closely related, in which the latter is also a complete synthesis, but inserted inside another, bigger, program that makes use pf the synthetized function.
 
@@
 synthesis of partial programs. The implementations are in Haskell, a functional language, since it's easier to reason about inference rules that way.
+
+\mypara{Evaluation}
+Validation and evaluation of our work is fairly simple. Simple questions such as \emph{does it typecheck}? \emph{can you synth}? \emph{how fast}?
+are the core of our evaluation. In more concrete terms, we can compare our times to state of the art synthetizers, present a list of working examples
+of programs synthesed by our synth, and according benchmarks. On the eventual addition of recursive functions, refinement types, and interactive synthesis,
+we'll consider, respectively, other examples to showcase recursive functions synthesis, updated times for benchmarks with refinement types, and a display of
+the possible user interaction in the user process, and the resulting function following it.
 

### 2021-05-09 article/article.tex
touch ups

--- 2021-05-09 article/article.tex
@@
 
-\section{Background}
+\section{Background}\label{sec:background}
 
@@
 complex, and so they provide solutions which adapt Hodas and Miller's
-resource management approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09} -- the model we'll be using.
+resource management
+approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
+-- the model we will be using.
 
-They also use focusing as a solution to trim down search space and
-to ensure that synthesis only produces well-typed programs, and
-our solution will use the same literature as a base.  However, since
-their underlying logic is \emph{modal} rather than purely
-\emph{linear}, it lacks a clear correspondence with concurrent
-session-typed programs~\cite{}, which is a crucial avenue of future
-work. Moreover, their use grading effectively requires constraint solving to
-be integrated with the synthesis procedure, which can limit the
-effectiveness of the overall approach as one scales to more
-sophisticated settings (e.g.~refinement types).
+They also use focusing as a solution to trim down search space and to
+ensure that synthesis only produces well-typed programs, and our
+solution will use the same literature as a base.  However, since their
+underlying logic is \emph{modal} rather than purely \emph{linear}, it
+lacks a clear correspondence with concurrent session-typed
+programs~\cite{DBLP:journals/mscs/CairesPT16,DBLP:conf/concur/CairesP10},
+which is a crucial avenue of future work. Moreover, their use grading
+effectively requires constraint solving to be integrated with the
+synthesis procedure, which can limit the effectiveness of the overall
+approach as one scales to more sophisticated settings (e.g.~refinement
+types).
 
@@
 
-We start with the definition of our type system and simply-typed functional language
-with linear types ($\lolli$, $\tensor$, $1$, $\with$, $\oplus$, $\bang$, $Bool?$).
-Through the Curry-Howard correspondence, we'll formulate the synthesis of a
-linear type in our functional lanugage as proof search in linear logic.
-Since \emph{natural deduction} is ``natural'' in practice, and thus, harder to streamline and reason about,
-we'll use the equivalent \emph{sequent calculus} to formulate our inference rules.
-In sequent calculus, all our rules can be read in a \emph{bottom-up} manner, simplifying the reasoning for the synthetizer.
-Therafter, a way to reason about our resource contexts is needed, and the Hodas and Miller's? input-output model~\cite{} comes into play.
-According to it, instead of a single context that we must split into parts when proving the judgments
-above the inference rule line, we'll utilize an \emph{input} and an \emph{output} context -- to prove $A\tensor B$ with $\Delta$, prove $A$ with $\Delta$, consuming the needed resources, and output a $\Delta'$, prove $B$ with $\Delta'$ and the output the remaining resources in $\Delta''$, and finally, after having proven $A\tensor B$, output $\Delta''$, which might be used by another judgment.
-With the input/output contexts, and the sequent calculus formulation of the inference rules, we're on a good path to synthetize programs.
-However, even though proof-search here might seem direct, it still encompasses a lot of non-determinism, e.g., which rule to apply at a certain point?
-To this we add focusing~\cite{} -- a method created to trim down the search space of valid proofs in a linear logic, by conducting the
-proof-search first with invertible rules, and then by focusing where no more direct (invertible) rules can be applied.
-With focusing we tame some of non-determinism from the synthesis aspect. \todo{talvez algo mais sobre focusing aqui.}
-We might, later on, add refinement types also as a way to constrain the valid programs space.
-Additionally, we might need to discuss strategies to synthetize recursive functions as well.
+We start with the definition of our type system and simply-typed
+functional language with linear function, sum and product types
+($\lolli$, $\tensor$, $\one$, $\with$, $\oplus$, $\bang$).  Through
+the Curry-Howard correspondence, we'll formulate the synthesis of
+typed programs in our functional language as bottom-up proof-search in
+linear logic.
+
+Since proof rules in \emph{natural deduction} do not entail a
+proof-search strategy, we will use the equivalent \emph{sequent
+  calculus} system to formulate our typing rules in a way that is
+amenable to proof search.
+%
+In sequent calculus, rules can be naturally understood in a
+\emph{bottom-up} manner, simplifying the reasoning for the
+synthesizer. Moreover, the linear nature of our system requires a way
+to algorithmically reason about resource contexts: for instance, if we
+read rule $\tensor I$ from Section~\ref{sec:background} bottom-up, we
+must non-deterministically guess how to correctly split ambient
+resources into $\Delta_1$ and $\Delta_2$ to separately prove $A$ and
+$B$. To solve this issue we will adopt the resource-management
+approach
+of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09}.
+Instead of requiring non-deterministic guesses of resource splits
+during proof search, we will make use of an \emph{input} and an
+\emph{output} context, leading to the following general strategy: to
+prove $A\tensor B$ using resources $\Delta$, prove $A$ with $\Delta$,
+consuming some subset of $\Delta$, and produce as output the leftover
+resources $\Delta'$; prove $B$ using $\Delta'$ and then output the
+remaining resources $\Delta''$; finally, after having proven
+$A\tensor B$, output $\Delta''$, for subsequent derivations.
+
+With the use of input/output contexts, and the sequent calculus
+formulation of the inference rules, we draw closer to an effective
+strategy for bottom-up proof search and so of program synthesis.
+%
+
+However, the search process still encompasses significant
+non-determinism, e.g., at any given point, many proof rules are
+applicable in general. To address this we draw on the technique of
+focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} --
+a method created to trim down the search space of valid proofs in
+linear logic, by eagerly applying invertible rules (i.e.~rules whose
+conclusion implies the premises), and then by ``focusing'' on a single
+connective when no more direct (invertible) rules can be applied. If
+the search is not successful, the procedure backtracks and another
+connective is chosen as the focus.
+%
+Focusing eliminates all of the
+``don't care'' non-determinism from proof search, since the order in
+which invertible rules are applied does not affect the outcome of the
+search, leaving only the non-determinism that pertains to unknowns (or
+``don't know'' non-determinism), identifying precisely the points at
+which backtracking search is necessary.
+
+With the basic framework in place, we will consider additional
+(simply-typed) language constructs and synthesis of program fragments
+or \emph{partial synthesis} -- given a program with a typed hole,
+synthesize a term to fill out the hole accordingly. We will
+subsequently also consider synthesis of recursive functions and,
+might, later on, add refinement types also as a way to constrain the
+valid programs space.
 

### 2021-05-09 article/article.tex
finish background

@@
 development process. Type-based synthesis leverages rich types as a
-way of pruning the search space, and by using types gives the user as
-a more ``familiar'' specification. For instance, the type
+way of pruning the search space, and by using types the user 
+is given a more ``familiar'' specification. For instance, the type
 $\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
@@
 $\lambda$-calculus and propositional intuitionistic logic. It extends
-to the realms of polymorphism (higher-order logic \todo{review -
-  \url{https://en.wikipedia.org/wiki/Lambda_cube}} second-order
-propositional calculus), dependent types (first-order predicate
+to the realms of polymorphism (second-order
+logic), dependent types (first-order
 logic), and various other extensions of natural deduction and
@@
 \]
+Again, through the Curry-Howard correspondence, we can view the process
+of finding a proof of a proposition in this linear logic as the process of synthesizing a
+linear functional program of a given type.
 
-\todo{Through the Curry-Howard correspondence, we can view the process
-  of finding a proof of a proposition as the process of synthesizing a
-program of a given type.}
+\mypara{Resource-Management}
+Coming from linear logic, we now have a new problem to solve:
+When searching for a proof, how do we break the
+contexts in the necessary way to allocate the correct resources
+to prove each judgment the judgment we're proving depends on?
+This is where the idea of \emph{input/output} contexts come in.
+To solve this issue we will adopt the resource-management
+approach
+of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09}.
+Instead of requiring non-deterministic guesses of resource splits
+during proof search, we will make use of an \emph{input} and an
+\emph{output} context, leading to the following general strategy: to
+prove $A\tensor B$ using resources $\Delta$, prove $A$ with $\Delta$,
+consuming some subset of $\Delta$, and produce as output the leftover
+resources $\Delta'$; prove $B$ using $\Delta'$ and then output the
+remaining resources $\Delta''$; finally, after having proven
+$A\tensor B$, output $\Delta''$, for subsequent derivations.
+
+\mypara{Sequent Calculus}
+Inference rules in natural deduction (the ones we have considered so far)
+are complicated to use in a proof-search strategy.
+A more suited candidate is the equivalent \emph{sequent calculus} system
+in which \emph{all} inference rules are read naturally in a \emph{bottom-up} manner.
+The introduction and elimination rules from natural deduction disappear,
+and its place is taken by right and left rules, respectively.
+Right rules and introduction rules are almost identical, since introduction rules were
+already read bottom-up. The inference rule for implication introduction ($\supset$ is used instead of $\rightarrow$), seen above in natural deduction under the
+\emph{propositions as types} subsection, is the following right rule in sequent calculus:
+\[
+    \infer*[right=($\supset R$)]
+    {\Gamma, u: A \vdash M : B}
+    {\Gamma \vdash \lambda u. M : A \supset B}
+\]
+Left rules try to capture the same idea as elimination rules in natural deduction, but are altered
+to reflect it bottom-up, instead of top-down (usual ``orientation'' of elimination rules).
+The inference rule for the also seen above implication elimination, is the following left rule here:
+\[
+    \infer*[right=($\supset L$)]
+    {\Gamma , u: A \supset B \vdash M : A \supset B \and \Gamma \vdash N : A}
+    {\Gamma \vdash M\,N : B}
+\]
+We'll define our inference rules for synthesis under the sequent calculus system,
+for it simplifies the work to be done in the synthetizer.
+
+\mypara{Focusing}
+Even with everything mentioned so far, non-determinism still pertains proof-search in logic, e,g., at any given point, many proof rules are applicable in general.
+Focusing has been previously studied as a way to conduct proof search in linear logic~\cite{}
+--
+a method created to trim down the search space of valid proofs in
+linear logic, by eagerly applying invertible rules (i.e.~rules whose
+conclusion implies the premises), and then by ``focusing'' on a single
+connective when no more direct (invertible) rules can be applied. If
+the search is not successful, the procedure backtracks and another
+connective is chosen as the focus.
+%
+Focusing eliminates all of the
+``don't care'' non-determinism from proof search, since the order in
+which invertible rules are applied does not affect the outcome of the
+search, leaving only the non-determinism that pertains to unknowns (or
+``don't know'' non-determinism), identifying precisely the points at
+which backtracking search is necessary.
 
@@
 \section{Related Work}
-
-% \todo[inline]{Os~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
-%   como ``estrategia geral'' de ``ler'' um sistema de tipos +
-%   engenharia como um sistema de sintese de programas tipificados;
-%   Aqui~\cite{DBLP:conf/pldi/PolikarpovaKS16} estamos num contexto
-%   ``mais avançado'' (polimorfismo e tipos refinados -- ``logica de
-%   primeira ordem'') -- tipos ricos -- trabalho aspiracional,
-%   tratamento de funcao recursivas, tratamento de tipos de dados
-%   indutivos/algebricos; Aqui~\cite{DBLP:conf/lopstr/HughesO20},
-%   sintese para graded modal types, tem a ver com ``linearidade'' kind
-%   of, abordagem baseado em focusing (portanto Curry-Howard++), no caso
-%   deles acaba por ser uma lógica ``modal'' vs linear -- não permite
-%   futuramente explorar relações com e.g. concorrência, análise de
-%   recursos.}
 
@@
 
-% (no recursion   -- in that work, no
-%   obvious connection of grading with
-%   concurrency)
-
-% Refinement types provide very precise specifications to assist the
-% synthesis process.
-
-% \item Synthesis of heap-manipulating
-%   programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}
-
-% Less related, but more evidence of this idea of turning ``checking''
-% systems into synthesis frameworks.
-
 % \item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}
-
 % This resource-guided means something a bit different. Programs satisfy
@@
 resources into $\Delta_1$ and $\Delta_2$ to separately prove $A$ and
-$B$. To solve this issue we will adopt the resource-management
-approach
-of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09}.
-Instead of requiring non-deterministic guesses of resource splits
-during proof search, we will make use of an \emph{input} and an
-\emph{output} context, leading to the following general strategy: to
-prove $A\tensor B$ using resources $\Delta$, prove $A$ with $\Delta$,
-consuming some subset of $\Delta$, and produce as output the leftover
-resources $\Delta'$; prove $B$ using $\Delta'$ and then output the
-remaining resources $\Delta''$; finally, after having proven
-$A\tensor B$, output $\Delta''$, for subsequent derivations.
+$B$. The resource-management approach we'll adopt to solve this has been 
+previously explained in more detail.
 
@@
 However, the search process still encompasses significant
-non-determinism, e.g., at any given point, many proof rules are
-applicable in general. To address this we draw on the technique of
-focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} --
-a method created to trim down the search space of valid proofs in
-linear logic, by eagerly applying invertible rules (i.e.~rules whose
-conclusion implies the premises), and then by ``focusing'' on a single
-connective when no more direct (invertible) rules can be applied. If
-the search is not successful, the procedure backtracks and another
-connective is chosen as the focus.
-%
-Focusing eliminates all of the
-``don't care'' non-determinism from proof search, since the order in
-which invertible rules are applied does not affect the outcome of the
-search, leaving only the non-determinism that pertains to unknowns (or
-``don't know'' non-determinism), identifying precisely the points at
-which backtracking search is necessary.
+non-determinism. To address this we draw on the technique of
+focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as seen before.
 
@@
 
-The full and partial synthesis mentioned are closely related, in which the latter is also a complete synthesis, but inserted inside another, bigger, program that makes use pf the synthetized function.
-
 \mypara{Work Plan}
@@
 and to produce (possibly interactively) a function or more (possibly recursive) with that same type.
-Besides the synth, we've already created a parser and a type-checker -- more than good preparation for the
+We've already created a parser and a type-checker -- more than good preparation for the
 development of the synthetizer, they will be useful to \emph{validate} the functions synthetized, and to implement the

### 2021-05-09 article/article.tex
touch up

@@
 Again, through the Curry-Howard correspondence, we can view the process
-of finding a proof of a proposition in this linear logic as the process of synthesizing a
-linear functional program of a given type.
+of finding a proof of a proposition in linear logic as the process of synthesizing a
+linear functional program of the given type.
 
 \mypara{Resource-Management}
-Coming from linear logic, we now have a new problem to solve:
-When searching for a proof, how do we break the
-contexts in the necessary way to allocate the correct resources
-to prove each judgment the judgment we're proving depends on?
-This is where the idea of \emph{input/output} contexts come in.
+Be it from the perspective of proof \emph{checking} (i.e.~type checking) or
+proof \emph{search} linear logic poses a key challenge when compared
+to the non-linear setting:
+When constructing a derivation (bottom-up), we are seemingly forced to
+guess how to correctly split the linear context such that the
+sub-derivations have access to the correct resources (e.g.~the
+$\tensor I$ rule above).
+%
+%
+
 To solve this issue we will adopt the resource-management
 approach
-of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09}.
+of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09},
+which generalizes the judgment  from $\Delta \vdash M : A$ to
+$\Delta_I \ \Delta_O \vdash M : A$, where $\Delta_I$ is an input
+context and $\Delta_O$ is an output context.
 Instead of requiring non-deterministic guesses of resource splits
-during proof search, we will make use of an \emph{input} and an
-\emph{output} context, leading to the following general strategy: to
-prove $A\tensor B$ using resources $\Delta$, prove $A$ with $\Delta$,
+during proof search, we track which resources are used and which are
+remaining via the two contexts, leading to the following general strategy: to
+prove $A\tensor B$ using (input) resources $\Delta$, prove $A$ with
+input context $\Delta$,
 consuming some subset of $\Delta$, and produce as output the leftover
-resources $\Delta'$; prove $B$ using $\Delta'$ and then output the
+resources $\Delta'$; prove $B$ using $\Delta'$ as its input context and then output the
 remaining resources $\Delta''$; finally, after having proven
@@
 
-\mypara{Sequent Calculus}
-Inference rules in natural deduction (the ones we have considered so far)
-are complicated to use in a proof-search strategy.
-A more suited candidate is the equivalent \emph{sequent calculus} system
-in which \emph{all} inference rules are read naturally in a \emph{bottom-up} manner.
-The introduction and elimination rules from natural deduction disappear,
-and its place is taken by right and left rules, respectively.
-Right rules and introduction rules are almost identical, since introduction rules were
-already read bottom-up. The inference rule for implication introduction ($\supset$ is used instead of $\rightarrow$), seen above in natural deduction under the
-\emph{propositions as types} subsection, is the following right rule in sequent calculus:
+\mypara{Sequent Calculus} Inference rules in natural deduction (the
+ones we have considered so far) are ill-suited for bottom-up
+proof-search since elimination rules work top-down and introduction
+rules work bottom-up.  A more suited candidate is the equivalent
+\emph{sequent calculus} system in which \emph{all} inference rules can
+be understood naturally in a \emph{bottom-up} manner.  The
+introduction and elimination rules from natural deduction disappear,
+and their place is taken by right and left rules, respectively.
+%
+Right rules correspond exactly to the introduction rules of natural
+deduction, which were already understood bottom-up. The inference rule
+for implication introduction ($\supset$ is used instead of
+$\rightarrow$), seen above in natural deduction under the
+\emph{propositions as types} subsection, corresponds to the following
+right rule in sequent calculus:
 \[
@@
   \]
-Left rules try to capture the same idea as elimination rules in natural deduction, but are altered
-to reflect it bottom-up, instead of top-down (usual ``orientation'' of elimination rules).
-The inference rule for the also seen above implication elimination, is the following left rule here:
+
+Intuitively, left rules act as the elimination rules of natural deduction, but are altered
+to work bottom-up, instead of top-down.
+The inference rule for the also seen above implication elimination, is
+as follows:
 \[
     \infer*[right=($\supset L$)]
-    {\Gamma , u: A \supset B \vdash M : A \supset B \and \Gamma \vdash N : A}
-    {\Gamma \vdash M\,N : B}
+    {\Gamma , u:B\vdash M : C \and \Gamma, u{:}A\supset B \vdash N : A}
+    {\Gamma, u{:}A\supset B\vdash M\{(u\,N)/u\} : C}
   \]
-We'll define our inference rules for synthesis under the sequent calculus system,
-for it simplifies the work to be done in the synthetizer.
+  As implied by their name, left rules define how to decompose or make
+  use of a connective on the left of the turnstile $\vdash$. To use an
+  assumption of $A\supset B$ while attempting to show some proposition
+  $C$ we produce a proof of $A$, which allows us to use an assumption
+  of $B$ to prove $C$. In terms of the corresponding $\lambda$-terms,
+  the $\supset L$ rule corresponds to applying the variable $u$ to the
+  argument $N$ but potentially deep in the structure of $M$.
+
+  We'll define our inference rules for synthesis under the sequent
+  calculus system, for it simplifies the work to be done in the
+  synthesizer.
 
   \mypara{Focusing}
-Even with everything mentioned so far, non-determinism still pertains proof-search in logic, e,g., at any given point, many proof rules are applicable in general.
-Focusing has been previously studied as a way to conduct proof search in linear logic~\cite{}
---
-a method created to trim down the search space of valid proofs in
-linear logic, by eagerly applying invertible rules (i.e.~rules whose
-conclusion implies the premises), and then by ``focusing'' on a single
-connective when no more direct (invertible) rules can be applied. If
-the search is not successful, the procedure backtracks and another
-connective is chosen as the focus.
-%
-Focusing eliminates all of the
-``don't care'' non-determinism from proof search, since the order in
-which invertible rules are applied does not affect the outcome of the
-search, leaving only the non-determinism that pertains to unknowns (or
-``don't know'' non-determinism), identifying precisely the points at
-which backtracking search is necessary.
+
+  Even with everything mentioned so far, non-determinism is still
+  present proof-search in logic, e,g., at any given point, many proof
+  rules are applicable in general. The technique of focusing~\cite{}
+  has been previously studied as a way to discipline proof search in
+  linear logic-- a method created to trim down the search space of
+  valid proofs in linear logic, by eagerly applying invertible rules
+  (i.e.~rules whose conclusion implies the premises), and then by
+  ``focusing'' on a single connective when no more direct (invertible)
+  rules can be applied, that is, only applying rules that breakdown
+  the connective under focus or its subformulas. If the search is not
+  successful, the procedure backtracks and another connective is
+  chosen as the focus.
+
+  Focusing eliminates all of the ``don't care'' non-determinism from
+  proof search, since the order in which invertible rules are applied
+  does not affect the outcome of the search, leaving only the
+  non-determinism that pertains to unknowns (or ``don't know''
+  non-determinism), identifying precisely the points at which
+  backtracking search is necessary.
 
@@
 %
-
+%
 However, the search process still encompasses significant
 non-determinism. To address this we draw on the technique of
-focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as seen before.
+focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as
+discussed above.
 
@@
 
-Putting together everything above, and the more detailed information in the background section -- linear types, sequent calculus, resource management, and focusing,
-we have a set of inference rules (written out in the appendix) that define how the synthesis process will progress.
+Putting together everything above, and the more detailed information
+in the background section -- linear types, sequent calculus, resource
+management, and focusing, we have a set of inference rules (written
+out in the appendix) that define how the synthesis process will
+progress.
 
-\mypara{Work Plan}
-The end goal is to have a synthetizer capable of taking a type (possibly refined) as input,
-and to produce (possibly interactively) a function or more (possibly recursive) with that same type.
-We've already created a parser and a type-checker -- more than good preparation for the
-development of the synthetizer, they will be useful to \emph{validate} the functions synthetized, and to implement the
-synthesis of partial programs. The implementations are in Haskell, a functional language, since it's easier to reason about inference rules that way.
+\subsection{Work Plan}
 
-\mypara{Evaluation}
-Validation and evaluation of our work is fairly simple. Simple questions such as \emph{does it typecheck}? \emph{can you synth}? \emph{how fast}?
-are the core of our evaluation. In more concrete terms, we can compare our times to state of the art synthetizers, present a list of working examples
-of programs synthesed by our synth, and according benchmarks. On the eventual addition of recursive functions, refinement types, and interactive synthesis,
-we'll consider, respectively, other examples to showcase recursive functions synthesis, updated times for benchmarks with refinement types, and a display of
-the possible user interaction in the user process, and the resulting function following it.
+The end goal is to have a synthetizer capable of
+taking a type (possibly refined) as input, and to produce (possibly
+interactively) one or more functions with that type, if any exist.
+
+As mentioned above, we will begin with (both partial and full)
+synthesis for the linear $\lambda$-calculus. We will then extend the
+language with additional type constructs (e.g. base types, records,
+algebraic data types) and consider synthesis in that setting. Finally,
+we will consider synthesis of recursive functions, and,
+time-permitting, we will consider simple forms of type refinements
+(e.g. arithmetic refinements).
+
+We've already implemented a parser and a type-checker for a linear
+$\lambda$-calculus using the resource management technique described
+in previous section -- more than
+good preparation for the development of the synthesizer, they will be
+useful to \emph{validate} the synthesized functions, and to enable
+the synthesis of partial programs.
+%
+The implementations have been carried out in Haskell since it's easier
+to model and implement inference rules in a functional setting.
+
+\mypara{Evaluation} Validation and evaluation of our work is fairly
+simple. Simple questions such as \emph{does it typecheck}? \emph{can
+  we synthesize}? \emph{how fast}?  are the core of our evaluation. In
+more concrete terms, we can compare our times to state of the art
+synthetizers, present a list of working examples of programs synthesized
+by our framework, and corresponding benchmarks.
+
+On the eventual addition of recursive functions, refinement types, and
+interactive synthesis, we'll consider, respectively, other examples to
+showcase recursive functions synthesis, updated times for benchmarks
+with refinement types, and a display of the possible user interaction
+in the user process, and the resulting function following it.
 

### 2021-05-09 article/article.tex
typo

@@
 
-\mypara{Program Synthesis from Polymorfic Refinement Types} The
+\mypara{Program Synthesis from Polymorphic Refinement Types} The
 work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis

### 2021-05-09 article/article.tex
tweaks

--- 2021-05-09 article/article.tex
@@
 
-% \section{Separate}
-
-% More about the concepts than citing concrete works (but citations are
-% needed!):
-% \begin{enumerate}
-% \item PL/Type systems in a ``formal'' sense (inference rules, etc) [DONE]
-% \item Propositions as types: Proofs as programs, proof search as
-%   synthesis, etc. [MOSTLY DONE]
-% \item Linear logic: a ``resource aware logic'' [ONGOING]
-% \item Proof search in LL: challenges, focusing as a ``solution'' [ONGOING]
-% \item Linear $\lambda$-calculus: what does it look like.
-% \item Type-based synthesis: ``inverting'' a type system.
-% \end{enumerate}
-
 \section{Related Work}
@@
 The end goal is to have a synthetizer capable of taking a type (possibly refined) as input,
-and to produce (possibly interactively) a function or more (possibly recursive) with that same type.
+that produces (possibly interactively) a function or more (possibly recursive) with that same type.
 We've already created a parser and a type-checker -- more than good preparation for the
@@
 synthesis of partial programs. The implementations are in Haskell, a functional language, since it's easier to reason about inference rules that way.
+Our work began
 

### 2021-05-09 article/article.tex

--- 2021-05-09 article/article.tex
@@
 
+This is the theory we're interested in and have studied so far,
+which will allow us to build the actual synthetizer. From here,
+our work plan gives an overview of the steps we'll take.
+
 \subsection{Work Plan}
@@
 $\lambda$-calculus using the resource management technique described
-in previous section -- more than
+in the previous section -- more than
 good preparation for the development of the synthesizer, they will be
@@
 
+\mypara{Step by step} Our work so far has been constantly accompanied
+by somewhat dense theoretical material, state of the art,
+which we must understand to overcome challenges in these fields,
+and build from the start a more scalable synthetizer -- we want
+to display concrete results to prove the feasability of synthetising
+linear types, and thus possibility for future application to related fields,
+such as session types for message passing concurrency.
+
+Apart from the theory, when we started studying type systems, we created
+a simple type-checker for the simply-typed $\lambda$-calculus.
+While learning about linear type systems, and the simply-typed linear
+$\lambda$-calculus, we developed a parser, a desugar module, and an
+evaluation module to write simple programs in the language;
+culminating in writing the type-checker for linear types. At this point,
+where we stand as of now, we can parse, type-check, and evaluate this
+small linear functional language.
+The next step is to start developing the actual synthetizer, embedding it
+with all the techniques we've seen. Following work will be done to
+augment usability, expand the language domain, and add features such as the
+ones mentioned above.
+
 \mypara{Evaluation} Validation and evaluation of our work is fairly
-simple. Simple questions such as \emph{does it typecheck}? \emph{can
+easy. Simple questions such as \emph{does it typecheck}? \emph{can
   we synthesize}? \emph{how fast}?  are the core of our evaluation. In
@@
 
-% \section{Bonus}
-
-% \[
-%   \begin{array}{lcl}
-%     A, B & ::= & A \tensor B \mid A \lolli B \mid A \with B \mid A
-%                  \oplus B \mid \bang A \mid \one \mid \dots\\[1ex]
-%     M,N & ::= & \lambda x{:}A.M\\
-%          & \mid & M\,N\\
-%          & \mid & (M \tensor N)\\
-%          & \mid & \llet{x\tensor y = M}{N} \\
-%          & \mid & x \\
-%          & \mid & \dots\\
-%     \end{array}
-% \]
-
-% \[
-%   \infer[$(\lolli\! I)$]
-%   {\Delta , x{:}A \vdash M : B}
-%   {\Delta \vdash \lambda x {:} A . M : A \lolli B}
-%   \quad
-%   \infer[$(\lolli\! E)$]
-%   {\Delta_1 \vdash M : A \lolli B \and \Delta_2 \vdash N : A}
-%   {\Delta_1, \Delta_2 \vdash M\,N : B}
-% \]
-
-% \[
-%   \infer*[left=($\tensor I$)]
-%   {\Delta_1 \vdash M : A \and \Delta_2 \vdash N : B}
-%   {\Delta_1 , \Delta_2 \vdash (M \tensor N) : A \tensor B}
-%   \quad
-%   \infer*[right=($\tensor E$)]
-%   {\Delta_1 \vdash M : A \tensor B \and \Delta_2 , x{:}A, y{:}B\vdash
-%     N : C }
-%   {\Delta_1 , \Delta_2\vdash \llet{x\tensor y = M}{N} : C }
-% \]
-
 \bibliographystyle{splncs04}

### 2021-05-09 article/article.tex
add citations

@@
 \usepackage{hyperref}
-\usepackage{todonotes}
 
@@
 synthesis frameworks use rich type systems such as refinement
-types~\cite{} and polymorphic types~\cite{}.
+types~\cite{DBLP:conf/pldi/PolikarpovaKS16} and polymorphic types~\cite{DBLP:conf/pldi/PolikarpovaKS16}.
 
@@
 such as concurrent programming (e.g. session types for message passing
-concurrency~\cite{}), and to memory-management (e.g.~Rust's ownership
+concurrency~\cite{DBLP:journals/mscs/CairesPT16}), and to memory-management (e.g.~Rust's ownership
 types).
@@
 The simply-typed $\lambda$-calculus is a typed core functional
-language~\cite{} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
+language~\cite{10.5555/509043} that captures the essence of a type system in a simple and familiar environment. Its syntax consists of
 functional abstraction, written $\lambda x{:}A.M$, denoting
@@
 present proof-search in logic, e,g., at any given point, many proof
-  rules are applicable in general. The technique of focusing~\cite{}
+rules are applicable in general. The technique of focusing~\cite{10.1093/logcom/2.3.297}
 has been previously studied as a way to discipline proof search in
@@
 follows that a lot of literature is available to inspire and
-complement our work. Most works~\cite{} follow some variation of the
+complement our work. Most works~\cite{DBLP:conf/lopstr/HughesO20,DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16} follow some variation of the
 synthesis-as-proof-search approach, however, the process is novel for
@@
 common patterns we might want to follow are the use of type
-refinements~\cite{} and the support for synthesis of recursive
-functions~\cite{}.  These works share some common ground, and base
+refinements~\cite{DBLP:conf/pldi/PolikarpovaKS16} and the support for synthesis of recursive
+functions~\cite{DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15}.  These works share some common ground, and base
 themselves of the same works we do -- most famously the Curry-Howard
 correspondence, but also, for example, works on focusing-based
-proof-search~\cite{focusing}.
+proof-search~\cite{10.1093/logcom/2.3.297}.
 
@@
 auxiliar technique to trim down the program synthesis vast search
-space.  \todo[inline]{more introduction to the paper}.  A good use is
+space. A good use is
 made of the general strategy of using a type system to generate

### 2021-05-09 article/article.tex
merge

--- 2021-05-09 article/article.tex
@@
 examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
-types~\cite{DBLP:conf/lopstr/HughesO20}, ...).  Regardless of the kind
+types~\cite{DBLP:conf/lopstr/HughesO20}).  Regardless of the kind
 of specification, program synthesizers have to deal with two main

### 2021-05-09 article/article.tex

@@
 present proof-search in logic, e,g., at any given point, many proof
-rules are applicable in general. The technique of focusing~\cite{10.1093/logcom/2.3.297}
+rules are applicable in general. The technique of focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05}
 has been previously studied as a way to discipline proof search in
-linear logic-- a method created to trim down the search space of
+linear logic -- a method created to trim down the search space of
 valid proofs in linear logic, by eagerly applying invertible rules
@@
 \mypara{Step by step} Our work so far has been constantly accompanied
-by somewhat dense theoretical material, state of the art,
-which we must understand to overcome challenges in these fields,
-and build from the start a more scalable synthetizer -- we want
-to display concrete results to prove the feasability of synthetising
-linear types, and thus possibility for future application to related fields,
-such as session types for message passing concurrency.
+by the study of somewhat dense theoretical material, state of the art,
+which we must understand to overcome challenges in these fields, in
+order to develop a more scalable synthetizer -- we want to display
+concrete results to prove the feasability of program synthesis from
+linear types, and thus open the possibility for future application to
+related fields, such as program synthesis based on session types for
+message-passing concurrency.
 
-Apart from the theory, when we started studying type systems, we created
-a simple type-checker for the simply-typed $\lambda$-calculus.
+Apart from the theory, when we started studying type systems, we
+created a simple type-checker for the simply-typed $\lambda$-calculus.
 While learning about linear type systems, and the simply-typed linear
-$\lambda$-calculus, we developed a parser, a desugar module, and an
-evaluation module to write simple programs in the language;
-culminating in writing the type-checker for linear types. At this point,
-where we stand as of now, we can parse, type-check, and evaluate this
-small linear functional language.
-The next step is to start developing the actual synthetizer, embedding it
-with all the techniques we've seen. Following work will be done to
-augment usability, expand the language domain, and add features such as the
-ones mentioned above.
+$\lambda$-calculus, we developed a parser, a desugaring module, and an
+evaluation module to write, analyze and execute simple programs in the
+language; culminating in writing a type-checker for the linear
+$\lambda$-calculus. At this point, we can parse, typecheck, and
+execute this small linear functional language.  The next step is to
+start developing the actual synthetizer, embedding it with all the
+techniques we've seen. Following work will be done to augment
+usability, expand the language domain, and add features such as those
+mentioned above.
 

### 2021-07-08 article/article.tex
write overview start

@@
 \newcommand{\llet}[2]{\mathsf{let}\,#1\,\mathsf{in}\,#2}
+
+\newcommand{\synname}{\emph{SILI}}
 
@@
 %
-This work explores the process of synthesizing linear functional
+This work explores the synthesis of linear functional
 programs from types based on linear logic propositions (i.e.~linear
-types) by leveraging the Curry-Howard correspondence.  The
-correspondence states that propositions in a logic have a direct
+types) by leveraging the Curry-Howard correspondence. It
+states that propositions in a logic have a direct
 mapping to types, and well-typed programs correspond to proofs of
 those propositions.  As such, having a type be a proposition in linear
-logic, we can relate a proof of that proposition directly to a linear
+logic, we can relate a proof of it directly to a linear
 functional program — finding a proof is finding a program with that
-type. Thus, we formulate synthesis as proof search in linear logic,
-which allows us to inform our work with approaches from the proof
-search literature.  Linear types differ from more traditional types in
+type. Formulating synthesis as proof search in linear logic
+allows us to inform our work with approaches from the relevant 
+literature.
+
+Linear types differ from more traditional types in
 that they constrain resource usage in programs by \emph{statically}
@@
 
-\mypara{Goals}
- % Goals
-In the end, we intend to be able to do full and partial synthesis of
-well-typed programs. Full synthesis consists of the production of a
-function (or set of) satisfying the specification; partial synthesis
-is the ``completion'' of a partial program (i.e. a function with a
-\emph{hole} in it). The work will start from a small core linear --
-i.e. \emph{resource-aware} functional language, building up to
-recursive types/functions, with potential other avenues of further
-extension. 
-%
-We will evaluate the work through
-expressiveness benchmarks (i.e. can we synthesize a term of type $T$?) and time
-measurements (i.e. how fast can we synthesize a term of type $T$?').
+%\mypara{Goals}
+% % Goals
+%In the end, we intend to be able to do full and partial synthesis of
+%well-typed programs. Full synthesis consists of the production of a
+%function (or set of) satisfying the specification; partial synthesis
+%is the ``completion'' of a partial program (i.e. a function with a
+%\emph{hole} in it). The work will start from a small core linear --
+%i.e. \emph{resource-aware} functional language, building up to
+%recursive types/functions, with potential other avenues of further
+%extension. 
+%%
+%We will evaluate the work through
+%expressiveness benchmarks (i.e. can we synthesize a term of type $T$?) and time
+%measurements (i.e. how fast can we synthesize a term of type $T$?').
 
@@
 Inference rules in natural deduction are categorized as introduction
-or elimination rules, these correspond, respectively, to rules for
+or elimination rules; these correspond, respectively, to rules for
 constructors and destructors in programming languages (e.g. function
@@
 but where only one of the two resources may be used. Similarly,
-implication becomes into linear implication, written $A\lolli B$,
+implication becomes linear implication, written $A\lolli B$,
 denoting a resource that will consume (exactly one) resource $A$ to
@@
 augments the form of the judgment and it's the one that ensures
-logical strength -- i.e. we're able to translate intuitionistic logic
+logical strength i.e. we're able to translate intuitionistic logic
 into linear logic.  The proposition $\bang A$ (read \emph{of course}
@@
 
+\mypara{Sequent Calculus} Inference rules in natural deduction (the
+ones we have considered so far) are ill-suited for bottom-up
+proof-search since elimination rules work top-down and introduction
+rules work bottom-up.  A more suited candidate is the equivalent
+\emph{sequent calculus} system in which \emph{all} inference rules can
+be understood naturally in a \emph{bottom-up} manner.  The
+introduction and elimination rules from natural deduction disappear,
+and their place is taken by right and left rules, respectively.
+%
+Right rules correspond exactly to the introduction rules of natural
+deduction, which were already understood bottom-up. The inference rule
+for implication introduction ($\supset$ is used instead of
+$\rightarrow$), seen above in natural deduction under the
+\emph{propositions as types} subsection, corresponds to the following
+right rule in sequent calculus:
+\[
+    \infer*[right=($\supset R$)]
+    {\Gamma, u: A \vdash M : B}
+    {\Gamma \vdash \lambda u. M : A \supset B}
+  \]
+
+Intuitively, left rules act as the elimination rules of natural deduction, but are altered
+to work bottom-up, instead of top-down.
+The inference rule for the also seen above implication elimination is
+as follows:
+\[
+    \infer*[right=($\supset L$)]
+    {\Gamma , u:B\vdash M : C \and \Gamma, u{:}A\supset B \vdash N : A}
+    {\Gamma, u{:}A\supset B\vdash M\{(u\,N)/u\} : C}
+\]
+As implied by their name, left rules define how to decompose or make
+use of a connective on the left of the turnstile $\vdash$. To use an
+assumption of $A\supset B$ while attempting to show some proposition
+$C$ we produce a proof of $A$, which allows us to use an assumption
+of $B$ to prove $C$. In terms of the corresponding $\lambda$-terms,
+the $\supset L$ rule corresponds to applying the variable $u$ to the
+argument $N$ but potentially deep in the structure of $M$.
+
+We'll define our inference rules for synthesis under the sequent
+calculus system, for it simplifies the work to be done in the
+synthesizer.
+
 \mypara{Resource-Management}
@@
 
-\mypara{Sequent Calculus} Inference rules in natural deduction (the
-ones we have considered so far) are ill-suited for bottom-up
-proof-search since elimination rules work top-down and introduction
-rules work bottom-up.  A more suited candidate is the equivalent
-\emph{sequent calculus} system in which \emph{all} inference rules can
-be understood naturally in a \emph{bottom-up} manner.  The
-introduction and elimination rules from natural deduction disappear,
-and their place is taken by right and left rules, respectively.
-%
-Right rules correspond exactly to the introduction rules of natural
-deduction, which were already understood bottom-up. The inference rule
-for implication introduction ($\supset$ is used instead of
-$\rightarrow$), seen above in natural deduction under the
-\emph{propositions as types} subsection, corresponds to the following
-right rule in sequent calculus:
-\[
-    \infer*[right=($\supset R$)]
-    {\Gamma, u: A \vdash M : B}
-    {\Gamma \vdash \lambda u. M : A \supset B}
-  \]
-
-Intuitively, left rules act as the elimination rules of natural deduction, but are altered
-to work bottom-up, instead of top-down.
-The inference rule for the also seen above implication elimination, is
-as follows:
-\[
-    \infer*[right=($\supset L$)]
-    {\Gamma , u:B\vdash M : C \and \Gamma, u{:}A\supset B \vdash N : A}
-    {\Gamma, u{:}A\supset B\vdash M\{(u\,N)/u\} : C}
-\]
-As implied by their name, left rules define how to decompose or make
-use of a connective on the left of the turnstile $\vdash$. To use an
-assumption of $A\supset B$ while attempting to show some proposition
-$C$ we produce a proof of $A$, which allows us to use an assumption
-of $B$ to prove $C$. In terms of the corresponding $\lambda$-terms,
-the $\supset L$ rule corresponds to applying the variable $u$ to the
-argument $N$ but potentially deep in the structure of $M$.
-
-We'll define our inference rules for synthesis under the sequent
-calculus system, for it simplifies the work to be done in the
-synthesizer.
-
 \mypara{Focusing}
 
-Even with everything mentioned so far, non-determinism is still
-present proof-search in logic, e,g., at any given point, many proof
+Even with everything mentioned so far, non-determinism is still very
+present in proof-search, e.g. at any given point, many proof
 rules are applicable in general. The technique of focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05}
@@
 non-determinism), identifying precisely the points at which
-backtracking search is necessary.
+backtracking is necessary.
+
+\section{Overview}
+
+At the core of the synthesis process, and of the developed synthetiser \synname,
+is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
+resource management. It connects the concepts seen above -- a type system
+consisting of inference rules in sequent calculus for linear logic connectives;
+with resource management, and focusing, as formalized techniques to,
+respectively, solve resource related non-determinism, and to eliminate from the
+proof search don't-care non-determinism.  Guided solely by this system's formal
+representation\ref{sec:formal_system}, we can (and did) create a very robust
+synthetiser for a select part of the linear types that's able to synth
+programs successfully from e.g. the major linear logic equivalences. For example,
+given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
+the program $\lambda a \rightarrow \lambda b \rightarrow$ let $c \tensor d$ =
+$b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.
+
+The \synname\ programming language is naturally born alongside the synthetiser
+to create an environment for it. Synth goals are inserted in the program by use
+of a \emph{synth} keyword or syntatic \emph{mark}, indicating a program should
+be generated for a given, or inferred, type. Synth results are functions in this
+language's syntax. Additionally, it opens the possibility of synthetisis with
+context, that is, with knowledge of other functions and eventual structures
+defined in the same program.
+
+However, a programming language only with the canonical linear types lacks
+expressiveness. To enhance \synname\ and its language, and consequently synth
+more interesting programs -- and thus empirically prove the feasability of
+more relevant synthesis in a linear context, we extend the syntax and type
+system with richer types: -- polimorfic, refinements, adts parametrizáveis...  todo
+
+\section{Architecture}
+
+\subsection{Implementation}
+
+A implementação foi feita em haskell e o código está disponível em X e há um
+live demo com exemplos pré-feitos em Y?
+
+\section{Formal System}\label{sec:formal_system}
 
@@
 
-% Expand on earlier points. Start from a functional language with linear
-% types ($\tensor$, $\lolli$, $\oplus$, $\bang$); build on it with more
-% ``stuff''. General techniques, drawn from proof theory via props as
-% types: explore focusing to tame non-determinism/search space. Partial
-% synthesis is still proof search! Add other techniques as we go along
-% (e.g. for recursion we need to constrain recursive calls).
+% Expand on earlier points. Start from a functional language with linear types
+% ($\tensor$, $\lolli$, $\oplus$, $\bang$); build on it with more ``stuff''.
+% General techniques, drawn from proof theory via props as types: explore
+% focusing to tame non-determinism/search space. Partial synthesis is still
+% proof search! Add other techniques as we go along (e.g. for recursion we need
+% to constrain recursive calls).
 
-% Mention that you've already implemented a type-checker for this
-% (useful as a prelim. Exercise but also later, for \emph{validation}).
+% Mention that you've already implemented a type-checker for this (useful as a
+% prelim. Exercise but also later, for \emph{validation}).
 
-% Validation and evaluation: validation is as simple as ``does it typecheck''? can you synth? how fast?
+% Validation and evaluation: validation is as simple as ``does it typecheck''?
+% can you synth? how fast?
+
+\section{Evaluation}
+
+% Adicionar já exemplos do que consigo
 

### 2021-07-08 STLLC/RefinementTypes.hs

@@ -5,11 +5,11 @@ nat :: x { Int | x > 0 };
 nat = add 6 3;
 
 
-wplus :: (x { Int } -> y { Int | x == y } -> z { Int | x >= y && x == y });
+wplus :: (x { Int } -> y { Int | x == y } -> z { Int | z >= x + y });
 wplus x y = add x y;
 
 
-inc :: (x { Int | x > 0 } -> z { Int | z == x + 1 });
+inc :: (x { Int | x >= 0 } -> z { Int | z == x + 1 && x == 0 });
 inc = add 1;
 
 

### 2021-07-09 article/article.tex
Write formal system start

@@
 
+Escrever introdução melhor que tenha isto como seguimento...
+
 At the core of the synthesis process, and of the developed synthetiser \synname,
@@
 However, a programming language only with the canonical linear types lacks
-expressiveness. To enhance \synname\ and its language, and consequently synth
-more interesting programs -- and thus empirically prove the feasability of
-more relevant synthesis in a linear context, we extend the syntax and type
-system with richer types: -- polimorfic, refinements, adts parametrizáveis...  todo
+expressiveness. To enhance \synname\ and its language, synth more interesting
+programs -- and thus empirically proving the feasability of more relevant
+synthesis in a linear context, we extend the syntax and type system with
+recursiveness and richer
+types: -- polimorfic, refinements, adts parametrizáveis...  todo
 
@@
 \section{Formal System}\label{sec:formal_system}
+
+The development of \synname\ was initially found on an almost direct reading of
+concrete inference rules and theoretical principles; and work on its environment
+was also guided by formal systems and methods (e.g. the Hindley-Milner type
+inference~\cite{}).  However, with the addition of recursion and richer types to the synth and
+language, formal guidelines to support the implementation become scarcer, or
+more complex. Accordingly, rather than derived from explicit rules,
+solutions appear first to tame the extensions' inherent complex
+problems (such as infinite recursion) and from them emerge implicit rules.
+
+With this section we intend to formalize the implicit rules that guide the
+synthesis of more expressive programs, alongside the already explicit rules that
+model the core of the synth process, resulting in a \emph{sound} set of rules
+that hopes to serve as a source of thoughts on taming synthesis of recursive
+functions with richer types in a linear context, as well as describe the system
+in enough detail for the synthetiser to be reproducible by a theory-driven
+different implementation.
+
+Note that \emph{sound} and \emph{complete} mean two very different
+things, and that this set of rules is only sound, meaning it might generate
+correct results but simultaneously hiding an infinite amount of valid programs,
+and that it reflects the quite subjective trade-offs we decided on, and that
+different synths might take completely different approaches.
 

### 2021-07-09 article/article.tex
Write focusing rules and core formal system part

--- 2021-07-08 article/article.tex
@@
 
-\mypara{Resource-Management}
+\mypara{Resource-Management}\label{sec:background_resource_management}
 Be it from the perspective of proof \emph{checking} (i.e.~type checking) or
@@
 
-\mypara{Focusing}
+\mypara{Focusing}\label{sec:background_focusing}
 
@@
 \section{Overview}
+
+Escrever introdução melhor que tenha isto como seguimento...
 
@@
 
+\subsection{The \synname\ language}
+
 \subsection{Implementation}
 
-A implementação foi feita em haskell e o código está disponível em X e há um
-live demo com exemplos pré-feitos em Y?
+% A implementação foi feita em haskell e o código está disponível em X e há um
+% live demo com exemplos pré-feitos em Y?
 
 \section{Formal System}\label{sec:formal_system}
+
+The development of \synname\ was initially found on an almost direct reading of
+concrete inference rules and theoretical principles (~\cite{frank pfenning notes}); and work on its environment
+was also guided by formal systems and methods (e.g. the Hindley-Milner type
+inference~\cite{}).  However, with the addition of recursion and richer types to
+the synth and language, formal guidelines to support the implementation become
+scarcer, or more complex. Accordingly, rather than derived from explicit rules,
+solutions appear first to tame the extensions' inherent complex problems (such
+as infinite recursion), and implicit rules come only as a byproduct.
+
+With this section we intend to formalize the implicit rules that guide the
+synthesis of more expressive specifications, alongside the already explicit
+rules that model the core of the synth process, resulting in a \emph{sound} set
+of rules that hopes to serve as a source of ideas on taming synthesis of
+recursive functions with richer types in a linear context, as well as describe
+the system in enough detail for the synthetiser to be reproducible by a
+theory-driven implementation.
+
+Even so, note that \emph{soundness} and \emph{completeness} have two very
+different meanings, and our \emph{sound} set of rules only guarantees we won't
+synth incorrect programs, but still hides a possibly infinite number of correct
+proofs; and that the valid programs derivable through it reflect the somewhat
+subjective trade-offs we decided on. Different choices and approaches outside
+the core might lead to completely distinct synths and spaces of valid programs.
+
+\mypara{Core Rules} The core system comprises of the sequent calculus rules for
+linear logic connectives with resource-management and focusing. As seen
+previously\ref{sec:background}, focusing separates the proof search into two
+phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
+focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
+apply non-invertible rules, staying focused until either the focus is again an
+asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
+invertible rules, a third, ordered, context ($\Omega$) is needed to hold
+propositions resulting from decomposing connectives, that we'll later try to
+invert, moving them to the linear ($\Delta$) context when we fail to.
+
+Putting together linear logic, the Curry-Howard correspondence, resource
+management, and focusing, we get the following formal system (heavily inspired
+by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
+first rule is read: to synth a program of type $A \lolli B$ while inverting
+right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
+context $\Delta$, and ordered context $\Omega$, synth a program of type $B$ with
+an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
+resulting in the program $M$ and in the output\ref{sec:background} linear
+context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
+resulting program is $\lambda x \rightarrow M$ and the output linear context is
+$\Delta'$.
+
+We begin with the right invertible rules:
+
+\begin{mathpar}
+
+    % -o R
+    \infer*[right=($\lolli R$)]
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
+    \not\in \Delta'}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x \rightarrow M : A
+    \lolli B \Uparrow}
+
+\and
+
+    % & R
+    \infer*[right=($\with R$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
+    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A
+    \with B \Uparrow}
+
+\end{mathpar}
+
+When we reach a non-invertible proposition on the right, we start inverting the
+$\Omega$ context. The rule to transition to inversion on the left is:
+
+\begin{mathpar}
+    \infer*[right=($\Uparrow R$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C \and C\ \textrm{not
+    right asynchronous}}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash M : C \Uparrow}
+\end{mathpar}
+
+Followed by the left invertible rules:
+
+% * L
+\begin{mathpar}
+
+    \infer*[right=($\tensor L$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega, y{:}A, z{:}B \Uparrow\ \vdash M : C
+    \and y,z \not\in \Delta'}
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \tensor B \Uparrow\ \vdash\
+    \textrm{let}\ y \tensor z = x\ \textrm{in}\ M : C}
+
+\and
+
+    \infer*[right=($1 L$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}1 \Uparrow\ \vdash\ \textrm{let}\
+    \star =
+    x\ \textrm{in}\ M : C}
+
+\and
+
+    \mprset{flushleft}
+    \infer*[right=($\oplus L$)]
+    {
+    \Gamma ; \Delta/ \Delta' ; \Omega, y{:}A \Uparrow\ \vdash M : C \and
+    y \not\in \Delta' \\
+    \Gamma ; \Delta/ \Delta'' ; \Omega, z{:}B \Uparrow\ \vdash N : C \\
+    z \not\in \Delta'' \\
+    \Delta' = \Delta''
+    }
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \oplus B \Uparrow\ \vdash\
+    \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y \rightarrow M\ |\
+    \textrm{inr}\ z \rightarrow N : C}
+
+\and
+
+    % {TODO: Nao ha restrições sobre este? no código tinha que x não podia estar em
+    % delta mas não faz muito sentido hah}
+    \infer*[right=($\bang L$)]
+    {\Gamma, y{:}A ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
+    {\Gamma ; \Delta/\Delta ; \Omega, x{:}\bang A \Uparrow\ \vdash\
+    \textrm{let}\ \bang y = x\ \textrm{in}\ M : C}
+
+\end{mathpar}
+
+When we find a synchronous (i.e. non-invertible) proposition in $\Omega$,
+we simply move it to the linear context $\Delta$, and keep inverting on the left:
+
+\begin{mathpar}
+    \infer*[right=($\Uparrow L$)]
+    {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
+    \textrm{not left asynchronous}}
+    {\Gamma; \Delta/\Delta'; \Omega, A \Uparrow\ \vdash C}
+\end{mathpar}
+
+After inverting all the asynchronous propositions in $\Omega$ we'll reach a state
+where there are no more propositions to invert ($\Gamma'; \Delta'; \cdot
+\Uparrow\ \vdash C$). At this point, we want to \emph{focus} on a proposition.
+The focus object can be any of the three: the proposition on the right (the
+goal), a proposition from the linear $\Delta$ context, or a proposition from the
+unrestricted $\Gamma$ context. For these options we have three \emph{decision}
+rules:
+
+\begin{mathpar}
+    \infer*[right=(decideR)]
+    {\Gamma; \Delta/\Delta' \vdash C \Downarrow \and C\ \textrm{not atomic}}
+    {\Gamma; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
+\and
+    \infer*[right=(decideL)]
+    {\Gamma; \Delta/\Delta' ; A \Downarrow\ \vdash C}
+    {\Gamma; \Delta, A/\Delta';\cdot \Uparrow\ \vdash C}
+\and
+    \infer*[right=(decideL!)]
+    {\Gamma, A; \Delta/\Delta' ; A \Downarrow\ \vdash C}
+    {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
+\end{mathpar}
+
+The decision rules are naturally followed by either left or right focus rules:
+
+\begin{mathpar}
+    \infer*[right=($\tensor R$)]
+    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
+    : B \Downarrow}
+    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
+\and
+    \infer*[right=($1 R$)]
+    {\and}
+    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
+\and
+    \infer*[right=($\oplus R_1$)]
+    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
+\and
+    \infer*[right=($\oplus R_2$)]
+    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
+\and
+    \infer*[right=($\bang R$)]
+    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
+    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
+\and
+%TODO: eu tenho uma assertion que garante que y não sai com Delta', mas agora
+%parece-me inutil?
+    \infer*[right=($\lolli L$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
+    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
+    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
+\and
+    \infer*[right=($\with L_1$)]
+    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
+\and
+    \infer*[right=($\with L_2$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
+\end{mathpar}
+
+Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
+asynchronous. If we're left focused on an atomic proposition we either instance
+the goal or fail. Otherwise the left focus is asynchronous and we can start
+inverting it. If we're right focused on a proposition that isn't right
+synchronous, we switch to inversion as well. Three rules model these conditions:
+
+\begin{mathpar}
+    \infer*[right=(init)]
+    {\and}
+    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
+\and
+    \infer*[right=($\Downarrow L$)]
+    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
+    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
+\and
+    \infer*[right=($\Downarrow R$)]
+    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
+    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
+\end{mathpar}
+
+The rules written above together make the core of our synthetiser. Ahead, we'll
+present new rules that align and build on top of these to synth recursive
+programs from more expressive (richer) types.
+
+\mypara{Recursion}
 

### 2021-07-09 STLLC/List.hs

@@ -4,22 +4,22 @@ data Maybe a = Nothing | Just a;
 
 
 
-#synth na :: (List Int -> List Int);
+synth na :: (List Int -> List Int);
 
 
-#synth singleton :: (a -> List a);
+synth singleton :: (a -> List a);
 
-#synth append :: (List a -> List a -> List a);
+synth append :: (List a -> List a -> List a);
 
-#synth map :: ((!(a -> b)) -> List a -> List b);
+synth map :: ((!(a -> b)) -> List a -> List b);
 
-#synth foldl :: (!(b -> a -> b) -> b -> List a -> b);
+synth foldl :: (!(b -> a -> b) -> b -> List a -> b);
 
-#synth uncons :: (List a -> Maybe (a * List a));
+synth uncons :: (List a -> Maybe (a * List a));
 
 # is wrong, synth is equal to foldl c parametros trocados -- escrever
-#synth foldr :: (!(a -> b -> b) -> b -> List a -> b);
+synth foldr :: (!(a -> b -> b) -> b -> List a -> b);
 
-#synth insert :: (a -> List a -> List a)
+synth insert :: (a -> List a -> List a)
 
-synth concat :: (List (List a) -> List a);
+#synth concat :: (List (List a) -> List a);

### 2021-07-09 STLLC/app/Synth.hs

@@ -323,7 +323,7 @@ synth c@(g, d, (n, Plus a b):o) t = addtrace (LeftPlus c (Plus a b) t) $ do
     (expa, d') <- memoizesynth synth (g, d, (n1, a):o) t
     (expb, d'')  <- memoizesynth synth (g, d, (n2, b):o) t
     guard (d' == d'')
-    guard ((n1 `notElem` map fst d') && (n2 `notElem` map fst d'))
+    guard ((n1 `notElem` map fst d') && (n2 `notElem` map fst d''))
     return (CaseOfPlus (Var n) n1 expa n2 expb, d')
 
 ---- sumL
@@ -377,7 +377,7 @@ synth c@(g, d, p@(n, ADT tyn tps):o) t = (lift (lift ask) >>= \(res,_,_,_,_) -> 
 synth c@(g, d, (n, Bang a):o) t = addtrace (LeftBang c (Bang a) t) $ do
     nname <- fresh
     (exp, d') <- memoizesynth synth ((nname, Right a):g, d, o) t
-    guard (nname `notElem` map fst d')
+    guard (nname `notElem` map fst d') -- Useless?
     return (LetBang nname (Var n) exp, d')
 
 ---- refinementL
@@ -582,9 +582,7 @@ focus c goal =
         focus' (Just (n, Fun a b)) c@(g, d) goal = addtrace (LeftFun c (Fun a b) goal) $ do
             nname <- fresh
             (expb, d') <- memoizefocus' focus' (Just (nname, b)) c goal
-            guard (nname `notElem` map fst d')
-            -- subs <- getsubs
-            -- (expa, d'') <- synth (g, apply subs d', []) (apply subs a) -- TODO?
+            guard (nname `notElem` map fst d') -- TODO: Useless? make sure the focusL rule makes sure the proposition is not left synchronous
             (expa, d'') <- memoizesynth synth (g, d', []) a
             return (substitute nname (App (Var n) expa) expb, d'')
             
@@ -593,13 +591,13 @@ focus c goal =
             do
                 nname <- fresh
                 (lf, d') <- memoizefocus' focus' (Just (nname, a)) c goal
-                guard (nname `notElem` map fst d')
+                guard (nname `notElem` map fst d') -- TODO: useless :)
                 return (substitute nname (Fst (Var n)) lf, d')
             <|>
             do
                 nname <- fresh
                 (rt, d') <- memoizefocus' focus' (Just (nname, b)) c goal
[…]

### 2021-07-10 article/article.tex
Write simple ADTs, trivial recursion, debugging

@@
 \newcommand{\synname}{\emph{SILI}}
+\newcommand{\te}[1]{\textrm{\emph{#1}}}
 
@@
 
+\mypara{Debugging} The rules described in the formal
+system\ref{sec:formal_system} to handle infinite recursion might seem obvious in
+retrospective, however, when faced with a program that doesn't terminate, what
+went wrong is not so clear. While we were able to figure out a solution for the
+first couple of "infinite examples", the following problems were much harder,
+taking sometimes full days to solve (and as of now, some still remain unsolved,
+e.g. List (List a) -o List a). To make the debugging experience better, we
+developed a tracing system that prints information whenever a rule is applied,
+alongside the stack of rules applied up until that one. This improved our
+debugging speed considerably, especially when put together with a correct
+derivation done by hand, with which we could match what was happening against
+what was supposed to happen.
+
+Unfortunately, it seems that sometimes in the process of backtracking, stack
+information is "lost", and the trace logs lose useability.
+
 % A implementação foi feita em haskell e o código está disponível em X e há um
@@
 invertible rules, a third, ordered, context ($\Omega$) is needed to hold
-propositions resulting from decomposing connectives, that we'll later try to
-invert, moving them to the linear ($\Delta$) context when we fail to.
+propositions resulting from decomposing connectives; propositions we'll later try to
+invert, moving them to the linear context ($\Delta$) when we fail to.
 
@@
 an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
-resulting in the program $M$ and in the output\ref{sec:background} linear
+resulting in the program $M$ and output\ref{sec:background} linear
 context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
@@
     \infer*[right=($1 R$)]
-    {\and}
+    {  }
     {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
@@
     \infer*[right=(init)]
-    {\and}
+    {  }
     {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
@@
 
-\mypara{Recursion}
+\mypara{Recursion} The main idea behind synthesis of recursive programs is the
+labeling of the main goal and the addition of its type, under that name, to the
+unrestricted context. That is, to synth a function of type $A \lolli B$ named
+\emph{foo}, the initial judgment can be written as
+
+\begin{mathpar}
+    \infer
+    {\vdots}
+    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
+    A \lolli B \Uparrow}
+\end{mathpar}
+and, by definition, all subsequent inference rules will have
+($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.
+
+We can force the usage of the recursive call by adding it not only to the
+unrestricted context, but to the linear one as well.
+
+% TODO: E como evitamos coisas trivialmente estúpidas?
+
+\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
+is a named sum type of any type, i.e. a named type instanceable by one of many
+tags (also known as constructors) that take exactly one fixed type, which might
+be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
+for constructors with an arbitrary number of parameters. In the \synname\
+language, the programmer can define custom ADTs; as an example, we show the
+definition of an ADT which might be constructed with zero, one, or two
+propositions of type $A$, using its syntax: $\textrm{data Container} =
+\textrm{None}\ 1\ \vert\ \textrm{One}\ A\ \vert\ \textrm{Two}\ (A * A)$.
+
+The semantics of the simple ADT relate to those of the plus ($\oplus$) type --
+both are additive disjunctions.  A proof of an ADT requires only a proof of one
+of the types that a constructor takes, similar to the way $\oplus$ requires only
+proof of either the left or right type it consists of. Analogously, a proof of
+$C$ by deconstructing an ADT stipulates that all types that can construct the
+ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
+the inference rules for a simple ADT mimic the $\oplus$ rules, for an
+arbitrary ($> 1$) number of types, rather than exactly two. Therefore, there's one
+left rule for ADTs, and an arbitrary number of right rules, one for each
+constructor:
+
+\begin{mathpar}
+    \infer*[right=(adtR)]
+    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
+\and
+    \mprset{flushleft}
+    \infer*[right=(adtL)]
+    { \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C \and
+    y_1 \not\in \Delta'_1
+    \\ \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
+    \\ y_2 \not\in \Delta'_2
+    \\\\ \\ \dots
+    \\ \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
+    \\ y_n \not\in \Delta'_n
+    \\ \Delta'_1 = \Delta'_2 = \dots = \Delta'_n }
+    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
+    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
+    \rightarrow M_n : C}
+\end{mathpar}
+
+where the ADT \emph{Foo} and its definition stand for any ADT with form
+\[
+    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
+\dots\ \vert\ \textrm{Bar}_n\ X_n
+\]
+
+A more general (and interesting) formulation of ADTs says an ADT can be
+recursive (or "indutively defined"), meaning the type a constructor needs might
+be the type the constructor constructs. Seemingly small, this change has a big
+impact in the synthesis process. Take, for instance, the ADT defined as
+$\textrm{data Foo} = \textrm{Bar Foo}$, the synth goal $\te{Foo}
+\lolli C$, and its derivation:
+
+\begin{mathpar}
+    \infer*[right=($\lolli R$)]
+    {
+        \infer*[right=($\Uparrow R$)]
+        {
+            \infer*[right=(adtL)]
+            {
+                \infer*[right=(adtL)]
+                {
+                    \infer*[right=(adtL)]
+                    {\dots}
+                    {\Gamma; \Delta/\Delta'; \Omega, z{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ z\ 
+                    \textrm{of Bar}\ k \rightarrow \dots : C}
+                }
+                {\Gamma; \Delta/\Delta'; \Omega, y{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ y\ 
+                \textrm{of Bar}\ z \rightarrow \dots : C}
+            }
+            {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ x\ 
+            \textrm{of Bar}\ y \rightarrow \dots : C}
+        }
+        {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \vdash \dots : C
+        \Uparrow}
+    }
+    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
+    \te{Foo} \lolli C \Uparrow}
+\end{mathpar}
+
+Using our current system, we are to apply an infinite number of times the left
+ADT rule, never reaching a final state for the proof. Simetrically, the
+derivation for the synth goal $\te{Foo}$ doesn't have an end either.
+
+\begin{mathpar}
+    \infer*[left=(adtR)]
+    {
+        \infer*[left=(adtR)]
+        {
+            \infer*[left=(adtR)]
+            {
+                \infer*[left=(adtR)]
+                {\dots}
+                {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
+            }
+            {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
+        }
+        {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
+    }
+    {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo}
+    \Downarrow}
+\end{mathpar}
+
+\mypara{Polimorfic Types}
 

### 2021-07-11 article/article.tex
Write on recursive ADTs

@@
 \newcommand{\synname}{\emph{SILI}}
+\def\Rho{P}
 \newcommand{\te}[1]{\textrm{\emph{#1}}}
@@
 
+This section presents the rules while explaining them, modifying some along the
+process. For a
+
 \mypara{Core Rules} The core system comprises of the sequent calculus rules for
@@
     {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
-    \not\in \Delta'}
+    \notin \Delta'}
     {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x \rightarrow M : A
@@
     \infer*[right=($\Uparrow R$)]
-    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C \and C\ \textrm{not
+    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
     right asynchronous}}
-    {\Gamma ; \Delta/\Delta' ; \Omega \vdash M : C \Uparrow}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
 \end{mathpar}
@@
     {\Gamma ; \Delta/ \Delta' ; \Omega, y{:}A, z{:}B \Uparrow\ \vdash M : C
-    \and y,z \not\in \Delta'}
+    \and y,z \notin \Delta'}
     {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \tensor B \Uparrow\ \vdash\
@@
     \Gamma ; \Delta/ \Delta' ; \Omega, y{:}A \Uparrow\ \vdash M : C \and
-    y \not\in \Delta' \\
+    y \notin \Delta' \\
     \Gamma ; \Delta/ \Delta'' ; \Omega, z{:}B \Uparrow\ \vdash N : C \\
-    z \not\in \Delta'' \\
+    z \notin \Delta'' \\
     \Delta' = \Delta''
@@
     \infer
-    {\vdots}
+    {\dots}
     {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
@@
 
-We can force the usage of the recursive call by adding it not only to the
+We can also force the usage of the recursive call by adding it not only to the
 unrestricted context, but to the linear one as well.
@@
     \infer*[right=(adtL)]
-    { \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C \and
-    y_1 \not\in \Delta'_1
-    \\ \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
-    \\ y_2 \not\in \Delta'_2
-    \\\\ \\ \dots
-    \\ \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
-    \\ y_n \not\in \Delta'_n
-    \\ \Delta'_1 = \Delta'_2 = \dots = \Delta'_n }
+    {
+        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
+        \and
+        y_1 \notin \Delta'_1
+        \\
+        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
+        \\
+        y_2 \notin \Delta'_2
+        \\\\
+        \\ \dots
+        \\
+        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
+        \\
+        y_n \notin \Delta'_n
+        \\
+        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
+    }
     {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
@@
 
-where the ADT \emph{Foo} and its definition stand for any ADT with form
+where the ADT \emph{Foo} and its constructors stand for any ADT with form
 \[
@@
 
+To handle this situation, we impede the decomposition of an ADT in subsequent
+proofs of its branches, and to impede the construction of an ADT in a subsequent
+proof of a type to construct it. For this, we need two more contexts, $\Rho_C$
+for restrictions on construction and $\Rho_D$ for restrictions on
+deconstruction. For convenience, they are represented by a single $\Rho$ if
+unused. All rules must trivially propagate these. The ADT rules are explicitly
+modified to take this into consideration:
+
+\begin{mathpar}
+    \infer*[right=(adtR)]
+    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
+    \te{Foo} \notin \Rho_C}
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
+\and
+    \mprset{flushleft}
+    \infer*[right=(adtL)]
+    {
+        \te{Foo} \notin \Rho_D
+        \and
+        \Delta'_1 = \dots = \Delta'_n 
+        \\
+        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
+        \and
+        y_1 \notin \Delta'_1
+        \\\\
+        \\ \dots
+        \\\\
+        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
+        \and
+        y_n \notin \Delta'_n
+    }
+    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
+    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
+    \rightarrow M_n : C}
+\end{mathpar}
+where
+\begin{mathpar}
+    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
+    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C
+
+    \Rho'_D = \textrm{likewise}
+\end{mathpar}
+
+These modifications prevent the derivation from not ending for the two seen
+situations, however, they also greatly limit the derivable valid programs space,
+leaving the synthetiser pretty much unable to synth specifications with
+recursive types, which is so undesirable that we add two rules to complement the
+restrictions on construction and destruction of recursive types.
+
+First, after decomposing a recursive ADT, we must consume all constructor
+parameters, and since we can't decompose the self type any further because of a
+restriction, we should be able to use it from the linear context; even though,
+non-restricted types are not to take that course since they weren't unnaturally
+restricted, and because it would result in additional non-determinism and
+infinite recursion. Second, without any additional rules, an ADT in the linear
+context will loop back to the inversion context, and then loop again, going back and
+forth between the two contexts for infinity; instead, when focused on an ADT, we should
+either instance the goal (provided they're the same type), fail if the
+decomposition is restricted, or switch to inversion if it's not. The three
+following rules ensure this:
+
+\begin{mathpar}
+    \infer*[right=(adt$\Uparrow$L)]
+    {
+        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
+        \and
+        \te{Foo} \in \Rho_D
+    }
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
+    \and
+    \infer*[right=(adt-init)]
+    {  }
+    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
+    \and
+    \infer*[right=(adt$\Downarrow$L)]
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
+    \te{Foo} \and \te{Foo} \notin \Rho_D}
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
+\end{mathpar}
+
+% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
+% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
+% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
+% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
+% é agora inutil. Talvez seja demais adicionar à regra
+
+conclude...
+
 \mypara{Polimorfic Types}

### 2021-07-11 STLLC/app/Synth.hs

@@ -350,7 +350,7 @@ synth c@(g, d, p@(n, ADT tyn tps):o) t = (lift (lift ask) >>= \(res,_,_,_,_) -> 
                 (if isrectype -- TODO: make restrictions not dependent
                    then addrestriction DeconstructADT (ADT tyn tps)           -- For recursive types, restrict deconstruction of this type in further computations
                    else id) $
-                   (if t /= ADT tyn tps
+                   (if t /= ADT tyn tps -- TODO!!! para que exemplos preciso disto?? talvez já não seja preciso com as modificações que fiz entretanto
                        then addrestriction ConstructADT (ADT tyn tps)         -- If the goal is anything but the recursive type, restrict construction of type to avoid "stupid functions"
                        else id) $
                         case vtype of
@@ -363,7 +363,6 @@ synth c@(g, d, p@(n, ADT tyn tps):o) t = (lift (lift ask) >>= \(res,_,_,_,_) -> 
                             (exp, d') <- memoizesynth synth (g, d, (varid, argty):o) t
                             guard $ varid `notElem` map fst d'
                             return (name, varid, exp, d')
-                  -- TODO: polymorphic ADT
                 ) adtds
             let (n1, varid1, e1, d1') = head ls
             guard $ all ((== d1') . (\(_,_,_,c) -> c)) (tail ls) -- all resulting contexts are the same
@@ -371,7 +370,7 @@ synth c@(g, d, p@(n, ADT tyn tps):o) t = (lift (lift ask) >>= \(res,_,_,_,_) -> 
     ) <|> (
     lift (lift ask) >>= \(res,_,_,_,_) -> addtrace (MoveLeftADT res c (ADT tyn tps) t) $ do
     checkrestrictions DeconstructADT (ADT tyn tps) >>= guard . not          -- ADT with a restriction on deconstruction might still be useful by being instantiated while focused -- e.g. a Tensor was deconstructed asynchronously but the ADT has a deconstruct restriction -- it shouldn't fail, yet it shouldn't deconstruct either -- this option covers that case (Similar to "Move To Delta" but for ADTs that we cannot deconstruct any further)
-    memoizesynth synth (g, p:d, o) t)                                                    -- So if we failed above because a restriction didn't allow us to invert this ADT, try using the hypothesis in the linear context -- it won't loop back here because the DeconstructADT
[…]

### 2021-07-11 article/article.tex
Write polymorphism

@@
 We begin with the right invertible rules:
-
 \begin{mathpar}
@@
 $\Omega$ context. The rule to transition to inversion on the left is:
-
 \begin{mathpar}
@@
 Followed by the left invertible rules:
-
-% * L
 \begin{mathpar}
@@
 we simply move it to the linear context $\Delta$, and keep inverting on the left:
-
 \begin{mathpar}
@@
 rules:
-
 \begin{mathpar}
@@
 The decision rules are naturally followed by either left or right focus rules:
-
 \begin{mathpar}
@@
 \end{mathpar}
-
 Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
@@
 synchronous, we switch to inversion as well. Three rules model these conditions:
-
 \begin{mathpar}
@@
 \end{mathpar}
-
 The rules written above together make the core of our synthetiser. Ahead, we'll
@@
 \emph{foo}, the initial judgment can be written as
-
 \begin{mathpar}
@@
 ($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.
-
 We can also force the usage of the recursive call by adding it not only to the
@@
 constructor:
-
 \begin{mathpar}
@@
 \end{mathpar}
-
 where the ADT \emph{Foo} and its constructors stand for any ADT with form
@@
 \lolli C$, and its derivation:
-
 \begin{mathpar}
@@
 \end{mathpar}
-
 Using our current system, we are to apply an infinite number of times the left
@@
 derivation for the synth goal $\te{Foo}$ doesn't have an end either.
-
 \begin{mathpar}
@@
 \end{mathpar}
-
 To handle this situation, we impede the decomposition of an ADT in subsequent
@@
 modified to take this into consideration:
-
 \begin{mathpar}
@@
 \end{mathpar}
-
 These modifications prevent the derivation from not ending for the two seen
@@
 leaving the synthetiser pretty much unable to synth specifications with
-recursive types, which is so undesirable that we add two rules to complement the
+recursive types, which is so undesirable that we add three rules to complement the
 restrictions on construction and destruction of recursive types.
@@
 forth between the two contexts for infinity; instead, when focused on an ADT, we should
-either instance the goal (provided they're the same type), fail if the
-decomposition is restricted, or switch to inversion if it's not. The three
+either instance the goal (provided they're the same type), or switch to
+inversion if and only if its decomposition isn't restricted. The three
 following rules ensure this:
-
 \begin{mathpar}
@@
 \end{mathpar}
-
 % TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
@@
 % é agora inutil. Talvez seja demais adicionar à regra
-
-conclude...
-
-\mypara{Polimorfic Types}
+The ADT rules draw closer to a final form, but will still need a revision
+after the introduction of polymorfism and polymorfic parametrization on ADTs.
+
+\mypara{Polimorfic Types} A polymorfic specification is a type $X$ and a set of
+bound variables that quantify universally the type variables in $X$, called a
+\emph{scheme} ($\forall \overline{\alpha}.\ \tau$). Synthesis of a scheme is straightforward.
+First, \textbf{type variables are considered atomic types}, then, instantiate
+the bound vars of the scheme as described by the Hindley-Milner inference
+method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
+each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli \alpha$ could be instanced to
+$\alpha0 \lolli \alpha0$. We add a rule for this:
+\begin{mathpar}
+    \infer*[right=($\forall R$)]
+    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
+    \overline{\alpha}.\ \tau
+    \sqsubseteq \tau'}
+    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
+    \tau \Uparrow}
+\end{mathpar}
+where
+\[
+    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
+    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
+    \overline{\alpha}.\ \tau
+\]
+As such, the construction of a derivation in which, to some types, no specific rules can be applied (except for the \textsc{init} rule,
+a type variable can instance a goal that's a type variable of the same name),
+corresponds to the synthesis of a program where some expressions are treated
+agnostically -- nothing constrains their type, meaning the program is
+polymorphic * TODO: rewrite this?.
+
+The interesting challenge of polymorphism in synthesis is the usage of schemes
+from the unrestricted context.  To begin with, $\Gamma$ now holds both types and
+schemes. Consequently, after the rule \textsc{decideLeft!} is applied, we become
+left focused on either a type or a scheme. Since left focus on a type is already
+well formulated, only rules for left focus on a scheme are needed. Our algorithm
+instances bound type vars of the focused scheme with existential type variables,
+and the instanced type becomes the left focus. Inspired by the Hindley-Milner
+system, we also generate inference constraints on the existential type variables
+(postponing the decision of what type it should be to be used in the proof), and
+collect them in a new constraints context (TODO: should be a substitution not a
+context?) $\Theta$ that is propagated across derivation branches the same way
+the linear context is (by havin an input and output context ($\Theta/\Theta'$)).
+In contrast to HM, everytime a constraint is added it is solved against all
+other constraints -- a branch of the proof search is desired to fail as soon as
+possible; and more, two different existential vars do not unify against
+one another ($?\alpha \mapsto ?\beta$ fails). Note that we instance the scheme with \emph{existential} type
+variables ($\alpha$) rather than just type variables since the latter represent
+universal types in the synth, and the first represent an existential instance of
+a scheme, that might inclusively constraint normal type variables. Additionally,
+it's required that all existential type variables were assigned a type, and
+provided they were, we can forget the constraints generated from the usage of
+the scheme. These concepts are formalized with the rules:
+\begin{mathpar}
+    \infer*[right=($\forall L$)]
+    {
+        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
+        \\
+        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
+        \\
+        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
+    }
+    {\Theta/\Theta; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
+    \and
+    \infer*[right=($?L$)]
+    {\textsc{solve}(?\alpha
+    \mapsto C, \Theta)}
+    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
+    x{:}?\alpha \Downarrow\ \vdash x : C}
+    \and
+    \infer*[right=($\Downarrow L?$)]
+    {\textsc{solve}(?\alpha
+    \mapsto A, \Theta)}
+    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
+    x{:}A \Downarrow\ \vdash x : ?\alpha}
+\end{mathpar}
+where
+\begin{mathpar}
+    \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'\ \textrm{means type}\
+    \tau'\ \textrm{is an \emph{existential instantiation} of scheme}\ \forall
+    \overline{\alpha}.\ \tau
+    \and
+    \textrm{ftv}_E(\tau')\ \textrm{is the set of free \emph{existential} type variables in type}\ \tau'
+    \and
+    ?\alpha \mapsto \tau_x\ \textrm{is a mapping from \emph{existential} type}\ ?\alpha\
+    \textrm{to type}\ \tau_x
+    \\
+    \textsc{solve}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
+    \textrm{can be solved with those in}\ \Theta
+\end{mathpar}
 
@@
 
-\section{Goals and Work Plan}
-
-We start with the definition of our type system and simply-typed
-functional language with linear function, sum and product types
-($\lolli$, $\tensor$, $\one$, $\with$, $\oplus$, $\bang$).  Through
-the Curry-Howard correspondence, we'll formulate the synthesis of
-typed programs in our functional language as bottom-up proof-search in
-linear logic.
-
-Since proof rules in \emph{natural deduction} do not entail a
-proof-search strategy, we will use the equivalent \emph{sequent
-calculus} system to formulate our typing rules in a way that is
-amenable to proof search.
-%
-In sequent calculus, rules can be naturally understood in a
-\emph{bottom-up} manner, simplifying the reasoning for the
-synthesizer. Moreover, the linear nature of our system requires a way
-to algorithmically reason about resource contexts: for instance, if we
-read rule $\tensor I$ from Section~\ref{sec:background} bottom-up, we
-must non-deterministically guess how to correctly split ambient
-resources into $\Delta_1$ and $\Delta_2$ to separately prove $A$ and
-$B$. The resource-management approach we'll adopt to solve this has been 
-previously explained in more detail.
-
-With the use of input/output contexts, and the sequent calculus
-formulation of the inference rules, we draw closer to an effective
-strategy for bottom-up proof search and so of program synthesis.
-%
-%
-However, the search process still encompasses significant
-non-determinism. To address this we draw on the technique of
-focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as
-discussed above.
-
-With the basic framework in place, we will consider additional
-(simply-typed) language constructs and synthesis of program fragments
-or \emph{partial synthesis} -- given a program with a typed hole,
-synthesize a term to fill out the hole accordingly. We will
-subsequently also consider synthesis of recursive functions and,
-might, later on, add refinement types also as a way to constrain the
-valid programs space.
-
-Putting together everything above, and the more detailed information
-in the background section -- linear types, sequent calculus, resource
-management, and focusing, we have a set of inference rules (written
-out in the appendix) that define how the synthesis process will
-progress.
-
-This is the theory we're interested in and have studied so far,
-which will allow us to build the actual synthetizer. From here,
-our work plan gives an overview of the steps we'll take.
-
-\subsection{Work Plan}
-
-The end goal is to have a synthetizer capable of
-taking a type (possibly refined) as input, and to produce (possibly
-interactively) one or more functions with that type, if any exist.
-
-As mentioned above, we will begin with (both partial and full)
-synthesis for the linear $\lambda$-calculus. We will then extend the
-language with additional type constructs (e.g. base types, records,
-algebraic data types) and consider synthesis in that setting. Finally,
-we will consider synthesis of recursive functions, and,
-time-permitting, we will consider simple forms of type refinements
-(e.g. arithmetic refinements).
-
-We've already implemented a parser and a type-checker for a linear
-$\lambda$-calculus using the resource management technique described
-in the previous section -- more than
-good preparation for the development of the synthesizer, they will be
-useful to \emph{validate} the synthesized functions, and to enable
-the synthesis of partial programs.
-%
-The implementations have been carried out in Haskell since it's easier
-to model and implement inference rules in a functional setting.
-
-\mypara{Step by step} Our work so far has been constantly accompanied
-by the study of somewhat dense theoretical material, state of the art,
-which we must understand to overcome challenges in these fields, in
-order to develop a more scalable synthetizer -- we want to display
-concrete results to prove the feasability of program synthesis from
-linear types, and thus open the possibility for future application to
-related fields, such as program synthesis based on session types for
-message-passing concurrency.
-
-Apart from the theory, when we started studying type systems, we
-created a simple type-checker for the simply-typed $\lambda$-calculus.
-While learning about linear type systems, and the simply-typed linear
-$\lambda$-calculus, we developed a parser, a desugaring module, and an
-evaluation module to write, analyze and execute simple programs in the
-language; culminating in writing a type-checker for the linear
-$\lambda$-calculus. At this point, we can parse, typecheck, and
-execute this small linear functional language.  The next step is to
-start developing the actual synthetizer, embedding it with all the
-techniques we've seen. Following work will be done to augment
-usability, expand the language domain, and add features such as those
-mentioned above.
+%\section{Goals and Work Plan}
+
+%We start with the definition of our type system and simply-typed
+%functional language with linear function, sum and product types
+%($\lolli$, $\tensor$, $\one$, $\with$, $\oplus$, $\bang$).  Through
+%the Curry-Howard correspondence, we'll formulate the synthesis of
+%typed programs in our functional language as bottom-up proof-search in
+%linear logic.
+
+%Since proof rules in \emph{natural deduction} do not entail a
+%proof-search strategy, we will use the equivalent \emph{sequent
+%calculus} system to formulate our typing rules in a way that is
+%amenable to proof search.
+%%
+%In sequent calculus, rules can be naturally understood in a
+%\emph{bottom-up} manner, simplifying the reasoning for the
+%synthesizer. Moreover, the linear nature of our system requires a way
+%to algorithmically reason about resource contexts: for instance, if we
+%read rule $\tensor I$ from Section~\ref{sec:background} bottom-up, we
+%must non-deterministically guess how to correctly split ambient
+%resources into $\Delta_1$ and $\Delta_2$ to separately prove $A$ and
+%$B$. The resource-management approach we'll adopt to solve this has been 
+%previously explained in more detail.
+
+%With the use of input/output contexts, and the sequent calculus
+%formulation of the inference rules, we draw closer to an effective
+%strategy for bottom-up proof search and so of program synthesis.
+%%
+%%
+%However, the search process still encompasses significant
+%non-determinism. To address this we draw on the technique of
+%focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as
+%discussed above.
+
+%With the basic framework in place, we will consider additional
+%(simply-typed) language constructs and synthesis of program fragments
+%or \emph{partial synthesis} -- given a program with a typed hole,
+%synthesize a term to fill out the hole accordingly. We will
+%subsequently also consider synthesis of recursive functions and,
+%might, later on, add refinement types also as a way to constrain the
+%valid programs space.
+
+%Putting together everything above, and the more detailed information
+%in the background section -- linear types, sequent calculus, resource
+%management, and focusing, we have a set of inference rules (written
+%out in the appendix) that define how the synthesis process will
+%progress.
+
+%This is the theory we're interested in and have studied so far,
+%which will allow us to build the actual synthetizer. From here,
+%our work plan gives an overview of the steps we'll take.
+
+%\subsection{Work Plan}
+
+%The end goal is to have a synthetizer capable of
+%taking a type (possibly refined) as input, and to produce (possibly
+%interactively) one or more functions with that type, if any exist.
+
+%As mentioned above, we will begin with (both partial and full)
+%synthesis for the linear $\lambda$-calculus. We will then extend the
+%language with additional type constructs (e.g. base types, records,
+%algebraic data types) and consider synthesis in that setting. Finally,
+%we will consider synthesis of recursive functions, and,
+%time-permitting, we will consider simple forms of type refinements
+%(e.g. arithmetic refinements).
+
+%We've already implemented a parser and a type-checker for a linear
+%$\lambda$-calculus using the resource management technique described
+%in the previous section -- more than
+%good preparation for the development of the synthesizer, they will be
+%useful to \emph{validate} the synthesized functions, and to enable
+%the synthesis of partial programs.
+%%
+%The implementations have been carried out in Haskell since it's easier
+%to model and implement inference rules in a functional setting.
+
+%\mypara{Step by step} Our work so far has been constantly accompanied
+%by the study of somewhat dense theoretical material, state of the art,
+%which we must understand to overcome challenges in these fields, in
+%order to develop a more scalable synthetizer -- we want to display
+%concrete results to prove the feasability of program synthesis from
+%linear types, and thus open the possibility for future application to
+%related fields, such as program synthesis based on session types for
+%message-passing concurrency.
+
+%Apart from the theory, when we started studying type systems, we
+%created a simple type-checker for the simply-typed $\lambda$-calculus.
+%While learning about linear type systems, and the simply-typed linear
+%$\lambda$-calculus, we developed a parser, a desugaring module, and an
+%evaluation module to write, analyze and execute simple programs in the
+%language; culminating in writing a type-checker for the linear
+%$\lambda$-calculus. At this point, we can parse, typecheck, and
+%execute this small linear functional language.  The next step is to
+%start developing the actual synthetizer, embedding it with all the
+%techniques we've seen. Following work will be done to augment
+%usability, expand the language domain, and add features such as those
+%mentioned above.
 

### 2021-07-11 article/article.tex
Write architecture and implementation start

@@
 % General motivation?
-Program synthesis is an automated or semi-automated process of
-deriving a program (i.e.~generating code) from a high-level
-specification.  
+Program synthesis is an automated or semi-automated process of deriving a
+program (i.e.~generating code) from a high-level specification.
 %
-Synthesis can be seen as a means to improve a programmer's
-productivity and/or program correctness (i.e. through suggestion
-and/or autocompletion), or as a tool to automate certain parts of the
-programming process (e.g. in the same way AI might generate some
-boilerplate text to aid a journalist, rather than to replace the
-journalist outright), just to name a few examples.
+Synthesis can be seen as a means to improve a programmer's productivity and/or
+program correctness (i.e. through suggestion and/or autocompletion), or as a
+tool to automate certain parts of the programming process, to name a few
+examples.
 
@@
 examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
-types~\cite{DBLP:conf/lopstr/HughesO20}).  Regardless of the kind
-of specification, program synthesizers have to deal with two main
-inherent sources of complexity: searching over a vast space of
-(potentially) valid programs, and interpreting user intent.
+types~\cite{DBLP:conf/lopstr/HughesO20}).  Regardless of the kind of
+specification, program synthesizers have to deal with two main inherent sources
+of complexity: searching over a vast space of (potentially) valid programs, and
+interpreting user intent.
 %
-In this work we will explore type-based synthesis where specifications
-take the form of linear types, which are types that limit resource
-usage in programs.  Concretely, we will explore the proof-theoretic
-foundations of linear types via the Curry-Howard correspondence with
-linear logic, viewing program synthesis under the lens of (linear
-logic) bottom-up proof search.
+In this work we'll explore type-based synthesis where specifications take the
+form of linear types, which limit resource usage in programs, in combination
+with polymorphism.  Concretely, we will explore the proof-theoretic foundations
+of linear types via the Curry-Howard correspondence with linear logic, viewing
+program synthesis under the lens of (linear logic) bottom-up proof search.
 
@@
 way of pruning the search space, and by using types the user 
-is given a more ``familiar'' specification. For instance, the type
+is given a more ``familiar'' specification interface. For instance, the type
 $\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
@@
 programs from types based on linear logic propositions (i.e.~linear
-types) by leveraging the Curry-Howard correspondence. It
-states that propositions in a logic have a direct
-mapping to types, and well-typed programs correspond to proofs of
-those propositions.  As such, having a type be a proposition in linear
-logic, we can relate a proof of it directly to a linear
-functional program — finding a proof is finding a program with that
-type. Formulating synthesis as proof search in linear logic
-allows us to inform our work with approaches from the relevant 
-literature.
+types) by leveraging the Curry-Howard correspondence.
+
+% It
+% states that propositions in a logic have a direct
+% mapping to types, and well-typed programs correspond to proofs of
+% those propositions.  As such, having a type be a proposition in linear
+% logic, we can relate a proof of it directly to a linear
+% functional program — finding a proof is finding a program with that
+% type. Formulating synthesis as proof search in linear logic
+% allows us to inform our work with approaches from the relevant 
+% literature.
 
@@
 %measurements (i.e. how fast can we synthesize a term of type $T$?').
+
+\section{Overview}
+
+Escrever introdução melhor que tenha isto como seguimento...e que siga duma
+introdução extendida
+
+At the core of the synthesis process, and of the developed synthetiser \synname,
+is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
+resource management. It connects the concepts seen in
+background~\ref{sec:background} -- a type system
+consisting of inference rules in sequent calculus for linear logic connectives;
+with resource management, and focusing, as formalized techniques to,
+respectively, solve resource related non-determinism, and to eliminate from the
+proof search don't-care non-determinism.  Guided solely by this system's formal
+representation\ref{sec:formal_system}, we can (and did) create a very robust
+synthetiser for a select part of the linear types that's able to synth
+programs successfully from e.g. the major linear logic equivalences. For example,
+given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
+the program $\lambda a \rightarrow \lambda b \rightarrow$ let $c \tensor d$ =
+$b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.
+
+The \synname\ programming language is naturally born alongside the synthetiser
+to create an environment for it. Synth goals are inserted in the program by use
+of a \emph{synth} keyword or syntatic \emph{mark}, indicating a program should
+be generated for a given, or inferred, type. Synth results are functions in this
+language's syntax. Additionally, it opens the possibility of synthetisis with
+context, that is, with knowledge of other functions and eventual structures
+defined in the same program.
+
+However, a programming language only with the canonical linear types lacks
+expressiveness. To enhance \synname\ and its language, synth more interesting
+programs -- and thus empirically proving the feasability of more relevant
+synthesis in a linear context, we extend the syntax and type system with
+recursiveness and richer
+types: -- polimorfic, refinements, adts parametrizáveis...  todo
+
+\section{Formal System}\label{sec:formal_system}
+
+The development of \synname\ was initially found on an almost direct reading of
+concrete inference rules and theoretical principles (~\cite{frank pfenning notes}); and work on its environment
+was also guided by formal systems and methods (e.g. the Hindley-Milner type
+inference~\cite{}).  However, with the addition of recursion and richer types to
+the synth and language, formal guidelines to support the implementation become
+scarcer, or more complex. Accordingly, rather than derived from explicit rules,
+solutions appear first to tame the extensions' inherent complex problems (such
+as infinite recursion), and implicit rules come only as a byproduct.
+
+With this section we intend to formalize the implicit rules that guide the
+synthesis of more expressive specifications, alongside the already explicit
+rules that model the core of the synth process, resulting in a \emph{sound} set
+of rules that hopes to serve as a source of ideas on taming synthesis of
+recursive functions with richer types in a linear context, as well as describe
+the system in enough detail for the synthetiser to be reproducible by a
+theory-driven implementation.
+
+Even so, note that \emph{soundness} and \emph{completeness} have two very
+different meanings, and our \emph{sound} set of rules only guarantees we won't
+synth incorrect programs, but still hides a possibly infinite number of correct
+proofs; and that the valid programs derivable through it reflect the somewhat
+subjective trade-offs we decided on. Different choices and approaches outside
+the core might lead to completely distinct synths and spaces of valid programs.
+
+This section presents the rules while explaining them, modifying some along the
+process. For a
+
+\mypara{Core Rules} The core system comprises of the sequent calculus rules for
+linear logic connectives with resource-management and focusing. As explained in\ref{sec:background}, focusing separates the proof search into two
+phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
+focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
+apply non-invertible rules, staying focused until either the focus is again an
+asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
+invertible rules, a third, ordered, context ($\Omega$) is needed to hold
+propositions resulting from decomposing connectives; propositions we'll later try to
+invert, moving them to the linear context ($\Delta$) when we fail to.
+
+Putting together linear logic, the Curry-Howard correspondence, resource
+management, and focusing, we get the following formal system (heavily inspired
+by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
+first rule is read: to synth a program of type $A \lolli B$ while inverting
+right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
+context $\Delta$, and ordered context $\Omega$, synth a program of type $B$ with
+an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
+resulting in the program $M$ and output\ref{sec:background} linear
+context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
+resulting program is $\lambda x \rightarrow M$ and the output linear context is
+$\Delta'$.
+
+We begin with the right invertible rules:
+\begin{mathpar}
+
+    % -o R
+    \infer*[right=($\lolli R$)]
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
+    \notin \Delta'}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x \rightarrow M : A
+    \lolli B \Uparrow}
+
+\and
+
+    % & R
+    \infer*[right=($\with R$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
+    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A
+    \with B \Uparrow}
+
+\end{mathpar}
+
+When we reach a non-invertible proposition on the right, we start inverting the
+$\Omega$ context. The rule to transition to inversion on the left is:
+\begin{mathpar}
+    \infer*[right=($\Uparrow R$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
+    right asynchronous}}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
+\end{mathpar}
+
+Followed by the left invertible rules:
+\begin{mathpar}
+
+    \infer*[right=($\tensor L$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega, y{:}A, z{:}B \Uparrow\ \vdash M : C
+    \and y,z \notin \Delta'}
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \tensor B \Uparrow\ \vdash\
+    \textrm{let}\ y \tensor z = x\ \textrm{in}\ M : C}
+
+\and
+
+    \infer*[right=($1 L$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}1 \Uparrow\ \vdash\ \textrm{let}\
+    \star =
+    x\ \textrm{in}\ M : C}
+
+\and
+
+    \mprset{flushleft}
+    \infer*[right=($\oplus L$)]
+    {
+    \Gamma ; \Delta/ \Delta' ; \Omega, y{:}A \Uparrow\ \vdash M : C \and
+    y \notin \Delta' \\
+    \Gamma ; \Delta/ \Delta'' ; \Omega, z{:}B \Uparrow\ \vdash N : C \\
+    z \notin \Delta'' \\
+    \Delta' = \Delta''
+    }
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \oplus B \Uparrow\ \vdash\
+    \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y \rightarrow M\ |\
+    \textrm{inr}\ z \rightarrow N : C}
+
+\and
+
+    % {TODO: Nao ha restrições sobre este? no código tinha que x não podia estar em
+    % delta mas não faz muito sentido hah}
+    \infer*[right=($\bang L$)]
+    {\Gamma, y{:}A ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
+    {\Gamma ; \Delta/\Delta ; \Omega, x{:}\bang A \Uparrow\ \vdash\
+    \textrm{let}\ \bang y = x\ \textrm{in}\ M : C}
+
+\end{mathpar}
+
+When we find a synchronous (i.e. non-invertible) proposition in $\Omega$,
+we simply move it to the linear context $\Delta$, and keep inverting on the left:
+\begin{mathpar}
+    \infer*[right=($\Uparrow L$)]
+    {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
+    \textrm{not left asynchronous}}
+    {\Gamma; \Delta/\Delta'; \Omega, A \Uparrow\ \vdash C}
+\end{mathpar}
+
+After inverting all the asynchronous propositions in $\Omega$ we'll reach a state
+where there are no more propositions to invert ($\Gamma'; \Delta'; \cdot
+\Uparrow\ \vdash C$). At this point, we want to \emph{focus} on a proposition.
+The focus object can be any of the three: the proposition on the right (the
+goal), a proposition from the linear $\Delta$ context, or a proposition from the
+unrestricted $\Gamma$ context. For these options we have three \emph{decision}
+rules:
+\begin{mathpar}
+    \infer*[right=(decideR)]
+    {\Gamma; \Delta/\Delta' \vdash C \Downarrow \and C\ \textrm{not atomic}}
+    {\Gamma; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
+\and
+    \infer*[right=(decideL)]
+    {\Gamma; \Delta/\Delta' ; A \Downarrow\ \vdash C}
+    {\Gamma; \Delta, A/\Delta';\cdot \Uparrow\ \vdash C}
+\and
+    \infer*[right=(decideL!)]
+    {\Gamma, A; \Delta/\Delta' ; A \Downarrow\ \vdash C}
+    {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
+\end{mathpar}
+
+The decision rules are naturally followed by either left or right focus rules:
+\begin{mathpar}
+    \infer*[right=($\tensor R$)]
+    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
+    : B \Downarrow}
+    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
+\and
+    \infer*[right=($1 R$)]
+    {  }
+    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
+\and
+    \infer*[right=($\oplus R_1$)]
+    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
+\and
+    \infer*[right=($\oplus R_2$)]
+    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
+\and
+    \infer*[right=($\bang R$)]
+    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
+    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
+\and
+%TODO: eu tenho uma assertion que garante que y não sai com Delta', mas agora
+%parece-me inutil?
+    \infer*[right=($\lolli L$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
+    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
+    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
+\and
+    \infer*[right=($\with L_1$)]
+    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
+\and
+    \infer*[right=($\with L_2$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
+\end{mathpar}
+Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
+asynchronous. If we're left focused on an atomic proposition we either instance
+the goal or fail. Otherwise the left focus is asynchronous and we can start
+inverting it. If we're right focused on a proposition that isn't right
+synchronous, we switch to inversion as well. Three rules model these conditions:
+\begin{mathpar}
+    \infer*[right=(init)]
+    {  }
+    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
+\and
+    \infer*[right=($\Downarrow L$)]
+    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
+    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
+\and
+    \infer*[right=($\Downarrow R$)]
+    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
+    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
+\end{mathpar}
+The rules written above together make the core of our synthetiser. Ahead, we'll
+present new rules that align and build on top of these to synth recursive
+programs from more expressive (richer) types.
+
+\mypara{Recursion} The main idea behind synthesis of recursive programs is the
+labeling of the main goal and the addition of its type, under that name, to the
+unrestricted context. That is, to synth a function of type $A \lolli B$ named
+\emph{foo}, the initial judgment can be written as
+\begin{mathpar}
+    \infer
+    {\dots}
+    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
+    A \lolli B \Uparrow}
+\end{mathpar}
+and, by definition, all subsequent inference rules will have
+($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.
+We can also force the usage of the recursive call by adding it not only to the
+unrestricted context, but to the linear one as well.
+
+% TODO: E como evitamos coisas trivialmente estúpidas?
+
+\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
+is a named sum type of any type, i.e. a named type instanceable by one of many
+tags (also known as constructors) that take exactly one fixed type, which might
+be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
+for constructors with an arbitrary number of parameters. In the \synname\
+language, the programmer can define custom ADTs; as an example, we show the
+definition of an ADT which might be constructed with zero, one, or two
+propositions of type $A$, using its syntax: $\textrm{data Container} =
+\textrm{None}\ 1\ \vert\ \textrm{One}\ A\ \vert\ \textrm{Two}\ (A * A)$.
+
+The semantics of the simple ADT relate to those of the plus ($\oplus$) type --
+both are additive disjunctions.  A proof of an ADT requires only a proof of one
+of the types that a constructor takes, similar to the way $\oplus$ requires only
+proof of either the left or right type it consists of. Analogously, a proof of
+$C$ by deconstructing an ADT stipulates that all types that can construct the
+ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
+the inference rules for a simple ADT mimic the $\oplus$ rules, for an
+arbitrary ($> 1$) number of types, rather than exactly two. Therefore, there's one
+left rule for ADTs, and an arbitrary number of right rules, one for each
+constructor:
+\begin{mathpar}
+    \infer*[right=(adtR)]
+    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
+\and
+    \mprset{flushleft}
+    \infer*[right=(adtL)]
+    {
+        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
+        \and
+        y_1 \notin \Delta'_1
+        \\
+        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
+        \\
+        y_2 \notin \Delta'_2
+        \\\\
+        \\ \dots
+        \\
+        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
+        \\
+        y_n \notin \Delta'_n
+        \\
+        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
+    }
+    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
+    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
+    \rightarrow M_n : C}
+\end{mathpar}
+where the ADT \emph{Foo} and its constructors stand for any ADT with form
+\[
+    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
+\dots\ \vert\ \textrm{Bar}_n\ X_n
+\]
+
+A more general (and interesting) formulation of ADTs says an ADT can be
+recursive (or "indutively defined"), meaning the type a constructor needs might
+be the type the constructor constructs. Seemingly small, this change has a big
+impact in the synthesis process. Take, for instance, the ADT defined as
+$\textrm{data Foo} = \textrm{Bar Foo}$, the synth goal $\te{Foo}
+\lolli C$, and its derivation:
+\begin{mathpar}
+    \infer*[right=($\lolli R$)]
+    {
+        \infer*[right=($\Uparrow R$)]
+        {
+            \infer*[right=(adtL)]
+            {
+                \infer*[right=(adtL)]
+                {
+                    \infer*[right=(adtL)]
+                    {\dots}
+                    {\Gamma; \Delta/\Delta'; \Omega, z{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ z\ 
+                    \textrm{of Bar}\ k \rightarrow \dots : C}
+                }
+                {\Gamma; \Delta/\Delta'; \Omega, y{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ y\ 
+                \textrm{of Bar}\ z \rightarrow \dots : C}
+            }
+            {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ x\ 
+            \textrm{of Bar}\ y \rightarrow \dots : C}
+        }
+        {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \vdash \dots : C
+        \Uparrow}
+    }
+    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
+    \te{Foo} \lolli C \Uparrow}
+\end{mathpar}
+Using our current system, we are to apply an infinite number of times the left
+ADT rule, never reaching a final state for the proof. Simetrically, the
+derivation for the synth goal $\te{Foo}$ doesn't have an end either.
+\begin{mathpar}
+    \infer*[left=(adtR)]
+    {
+        \infer*[left=(adtR)]
+        {
+            \infer*[left=(adtR)]
+            {
+                \infer*[left=(adtR)]
+                {\dots}
+                {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
+            }
+            {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
+        }
+        {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
+    }
+    {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo}
+    \Downarrow}
+\end{mathpar}
+To handle this situation, we impede the decomposition of an ADT in subsequent
+proofs of its branches, and to impede the construction of an ADT in a subsequent
+proof of a type to construct it. For this, we need two more contexts, $\Rho_C$
+for restrictions on construction and $\Rho_D$ for restrictions on
+deconstruction. For convenience, they are represented by a single $\Rho$ if
+unused. All rules must trivially propagate these. The ADT rules are explicitly
+modified to take this into consideration:
+\begin{mathpar}
+    \infer*[right=(adtR)]
+    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
+    \te{Foo} \notin \Rho_C}
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
+\and
+    \mprset{flushleft}
+    \infer*[right=(adtL)]
+    {
+        \te{Foo} \notin \Rho_D
+        \and
+        \Delta'_1 = \dots = \Delta'_n 
+        \\
+        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
+        \and
+        y_1 \notin \Delta'_1
+        \\\\
+        \\ \dots
+        \\\\
+        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
+        \and
+        y_n \notin \Delta'_n
+    }
+    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
+    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
+    \rightarrow M_n : C}
+\end{mathpar}
+where
+\begin{mathpar}
+    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
+    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C
+
+    \Rho'_D = \textrm{likewise}
+\end{mathpar}
+These modifications prevent the derivation from not ending for the two seen
+situations, however, they also greatly limit the derivable valid programs space,
+leaving the synthetiser pretty much unable to synth specifications with
+recursive types, which is so undesirable that we add three rules to complement the
+restrictions on construction and destruction of recursive types.
+
+First, after decomposing a recursive ADT, we must consume all constructor
+parameters, and since we can't decompose the self type any further because of a
+restriction, we should be able to use it from the linear context; even though,
+non-restricted types are not to take that course since they weren't unnaturally
+restricted, and because it would result in additional non-determinism and
+infinite recursion. Second, without any additional rules, an ADT in the linear
+context will loop back to the inversion context, and then loop again, going back and
+forth between the two contexts for infinity; instead, when focused on an ADT, we should
+either instance the goal (provided they're the same type), or switch to
+inversion if and only if its decomposition isn't restricted. The three
+following rules ensure this:
+\begin{mathpar}
+    \infer*[right=(adt$\Uparrow$L)]
+    {
+        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
+        \and
+        \te{Foo} \in \Rho_D
+    }
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
+    \and
+    \infer*[right=(adt-init)]
+    {  }
+    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
+    \and
+    \infer*[right=(adt$\Downarrow$L)]
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
+    \te{Foo} \and \te{Foo} \notin \Rho_D}
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
+\end{mathpar}
+% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
+% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
+% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
+% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
+% é agora inutil. Talvez seja demais adicionar à regra
+The ADT rules draw closer to a final form, but will still need a revision
+after the introduction of polymorfism and polymorfic parametrization on ADTs.
+
+\mypara{Polimorfic Types} A polymorfic specification is a type $X$ and a set of
+bound variables that quantify universally the type variables in $X$, called a
+\emph{scheme} ($\forall \overline{\alpha}.\ \tau$). Synthesis of a scheme is straightforward.
+First, \textbf{type variables are considered atomic types}, then, instantiate
+the bound vars of the scheme as described by the Hindley-Milner inference
+method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
+each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli \alpha$ could be instanced to
+$\alpha0 \lolli \alpha0$. We add a rule for this:
+\begin{mathpar}
+    \infer*[right=($\forall R$)]
+    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
+    \overline{\alpha}.\ \tau
+    \sqsubseteq \tau'}
+    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
+    \tau \Uparrow}
+\end{mathpar}
+where
+\[
+    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
+    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
+    \overline{\alpha}.\ \tau
+\]
+As such, the construction of a derivation in which, to some types, no specific rules can be applied (except for the \textsc{init} rule,
+a type variable can instance a goal that's a type variable of the same name),
+corresponds to the synthesis of a program where some expressions are treated
+agnostically -- nothing constrains their type, meaning the program is
+polymorphic * TODO: rewrite this?.
+
+The interesting challenge of polymorphism in synthesis is the usage of schemes
+from the unrestricted context.  To begin with, $\Gamma$ now holds both types and
+schemes. Consequently, after the rule \textsc{decideLeft!} is applied, we become
+left focused on either a type or a scheme. Since left focus on a type is already
+well formulated, only rules for left focus on a scheme are needed. Our algorithm
+instances bound type vars of the focused scheme with existential type variables,
+and the instanced type becomes the left focus. Inspired by the Hindley-Milner
+system, we also generate inference constraints on the existential type variables
+(postponing the decision of what type it should be to be used in the proof), and
+collect them in a new constraints context (TODO: should be a substitution not a
+context?) $\Theta$ that is propagated across derivation branches the same way
+the linear context is (by havin an input and output context ($\Theta/\Theta'$)).
+In contrast to HM, everytime a constraint is added it is solved against all
+other constraints -- a branch of the proof search is desired to fail as soon as
+possible; and more, two different existential vars do not unify against
+one another ($?\alpha \mapsto ?\beta$ fails). Note that we instance the scheme with \emph{existential} type
+variables ($\alpha$) rather than just type variables since the latter represent
+universal types in the synth, and the first represent an existential instance of
+a scheme, that might inclusively constraint normal type variables. Additionally,
+it's required that all existential type variables were assigned a type, and
+provided they were, we can forget the constraints generated from the usage of
+the scheme. These concepts are formalized with the rules:
+\begin{mathpar}
+    \infer*[right=($\forall L$)]
+    {
+        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
+        \\
+        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
+        \\
+        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
+    }
+    {\Theta/\Theta; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
+    \and
+    \infer*[right=($?L$)]
+    {\textsc{solve}(?\alpha
+    \mapsto C, \Theta)}
+    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
+    x{:}?\alpha \Downarrow\ \vdash x : C}
+    \and
+    \infer*[right=($\Downarrow L?$)]
+    {\textsc{solve}(?\alpha
+    \mapsto A, \Theta)}
+    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
+    x{:}A \Downarrow\ \vdash x : ?\alpha}
+\end{mathpar}
+where
+\begin{mathpar}
+    \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'\ \textrm{means type}\
+    \tau'\ \textrm{is an \emph{existential instantiation} of scheme}\ \forall
+    \overline{\alpha}.\ \tau
+    \and
+    \textrm{ftv}_E(\tau')\ \textrm{is the set of free \emph{existential} type variables in type}\ \tau'
+    \and
+    ?\alpha \mapsto \tau_x\ \textrm{is a mapping from \emph{existential} type}\ ?\alpha\
+    \textrm{to type}\ \tau_x
+    \\
+    \textsc{solve}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
+    \textrm{can be solved with those in}\ \Theta
+\end{mathpar}
+
+\mypara{Polymorfic ADTs}
+
+\mypara{Refinement Types}
+
+\section{Architecture}
+
+The \synname\ synthetiser operates as part of the pipeline that processes a
+full program in the \synname\ language, and is called to generate parts marked
+(by the syntax \emph{\{\{ ... \}\}}) for
+synthesis of said program. The main pipeline consists of:
+\[
+    \textrm{Parsing} \rightarrow \textrm{Desugaring} \rightarrow \textrm{Inference} \rightarrow \textrm{Synthesis}
+\]
+
+The parsing module converts ADT declarations and top level functions written in
+the language's syntax to a list of abstract syntax trees (ASTs) understood by the rest of the
+pipeline.
+
+The desugaring module converts the frontend AST to a core AST without syntatic
+sugar and using the locally-nameless~\cite{locally nameless} representation for
+all bound variables. This way inference can be done without worrying about name
+conflicts.
+
+The inference module will traverse the list of ASTs in order (top-down), and
+infer the type of all functions defined by the user, checking the inferred
+type against a possible user given type annotation. Whenever a synth mark is
+found in the AST, all the inferrence context up to that point is added to the
+mark, and if not explicit, its type is inferred, essentially defining the synth
+context and synth goal.
+
+The synthesis module iterates over the synth marks and runs the synthetiser with
+its context and goal for each, minimizing the resulting expression, and
+replacing the mark in the frontend AST with the synthed program.
+
+\subsection{Implementation}
+
+\mypara{General} The complete implementation of the synthetiser and environment
+pipeline was done in Haskell...
+
+\mypara{Backtracking} Even though the synthetiser formal system helps, there's
+still non-determinism in the synth process. At diverse times in the proof
+construction we must choose one of multiple rules to apply, and in the event of
+not being able to construct the proof following that choice, we want to
+backtrack to the decision point and attempt to create a proof taking a different
+route. To this effect, our prototype synthetiser depends on \emph{logict}, "a backtracking logic-programming
+monad"~\cite{logict}.
+
+\mypara{SMT Solving} To typecheck and synth refinement types and dependent
+functions, we need to contact an SMT solver to check satisfiability. For this,
+we use the library \emph{sbv}~\cite{sbv}. Unfortunately, this library does not
+support the usage of the SMT solvers to satisfy and output more complicated
+first-order logic solutions -- since this is needed to synth refinement types
+when right focused, we instead use unsafe library primitives to communicate with
+the solver directly, constructing the logical formulas ourselves. For 
+satisfying formulas, for instance, when typechecking, or left focused on a
+refinement type, we use the library api successfully to get a veridct on
+satisfiability.
+
+\mypara{Debugging} The rules described in the formal
+system\ref{sec:formal_system} to handle infinite recursion might seem obvious in
+retrospective, however, when faced with a program that doesn't terminate, what
+went wrong is not so clear. While we were able to figure out a solution for the
+first couple of "infinite examples", the following problems were much harder,
+taking sometimes full days to solve (and as of now, some still remain unsolved,
+e.g. List (List a) -o List a). To make the debugging experience better, we
+developed a tracing system that prints information whenever a rule is applied,
+alongside the stack of rules applied up until that one. This improved our
+debugging speed considerably, especially when put together with a correct
+derivation done by hand, with which we could match what was happening against
+what was supposed to happen.
+
+Unfortunately, it seems that sometimes in the process of backtracking, stack
+information is "lost", and the trace logs lose useability.
+
+\mypara{Interface} A simple, syntax-highlighted, web interface was developed
+alongside a server to make possible experimenting with the \synname\ 
+language without having to download and compile the complete toolchain. The web
+interface is available from the repository~\cite{github repo}.
+
+% A implementação foi feita em haskell e o código está disponível em X e há um
+% live demo com exemplos pré-feitos em Y?
+
+\section{Related Work}
+
+Type-based program synthesis is a vast field of study and so it
+follows that a lot of literature is available to inspire and
+complement our work. Most works~\cite{DBLP:conf/lopstr/HughesO20,DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16} follow some variation of the
+synthesis-as-proof-search approach, however, the process is novel for
+each, due to a variety of different rich types explored, and their
+respective corresponding logics and programming languages; or nuances
+of the synthesis process itself, such as complementing types with
+program examples; or even the programming paradigm of the output
+produced (e.g. generating heap manipulating
+programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).  Some other
+common patterns we might want to follow are the use of type
+refinements~\cite{DBLP:conf/pldi/PolikarpovaKS16} and the support for synthesis of recursive
+functions~\cite{DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15}.  These works share some common ground, and base
+themselves of the same works we do -- most famously the Curry-Howard
+correspondence, but also, for example, works on focusing-based
+proof-search~\cite{10.1093/logcom/2.3.297}.
+
+\mypara{Type-and-Example-Directed Program Synthesis} This
+work~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
+explores a purely functional, recursive program synthesis technique
+based on types/proof search (as we intend to do), with examples as an
+auxiliar technique to trim down the program synthesis vast search
+space. A good use is
+made of the general strategy of using a type system to generate
+programs (by ``inverting'' the type system), rather than using it to
+type-check one.  Moreover, the paper uses a data structure called
+``refinement tree'' in conjunction with the extra information provided
+by the examples to allow for efficient synthesis of non-trivial
+programs. Overall it employs good engineering to achieve type-based
+synthesis of functional typed programs, with at least two additions
+that we might want to follow too (type refinements and recursive
+functions).
+
+% or how to turn your type system upside down ~ não apagar a Kubrick reference? :)
+
+\mypara{Program Synthesis from Polymorphic Refinement Types} The
+work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis
+recursive functional programs but in a more ``advanced'' setting. The
+specification is a combination of two rich forms of types: polymorphic and
+refinement types (which correspond to a first-order logic through the
+Curry-Howard isomorphism) -- this is an interesting combination that offers
+``expressive power and decidability, which enables automatic
+verification--and hence synthesis--of nontrivial programs''.  Their
+approach to refinement types consists a new algorithm that supports
+decomposition of the refinement specification, allowing for separation
+between the language of specification and programs and making the
+approach amenable to compositional synthesis techniques.
+
+\mypara{Resourceful Program Synthesis from Graded Linear Types} The
+work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
+approach very similar to our own.  The work employs so-called graded
+modal types, which is an refinement of pure linear types -- a more
+\emph{fine-grained} version, since it allows for quantitative
+specification of resource usage, in contrast to ours either
+\emph{linear} or \emph{unrestricted} (via the linear logic
+exponential) use of assumptions.  Their resource management is more
+complex, and so they provide solutions which adapt Hodas and Miller's
+resource management
+approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
+-- the model we will be using.
+
+They also use focusing as a solution to trim down search space and to
+ensure that synthesis only produces well-typed programs, and our
+solution will use the same literature as a base.  However, since their
+underlying logic is \emph{modal} rather than purely \emph{linear}, it
+lacks a clear correspondence with concurrent session-typed
+programs~\cite{DBLP:journals/mscs/CairesPT16,DBLP:conf/concur/CairesP10},
+which is a crucial avenue of future work. Moreover, their use grading
+effectively requires constraint solving to be integrated with the
+synthesis procedure, which can limit the effectiveness of the overall
+approach as one scales to more sophisticated settings (e.g.~refinement
+types).
+
+% \item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}
+% This resource-guided means something a bit different. Programs satisfy
+% a functional specification and a symbolic resource bound in the sense
+% of amortized analysis, but can provide technical insights.
+
+%\section{Goals and Work Plan}
+
+%We start with the definition of our type system and simply-typed
+%functional language with linear function, sum and product types
+%($\lolli$, $\tensor$, $\one$, $\with$, $\oplus$, $\bang$).  Through
+%the Curry-Howard correspondence, we'll formulate the synthesis of
+%typed programs in our functional language as bottom-up proof-search in
+%linear logic.
+
+%Since proof rules in \emph{natural deduction} do not entail a
+%proof-search strategy, we will use the equivalent \emph{sequent
+%calculus} system to formulate our typing rules in a way that is
+%amenable to proof search.
+%%
+%In sequent calculus, rules can be naturally understood in a
+%\emph{bottom-up} manner, simplifying the reasoning for the
+%synthesizer. Moreover, the linear nature of our system requires a way
+%to algorithmically reason about resource contexts: for instance, if we
+%read rule $\tensor I$ from Section~\ref{sec:background} bottom-up, we
+%must non-deterministically guess how to correctly split ambient
+%resources into $\Delta_1$ and $\Delta_2$ to separately prove $A$ and
+%$B$. The resource-management approach we'll adopt to solve this has been 
+%previously explained in more detail.
+
+%With the use of input/output contexts, and the sequent calculus
+%formulation of the inference rules, we draw closer to an effective
+%strategy for bottom-up proof search and so of program synthesis.
+%%
+%%
+%However, the search process still encompasses significant
+%non-determinism. To address this we draw on the technique of
+%focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as
+%discussed above.
+
+%With the basic framework in place, we will consider additional
+%(simply-typed) language constructs and synthesis of program fragments
+%or \emph{partial synthesis} -- given a program with a typed hole,
+%synthesize a term to fill out the hole accordingly. We will
+%subsequently also consider synthesis of recursive functions and,
+%might, later on, add refinement types also as a way to constrain the
+%valid programs space.
+
+%Putting together everything above, and the more detailed information
+%in the background section -- linear types, sequent calculus, resource
+%management, and focusing, we have a set of inference rules (written
+%out in the appendix) that define how the synthesis process will
+%progress.
+
+%This is the theory we're interested in and have studied so far,
+%which will allow us to build the actual synthetizer. From here,
+%our work plan gives an overview of the steps we'll take.
+
+%\subsection{Work Plan}
+
+%The end goal is to have a synthetizer capable of
+%taking a type (possibly refined) as input, and to produce (possibly
+%interactively) one or more functions with that type, if any exist.
+
+%As mentioned above, we will begin with (both partial and full)
+%synthesis for the linear $\lambda$-calculus. We will then extend the
+%language with additional type constructs (e.g. base types, records,
+%algebraic data types) and consider synthesis in that setting. Finally,
+%we will consider synthesis of recursive functions, and,
+%time-permitting, we will consider simple forms of type refinements
+%(e.g. arithmetic refinements).
+
+%We've already implemented a parser and a type-checker for a linear
+%$\lambda$-calculus using the resource management technique described
+%in the previous section -- more than
+%good preparation for the development of the synthesizer, they will be
+%useful to \emph{validate} the synthesized functions, and to enable
+%the synthesis of partial programs.
+%%
+%The implementations have been carried out in Haskell since it's easier
+%to model and implement inference rules in a functional setting.
+
+%\mypara{Step by step} Our work so far has been constantly accompanied
+%by the study of somewhat dense theoretical material, state of the art,
+%which we must understand to overcome challenges in these fields, in
+%order to develop a more scalable synthetizer -- we want to display
+%concrete results to prove the feasability of program synthesis from
+%linear types, and thus open the possibility for future application to
+%related fields, such as program synthesis based on session types for
+%message-passing concurrency.
+
+%Apart from the theory, when we started studying type systems, we
+%created a simple type-checker for the simply-typed $\lambda$-calculus.
+%While learning about linear type systems, and the simply-typed linear
+%$\lambda$-calculus, we developed a parser, a desugaring module, and an
+%evaluation module to write, analyze and execute simple programs in the
+%language; culminating in writing a type-checker for the linear
+%$\lambda$-calculus. At this point, we can parse, typecheck, and
+%execute this small linear functional language.  The next step is to
+%start developing the actual synthetizer, embedding it with all the
+%techniques we've seen. Following work will be done to augment
+%usability, expand the language domain, and add features such as those
+%mentioned above.
+
+\mypara{Evaluation} Validation and evaluation of our work is fairly
+easy. Simple questions such as \emph{does it typecheck}? \emph{can
+  we synthesize}? \emph{how fast}?  are the core of our evaluation. In
+more concrete terms, we can compare our times to state of the art
+synthetizers, present a list of working examples of programs synthesized
+by our framework, and corresponding benchmarks.
+
+On the eventual addition of recursive functions, refinement types, and
+interactive synthesis, we'll consider, respectively, other examples to
+showcase recursive functions synthesis, updated times for benchmarks
+with refinement types, and a display of the possible user interaction
+in the user process, and the resulting function following it.
+
+% Expand on earlier points. Start from a functional language with linear types
+% ($\tensor$, $\lolli$, $\oplus$, $\bang$); build on it with more ``stuff''.
+% General techniques, drawn from proof theory via props as types: explore
+% focusing to tame non-determinism/search space. Partial synthesis is still
+% proof search! Add other techniques as we go along (e.g. for recursion we need
+% to constrain recursive calls).
+
+% Mention that you've already implemented a type-checker for this (useful as a
+% prelim. Exercise but also later, for \emph{validation}).
+
+% Validation and evaluation: validation is as simple as ``does it typecheck''?
+% can you synth? how fast?
+
+\section{Evaluation}
+
+% Adicionar já exemplos do que consigo
 
@@
 
-\mypara{Resource-Management}\label{sec:background_resource_management}
+\mypara{Resource-Management}
 Be it from the perspective of proof \emph{checking} (i.e.~type checking) or
@@
 
-\mypara{Focusing}\label{sec:background_focusing}
+\mypara{Focusing}
 
@@
 
-\section{Overview}
-
-Escrever introdução melhor que tenha isto como seguimento...
-
-At the core of the synthesis process, and of the developed synthetiser \synname,
-is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
-resource management. It connects the concepts seen above -- a type system
-consisting of inference rules in sequent calculus for linear logic connectives;
-with resource management, and focusing, as formalized techniques to,
-respectively, solve resource related non-determinism, and to eliminate from the
-proof search don't-care non-determinism.  Guided solely by this system's formal
-representation\ref{sec:formal_system}, we can (and did) create a very robust
-synthetiser for a select part of the linear types that's able to synth
-programs successfully from e.g. the major linear logic equivalences. For example,
-given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
-the program $\lambda a \rightarrow \lambda b \rightarrow$ let $c \tensor d$ =
-$b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.
-
-The \synname\ programming language is naturally born alongside the synthetiser
-to create an environment for it. Synth goals are inserted in the program by use
-of a \emph{synth} keyword or syntatic \emph{mark}, indicating a program should
-be generated for a given, or inferred, type. Synth results are functions in this
-language's syntax. Additionally, it opens the possibility of synthetisis with
-context, that is, with knowledge of other functions and eventual structures
-defined in the same program.
-
-However, a programming language only with the canonical linear types lacks
-expressiveness. To enhance \synname\ and its language, synth more interesting
-programs -- and thus empirically proving the feasability of more relevant
-synthesis in a linear context, we extend the syntax and type system with
-recursiveness and richer
-types: -- polimorfic, refinements, adts parametrizáveis...  todo
-
-\section{Architecture}
-
-\subsection{The \synname\ language}
-
-\subsection{Implementation}
-
-\mypara{Debugging} The rules described in the formal
-system\ref{sec:formal_system} to handle infinite recursion might seem obvious in
-retrospective, however, when faced with a program that doesn't terminate, what
-went wrong is not so clear. While we were able to figure out a solution for the
-first couple of "infinite examples", the following problems were much harder,
-taking sometimes full days to solve (and as of now, some still remain unsolved,
-e.g. List (List a) -o List a). To make the debugging experience better, we
-developed a tracing system that prints information whenever a rule is applied,
-alongside the stack of rules applied up until that one. This improved our
-debugging speed considerably, especially when put together with a correct
-derivation done by hand, with which we could match what was happening against
-what was supposed to happen.
-
-Unfortunately, it seems that sometimes in the process of backtracking, stack
-information is "lost", and the trace logs lose useability.
-
-% A implementação foi feita em haskell e o código está disponível em X e há um
-% live demo com exemplos pré-feitos em Y?
-
-\section{Formal System}\label{sec:formal_system}
-
-The development of \synname\ was initially found on an almost direct reading of
-concrete inference rules and theoretical principles (~\cite{frank pfenning notes}); and work on its environment
-was also guided by formal systems and methods (e.g. the Hindley-Milner type
-inference~\cite{}).  However, with the addition of recursion and richer types to
-the synth and language, formal guidelines to support the implementation become
-scarcer, or more complex. Accordingly, rather than derived from explicit rules,
-solutions appear first to tame the extensions' inherent complex problems (such
-as infinite recursion), and implicit rules come only as a byproduct.
-
-With this section we intend to formalize the implicit rules that guide the
-synthesis of more expressive specifications, alongside the already explicit
-rules that model the core of the synth process, resulting in a \emph{sound} set
-of rules that hopes to serve as a source of ideas on taming synthesis of
-recursive functions with richer types in a linear context, as well as describe
-the system in enough detail for the synthetiser to be reproducible by a
-theory-driven implementation.
-
-Even so, note that \emph{soundness} and \emph{completeness} have two very
-different meanings, and our \emph{sound} set of rules only guarantees we won't
-synth incorrect programs, but still hides a possibly infinite number of correct
-proofs; and that the valid programs derivable through it reflect the somewhat
-subjective trade-offs we decided on. Different choices and approaches outside
-the core might lead to completely distinct synths and spaces of valid programs.
-
-This section presents the rules while explaining them, modifying some along the
-process. For a
-
-\mypara{Core Rules} The core system comprises of the sequent calculus rules for
-linear logic connectives with resource-management and focusing. As seen
-previously\ref{sec:background}, focusing separates the proof search into two
-phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
-focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
-apply non-invertible rules, staying focused until either the focus is again an
-asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
-invertible rules, a third, ordered, context ($\Omega$) is needed to hold
-propositions resulting from decomposing connectives; propositions we'll later try to
-invert, moving them to the linear context ($\Delta$) when we fail to.
-
-Putting together linear logic, the Curry-Howard correspondence, resource
-management, and focusing, we get the following formal system (heavily inspired
-by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
-first rule is read: to synth a program of type $A \lolli B$ while inverting
-right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
-context $\Delta$, and ordered context $\Omega$, synth a program of type $B$ with
-an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
-resulting in the program $M$ and output\ref{sec:background} linear
-context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
-resulting program is $\lambda x \rightarrow M$ and the output linear context is
-$\Delta'$.
-
-We begin with the right invertible rules:
-\begin{mathpar}
-
-    % -o R
-    \infer*[right=($\lolli R$)]
-    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
-    \notin \Delta'}
-    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x \rightarrow M : A
-    \lolli B \Uparrow}
-
-\and
-
-    % & R
-    \infer*[right=($\with R$)]
-    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
-    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
-    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A
-    \with B \Uparrow}
-
-\end{mathpar}
-
-When we reach a non-invertible proposition on the right, we start inverting the
-$\Omega$ context. The rule to transition to inversion on the left is:
-\begin{mathpar}
-    \infer*[right=($\Uparrow R$)]
-    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
-    right asynchronous}}
-    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
-\end{mathpar}
-
-Followed by the left invertible rules:
-\begin{mathpar}
-
-    \infer*[right=($\tensor L$)]
-    {\Gamma ; \Delta/ \Delta' ; \Omega, y{:}A, z{:}B \Uparrow\ \vdash M : C
-    \and y,z \notin \Delta'}
-    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \tensor B \Uparrow\ \vdash\
-    \textrm{let}\ y \tensor z = x\ \textrm{in}\ M : C}
-
-\and
-
-    \infer*[right=($1 L$)]
-    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
-    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}1 \Uparrow\ \vdash\ \textrm{let}\
-    \star =
-    x\ \textrm{in}\ M : C}
-
-\and
-
-    \mprset{flushleft}
-    \infer*[right=($\oplus L$)]
-    {
-    \Gamma ; \Delta/ \Delta' ; \Omega, y{:}A \Uparrow\ \vdash M : C \and
-    y \notin \Delta' \\
-    \Gamma ; \Delta/ \Delta'' ; \Omega, z{:}B \Uparrow\ \vdash N : C \\
-    z \notin \Delta'' \\
-    \Delta' = \Delta''
-    }
-    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \oplus B \Uparrow\ \vdash\
-    \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y \rightarrow M\ |\
-    \textrm{inr}\ z \rightarrow N : C}
-
-\and
-
-    % {TODO: Nao ha restrições sobre este? no código tinha que x não podia estar em
-    % delta mas não faz muito sentido hah}
-    \infer*[right=($\bang L$)]
-    {\Gamma, y{:}A ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
-    {\Gamma ; \Delta/\Delta ; \Omega, x{:}\bang A \Uparrow\ \vdash\
-    \textrm{let}\ \bang y = x\ \textrm{in}\ M : C}
-
-\end{mathpar}
-
-When we find a synchronous (i.e. non-invertible) proposition in $\Omega$,
-we simply move it to the linear context $\Delta$, and keep inverting on the left:
-\begin{mathpar}
-    \infer*[right=($\Uparrow L$)]
-    {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
-    \textrm{not left asynchronous}}
-    {\Gamma; \Delta/\Delta'; \Omega, A \Uparrow\ \vdash C}
-\end{mathpar}
-
-After inverting all the asynchronous propositions in $\Omega$ we'll reach a state
-where there are no more propositions to invert ($\Gamma'; \Delta'; \cdot
-\Uparrow\ \vdash C$). At this point, we want to \emph{focus} on a proposition.
-The focus object can be any of the three: the proposition on the right (the
-goal), a proposition from the linear $\Delta$ context, or a proposition from the
-unrestricted $\Gamma$ context. For these options we have three \emph{decision}
-rules:
-\begin{mathpar}
-    \infer*[right=(decideR)]
-    {\Gamma; \Delta/\Delta' \vdash C \Downarrow \and C\ \textrm{not atomic}}
-    {\Gamma; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
-\and
-    \infer*[right=(decideL)]
-    {\Gamma; \Delta/\Delta' ; A \Downarrow\ \vdash C}
-    {\Gamma; \Delta, A/\Delta';\cdot \Uparrow\ \vdash C}
-\and
-    \infer*[right=(decideL!)]
-    {\Gamma, A; \Delta/\Delta' ; A \Downarrow\ \vdash C}
-    {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
-\end{mathpar}
-
-The decision rules are naturally followed by either left or right focus rules:
-\begin{mathpar}
-    \infer*[right=($\tensor R$)]
-    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
-    : B \Downarrow}
-    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
-\and
-    \infer*[right=($1 R$)]
-    {  }
-    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
-\and
-    \infer*[right=($\oplus R_1$)]
-    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
-    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
-\and
-    \infer*[right=($\oplus R_2$)]
-    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
-    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
-\and
-    \infer*[right=($\bang R$)]
-    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
-    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
-\and
-%TODO: eu tenho uma assertion que garante que y não sai com Delta', mas agora
-%parece-me inutil?
-    \infer*[right=($\lolli L$)]
-    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
-    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
-    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
-\and
-    \infer*[right=($\with L_1$)]
-    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
-    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
-\and
-    \infer*[right=($\with L_2$)]
-    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
-    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
-\end{mathpar}
-Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
-asynchronous. If we're left focused on an atomic proposition we either instance
-the goal or fail. Otherwise the left focus is asynchronous and we can start
-inverting it. If we're right focused on a proposition that isn't right
-synchronous, we switch to inversion as well. Three rules model these conditions:
-\begin{mathpar}
-    \infer*[right=(init)]
-    {  }
-    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
-\and
-    \infer*[right=($\Downarrow L$)]
-    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
-    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
-\and
-    \infer*[right=($\Downarrow R$)]
-    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
-    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
-\end{mathpar}
-The rules written above together make the core of our synthetiser. Ahead, we'll
-present new rules that align and build on top of these to synth recursive
-programs from more expressive (richer) types.
-
-\mypara{Recursion} The main idea behind synthesis of recursive programs is the
-labeling of the main goal and the addition of its type, under that name, to the
-unrestricted context. That is, to synth a function of type $A \lolli B$ named
-\emph{foo}, the initial judgment can be written as
-\begin{mathpar}
-    \infer
-    {\dots}
-    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
-    A \lolli B \Uparrow}
-\end{mathpar}
-and, by definition, all subsequent inference rules will have
-($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.
-We can also force the usage of the recursive call by adding it not only to the
-unrestricted context, but to the linear one as well.
-
-% TODO: E como evitamos coisas trivialmente estúpidas?
-
-\mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
-is a named sum type of any type, i.e. a named type instanceable by one of many
-tags (also known as constructors) that take exactly one fixed type, which might
-be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
-for constructors with an arbitrary number of parameters. In the \synname\
-language, the programmer can define custom ADTs; as an example, we show the
-definition of an ADT which might be constructed with zero, one, or two
-propositions of type $A$, using its syntax: $\textrm{data Container} =
-\textrm{None}\ 1\ \vert\ \textrm{One}\ A\ \vert\ \textrm{Two}\ (A * A)$.
-
-The semantics of the simple ADT relate to those of the plus ($\oplus$) type --
-both are additive disjunctions.  A proof of an ADT requires only a proof of one
-of the types that a constructor takes, similar to the way $\oplus$ requires only
-proof of either the left or right type it consists of. Analogously, a proof of
-$C$ by deconstructing an ADT stipulates that all types that can construct the
-ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
-the inference rules for a simple ADT mimic the $\oplus$ rules, for an
-arbitrary ($> 1$) number of types, rather than exactly two. Therefore, there's one
-left rule for ADTs, and an arbitrary number of right rules, one for each
-constructor:
-\begin{mathpar}
-    \infer*[right=(adtR)]
-    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
-    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
-\and
-    \mprset{flushleft}
-    \infer*[right=(adtL)]
-    {
-        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
-        \and
-        y_1 \notin \Delta'_1
-        \\
-        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
-        \\
-        y_2 \notin \Delta'_2
-        \\\\
-        \\ \dots
-        \\
-        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
-        \\
-        y_n \notin \Delta'_n
-        \\
-        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
-    }
-    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
-    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
-    \rightarrow M_n : C}
-\end{mathpar}
-where the ADT \emph{Foo} and its constructors stand for any ADT with form
-\[
-    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
-\dots\ \vert\ \textrm{Bar}_n\ X_n
-\]
-
-A more general (and interesting) formulation of ADTs says an ADT can be
-recursive (or "indutively defined"), meaning the type a constructor needs might
-be the type the constructor constructs. Seemingly small, this change has a big
-impact in the synthesis process. Take, for instance, the ADT defined as
-$\textrm{data Foo} = \textrm{Bar Foo}$, the synth goal $\te{Foo}
-\lolli C$, and its derivation:
-\begin{mathpar}
-    \infer*[right=($\lolli R$)]
-    {
-        \infer*[right=($\Uparrow R$)]
-        {
-            \infer*[right=(adtL)]
-            {
-                \infer*[right=(adtL)]
-                {
-                    \infer*[right=(adtL)]
-                    {\dots}
-                    {\Gamma; \Delta/\Delta'; \Omega, z{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ z\ 
-                    \textrm{of Bar}\ k \rightarrow \dots : C}
-                }
-                {\Gamma; \Delta/\Delta'; \Omega, y{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ y\ 
-                \textrm{of Bar}\ z \rightarrow \dots : C}
-            }
-            {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ x\ 
-            \textrm{of Bar}\ y \rightarrow \dots : C}
-        }
-        {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \vdash \dots : C
-        \Uparrow}
-    }
-    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
-    \te{Foo} \lolli C \Uparrow}
-\end{mathpar}
-Using our current system, we are to apply an infinite number of times the left
-ADT rule, never reaching a final state for the proof. Simetrically, the
-derivation for the synth goal $\te{Foo}$ doesn't have an end either.
-\begin{mathpar}
-    \infer*[left=(adtR)]
-    {
-        \infer*[left=(adtR)]
-        {
-            \infer*[left=(adtR)]
-            {
-                \infer*[left=(adtR)]
-                {\dots}
-                {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
-            }
-            {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
-        }
-        {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
-    }
-    {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo}
-    \Downarrow}
-\end{mathpar}
-To handle this situation, we impede the decomposition of an ADT in subsequent
-proofs of its branches, and to impede the construction of an ADT in a subsequent
-proof of a type to construct it. For this, we need two more contexts, $\Rho_C$
-for restrictions on construction and $\Rho_D$ for restrictions on
-deconstruction. For convenience, they are represented by a single $\Rho$ if
-unused. All rules must trivially propagate these. The ADT rules are explicitly
-modified to take this into consideration:
-\begin{mathpar}
-    \infer*[right=(adtR)]
-    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
-    \te{Foo} \notin \Rho_C}
-    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
-\and
-    \mprset{flushleft}
-    \infer*[right=(adtL)]
-    {
-        \te{Foo} \notin \Rho_D
-        \and
-        \Delta'_1 = \dots = \Delta'_n 
-        \\
-        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
-        \and
-        y_1 \notin \Delta'_1
-        \\\\
-        \\ \dots
-        \\\\
-        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
-        \and
-        y_n \notin \Delta'_n
-    }
-    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
-    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
-    \rightarrow M_n : C}
-\end{mathpar}
-where
-\begin{mathpar}
-    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
-    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C
-
-    \Rho'_D = \textrm{likewise}
-\end{mathpar}
-These modifications prevent the derivation from not ending for the two seen
-situations, however, they also greatly limit the derivable valid programs space,
-leaving the synthetiser pretty much unable to synth specifications with
-recursive types, which is so undesirable that we add three rules to complement the
-restrictions on construction and destruction of recursive types.
-
-First, after decomposing a recursive ADT, we must consume all constructor
-parameters, and since we can't decompose the self type any further because of a
-restriction, we should be able to use it from the linear context; even though,
-non-restricted types are not to take that course since they weren't unnaturally
-restricted, and because it would result in additional non-determinism and
-infinite recursion. Second, without any additional rules, an ADT in the linear
-context will loop back to the inversion context, and then loop again, going back and
-forth between the two contexts for infinity; instead, when focused on an ADT, we should
-either instance the goal (provided they're the same type), or switch to
-inversion if and only if its decomposition isn't restricted. The three
-following rules ensure this:
-\begin{mathpar}
-    \infer*[right=(adt$\Uparrow$L)]
-    {
-        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
-        \and
-        \te{Foo} \in \Rho_D
-    }
-    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
-    \and
-    \infer*[right=(adt-init)]
-    {  }
-    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
-    \and
-    \infer*[right=(adt$\Downarrow$L)]
-    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
-    \te{Foo} \and \te{Foo} \notin \Rho_D}
-    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
-\end{mathpar}
-% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
-% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
-% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
-% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
-% é agora inutil. Talvez seja demais adicionar à regra
-The ADT rules draw closer to a final form, but will still need a revision
-after the introduction of polymorfism and polymorfic parametrization on ADTs.
-
-\mypara{Polimorfic Types} A polymorfic specification is a type $X$ and a set of
-bound variables that quantify universally the type variables in $X$, called a
-\emph{scheme} ($\forall \overline{\alpha}.\ \tau$). Synthesis of a scheme is straightforward.
-First, \textbf{type variables are considered atomic types}, then, instantiate
-the bound vars of the scheme as described by the Hindley-Milner inference
-method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
-each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli \alpha$ could be instanced to
-$\alpha0 \lolli \alpha0$. We add a rule for this:
-\begin{mathpar}
-    \infer*[right=($\forall R$)]
-    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
-    \overline{\alpha}.\ \tau
-    \sqsubseteq \tau'}
-    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
-    \tau \Uparrow}
-\end{mathpar}
-where
-\[
-    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
-    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
-    \overline{\alpha}.\ \tau
-\]
-As such, the construction of a derivation in which, to some types, no specific rules can be applied (except for the \textsc{init} rule,
-a type variable can instance a goal that's a type variable of the same name),
-corresponds to the synthesis of a program where some expressions are treated
-agnostically -- nothing constrains their type, meaning the program is
-polymorphic * TODO: rewrite this?.
-
-The interesting challenge of polymorphism in synthesis is the usage of schemes
-from the unrestricted context.  To begin with, $\Gamma$ now holds both types and
-schemes. Consequently, after the rule \textsc{decideLeft!} is applied, we become
-left focused on either a type or a scheme. Since left focus on a type is already
-well formulated, only rules for left focus on a scheme are needed. Our algorithm
-instances bound type vars of the focused scheme with existential type variables,
-and the instanced type becomes the left focus. Inspired by the Hindley-Milner
-system, we also generate inference constraints on the existential type variables
-(postponing the decision of what type it should be to be used in the proof), and
-collect them in a new constraints context (TODO: should be a substitution not a
-context?) $\Theta$ that is propagated across derivation branches the same way
-the linear context is (by havin an input and output context ($\Theta/\Theta'$)).
-In contrast to HM, everytime a constraint is added it is solved against all
-other constraints -- a branch of the proof search is desired to fail as soon as
-possible; and more, two different existential vars do not unify against
-one another ($?\alpha \mapsto ?\beta$ fails). Note that we instance the scheme with \emph{existential} type
-variables ($\alpha$) rather than just type variables since the latter represent
-universal types in the synth, and the first represent an existential instance of
-a scheme, that might inclusively constraint normal type variables. Additionally,
-it's required that all existential type variables were assigned a type, and
-provided they were, we can forget the constraints generated from the usage of
-the scheme. These concepts are formalized with the rules:
-\begin{mathpar}
-    \infer*[right=($\forall L$)]
-    {
-        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
-        \\
-        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
-        \\
-        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
-    }
-    {\Theta/\Theta; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
-    \and
-    \infer*[right=($?L$)]
-    {\textsc{solve}(?\alpha
-    \mapsto C, \Theta)}
-    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
-    x{:}?\alpha \Downarrow\ \vdash x : C}
-    \and
-    \infer*[right=($\Downarrow L?$)]
-    {\textsc{solve}(?\alpha
-    \mapsto A, \Theta)}
-    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
-    x{:}A \Downarrow\ \vdash x : ?\alpha}
-\end{mathpar}
-where
-\begin{mathpar}
-    \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'\ \textrm{means type}\
-    \tau'\ \textrm{is an \emph{existential instantiation} of scheme}\ \forall
-    \overline{\alpha}.\ \tau
-    \and
-    \textrm{ftv}_E(\tau')\ \textrm{is the set of free \emph{existential} type variables in type}\ \tau'
-    \and
-    ?\alpha \mapsto \tau_x\ \textrm{is a mapping from \emph{existential} type}\ ?\alpha\
-    \textrm{to type}\ \tau_x
-    \\
-    \textsc{solve}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
-    \textrm{can be solved with those in}\ \Theta
-\end{mathpar}
-
-\section{Related Work}
-
-Type-based program synthesis is a vast field of study and so it
-follows that a lot of literature is available to inspire and
-complement our work. Most works~\cite{DBLP:conf/lopstr/HughesO20,DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16} follow some variation of the
-synthesis-as-proof-search approach, however, the process is novel for
-each, due to a variety of different rich types explored, and their
-respective corresponding logics and programming languages; or nuances
-of the synthesis process itself, such as complementing types with
-program examples; or even the programming paradigm of the output
-produced (e.g. generating heap manipulating
-programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).  Some other
-common patterns we might want to follow are the use of type
-refinements~\cite{DBLP:conf/pldi/PolikarpovaKS16} and the support for synthesis of recursive
-functions~\cite{DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15}.  These works share some common ground, and base
-themselves of the same works we do -- most famously the Curry-Howard
-correspondence, but also, for example, works on focusing-based
-proof-search~\cite{10.1093/logcom/2.3.297}.
-
-\mypara{Type-and-Example-Directed Program Synthesis} This
-work~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
-explores a purely functional, recursive program synthesis technique
-based on types/proof search (as we intend to do), with examples as an
-auxiliar technique to trim down the program synthesis vast search
-space. A good use is
-made of the general strategy of using a type system to generate
-programs (by ``inverting'' the type system), rather than using it to
-type-check one.  Moreover, the paper uses a data structure called
-``refinement tree'' in conjunction with the extra information provided
-by the examples to allow for efficient synthesis of non-trivial
-programs. Overall it employs good engineering to achieve type-based
-synthesis of functional typed programs, with at least two additions
-that we might want to follow too (type refinements and recursive
-functions).
-
-% or how to turn your type system upside down ~ não apagar a Kubrick reference? :)
-
-\mypara{Program Synthesis from Polymorphic Refinement Types} The
-work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis
-recursive functional programs but in a more ``advanced'' setting. The
-specification is a combination of two rich forms of types: polymorphic and
-refinement types (which correspond to a first-order logic through the
-Curry-Howard isomorphism) -- this is an interesting combination that offers
-``expressive power and decidability, which enables automatic
-verification--and hence synthesis--of nontrivial programs''.  Their
-approach to refinement types consists a new algorithm that supports
-decomposition of the refinement specification, allowing for separation
-between the language of specification and programs and making the
-approach amenable to compositional synthesis techniques.
-
-\mypara{Resourceful Program Synthesis from Graded Linear Types} The
-work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
-approach very similar to our own.  The work employs so-called graded
-modal types, which is an refinement of pure linear types -- a more
-\emph{fine-grained} version, since it allows for quantitative
-specification of resource usage, in contrast to ours either
-\emph{linear} or \emph{unrestricted} (via the linear logic
-exponential) use of assumptions.  Their resource management is more
-complex, and so they provide solutions which adapt Hodas and Miller's
-resource management
-approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
--- the model we will be using.
-
-They also use focusing as a solution to trim down search space and to
-ensure that synthesis only produces well-typed programs, and our
-solution will use the same literature as a base.  However, since their
-underlying logic is \emph{modal} rather than purely \emph{linear}, it
-lacks a clear correspondence with concurrent session-typed
-programs~\cite{DBLP:journals/mscs/CairesPT16,DBLP:conf/concur/CairesP10},
-which is a crucial avenue of future work. Moreover, their use grading
-effectively requires constraint solving to be integrated with the
-synthesis procedure, which can limit the effectiveness of the overall
-approach as one scales to more sophisticated settings (e.g.~refinement
-types).
-
-% \item Resource-guided synthesis \cite{DBLP:conf/pldi/KnothWP019}
-% This resource-guided means something a bit different. Programs satisfy
-% a functional specification and a symbolic resource bound in the sense
-% of amortized analysis, but can provide technical insights.
-
-%\section{Goals and Work Plan}
-
-%We start with the definition of our type system and simply-typed
-%functional language with linear function, sum and product types
-%($\lolli$, $\tensor$, $\one$, $\with$, $\oplus$, $\bang$).  Through
-%the Curry-Howard correspondence, we'll formulate the synthesis of
-%typed programs in our functional language as bottom-up proof-search in
-%linear logic.
-
-%Since proof rules in \emph{natural deduction} do not entail a
-%proof-search strategy, we will use the equivalent \emph{sequent
-%calculus} system to formulate our typing rules in a way that is
-%amenable to proof search.
-%%
-%In sequent calculus, rules can be naturally understood in a
-%\emph{bottom-up} manner, simplifying the reasoning for the
-%synthesizer. Moreover, the linear nature of our system requires a way
-%to algorithmically reason about resource contexts: for instance, if we
-%read rule $\tensor I$ from Section~\ref{sec:background} bottom-up, we
-%must non-deterministically guess how to correctly split ambient
-%resources into $\Delta_1$ and $\Delta_2$ to separately prove $A$ and
-%$B$. The resource-management approach we'll adopt to solve this has been 
-%previously explained in more detail.
-
-%With the use of input/output contexts, and the sequent calculus
-%formulation of the inference rules, we draw closer to an effective
-%strategy for bottom-up proof search and so of program synthesis.
-%%
-%%
-%However, the search process still encompasses significant
-%non-determinism. To address this we draw on the technique of
-%focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as
-%discussed above.
-
-%With the basic framework in place, we will consider additional
-%(simply-typed) language constructs and synthesis of program fragments
-%or \emph{partial synthesis} -- given a program with a typed hole,
-%synthesize a term to fill out the hole accordingly. We will
-%subsequently also consider synthesis of recursive functions and,
-%might, later on, add refinement types also as a way to constrain the
-%valid programs space.
-
-%Putting together everything above, and the more detailed information
-%in the background section -- linear types, sequent calculus, resource
-%management, and focusing, we have a set of inference rules (written
-%out in the appendix) that define how the synthesis process will
-%progress.
-
-%This is the theory we're interested in and have studied so far,
-%which will allow us to build the actual synthetizer. From here,
-%our work plan gives an overview of the steps we'll take.
-
-%\subsection{Work Plan}
-
-%The end goal is to have a synthetizer capable of
-%taking a type (possibly refined) as input, and to produce (possibly
-%interactively) one or more functions with that type, if any exist.
-
-%As mentioned above, we will begin with (both partial and full)
-%synthesis for the linear $\lambda$-calculus. We will then extend the
-%language with additional type constructs (e.g. base types, records,
-%algebraic data types) and consider synthesis in that setting. Finally,
-%we will consider synthesis of recursive functions, and,
-%time-permitting, we will consider simple forms of type refinements
-%(e.g. arithmetic refinements).
-
-%We've already implemented a parser and a type-checker for a linear
-%$\lambda$-calculus using the resource management technique described
-%in the previous section -- more than
-%good preparation for the development of the synthesizer, they will be
-%useful to \emph{validate} the synthesized functions, and to enable
-%the synthesis of partial programs.
-%%
-%The implementations have been carried out in Haskell since it's easier
-%to model and implement inference rules in a functional setting.
-
-%\mypara{Step by step} Our work so far has been constantly accompanied
-%by the study of somewhat dense theoretical material, state of the art,
-%which we must understand to overcome challenges in these fields, in
-%order to develop a more scalable synthetizer -- we want to display
-%concrete results to prove the feasability of program synthesis from
-%linear types, and thus open the possibility for future application to
-%related fields, such as program synthesis based on session types for
-%message-passing concurrency.
-
-%Apart from the theory, when we started studying type systems, we
-%created a simple type-checker for the simply-typed $\lambda$-calculus.
-%While learning about linear type systems, and the simply-typed linear
-%$\lambda$-calculus, we developed a parser, a desugaring module, and an
-%evaluation module to write, analyze and execute simple programs in the
-%language; culminating in writing a type-checker for the linear
-%$\lambda$-calculus. At this point, we can parse, typecheck, and
-%execute this small linear functional language.  The next step is to
-%start developing the actual synthetizer, embedding it with all the
-%techniques we've seen. Following work will be done to augment
-%usability, expand the language domain, and add features such as those
-%mentioned above.
-
-\mypara{Evaluation} Validation and evaluation of our work is fairly
-easy. Simple questions such as \emph{does it typecheck}? \emph{can
-  we synthesize}? \emph{how fast}?  are the core of our evaluation. In
-more concrete terms, we can compare our times to state of the art
-synthetizers, present a list of working examples of programs synthesized
-by our framework, and corresponding benchmarks.
-
-On the eventual addition of recursive functions, refinement types, and
-interactive synthesis, we'll consider, respectively, other examples to
-showcase recursive functions synthesis, updated times for benchmarks
-with refinement types, and a display of the possible user interaction
-in the user process, and the resulting function following it.
-
-% Expand on earlier points. Start from a functional language with linear types
-% ($\tensor$, $\lolli$, $\oplus$, $\bang$); build on it with more ``stuff''.
-% General techniques, drawn from proof theory via props as types: explore
-% focusing to tame non-determinism/search space. Partial synthesis is still
-% proof search! Add other techniques as we go along (e.g. for recursion we need
-% to constrain recursive calls).
-
-% Mention that you've already implemented a type-checker for this (useful as a
-% prelim. Exercise but also later, for \emph{validation}).
-
-% Validation and evaluation: validation is as simple as ``does it typecheck''?
-% can you synth? how fast?
-
-\section{Evaluation}
-
-% Adicionar já exemplos do que consigo
-
 \bibliographystyle{splncs04}

### 2021-07-12 article/article.tex
tweaks

@@
 
-    % {TODO: Nao ha restrições sobre este? no código tinha que x não podia estar em
-    % delta mas não faz muito sentido hah}
     \infer*[right=($\bang L$)]
     {\Gamma, y{:}A ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
-    {\Gamma ; \Delta/\Delta ; \Omega, x{:}\bang A \Uparrow\ \vdash\
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}\bang A \Uparrow\ \vdash\
     \textrm{let}\ \bang y = x\ \textrm{in}\ M : C}
@@
 \and
-%TODO: eu tenho uma assertion que garante que y não sai com Delta', mas agora
-%parece-me inutil?
     \infer*[right=($\lolli L$)]
@@
 programs from more expressive (richer) types.
-
-\mypara{Recursion} The main idea behind synthesis of recursive programs is the
-labeling of the main goal and the addition of its type, under that name, to the
-unrestricted context. That is, to synth a function of type $A \lolli B$ named
-\emph{foo}, the initial judgment can be written as
-\begin{mathpar}
-    \infer
-    {\dots}
-    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
-    A \lolli B \Uparrow}
-\end{mathpar}
-and, by definition, all subsequent inference rules will have
-($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.
-We can also force the usage of the recursive call by adding it not only to the
-unrestricted context, but to the linear one as well.
-
-% TODO: E como evitamos coisas trivialmente estúpidas?
 
@@
 
+\mypara{Recursion} The main idea behind synthesis of recursive programs is the
+labeling of the main goal and the addition of its type, under that name, to the
+unrestricted context. That is, to synth a function of type $A \lolli B$ named
+\emph{foo}, the initial judgment can be written as
+\begin{mathpar}
+    \infer
+    {\dots}
+    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
+    A \lolli B \Uparrow}
+\end{mathpar}
+and, by definition, all subsequent inference rules will have
+($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.
+We can also force the usage of the recursive call by adding it not only to the
+unrestricted context, but to the linear one as well.
+
+% TODO: E como evitamos coisas trivialmente estúpidas?
+
 \mypara{Polimorfic Types} A polymorfic specification is a type $X$ and a set of
@@
 (postponing the decision of what type it should be to be used in the proof), and
-collect them in a new constraints context (TODO: should be a substitution not a
-context?) $\Theta$ that is propagated across derivation branches the same way
+collect them in a new constraints context $\Theta$ that is propagated across derivation branches the same way
 the linear context is (by havin an input and output context ($\Theta/\Theta'$)).

### 2021-07-12 STLLC/Tree.hs

@@ -1,0 +1,16 @@
+data BST a = Nil | Node (a * BST a * BST a); 
+
+data Maybe a = Nothing | Just a;
+
+synth return :: (a -o BST a);
+
+synth insert :: (a -o BST a -o BST a);
+
+synth id :: (BST a -o BST a);
+
+synth map :: ((a -o b) -> BST a -o BST b);
+
+synth uncons :: (BST a -o Maybe (a * (BST a) * (BST a)));
+
+# fails after a long time
+# synth foldl :: ((b -o a -o b) -> b -o BST a -o b);

### 2021-07-12 STLLC/app/Synth.hs

@@ -580,7 +580,6 @@ focus c goal =
         focus' (Just (n, Fun a b)) c@(g, d) goal = addtrace (LeftFun c (Fun a b) goal) $ do
             nname <- fresh
             (expb, d') <- memoizefocus' focus' (Just (nname, b)) c goal
-            guard (nname `notElem` map fst d') -- TODO: Useless? make sure the focusL rule makes sure the proposition is not left synchronous
             (expa, d'') <- memoizesynth synth (g, d', []) a
             return (substitute nname (App (Var n) expa) expb, d'')
             
@@ -589,13 +588,11 @@ focus c goal =
             do
                 nname <- fresh
                 (lf, d') <- memoizefocus' focus' (Just (nname, a)) c goal
-                guard (nname `notElem` map fst d') -- TODO: useless :)
                 return (substitute nname (Fst (Var n)) lf, d')
             <|>
             do
                 nname <- fresh
                 (rt, d') <- memoizefocus' focus' (Just (nname, b)) c goal
-                guard (nname `notElem` map fst d') --TODO: Useless
                 return (substitute nname (Snd (Var n)) rt, d')
 
         ---- ∃L (?)

### 2021-07-12 STLLC/RefinementTypes.hs
WIP: Left focus refinement types

@@ -4,6 +4,7 @@ nat = add 6 3;
 
 sum2 = add 6 3;
 
+sum3 = add 3;
 
 wplus :: (x { Int } -o y { Int | x == y } -o z { Int | z >= x + y });
 wplus x y = add x y;

### 2021-07-12 STLLC/SynthRef.hs

@@ -1,19 +1,35 @@
 
-synth plus :: (x {Int} -o y {Int} -o z { Int | z == x + y });
 
-synth inc :: (x {Int} -o y { Int | y == x + 1 });
+# Right Focus Synth
 
-synth deq :: (x {Int} -o y { Int | y == x - 1 });
+#synth plus :: (x {Int} -o y {Int} -o z { Int | z == x + y });
+
+#synth inc :: (x {Int} -o y { Int | y == x + 1 });
+
+#synth deq :: (x {Int} -o y { Int | y == x - 1 });
 
 # is this correct?
-synth mm :: (x {Int} -o y {Int} -o z {Int} -o k { Int | k == x - y + z });
+#synth mm :: (x {Int} -o y {Int} -o z {Int} -o k { Int | k == x - y + z });
 
-synth plus20 :: (x { Int | x > 0 } -o y { Int | y == x + 20 });
+#synth plus20 :: (x { Int | x > 0 } -o y { Int | y == x + 20 });
 
-synth number :: (x { Int | x > 345 && x != 346 });
+#synth number :: (x { Int | x > 345 && x != 346 });
 
 # correct me
-synth add3 :: (x {Int} -o y {Int} -o z {Int} -o k {Int | k == x + y + z});
+#synth add3 :: (x {Int} -o y {Int} -o z {Int} -o k {Int | k == x + y + z});
 
 # comment me
 #synth impossible :: (x {Int} -o y {Int} -o z {Int} -o k { Int | k == x * y + z });
+
+
+# Left Focus Synth
+
+# how to avoid stupid functions so that add is attempted
+#synth s :: (x {Int | x > 5} -o y {Int | y > x} -o z {Int});
+
+
+#inc :: (x {Int} -o y {Int | y == x + 1});
+
+#synth shouldntfail :: (x {Int} -o y {Int | y > x});
+
+synth shouldntfail2 :: (x {Int} -o z {Int} -o y {Int | y > x + z});

### 2021-07-12 STLLC/app/CoreSyntax.hs

@@ -158,7 +158,7 @@ instance (Show Type) where
     show (Sum ts) = "+ { " ++ foldl (\p (s, t) -> p ++ s ++ " : " ++ show t ++ "; ") "" ts ++ "}"
     show (ADT n ts) = n ++ concatMap ((" " ++) . (\case t@ADT {} -> "(" ++ show t ++ ")"; t -> show t)) ts
     show (RefinementType n t _ Nothing) = n ++ " { " ++ show t ++ " }"
-    show (RefinementType n t l (Just p)) = n ++ " { " ++ show t ++ " | " ++ foldr (\(RefinementType _ _ _ mp) -> ((case mp of {Nothing -> ""; Just p' -> show p' ++ " => "}) ++)) "" l ++ show p ++ " }"
+    show (RefinementType n t l (Just p)) = n ++ " { " ++ show t ++ " | " ++ foldr (\(RefinementType _ _ _ mp) -> ((case mp of {Nothing -> ""; Just p' -> show p' ++ " => "}) ++)) "" l ++ show p ++ " }" ++ " LIST: " ++ show l
 
 
 instance (Show Scheme) where

### 2021-07-12 STLLC/app/Synth.hs

@@ -63,7 +63,7 @@ runSynth n (g, d) t st adtds = do
 
     where
         synthComplete (g, d, o) t = do
-            (e, d') <- memoizesynth synth (recf, d, o) t <|> memoizesynth synth (g, d, o) t -- first try synth only with the recursive function in the gamma context instead of the whole program
+            (e, d') <- memoizesynth synth (tail recf, d, o) t <|> memoizesynth synth (tail g, d, o) t -- first try synth only with the recursive function in the gamma context instead of the whole program
             guard $ null d'
             return (e, d')
 
@@ -139,8 +139,8 @@ addbinaryrestriction tag ty ty' = local (\ (rs, adtds, d, t, e) -> ((tag, Left (
 addtrace :: TraceTag -> Synth a -> Synth a
 addtrace t s = do
     (tstck, depth) <- gettracestack
-    -- trace ("D" ++ show depth ++ ": " ++ show t ++ "\n-- stack --\n" ++ unlines (map show tstck) ++ "\n") $
-    local (\(a,b,c,d,e) -> (a,b,c+1,t:d,e)) s
+    trace ("D" ++ show depth ++ ": " ++ show t) $ -- ++ "\n-- stack --\n" ++ unlines (map show tstck) ++ "\n") $
+        local (\(a,b,c,d,e) -> (a,b,c+1,t:d,e)) s
 
 
 gettracestack :: Synth ([TraceTag], Int)
@@ -285,7 +285,10 @@ synth :: Ctxt -> Type -> Synth (Expr, Delta)
 ---- -oR
 synth c@(г, d, o) (Fun a b) = addtrace (RightFun c (Fun a b)) $ do
     name <- fresh
-    (exp, d') <- memoizesynth synth (г, d, (name, a):o) b
+    let b' = if isRefType a
+                then traceShow (a, b, addRefinementToCtxs a b) $ addRefinementToCtxs a b    -- Everytime we decompose a dependent function, we must add it to the rest's refinement context... we need to change the core rules for this...
+                else b
+    (exp, d') <- memoizesynth synth (г, d, (name, a):o) b'
     guard (name `notElem` map fst d')
     return (Abs name (Just a) exp, d')
 
@@ -379,9 +382,11 @@ synth c@(g, d, (n, Bang a):o) t = addtrace (LeftBang c (Bang a) t) $ do
     return (LetBang nname (Var n) exp, d')
 
 ---- refinementL
[…]

### 2021-07-13 article/article.tex
WIP: Left focus refinement types
some notes

@@
 \usepackage{hyperref}
+\usepackage{todonotes}
 
@@
 \title{Synthesis of Linear Functional Programs}
-\author{Rodrigo Mesquita \and Bernardo Toninho}
-\date{April 2021}
+\author{Rodrigo Mesquita - Nº XXXXX \\
+    Supervisor: Bernardo Toninho}
+\date{July 2021}
 \institute{NOVA School of Science and Technology}
@@
 
-\section{Formal System}\label{sec:formal_system}
-
-The development of \synname\ was initially found on an almost direct reading of
-concrete inference rules and theoretical principles (~\cite{frank pfenning notes}); and work on its environment
-was also guided by formal systems and methods (e.g. the Hindley-Milner type
-inference~\cite{}).  However, with the addition of recursion and richer types to
-the synth and language, formal guidelines to support the implementation become
-scarcer, or more complex. Accordingly, rather than derived from explicit rules,
-solutions appear first to tame the extensions' inherent complex problems (such
-as infinite recursion), and implicit rules come only as a byproduct.
-
-With this section we intend to formalize the implicit rules that guide the
-synthesis of more expressive specifications, alongside the already explicit
-rules that model the core of the synth process, resulting in a \emph{sound} set
-of rules that hopes to serve as a source of ideas on taming synthesis of
-recursive functions with richer types in a linear context, as well as describe
-the system in enough detail for the synthetiser to be reproducible by a
-theory-driven implementation.
+\section{Synthesis with Linear Types, Polymorphism and Refinements}\label{sec:formal_system}
+\todo[inline]{Encurtar titulo, mas algo nesta onda que descreve
+  ``tudo''}
+
+\todo[inline]{Aqui parece-me que deves comecar por explicar o que vais
+  fazer nesta seccao. Sinto falta da sintaxe da linguagem,
+  possivelmente incrementalmente. Comecava por listar os tipos/termos
+  do lambda linear ``simples'', depois em cada subseccao acrescentava
+  as construcoes novas e como a sintese funciona para esse
+  contexto. Enfatizar que a novelty é a combinacao destas features todas.}
+
+The development of \synname\ was based on an almost direct reading of
+concrete inference rules and theoretical principles (~\cite{frank
+  pfenning notes});
+\todo{aqui queres dizer que começaste por te basear
+  num sistema de focusing para a logica linear (era sequer linear a
+  formulacao que comecaste a seguir?) O que são theoretical
+  principles?}
+and work on its environment was also guided by formal
+systems and methods (e.g.~Hindley-Milner type
+inference~\cite{}).\todo{esta frase parece-me desnecessaria. O que
+  queres dizer neste paragrafo, parece-me, é que partes de um sistema
+  focused da literatura que depois vais extender, inspirando te em
+  tecnicas standard de proof search / inferencia.}
+
+However, with the addition of recursion and
+richer types to the synthesis and language, formal guidelines
+\todo{nao sao formal guidelines, nao existem trabalhos a combinar
+  estas features todas, essa parte é importante. Além de que 
+ao incluir estas coisas deixamos de estar completamente em proof
+search, pois muitas destas coisas (e.g. recursao) nao sao
+necessariamente sound logicamente.}
+  Tenta usar termos 
+  especificos. Richer types? Richer como?} to support
+the implementation become scarcer, or more complex \todo{nao existem
+  ou sao complicados? ambos? citacao?}. Accordingly,
+rather than derived from explicit rules, solutions appear first to
+tame the extensions' inherent complex problems (such as infinite
+recursion), and implicit rules come only as a byproduct. \todo{ Nao
+  percebo o que queres dizer aqui. O que sao explicit rules e implicit
+  rules? Ou há regras ou nao há. O que estas a dizer é que inventaste
+  tecnicas de sintese para estes casos, que correspondem as regras que
+  vais apresentar. E tao explicito ou implicito como o que esta na literatura.}
+
+With this section we aim to formalize the implicit \todo{mais uma vez\dots} rules that guide
+the synthesis of more \todo{nao se faz sintese de specs, mas sim a
+  partir de specs} expressive specifications, alongside the already
+explicit rules that model the core of the synth \todo{synthesis} process, resulting in
+a \emph{sound} set of rules that hopes to serve as a source of ideas
+\todo{ nao me parece\dots As regras caracterizam o que esta
+  implementado, e esse o objectivo, nao e ser uma fonte de ideias\dots }
+on taming synthesis of recursive functions with richer types in a
+linear context, as well as describe the system in enough detail for
+the synthetiser to be reproducible by a theory-driven implementation.
 
@@
 satisfiability.
+
+\mypara{Memoization} ...
 

### 2021-07-13 article/article.tex
Benchmarks and -oR order

--- 2021-07-12 article/article.tex
@@
 \usepackage{hyperref}
+\usepackage{multirow}
 
@@
 
-This section presents the rules while explaining them, modifying some along the
-process. For a
-
 \mypara{Core Rules} The core system comprises of the sequent calculus rules for
@@
 right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
-context $\Delta$, and ordered context $\Omega$, synth a program of type $B$ with
+context $\Delta$, and inversion context $\Omega$, synth a program of type $B$ with
 an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
@@
 
-The decision rules are naturally followed by either left or right focus rules:
+The decision rules are followed by either left or right focus rules:
 \begin{mathpar}
@@
 
+Having said that, we won't allow the use of the recursive call in $\Gamma$ right
+away. If we did, every synth goal would have a trivial proof (a non-terminating
+function that just calls itself) that could even shadow other valid, relevant,
+solutions.  Instead, our framework only allows the use of recursion after having
+deconstructed a recursive ADT. To avoid complicating the rules, we'll add the
+invariant: the recursive hypothesis can only be \textbf{used in recursive branches of
+ADT deconstruction}, i.e. the recursive call should only take "smaller", recursive,
+hypothesis as arguments. To illustrate, having $\textrm{data List} =
+\textrm{Nil}\ \vert\ \textrm{Cons}\ (A * \textrm{List})$, recursion is only
+"available" when deconstructing ($\textrm {List} \vdash C$), and only in
+the \emph{Cons} branch. % (TODO: rewrite)
+
 % TODO: E como evitamos coisas trivialmente estúpidas?
@@
 
+\mypara{Memoization} ...
+
 \mypara{Debugging} The rules described in the formal
@@
 
-% Adicionar já exemplos do que consigo
+The benchmarks are separated into groups: theorems of linear logic, recursive
+ADTs, refinements. For each goal we analyze if the synthesis process terminated
+according to what was expected, the mean time and standard deviation, if the
+resulting program fullfills its intended purpose
+
+% TODO: Utilizamos qual argumento primeiro ??? para \omega com ++ ou : ?
+
+\begin{center}
+    \begin{tabular}{ |c|c|c|c|c| }
+        \hline
+        Group & Goal & Avg. time + $\sigma$ & Found & Is expected \\
+        \hline
+        \multirow{4}{4em}{Linear Logic Theor.} & uncurry & 1s & yes & yes \\ 
+        & curry & 2s & yes & yes \\ 
+        & call by name?? & 2s & yes & yes \\ 
+        & 0/1?? & 2s & yes & yes \\ 
+        \hline
+        \multirow{8}{4em}{List} & singleton & 3s & yes & yes \\ 
+        & insert & 2s & yes & unordered \\ 
+        & map & 1s & yes & yes \\ 
+        & append & 2s & yes & yes \\ 
+        & foldl & 2s & yes & got (foldl . reverse) \\ 
+        & foldr & 2s & yes & yes \\ 
+        & concat & $\infty$ & no & no \\ 
+        & uncons & 1s & yes & yes \\
+        \hline
+        \multirow{3}{4em}{Maybe} & return & 1s & yes & yes \\ 
+        & $>>=$ & 1s & yes & yes \\
+        & maybe & 1s & yes & yes \\
+        & list to maybe & 12s & no & yes \\
+        \hline
+        \multirow{5}{4em}{Binary Tree} & singleton & 1s & yes & yes \\ 
+        & insert & 1s & yes & unordered \\
+        & map & 1s & yes & yes \\
+        & append & 1s & yes & appends to rightmost leaf \\
+        & join node & 1s & yes & joins at rightmost leaf \\
+        \hline
+        \multirow{7}{4em}{Map} & singleton & 1s & yes & yes \\ 
+        & insert & 1s & yes & unordered \\
+        & insertWithKey & 1s & yes & ignores function \\
+        & union & 1s & yes & ignores keys \\
+        & mapAccum & 1s & no & ?? \\
+        & map & 1s & yes & yes \\
+        & mapKeys & 1s & yes & yes \\
+        \hline
+        \multirow{1}{4em}{Misc} & either & 1s & yes & yes \\ 
+        \hline
+    \end{tabular}
+\end{center}
 

### 2021-07-13 STLLC/Simple.hs

@@ -1,0 +1,2 @@
+
+synth x :: (a -o a);

### 2021-07-13 STLLC/app/CoreSyntax.hs

@@ -1,4 +1,4 @@
-{-# LANGUAGE LambdaCase #-}
+{-# LANGUAGE LambdaCase, DeriveGeneric, DeriveAnyClass #-}
 module CoreSyntax (CoreExpr(..), Type(..), Scheme(..), Name, CoreBinding(..),
     TypeBinding(..), Var(..), Mult(..), transformM, transform, trivialScheme,
     Literal(..), TyLiteral(..), isInType, trivialInt, isExistentialType,
@@ -9,6 +9,8 @@ import Control.Monad
 import Data.Maybe
 import Data.Hashable
 import Control.Applicative
+import GHC.Generics (Generic)
+import Control.DeepSeq
 import Data.Bifunctor
 import Data.Functor.Identity
 
@@ -78,7 +80,7 @@ data CoreExpr
     deriving (Eq)
 
 
-newtype Literal = LitInt Integer deriving (Eq, Ord)
+newtype Literal = LitInt Integer deriving (Eq, Ord, Generic, NFData)
 
 
 
@@ -86,10 +88,10 @@ data Var = Var
     { varMult :: Mult 
     , unVar   :: Scheme
     } deriving (Eq, Show, Ord)
-data Mult = Lin | Unr deriving (Eq, Show, Ord)
+data Mult = Lin | Unr deriving (Eq, Show, Ord, Generic, NFData)
 
 
-data Scheme = Forall [Int] Type deriving (Eq, Ord)
+data Scheme = Forall [Int] Type deriving (Eq, Ord, Generic, NFData)
 
 
 data Type
@@ -110,9 +112,9 @@ data Type
 
     | RefinementType Name Type [Type] (Maybe Predicate)
     
-    deriving (Eq, Ord)
+    deriving (Eq, Ord, Generic, NFData)
 
-data TyLiteral = TyInt deriving (Eq, Ord)
+data TyLiteral = TyInt deriving (Eq, Ord, Generic, NFData)
 
 data Predicate
     = PVar Name
@@ -121,7 +123,7 @@ data Predicate
     | Conditional Predicate Predicate Predicate
     | UnaryOp Name Predicate
     | BinaryOp Name Predicate Predicate
-    deriving (Eq, Ord)
+    deriving (Eq, Ord, Generic, NFData)
 
 
 

### 2021-07-13 STLLC/app/Program.hs

@@ -2,7 +2,8 @@
 module Program (Program(..), ADTD(..), trivialProgram, completeFrontendMarksCtx) where
 
 import qualified Data.Map as Map
-
+import GHC.Generics (Generic)
+import Control.DeepSeq
 
 import CoreSyntax
 import Util

### 2021-07-13 STLLC/app/Syntax.hs

@@ -1,7 +1,10 @@
+{-# LANGUAGE LambdaCase, DeriveGeneric, DeriveAnyClass #-}
 module Syntax (Binding(..), Expr(..), Pattern(..), transformM, transform) where 
 
 import Data.Maybe
 import Prelude hiding (Bool)
+import GHC.Generics (Generic)
+import Control.DeepSeq
 import Data.Functor.Identity
 
 
@@ -14,6 +17,7 @@ import CoreSyntax (Type, Scheme, Name, Literal(..))
 -------------------------------------------------------------------------------
 
 data Binding = Binding Name Expr
+    deriving (Generic, NFData)
 
 
 data Expr
@@ -62,6 +66,7 @@ data Expr
     -- Added sugar :)
 
     | UnrestrictedAbs Name (Maybe Type) Expr
+    deriving (Generic, NFData)
 
 
 data Pattern

### 2021-07-13 STLLC/app/Synth.hs

@@ -285,9 +285,9 @@ synth :: Ctxt -> Type -> Synth (Expr, Delta)
 synth c@(г, d, o) (Fun a b) = addtrace (RightFun c (Fun a b)) $ do
     name <- fresh
     let b' = if isRefType a
-                then traceShow (a, b, addRefinementToCtxs a b) $ addRefinementToCtxs a b    -- Everytime we decompose a dependent function, we must add it to the rest's refinement context... we need to change the core rules for this...
+                then addRefinementToCtxs a b    -- Everytime we decompose a dependent function, we must add it to the rest's refinement context... we need to change the core rules for this...
                 else b
-    (exp, d') <- memoizesynth synth (г, d, (name, a):o) b'
+    (exp, d') <- memoizesynth synth (г, d, o ++ [(name, a)]) b' -- TODO:! Não usar append ?
     guard (name `notElem` map fst d')
     return (Abs name (Just a) exp, d')
 

### 2021-07-13 STLLC/bench/list2.hs

@@ -1,0 +1,3 @@
+data List a = Nil | Cons (a * List a);
+
+synth insert :: (a -o List a -o List a);

### 2021-07-13 STLLC/bench/list5.hs

@@ -1,0 +1,5 @@
+data List a = Nil | Cons (a * List a);
+
+synth foldl :: ((b -o a -o b) -> b -o List a -o b);
+
+# results in (foldl . reverse)

### 2021-07-13 STLLC/bench/list6.hs

@@ -1,0 +1,3 @@
+data List a = Nil | Cons (a * List a);
+
+synth foldr :: ((a -o b -o b) -> b -o List a -o b);

### 2021-07-13 STLLC/bench/list8.hs

@@ -1,0 +1,5 @@
+data List a = Nil | Cons (a * List a);
+
+data Maybe a = Nothing | Just a;
+
+synth uncons :: (List a -o Maybe (a * List a));

### 2021-07-13 STLLC/bench/llt4.hs

@@ -1,0 +1,2 @@
+# 0 / 1?
+synth goal :: (!(!A -o !B) -o !(!B -o !C) -o !(!A -o C));

### 2021-07-13 STLLC/bench/map10.hs

@@ -1,0 +1,7 @@
+data Map a b = Tip | Bin (a * b * (Map a b) * (Map a b));
+data List a = Nil | Cons (a * List a);
+
+# doesn't work
+#synth union :: (Map b a -o Map b a -o Map b a);
+
+#synth assocs :: (Map b a -o List (b * a));

### 2021-07-13 STLLC/bench/map4.hs

@@ -1,0 +1,3 @@
+data Map a b = Tip | Bin (a * b * (Map a b) * (Map a b));
+
+synth insertWith :: ((b -o b -o b) -> a -o b -o Map a b -o Map a b);

### 2021-07-13 STLLC/bench/map6.hs

@@ -1,0 +1,3 @@
+data Map a b = Tip | Bin (a * b * (Map a b) * (Map a b));
+
+synth union :: (Map a b -o Map a b -o Map a b);

### 2021-07-13 STLLC/bench/map9.hs

@@ -1,0 +1,3 @@
+data Map a b = Tip | Bin (a * b * (Map a b) * (Map a b));
+
+synth mapKeys :: ((a -o b) -> Map a c -o Map b c);

### 2021-07-13 STLLC/bench/maybe1.hs

@@ -1,0 +1,3 @@
+data Maybe a = Nothing | Just a;
+
+synth return :: (a -o Maybe a);

### 2021-07-13 STLLC/bench/tree1.hs

@@ -1,0 +1,3 @@
+data BinTree a = Nil | Node (a * BinTree a * BinTree a);
+
+synth singleton :: (a -o BinTree a);

### 2021-07-13 STLLC/bench/tree2.hs

@@ -1,0 +1,4 @@
+data BinTree a = Nil | Node (a * BinTree a * BinTree a);
+
+synth insert :: (a -o BinTree a -o BinTree a);
+

### 2021-07-13 STLLC/bench/tree3.hs

@@ -1,0 +1,3 @@
+data BinTree a = Nil | Node (a * BinTree a * BinTree a);
+
+synth map :: ((a -o b) -> BinTree a -o BinTree b);

### 2021-07-13 STLLC/t.hs

@@ -1,2 +1,5 @@
-main :: forall a b c d . ((! (a -o b)) -o (((! (a -o b)) -o (d -o c)) -o (d -o c)));
-main e f g = let !h = e in f ! (λk -o h k) g;
+data List1 a = Nil1 | Cons1 (a * List1 a);
+
+data List2 a = Nil2 | Cons2 (a * List2 a);
+
+synth conv :: (List1 a -o List2 a);

### 2021-07-13 STLLC/t2.hs

@@ -1,0 +1,2 @@
+main :: forall a b c d . ((! (a -o b)) -o (((! (a -o b)) -o (d -o c)) -o (d -o c)));
+main e f g = let !h = e in f ! (λk -o h k) g;

### 2021-07-13 STLLC/tests/support/List.hs

@@ -2,16 +2,10 @@ data List a = Nil | Cons (a * List a);
 
 data Maybe a = Nothing | Just a;
 
-data Bool = True | False;
 
-
-
-# can't synth
-#synth isEmpty :: (List a -o Bool);
-# can't synth -- should it?
-# synth zip :: (!(a -o b -o c) -o List a -o List b -o List c)
 
 synth na :: (List Int -o List Int);
+
 
 synth singleton :: (a -o List a);
 

### 2021-07-13 STLLC/tests/support/Maybe.hs

@@ -1,0 +1,14 @@
+data Maybe a = Nothing | Just a;
+
+data List a = Nil | Cons (a * List a);
+
+
+synth return :: (a -o Maybe a);
+
+synth empty :: Maybe a;
+
+synth bind :: (Maybe a -o (a -o Maybe b) -> Maybe b);
+
+synth maybe :: (b -> (a -o b) -> Maybe a -o b);
+
+#synth listToMaybe :: (List a -o Maybe a);

### 2021-07-13 STLLC/tt.hs

@@ -1,4 +1,4 @@
+data List a = Nil | Cons (a * List a);
+data Either a b = Left a | Right b;
 
-# -- synth x :: a
-
-synth y :: (a -> b)
+synth partitionEithers :: ((a -o c) -> (b -o c) -> Either a b -o c);

### 2021-07-13 article/article.tex
notes

--- 2021-07-13 article/article.tex
@@
 search, pois muitas destas coisas (e.g. recursao) nao sao
-necessariamente sound logicamente.}
+necessariamente sound logicamente.
   Tenta usar termos 
@@
 \mypara{Core Rules} The core system comprises of the sequent calculus rules for
-linear logic connectives with resource-management and focusing. As explained in\ref{sec:background}, focusing separates the proof search into two
+linear logic connectives with resource-management and focusing. As
+explained in\ref{sec:background} \todo{se o background nao vai ser
+  incluido no documento principal, nao deves referenciar assim tao
+  directamente para o conteudo principal.}, focusing separates the proof search into two
 phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
@@
 
+\todo{Primeiro explica os varios judgments (as ``formas'' das
+  conclusoes das regras) e o que significam as varias componentes,
+  nomeadamente a parte da gestao de recursos.  Depois podes explicar
+  regras concretas como fazes a seguir. Tambem seria util dizer quais
+  os tipos que sao invertiveis a esquerda/direita, pois o leitor nao
+  faz ideia do que estas a dizer.}
+
 Putting together linear logic, the Curry-Howard correspondence, resource
@@
 an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
-resulting in the program $M$ and output\ref{sec:background} linear
+resulting in the program $M$ and output linear
 context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
@@
 $\Delta'$.
+
+\todo{a sintaxe $\lambda x \rightarrow M$ nao e mt standard, usa
+  $\lambda x . M$}
 
@@
 Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
-asynchronous. If we're left focused on an atomic proposition we either instance
+asynchronous. If we're left focused on an atomic proposition we either
+instance \todo{instantiate}
 the goal or fail. Otherwise the left focus is asynchronous and we can start
@@
 \end{mathpar}
-The rules written above together make the core of our synthetiser. Ahead, we'll
+The rules written above together make the core of our synthetiser. Next, we'll
 present new rules that align and build on top of these to synth recursive
@@
 \mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
-is a named sum type of any type, i.e. a named type instanceable by one of many
+is a named sum type of any type, i.e. a named type that can be instantiated by one of many
 tags (also known as constructors) that take exactly one fixed type, which might
@@
 
-The semantics of the simple ADT relate to those of the plus ($\oplus$) type --
+The semantics of the ADTs relate to those of the plus ($\oplus$) type --
 both are additive disjunctions.  A proof of an ADT requires only a proof of one
@@
 constructor:
+
+\todo[inline]{Não uses Foo e Bar. $T_1$, $T_2$ , $\dots$ para
+  tipos. $C_1$, $C_2$, $\dots$ para os construtores.}
+
 \begin{mathpar}
@@
 \end{mathpar}
+
 where the ADT \emph{Foo} and its constructors stand for any ADT with form
@@
 recursive (or "indutively defined"), meaning the type a constructor needs might
-be the type the constructor constructs. Seemingly small, this change has a big
+be the type the constructor constructs \todo{ Na escrita, repetition
+  does not legitimize. Constructors can take as arguments values of
+  the type they are defining, ou algo assim}. Seemingly small, this change has a big
 impact in the synthesis process. Take, for instance, the ADT defined as
@@
 Using our current system, we are to apply an infinite number of times the left
-ADT rule, never reaching a final state for the proof. Simetrically, the
-derivation for the synth goal $\te{Foo}$ doesn't have an end either.
+ADT rule, never reaching a final state for the proof. Symmetrically, the
+derivation for the synth goal $\te{Foo}$ is also infinite.
 \begin{mathpar}
@@
 \end{mathpar}
-To handle this situation, we impede the decomposition of an ADT in subsequent
+To account for this situation, we impede the decomposition of an ADT in subsequent
 proofs of its branches, and to impede the construction of an ADT in a subsequent
-proof of a type to construct it. For this, we need two more contexts, $\Rho_C$
-for restrictions on construction and $\Rho_D$ for restrictions on
+proof of a type to construct it. \todo{Não se percebe muito bem a
+  segunda parte da frase, rephrase!}
+%
+%
+For this, we need two more contexts, $\Rho_C$
+for restrictions \todo{constraints} on construction and $\Rho_D$ for restrictions on
 deconstruction. For convenience, they are represented by a single $\Rho$ if
@@
 \end{mathpar}
-These modifications prevent the derivation from not ending for the two seen
-situations, however, they also greatly limit the derivable valid programs space,
-leaving the synthetiser pretty much unable to synth specifications with
-recursive types, which is so undesirable that we add three rules to complement the
-restrictions on construction and destruction of recursive types.
+These modifications prevent the derivation from not ending
+\todo{infinite derivations} for the scenarios described above,
+however, they also greatly limit the space of derivable programs,
+leaving the synthetiser effectively unable to synth specifications
+with recursive types\todo{refrasear tendo em conta que se sintetiza a
+  partir de specs e nao specs}, which is so undesirable that we add three rules
+to complement the restrictions on construction and destruction of
+recursive types.
 
@@
 % é agora inutil. Talvez seja demais adicionar à regra
+
+\todo[inline]{esta frase nao me parece muito necessaria}
 The ADT rules draw closer to a final form, but will still need a revision
@@
 
-\mypara{Polimorfic Types} A polymorfic specification is a type $X$ and a set of
+\todo[inline]{Polymorphic}
+
+\mypara{Polymorphic Types} A polymorphic specification is a type $X$ \todo{T} and a set of
 bound variables that quantify universally the type variables in $X$, called a
-\emph{scheme} ($\forall \overline{\alpha}.\ \tau$). Synthesis of a scheme is straightforward.
+\emph{scheme} ($\forall \overline{\alpha}.\ \tau$).
+\todo[inline]{Não é um tipo e um conjunto de variáveis bound. É um
+  tipo da forma $\forall \overline{\alpha}.\ T$ em que $\dots$ }
+Synthesis of a scheme is straightforward. \todo{Diria que nao e
+  especialmente straightforward.}
 First, \textbf{type variables are considered atomic types}, then, instantiate
-the bound vars of the scheme as described by the Hindley-Milner inference
+the bound vars\todo{variables} of the scheme as described by the Hindley-Milner inference
 method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
-each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli \alpha$ could be instanced to
+each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha
+\lolli \alpha$ could be instanced \todo{instantiated} to
 $\alpha0 \lolli \alpha0$. We add a rule for this:
@@
 agnostically -- nothing constrains their type, meaning the program is
-polymorphic * TODO: rewrite this?.
-
-The interesting challenge of polymorphism in synthesis is the usage of schemes
-from the unrestricted context.  To begin with, $\Gamma$ now holds both types and
-schemes. Consequently, after the rule \textsc{decideLeft!} is applied, we become
-left focused on either a type or a scheme. Since left focus on a type is already
-well formulated, only rules for left focus on a scheme are needed. Our algorithm
-instances bound type vars of the focused scheme with existential type variables,
-and the instanced type becomes the left focus. Inspired by the Hindley-Milner
-system, we also generate inference constraints on the existential type variables
-(postponing the decision of what type it should be to be used in the proof), and
-collect them in a new constraints context $\Theta$ that is propagated across derivation branches the same way
-the linear context is (by havin an input and output context ($\Theta/\Theta'$)).
-In contrast to HM, everytime a constraint is added it is solved against all
-other constraints -- a branch of the proof search is desired to fail as soon as
-possible; and more, two different existential vars do not unify against
-one another ($?\alpha \mapsto ?\beta$ fails). Note that we instance the scheme with \emph{existential} type
-variables ($\alpha$) rather than just type variables since the latter represent
-universal types in the synth, and the first represent an existential instance of
-a scheme, that might inclusively constraint normal type variables. Additionally,
-it's required that all existential type variables were assigned a type, and
-provided they were, we can forget the constraints generated from the usage of
-the scheme. These concepts are formalized with the rules:
+polymorphic * TODO: rewrite this?.\todo[inline]{Use an example.}
+
+The interesting challenge of polymorphism in synthesis is the usage of
+schemes from the unrestricted context.  To begin with, $\Gamma$ now
+holds both types and schemes. Consequently, after the rule
+\textsc{decideLeft!} is applied, we become left focused on either a
+type or a scheme. Since left focus on a type is already well
+formulated, only rules for left focus on a scheme are needed. Our
+algorithm instances bound type vars of the focused scheme with
+existential type variables, and the instanced type becomes the left
+focus. Inspired by the Hindley-Milner system, we also generate
+inference constraints on the existential type variables (postponing
+the decision of what type it should be to be used in the proof), and
+collect them in a new constraints context $\Theta$ that is propagated
+across derivation branches the same way the linear context is (by
+havin an input and output context ($\Theta/\Theta'$)).  In contrast to
+HM, everytime a constraint is added it is solved against all other
+constraints -- a branch of the proof search is desired to fail as soon
+as possible; and more, two different existential vars do not unify
+against one another ($?\alpha \mapsto ?\beta$ fails) \todo{O que
+  significa $?\alpha$ ? Não foi explicado -- no Hindley-Milner nao ha
+  propriamente o problema de misturar variaveis existenciais com universais}. Note that we
+instance the scheme with \emph{existential} type variables ($\alpha$)
+rather than just type variables since the latter represent universal
+types in the synth, and the first represent an existential instance of
+a scheme, that might inclusively constraint normal type
+variables. Additionally, it's required that all existential type
+variables were assigned a type, and provided they were, we can forget
+the constraints generated from the usage of the scheme. These concepts
+are formalized with the rules:
 \begin{mathpar}
@@
 sugar and using the locally-nameless~\cite{locally nameless} representation for
-all bound variables. This way inference can be done without worrying about name
+all bound variables. \todo{Explica o que e isto} This way inference can be done without worrying about name
 conflicts.

### 2021-07-13 article/article.tex

@@
 \usepackage{hyperref}
+
 \usepackage{todonotes}
+
+\usepackage{multirow}
 
@@
 unrestricted context, but to the linear one as well.
+
+Having said that, we won't allow the use of the recursive call in $\Gamma$ right
+away. If we did, every synth goal would have a trivial proof (a non-terminating
+function that just calls itself) that could even shadow other valid, relevant,
+solutions.  Instead, our framework only allows the use of recursion after having
+deconstructed a recursive ADT. To avoid complicating the rules, we'll add the
+invariant: the recursive hypothesis can only be \textbf{used in recursive branches of
+ADT deconstruction}, i.e. the recursive call should only take "smaller", recursive,
+hypothesis as arguments. To illustrate, having $\textrm{data List} =
+\textrm{Nil}\ \vert\ \textrm{Cons}\ (A * \textrm{List})$, recursion is only
+"available" when deconstructing ($\textrm {List} \vdash C$), and only in
+the \emph{Cons} branch. % (TODO: rewrite)
 

### 2021-07-13 article/article.tex
Memoization & run benchmarks

--- 2021-07-13 article/article.tex
@@
 support the usage of the SMT solvers to satisfy and output more complicated
-first-order logic solutions -- since this is needed to synth refinement types
+first-order logic uninterpreted functions -- since this is needed to synth refinement types
 when right focused, we instead use unsafe library primitives to communicate with
@@
 
-\mypara{Memoization} ...
+\mypara{Memoization} Some proofs attempt to synth the same sub-goal with the
+same premises more than once. Since these proof attempts might be expensive
+performance-wise, and we can rely on the program's determinism (equal calls to
+synth result in equal outcomes), we add to our synth memoization, an
+optimization technique that stores function calls and returns cached values for
+equal inputs. For our synth, two calls are equivalent when contexts $\Theta,
+\Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and the
+goal, are equivalent. Because comparing all of these is slow, our implementation
+hashes the relevant values and uses the result as the key. Using LogicT, a
+branch we want to memoize might fail -- it's desireable to also record that a
+synthesis fails, as it happens more than successful results. We capture failure
+of a logict computation using the $ifte$ function and represent failure or
+success information with the Maybe type. We ran benchmarks to measure the impact
+of memoization on the performance. It's more noticeable in synthesis with
+multiple functions in the unrestricted context, or more complex
+specifications:
+%
+\begin{center}
+    \begin{tabular}{ |c|c|c|c|c| }
+        \hline
+        Goal & Time + $\sigma$ & TimeM + $\sigma$ \\
+        \hline
+        insertWithKey & 13.31 s ± 593.4 ms & 2.667 s ± 284.3 ms
+    \end{tabular}
+\end{center}
 

### 2021-07-13 STLLC/app/Main.hs

@@ -99,10 +99,8 @@ main = do
     (action:args) <- getArgs
     if action == "bench"
     then do
-        let benchmarks = map (\x -> "bench/" ++ x ++ ".hs") ["llt1", "llt2", "llt3", "llt4", "list1", "list2", "list3", "list4", "list5", "list6", "list8", "maybe1", "maybe2", "maybe3", "tree1", "tree2", "tree3", "tree4", "tree5", "map2", "map3", "map5", "map6", "map7", "map8", "map9"]
-        cprogs <- mapM maindesugarModule benchmarks
-        synthed <- mapM (synthMarksModule >=> \(Program _ bs _ _) -> return bs) cprogs
-        defaultMain [bgroup "synth" (zipWith (\bs n -> bench n $ nfIO $ return bs) synthed benchmarks)]
+        let benchmarks = map (\x -> "bench/" ++ x ++ ".hs") ["llt1", "llt2", "llt3", "llt4", "list1", "list2", "list3", "list4", "list5", "list6", "list8", "maybe1", "maybe2", "maybe3", "tree1", "tree2", "tree3", "tree4", "tree5", "map2", "map3", "map5", "map6", "map8", "map9"]
+        defaultMain [bgroup "synth" (map (\n -> bench n $ nfIO $ (mainsynthMarksModule >=> \(Program _ bs _ _) -> return bs) n) benchmarks)]
     else
         let arg =
                 (case args of

### 2021-07-14 article/article.tex
tweaks

@@
 
-With this section we aim to formalize the implicit \todo{mais uma vez\dots} rules that guide
-the synthesis of more \todo{nao se faz sintese de specs, mas sim a
-  partir de specs} expressive specifications, alongside the already
-explicit rules that model the core of the synth \todo{synthesis} process, resulting in
+With this section we aim to formalize the rules that guide
+synthesis from more expressive specifications, alongside the already
+well defined rules that model the core of the synthesis process, resulting in
 a \emph{sound} set of rules that hopes to serve as a source of ideas
 \todo{ nao me parece\dots As regras caracterizam o que esta
-  implementado, e esse o objectivo, nao e ser uma fonte de ideias\dots }
+  implementado, e esse o objectivo, nao e ser uma fonte de ideias\dots}
+  \todo{R: não parece ser sound ?}
 on taming synthesis of recursive functions with richer types in a
@@
 Putting together linear logic, the Curry-Howard correspondence, resource
-management, and focusing, we get the following formal system (heavily inspired
+management, and focusing, we get the following core formal system (heavily inspired
 by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
@@
 \end{mathpar}
-Eventually, the focus proposition will no longer be synchronous, i.e. it's atomic or
-asynchronous. If we're left focused on an atomic proposition we either
-instance \todo{instantiate}
-the goal or fail. Otherwise the left focus is asynchronous and we can start
-inverting it. If we're right focused on a proposition that isn't right
+%
+Eventually, the focus proposition will no longer be synchronous, i.e. it's
+atomic or asynchronous. If we're left focused on an atomic proposition we either
+instantiate the goal or fail. Otherwise the left focus is asynchronous and we
+can start inverting it. If we're right focused on a proposition that isn't right
 synchronous, we switch to inversion as well. Three rules model these conditions:
+%
 \begin{mathpar}
@@
     {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
-    {\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
 \and
@@
     }
-    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
-    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
+    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
+    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
     \rightarrow M_n : C}
@@
 
-where the ADT \emph{Foo} and its constructors stand for any ADT with form
+where the ADT $T$ and its constructors stand for any ADT with form
 \[
-    \textrm{data Foo} = \textrm{Bar}_1\ X_1\ \vert\ \textrm{Bar}_2\ X_2\ \vert\
-\dots\ \vert\ \textrm{Bar}_n\ X_n
+    \textrm{data}\ T = C_1\ X_1\ \vert\ C_2\ X_2\ \vert\
+\dots\ \vert\ C_n\ X_n
 \]
+
+% repetition does not legitimize :p
 
@@
 recursive (or "indutively defined"), meaning the type a constructor needs might
-be the type the constructor constructs \todo{ Na escrita, repetition
-  does not legitimize. Constructors can take as arguments values of
-  the type they are defining, ou algo assim}. Seemingly small, this change has a big
-impact in the synthesis process. Take, for instance, the ADT defined as
-$\textrm{data Foo} = \textrm{Bar Foo}$, the synth goal $\te{Foo}
-\lolli C$, and its derivation:
+be the type the constructor constructs \todo{ Na escrita, repetition does not
+legitimize. Constructors can take as arguments values of the type they are
+defining, ou algo assim}. Seemingly small, this change has a big impact in the
+synthesis process. Take, for instance, the ADT defined as $\textrm{data}\ T =
+C_1\ T$, the synth goal $T \lolli C$, and its derivation:
+%
 \begin{mathpar}
@@
                     {\dots}
-                    {\Gamma; \Delta/\Delta'; \Omega, z{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ z\ 
-                    \textrm{of Bar}\ k \rightarrow \dots : C}
+                    {\Gamma; \Delta/\Delta'; \Omega, z{:}T \Uparrow\ \vdash \textrm{case}\ z\ 
+                    \textrm{of}\ C_1\ k \rightarrow \dots : C}
                 }
-                {\Gamma; \Delta/\Delta'; \Omega, y{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ y\ 
-                \textrm{of Bar}\ z \rightarrow \dots : C}
+                {\Gamma; \Delta/\Delta'; \Omega, y{:}T \Uparrow\ \vdash \textrm{case}\ y\ 
+                \textrm{of}\ C_1\ z \rightarrow \dots : C}
             }
-            {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash \textrm{case}\ x\ 
-            \textrm{of Bar}\ y \rightarrow \dots : C}
+            {\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash \textrm{case}\ x\ 
+            \textrm{of}\ C_1\ y \rightarrow \dots : C}
         }
-        {\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \vdash \dots : C
+        {\Gamma; \Delta/\Delta'; \Omega, x{:}T \vdash \dots : C
         \Uparrow}
@@
     {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
-    \te{Foo} \lolli C \Uparrow}
-\end{mathpar}
+    T \lolli C \Uparrow}
+\end{mathpar}
+%
 Using our current system, we are to apply an infinite number of times the left
 ADT rule, never reaching a final state for the proof. Symmetrically, the
-derivation for the synth goal $\te{Foo}$ is also infinite.
+derivation for the synth goal $T$ is also infinite.
+%
 \begin{mathpar}
@@
                 {\dots}
-                {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
+                {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
             }
-            {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
+            {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
         }
-        {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo} \Downarrow}
+        {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
     }
-    {\Gamma; \Delta/\Delta'; \Omega \vdash \textrm{Bar}\ \dots : \te{Foo}
-    \Downarrow}
-\end{mathpar}
+    {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
+\end{mathpar}
+%
 To account for this situation, we impede the decomposition of an ADT in subsequent
-proofs of its branches, and to impede the construction of an ADT in a subsequent
-proof of a type to construct it. \todo{Não se percebe muito bem a
-  segunda parte da frase, rephrase!}
+proofs of its branches, and, simetrically, don't allow construction of an
+ADT when trying to synth an argument for its constructor. \todo{R: A segunda parte confusa era da frase de
+  cima?, está melhor?}
 %
@@
 For this, we need two more contexts, $\Rho_C$
-for restrictions \todo{constraints} on construction and $\Rho_D$ for restrictions on
+for constraints on construction and $\Rho_D$ for constraints on
 deconstruction. For convenience, they are represented by a single $\Rho$ if
@@
 modified to take this into consideration:
+%
 \begin{mathpar}
@@
     {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
-    \te{Foo} \notin \Rho_C}
-    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ \textrm{Bar}_n \ M : \te{Foo} \Downarrow}
+    T \notin \Rho_C}
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
 \and
@@
     {
-        \te{Foo} \notin \Rho_D
+        T \notin \Rho_D
         \and
@@
     }
-    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}\te{Foo} \Uparrow\
-    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ \textrm{Bar}_n\ y_n
+    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
+    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
     \rightarrow M_n : C}
@@
 \begin{mathpar}
-    \Rho'_C = \textrm{\textbf{if}}\ \te{Foo}\ \textrm{is recursive \textbf{then}}\ \Rho_C,
-    \te{Foo}\ \textrm{\textbf{else}}\ \Rho_C
+    \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
+    T\ \textrm{\textbf{else}}\ \Rho_C
 
@@
 \end{mathpar}
-These modifications prevent the derivation from not ending
-\todo{infinite derivations} for the scenarios described above,
-however, they also greatly limit the space of derivable programs,
-leaving the synthetiser effectively unable to synth specifications
-with recursive types\todo{refrasear tendo em conta que se sintetiza a
-  partir de specs e nao specs}, which is so undesirable that we add three rules
-to complement the restrictions on construction and destruction of
-recursive types.
+%
+These modifications prevent the infinite derivations in the scenarios described
+above, however, they also greatly limit the space of derivable programs, leaving
+the synthetiser effectively unable to synth from specifications with recursive
+types\todo{refrasear tendo em conta que se sintetiza a partir de specs e nao
+specs R: refrasear mas só com "from" não fica bem?}, which is so undesirable that we add three rules to complement the
+restrictions on construction and destruction of recursive types.
 
@@
     {
-        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}\te{Foo}/\Delta'; \Omega \Uparrow\ \vdash M : C
-        \and
-        \te{Foo} \in \Rho_D
+        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
+        \and
+        T \in \Rho_D
     }
-    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}\te{Foo} \Uparrow\ \vdash M : C}
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
     \and
@@
     {  }
-    {\Rho; \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash x : \te{Foo}}
+    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}
     \and
     \infer*[right=(adt$\Downarrow$L)]
-    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Uparrow\ \vdash M :
-    \te{Foo} \and \te{Foo} \notin \Rho_D}
-    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}\te{Foo} \Downarrow\ \vdash M : \te{Foo}}
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
+    T \and T \notin \Rho_D}
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
 \end{mathpar}
@@
 
-\todo[inline]{esta frase nao me parece muito necessaria}
-The ADT rules draw closer to a final form, but will still need a revision
-after the introduction of polymorfism and polymorfic parametrization on ADTs.
-
 \mypara{Recursion} The main idea behind synthesis of recursive programs is the
@@
 unrestricted context. That is, to synth a function of type $A \lolli B$ named
-\emph{foo}, the initial judgment can be written as
+\emph{f}, the initial judgment can be written as
 \begin{mathpar}
@@
     {\dots}
-    {\Gamma, \textrm{\emph{foo}}{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
+    {\Gamma, f{:}A \lolli B; \Delta/\Delta'; \Omega \vdash M :
     A \lolli B \Uparrow}
@@
 and, by definition, all subsequent inference rules will have
-($\textrm{\emph{foo}}{:}A \lolli B$) in the $\Gamma$ context too.
+($f{:}A \lolli B$) in the $\Gamma$ context too.
 We can also force the usage of the recursive call by adding it not only to the
@@
 function that just calls itself) that could even shadow other valid, relevant,
-solutions.  Instead, our framework only allows the use of recursion after having
+solutions.  Instead, our framework allows the use of recursion only after having
 deconstructed a recursive ADT. To avoid complicating the rules, we'll add the
@@
 "available" when deconstructing ($\textrm {List} \vdash C$), and only in
-the \emph{Cons} branch. % (TODO: rewrite)
-
-% TODO: E como evitamos coisas trivialmente estúpidas?
-
-\todo[inline]{Polymorphic}
-
-\mypara{Polymorphic Types} A polymorphic specification is a type $X$ \todo{T} and a set of
-bound variables that quantify universally the type variables in $X$, called a
-\emph{scheme} ($\forall \overline{\alpha}.\ \tau$).
-\todo[inline]{Não é um tipo e um conjunto de variáveis bound. É um
-  tipo da forma $\forall \overline{\alpha}.\ T$ em que $\dots$ }
-Synthesis of a scheme is straightforward. \todo{Diria que nao e
-  especialmente straightforward.}
+the \emph{Cons} branch. % (TODO: rewrite? O bold é meio estranho não?)
+
+\todo[inline]{Polymorphic R: ??}
+
+\mypara{Polymorphic Types} A polymorphic specification is a type of form
+$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of
+bound variables that quantify universally the type variables in $\tau$, which is
+also called a \emph{scheme}.
+Synthesis of a scheme comprises of turning it into a non-quantified type, and
+then treating its type variables agnostically. \todo{R: Melhor que
+straightforward?}
 First, \textbf{type variables are considered atomic types}, then, instantiate
-the bound vars\todo{variables} of the scheme as described by the Hindley-Milner inference
+the bound variables of the scheme as described by the Hindley-Milner inference
 method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
 each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha
-\lolli \alpha$ could be instanced \todo{instantiated} to
+\lolli \alpha$ could be instantiated to
 $\alpha0 \lolli \alpha0$. We add a rule for this:
@@
 \]
-As such, the construction of a derivation in which, to some types, no specific rules can be applied (except for the \textsc{init} rule,
-a type variable can instance a goal that's a type variable of the same name),
-corresponds to the synthesis of a program where some expressions are treated
-agnostically -- nothing constrains their type, meaning the program is
-polymorphic * TODO: rewrite this?.\todo[inline]{Use an example.}
+%
+As such, the construction of a derivation in which, to some types, no specific
+rules can be applied (except for the \textsc{init} rule, a type variable can
+instance a goal that's a type variable of the same name), corresponds to the
+synthesis of a program where some expressions are treated agnostically --
+nothing constrains their type -- meaning the program is polymorphic. The
+simplest example is the polymorphic function \emph{id} of type $\forall \alpha
+.\ \alpha \lolli \alpha$. The
+program synthed from that specification is $\lambda x \rightarrow x$, a lambda
+abstraction that doesn't constrain its parameter $x$ in any way. \todo{R :
+rewrite this? não estranho?}
 
@@
 goal, are equivalent. Because comparing all of these is slow, our implementation
-hashes the relevant values and uses the result as the key. Using LogicT, a
-branch we want to memoize might fail -- it's desireable to also record that a
-synthesis fails, as it happens more than successful results. We capture failure
+hashes the relevant values and uses the result as the key. A
+branch we want to memoize might fail, not providing any result -- we also want
+to record synthesis failure, as it happens more often than successful results. We capture failure
 of a logict computation using the $ifte$ function and represent failure or
 success information with the Maybe type. We ran benchmarks to measure the impact
-of memoization on the performance. It's more noticeable in synthesis with
+of memoization on the performance. It's especially noticeable in synthesis with
 multiple functions in the unrestricted context, or more complex
@@
         \hline
-        insertWithKey & 13.31 s ± 593.4 ms & 2.667 s ± 284.3 ms
+        insertWithKey & 13.31 s $\pm$ 593.4 ms & 2.667 s $\pm$ 284.3 ms \\
+        \hline
     \end{tabular}
@@
 taking sometimes full days to solve (and as of now, some still remain unsolved,
-e.g. List (List a) -o List a). To make the debugging experience better, we
+e.g. List (List a) $\lolli$ List a). To make the debugging experience better, we
 developed a tracing system that prints information whenever a rule is applied,

### 2021-07-14 article/article.tex
fixes according to prof

@@
 
+\usepackage{minted}
+
 \usepackage{todonotes}
@@
 
+% \usemintedstyle{lovelace}
+\usemintedstyle{murphy}
+
 \title{Synthesis of Linear Functional Programs}
-\author{Rodrigo Mesquita - Nº XXXXX \\
+\author{Rodrigo Mesquita - Nº 55902 \\
     Supervisor: Bernardo Toninho}
@@
 Synthesis can be seen as a means to improve a programmer's productivity and/or
-program correctness (i.e. through suggestion and/or autocompletion), or as a
+program correctness (e.g. through suggestion and/or autocompletion), or as a
 tool to automate certain parts of the programming process, to name a few
 examples.
+% TODO: Escrever um início melhor
 
@@
 form of linear types, which limit resource usage in programs, in combination
-with polymorphism.  Concretely, we will explore the proof-theoretic foundations
+with polymorphism and refinements.  Concretely, we will explore the proof-theoretic foundations
 of linear types via the Curry-Howard correspondence with linear logic, viewing
@@
 representation\ref{sec:formal_system}, we can (and did) create a very robust
-synthetiser for a select part of the linear types that's able to synth
+synthetiser for a select part of the linear types that's able to synthetise
 programs successfully from e.g. the major linear logic equivalences. For example,
 given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
-the program $\lambda a \rightarrow \lambda b \rightarrow$ let $c \tensor d$ =
+the program $\lambda a .\ \lambda b .$ let $c \tensor d$ =
 $b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.
@@
 However, a programming language only with the canonical linear types lacks
-expressiveness. To enhance \synname\ and its language, synth more interesting
+expressiveness. To enhance \synname\ and its language, synthetise more interesting
 programs -- and thus empirically proving the feasability of more relevant
@@
 
-\section{Synthesis with Linear Types, Polymorphism and Refinements}\label{sec:formal_system}
-\todo[inline]{Encurtar titulo, mas algo nesta onda que descreve
-  ``tudo''}
+\section{Synthesis Framework}\label{sec:formal_system}
+% with Linear Types, Polymorphism and Refinements
+
+% \todo[inline]{Encurtar titulo, mas algo nesta onda que descreve
+%   ``tudo''}
 
@@
 With this section we aim to formalize the rules that guide
-synthesis from more expressive specifications, alongside the already
-well defined rules that model the core of the synthesis process, resulting in
-a \emph{sound} set of rules that hopes to serve as a source of ideas
-\todo{ nao me parece\dots As regras caracterizam o que esta
-  implementado, e esse o objectivo, nao e ser uma fonte de ideias\dots}
-  \todo{R: não parece ser sound ?}
-on taming synthesis of recursive functions with richer types in a
-linear context, as well as describe the system in enough detail for
+the synthesis process for more expressive specifications, alongside the already
+well defined rules that model the core of the synthetiser, resulting in
+a \emph{sound} set of rules that characterizes our framework for synthesis of
+recursive functions from polymorphic and refinement types in a linear context
+% TODO: Better?....
+, and describes the system in enough detail for
 the synthetiser to be reproducible by a theory-driven implementation.
@@
 different meanings, and our \emph{sound} set of rules only guarantees we won't
-synth incorrect programs, but still hides a possibly infinite number of correct
+synthetise incorrect programs, but still hides a possibly infinite number of correct
 proofs; and that the valid programs derivable through it reflect the somewhat
@@
 \mypara{Core Rules} The core system comprises of the sequent calculus rules for
-linear logic connectives with resource-management and focusing. As
-explained in\ref{sec:background} \todo{se o background nao vai ser
-  incluido no documento principal, nao deves referenciar assim tao
-  directamente para o conteudo principal.}, focusing separates the proof search into two
+linear logic connectives with resource-management and focusing.
+\todo{Introdução para focusing...}
+Focusing separates the proof search into two
 phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
@@
 by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
-first rule is read: to synth a program of type $A \lolli B$ while inverting
+first rule is read: to synthetise a program of type $A \lolli B$ while inverting
 right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
-context $\Delta$, and inversion context $\Omega$, synth a program of type $B$ with
+context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
 an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
@@
 context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
-resulting program is $\lambda x \rightarrow M$ and the output linear context is
+resulting program is $\lambda x . M$ and the output linear context is
 $\Delta'$.
-
-\todo{a sintaxe $\lambda x \rightarrow M$ nao e mt standard, usa
-  $\lambda x . M$}
 
@@
     \notin \Delta'}
-    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x \rightarrow M : A
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x . M : A
     \lolli B \Uparrow}
@@
 The rules written above together make the core of our synthetiser. Next, we'll
-present new rules that align and build on top of these to synth recursive
+present new rules that align and build on top of these to synthetise recursive
 programs from more expressive (richer) types.
@@
 definition of an ADT which might be constructed with zero, one, or two
-propositions of type $A$, using its syntax: $\textrm{data Container} =
-\textrm{None}\ 1\ \vert\ \textrm{One}\ A\ \vert\ \textrm{Two}\ (A * A)$.
+propositions of type $A$, using its syntax: \mintinline{haskell}{data Container
+= None 1 | One A | Two (A * A)}
 
@@
 constructor:
-
-\todo[inline]{Não uses Foo e Bar. $T_1$, $T_2$ , $\dots$ para
-  tipos. $C_1$, $C_2$, $\dots$ para os construtores.}
 
@@
 where the ADT $T$ and its constructors stand for any ADT with form
-\[
-    \textrm{data}\ T = C_1\ X_1\ \vert\ C_2\ X_2\ \vert\
-\dots\ \vert\ C_n\ X_n
-\]
+\begin{minted}{haskell}
+    data T = C1 X1 | C2 X2 | ... | Cn Xn
+\end{minted}
 
@@
 A more general (and interesting) formulation of ADTs says an ADT can be
-recursive (or "indutively defined"), meaning the type a constructor needs might
-be the type the constructor constructs \todo{ Na escrita, repetition does not
-legitimize. Constructors can take as arguments values of the type they are
-defining, ou algo assim}. Seemingly small, this change has a big impact in the
-synthesis process. Take, for instance, the ADT defined as $\textrm{data}\ T =
-C_1\ T$, the synth goal $T \lolli C$, and its derivation:
+recursive (or "indutively defined"), meaning constructors can take as arguments
+values of the type they are defining. Seemingly small, this change has a big
+impact in the synthesis process. Take, for instance, the ADT defined as
+\mintinline{haskell}{data T = C1 T}, the synth goal $T \lolli C$, and its
+derivation:
 %
@@
     }
-    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x \rightarrow \dots :
+    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x . \dots :
     T \lolli C \Uparrow}
@@
 To account for this situation, we impede the decomposition of an ADT in subsequent
-proofs of its branches, and, simetrically, don't allow construction of an
-ADT when trying to synth an argument for its constructor. \todo{R: A segunda parte confusa era da frase de
-  cima?, está melhor?}
+proofs of its branches, and, symetrically, don't allow construction of an
+ADT when trying to synthetise an argument for its constructor.
 %
@@
 for constraints on construction and $\Rho_D$ for constraints on
-deconstruction. For convenience, they are represented by a single $\Rho$ if
+deconstruction. Together, they hold a list of ADTs that can not be constructed or
+deconstructed at some point in the proof. For convenience, they are represented by a single $\Rho$ if
 unused. All rules must trivially propagate these. The ADT rules are explicitly
@@
 above, however, they also greatly limit the space of derivable programs, leaving
-the synthetiser effectively unable to synth from specifications with recursive
-types\todo{refrasear tendo em conta que se sintetiza a partir de specs e nao
-specs R: refrasear mas só com "from" não fica bem?}, which is so undesirable that we add three rules to complement the
+the synthetiser effectively unable to synthetise from specifications with recursive
+types, which is undesirable, so we add three rules to complement the
 restrictions on construction and destruction of recursive types.
@@
 labeling of the main goal and the addition of its type, under that name, to the
-unrestricted context. That is, to synth a function of type $A \lolli B$ named
+unrestricted context. That is, to synthetise a function of type $A \lolli B$ named
 \emph{f}, the initial judgment can be written as
@@
 
-\todo[inline]{Polymorphic R: ??}
-
 \mypara{Polymorphic Types} A polymorphic specification is a type of form
-$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of
-bound variables that quantify universally the type variables in $\tau$, which is
-also called a \emph{scheme}.
-Synthesis of a scheme comprises of turning it into a non-quantified type, and
-then treating its type variables agnostically. \todo{R: Melhor que
-straightforward?}
-First, \textbf{type variables are considered atomic types}, then, instantiate
-the bound variables of the scheme as described by the Hindley-Milner inference
-method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names for
-each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha
-\lolli \alpha$ could be instantiated to
-$\alpha0 \lolli \alpha0$. We add a rule for this:
+$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of bound
+variables that quantify universally the type variables in $\tau$, which is also
+called a \emph{scheme}.  Synthesis of a scheme comprises of turning it into a
+non-quantified type, and then treating its type variables agnostically.  First,
+\textbf{type variables are considered atomic types}, then, instantiate the bound
+variables of the scheme as described by the Hindley-Milner inference
+method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names
+for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
+\alpha$ could be instantiated to $\alpha0 \lolli \alpha0$. We add a rule for
+this:
 \begin{mathpar}
@@
 .\ \alpha \lolli \alpha$. The
-program synthed from that specification is $\lambda x \rightarrow x$, a lambda
-abstraction that doesn't constrain its parameter $x$ in any way. \todo{R :
-rewrite this? não estranho?}
+program synthetised from that specification is $\lambda x . x$, a lambda
+abstraction that doesn't constrain its parameter $x$ in any way.
+% TODO: \todo{R : rewrite this? não estranho?}
 
@@
 havin an input and output context ($\Theta/\Theta'$)).  In contrast to
-HM, everytime a constraint is added it is solved against all other
+Hindley-Milner's inference, everytime a constraint is added it is solved against all other
 constraints -- a branch of the proof search is desired to fail as soon
@@
 types in the synth, and the first represent an existential instance of
-a scheme, that might inclusively constraint normal type
+a scheme, that might inclusively constrain normal type
 variables. Additionally, it's required that all existential type
@@
     \infer*[right=($?L$)]
-    {\textsc{solve}(?\alpha
+    {\textsc{unify}(?\alpha
     \mapsto C, \Theta)}
@@
     \infer*[right=($\Downarrow L?$)]
-    {\textsc{solve}(?\alpha
+    {\textsc{unify}(?\alpha
     \mapsto A, \Theta)}
@@
     \\
-    \textsc{solve}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
-    \textrm{can be solved with those in}\ \Theta
-\end{mathpar}
-
-\mypara{Polymorfic ADTs}
+    \textsc{unify}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
+    \textrm{can be unified with those in}\ \Theta
+\end{mathpar}
+
+\mypara{Polymorphic ADTs}
 
@@
 The desugaring module converts the frontend AST to a core AST without syntatic
-sugar and using the locally-nameless~\cite{locally nameless} representation for
-all bound variables. \todo{Explica o que e isto} This way inference can be done without worrying about name
-conflicts.
+sugar and using the locally-nameless representation~\cite{locally nameless}, in
+which all bound variables have no name, and instead are addressed using the
+number of binders traversed since the binding.
+% TODO: não ficou muito mais claro :\todo{Explica o que e isto}
+This way inference can be done without worrying about name conflicts.
 
@@
 \mypara{Backtracking} Even though the synthetiser formal system helps, there's
-still non-determinism in the synth process. At diverse times in the proof
+still non-determinism in the synthesis process. At diverse times in the proof
 construction we must choose one of multiple rules to apply, and in the event of
@@
 
-\mypara{SMT Solving} To typecheck and synth refinement types and dependent
+\mypara{SMT Solving} To typecheck and synthetise refinement types and dependent
 functions, we need to contact an SMT solver to check satisfiability. For this,
@@
 support the usage of the SMT solvers to satisfy and output more complicated
-first-order logic uninterpreted functions -- since this is needed to synth refinement types
+first-order logic uninterpreted functions -- since this is needed to synthetise refinement types
 when right focused, we instead use unsafe library primitives to communicate with
@@
 
-\mypara{Memoization} Some proofs attempt to synth the same sub-goal with the
-same premises more than once. Since these proof attempts might be expensive
+\mypara{Memoization} Some proofs attempt to synthetise the same sub-goal with
+the same premises more than once. Since these proof attempts might be expensive
 performance-wise, and we can rely on the program's determinism (equal calls to
-synth result in equal outcomes), we add to our synth memoization, an
+synthetise result in equal outcomes), we add to our synthetiser memoization, an
 optimization technique that stores function calls and returns cached values for
-equal inputs. For our synth, two calls are equivalent when contexts $\Theta,
-\Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and the
-goal, are equivalent. Because comparing all of these is slow, our implementation
-hashes the relevant values and uses the result as the key. A
-branch we want to memoize might fail, not providing any result -- we also want
-to record synthesis failure, as it happens more often than successful results. We capture failure
-of a logict computation using the $ifte$ function and represent failure or
-success information with the Maybe type. We ran benchmarks to measure the impact
-of memoization on the performance. It's especially noticeable in synthesis with
-multiple functions in the unrestricted context, or more complex
-specifications:
+equal inputs. For our synthetiser, two calls are equivalent when contexts
+$\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and
+the goal, are equivalent. Because comparing all of these is slow, our
+implementation hashes the relevant values and uses the result as the key.
+Furthermore, a branch we want to memoize might fail, not providing any result --
+we also want to record synthesis failure, as it happens more often than
+successful results. We capture failure of a logict computation using the $ifte$
+function and represent failure or success information with the Maybe type. We
+ran benchmarks to measure the impact of memoization on the performance. It's
+especially noticeable in synthesis with multiple functions in the unrestricted
+context, or more complex specifications:
 %
@@
 \end{center}
+%
+The details of the memoization are very subject to change
 

### 2021-07-14 STLLC/ListT.hs

@@ -1,0 +1,14 @@
+data List a = Nil | Cons (a * List a);
+
+data Tree a = Tip | Node (a * List (Tree a));
+
+
+
+#synth test :: ((A -o B) -> Tree A -o Tree B);
+
+synth concat :: (List (List Int) -o List Int);
+
+synth concat2 :: (List (List A) -o List A);
+
+synth zip :: ((A -o B -o C) -> List A -o List B -o List C);
+

### 2021-07-15 article/article.tex
Introduction first iteration

@@
 
-% \usemintedstyle{lovelace}
-\usemintedstyle{murphy}
-
+\usemintedstyle{manni}
+
+% TODO: Temos de ter o título assim? ficava mais fixe sem o número e sem o
+% supervisor :P
 \title{Synthesis of Linear Functional Programs}
@@
 Program synthesis is an automated or semi-automated process of deriving a
-program (i.e.~generating code) from a high-level specification.
-%
-Synthesis can be seen as a means to improve a programmer's productivity and/or
-program correctness (e.g. through suggestion and/or autocompletion), or as a
-tool to automate certain parts of the programming process, to name a few
-examples.
-% TODO: Escrever um início melhor
-
-Specifications can take many forms (e.g.  polymorphic refinement
-types~\cite{DBLP:conf/pldi/PolikarpovaKS16},
-examples~\cite{DBLP:conf/popl/FrankleOWZ16}, graded linear
-types~\cite{DBLP:conf/lopstr/HughesO20}).  Regardless of the kind of
-specification, program synthesizers have to deal with two main inherent sources
-of complexity: searching over a vast space of (potentially) valid programs, and
-interpreting user intent.
-%
-In this work we'll explore type-based synthesis where specifications take the
-form of linear types, which limit resource usage in programs, in combination
-with polymorphism and refinements.  Concretely, we will explore the proof-theoretic foundations
-of linear types via the Curry-Howard correspondence with linear logic, viewing
-program synthesis under the lens of (linear logic) bottom-up proof search.
-
-%
-
-\mypara{Type-Based Synthesis}
-% Type based synthesis
-Type-based synthesis uses types as a form of program specification,
-automatically or semi-automatically producing a program expression
-with the given type and so matching the specification.
-%
-In order to make type-based specifications more expressive, type-based
-synthesis frameworks use rich type systems such as refinement
-types~\cite{DBLP:conf/pldi/PolikarpovaKS16} and polymorphic types~\cite{DBLP:conf/pldi/PolikarpovaKS16}.
-
-Richer type systems allow for more precise types, which can
-statically eliminate various kinds of logical errors by making certain
-invalid program states ill-typed (e.g.,~a ``null aware'' type system,
-will ensure at compile-time that you cannot dereference a
-null-pointer). However, they can also be a burden -- the whole point
-of these type systems is to ensure that \emph{fewer} programs are
-deemed well-typed, which can pose additional challenges to the
-development process. Type-based synthesis leverages rich types as a
-way of pruning the search space, and by using types the user 
-is given a more ``familiar'' specification interface. For instance, the type
-$\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow \mathsf{Int}$
-specifies a (curried) function that takes two integers and produces an
-integer. Viewed as a specification, it is extremely imprecise (there
-are an infinite number of functions that satisfy this specification).
-However, the richer type $(x{:}\mathsf{Int}) \rightarrow
-(y{:}\mathsf{Int}) \rightarrow \{z{:}\mathsf{Int} \mid z = x+y\}$
-very precisely specifies a function that takes integer arguments $x$
-and $y$ and returns an integer $z$ that is the sum of $x$ and $y$.
-
-%
-This work explores the synthesis of linear functional
-programs from types based on linear logic propositions (i.e.~linear
-types) by leveraging the Curry-Howard correspondence.
+program, i.e.~generating code, from a high-level specification. Synthesis can be
+seen as a means to improve a programmer's productivity and program correctness
+(e.g. through suggestion and autocompletion).
+%
+Specifications can take many forms: natural language~\cite{chen2021evaluating},
+examples~\cite{DBLP:conf/popl/FrankleOWZ16}, polymorphic refinement
+types~\cite{DBLP:conf/pldi/PolikarpovaKS16}, graded linear
+types~\cite{DBLP:conf/lopstr/HughesO20}, \emph{etc.}, but, regardless the kind of
+specification, program synthetisers have to deal with two main inherent sources
+of complexity -- searching over a vast space of (potentially) valid programs,
+and interpreting user intent.
+
+Synthesis is said to be \emph{type-driven} when it uses types as a form of program
+specification and produces an expression whose type matches the specification.
+Type-driven synthesis frameworks usually leverage rich types as a way to make
+specifications more expressive and prune the valid programs search space, while
+maintaining a ``familiar'' specification interface (types) for the user.
+%
+Richer type systems allow for more precise types, which can statically eliminate
+various kinds of logical errors by making certain invalid program states
+ill-typed (e.g.,~a ``null aware'' type system will ensure at compile-time that
+you cannot dereference a null-pointer).
+%
+% However, they can also be a burden -- the whole point of these type systems is
+% to ensure that \emph{fewer} programs are deemed well-typed, which can pose
+% additional challenges to the development process.
+%
+For instance, the type $\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow
+\mathsf{Int}$ specifies a (curried) function that takes two integers and
+produces an integer. Viewed as a specification, it is extremely imprecise (there
+are an infinite number of functions that satisfy this specification).  However,
+the richer type $(x{:}\mathsf{Int}) \rightarrow (y{:}\mathsf{Int}) \rightarrow
+\{z{:}\mathsf{Int} \mid z = x+y\}$ very precisely specifies a function that
+takes integer arguments $x$ and $y$ and returns an integer $z$ that is the sum
+of $x$ and $y$.
+
+The focus of our work is on type-driven synthesis where specifications take the
+form of linear types.
+%
+% TODO: Deixar isto para mais à frente?  More concretely, we will explore the
+% proof-theoretic foundations of linear types via the Curry-Howard
+% correspondence with linear logic propositions, viewing program synthesis as
+% proof search in linear logic.  (Said correspondence maps \emph{types} to
+% \emph{propositions in a logic}, and \emph{programs} to \emph{proofs}).
+%
+Linear types differ from more traditional types in that they constrain resource
+usage in programs by \emph{statically} limiting the number of times certain
+resources can be used during their lifetime (linear resources must be used
+\emph{exactly once}).
+%
+They can be applied to resource-aware programming such as concurrent programming
+(e.g. session types for message passing
+concurrency~\cite{DBLP:journals/mscs/CairesPT16}), memory-management
+(e.g.~Rust's ownership types), safely updating-in-place mutable
+structures~\cite{Bernardy_2018}, enforcing protocols for external \textsc{api}s~\cite{Bernardy_2018}, to name a few.
+% TODO: Repensar esta parte dos exemplos de aplicação e de como podemos
+% relacionar o nosso trabalho com futuras aplicações nestes campos
+
+Despite their long-known potential (\emph{linear types can change the
+world!}\cite{Wadler90lineartypes}, \cite{DBLP:journals/mscs/CairesPT16,Bernardy_2018}) and,
+primarily, strong proof-theoretic
+foundations \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00}, synthesis of linear functional programs
+from linear types hasn't been previously explored in an extended context (e.g.
+with recursion, polymorphism, ...).
+%TODO: Sinto que faltam algumas palavras aqui..
+
+In this paper we present a framework for synthesis of linear types extended with
+recursive algebraic data types, parametric polymorphism and refinements.
+% TODO: extensible? proving feasability of... ? showing potential for ... ?
+
+First, we discuss the use of linear types as specifications, the outline of the
+synthesis process, and what that entails for the synthetiser -- ilustrated with
+concrete results (Section~\ref{sec:overview}). Then, we present the
+algorithm and formal system driving the synthesis
+(Section~\ref{sec:formal_system}). Aftwerwards, we describe the architecture 
+of our synthetiser named \synname, and examine technical details and challenges of the
+implementation (Section~\ref{sec:architecture}). Finally, we evaluate our work
+through expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks
+(i.e.~how fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
+
+% TODO: Why linear types?
+
+%%
+
+%%
+%This work explores the synthesis of linear functional
+%programs from types based on linear logic propositions (i.e.~linear
+%types) by leveraging the Curry-Howard correspondence.
 
@@
 % literature.
-
-Linear types differ from more traditional types in
-that they constrain resource usage in programs by \emph{statically}
-limiting the number of times certain resources can be used during
-their lifetime.  They can be applied to resource-aware programming
-such as concurrent programming (e.g. session types for message passing
-concurrency~\cite{DBLP:journals/mscs/CairesPT16}), and to memory-management (e.g.~Rust's ownership
-types).
 
@@
 
-\section{Overview}
-
-Escrever introdução melhor que tenha isto como seguimento...e que siga duma
-introdução extendida
+\section{Overview}\label{sec:overview}
+
+The \synname\ synthetiser combines the features of linear types, recursion,
+parametric polymorphism, recursive polymorphic? algebraic data types, and
+refinement types.
+
+The \synname\ language is an extension of the linear $\lambda$-calculus.
+
+... in proof search, our programs are correct by construction -- if
+we were able to synth it, then it typechecks.
+
+Linear types provide a more certain specification, when compared to simple
+traditional types. By requiring some types to be used linearly (consumed) we can
+assure, for example, the deconstruction of an ADT. Given the goal $\bang(a \lolli b)
+\lolli List\ a \lolli List\ b$, the program $\lambda x. \lambda y. Nil$ wouldn't
+be valid. To synthetise a valid program, $List\ a$
+would have to be deconstructed, and its constructor arguments used, resulting
+in the map function without relying on any additional information:
+
+\begin{minted}{haskell}
+map c d = let !e = c in
+    case d of
+          Nil  -> Nil
+        | Cons f ->
+            let g*h = f in Cons (e g, map (!e) h);
+\end{minted}
 
@@
 is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
-resource management. It connects the concepts seen in
-background~\ref{sec:background} -- a type system
-consisting of inference rules in sequent calculus for linear logic connectives;
-with resource management, and focusing, as formalized techniques to,
-respectively, solve resource related non-determinism, and to eliminate from the
-proof search don't-care non-determinism.  Guided solely by this system's formal
+resource management -- a type system consisting of inference rules in sequent
+calculus for linear logic connectives; with resource management, and focusing,
+as formalized techniques to, respectively, solve resource related
+non-determinism, and to eliminate from the proof search don't-care
+non-determinism.  Guided solely by this system's formal
 representation\ref{sec:formal_system}, we can (and did) create a very robust
 synthetiser for a select part of the linear types that's able to synthetise
-programs successfully from e.g. the major linear logic equivalences. For example,
-given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$,
-the program $\lambda a .\ \lambda b .$ let $c \tensor d$ =
-$b$ in $a$ $c$ $d$ (commonly known as \emph{uncurry}) is synthetised.
+programs successfully from e.g. the major linear logic equivalences. For
+example, given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B)
+\lolli C$, the program $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$
+$c$ $d$ (commonly known as \emph{uncurry}) is synthetised.
 
@@
 ADT deconstruction}, i.e. the recursive call should only take "smaller", recursive,
-hypothesis as arguments. To illustrate, having $\textrm{data List} =
-\textrm{Nil}\ \vert\ \textrm{Cons}\ (A * \textrm{List})$, recursion is only
+hypothesis as arguments. To illustrate, having \mintinline{haskell}{data List =
+Nil | Cons (A * List)}, recursion is only
 "available" when deconstructing ($\textrm {List} \vdash C$), and only in
@@
 
-\section{Architecture}
+\section{Architecture}\label{sec:architecture}
 
@@
 %
-The details of the memoization are very subject to change
+The implementation could be improved by better combining the
+memoization with the LogicT monad transformer (remembering decisions across
+backtracking branches) and by finetuning key-equality for polymorphic types.
 
@@
 
-\mypara{Evaluation} Validation and evaluation of our work is fairly
-easy. Simple questions such as \emph{does it typecheck}? \emph{can
-  we synthesize}? \emph{how fast}?  are the core of our evaluation. In
-more concrete terms, we can compare our times to state of the art
-synthetizers, present a list of working examples of programs synthesized
-by our framework, and corresponding benchmarks.
-
-On the eventual addition of recursive functions, refinement types, and
-interactive synthesis, we'll consider, respectively, other examples to
-showcase recursive functions synthesis, updated times for benchmarks
-with refinement types, and a display of the possible user interaction
-in the user process, and the resulting function following it.
+% \mypara{Evaluation} Validation and evaluation of our work is fairly
+% easy. Simple questions such as \emph{does it typecheck}? \emph{can
+%   we synthesize}? \emph{how fast}?  are the core of our evaluation. In
+% more concrete terms, we can compare our times to state of the art
+% synthetizers, present a list of working examples of programs synthesized
+% by our framework, and corresponding benchmarks.
+
+% On the eventual addition of recursive functions, refinement types, and
+% interactive synthesis, we'll consider, respectively, other examples to
+% showcase recursive functions synthesis, updated times for benchmarks
+% with refinement types, and a display of the possible user interaction
+% in the user process, and the resulting function following it.
 
@@
 
-\section{Evaluation}
+\section{Evaluation}\label{sec:evaluation}
 
@@
 
+Synthesis of
+
 \section{Background}\label{sec:background}

### 2021-07-15 article/article.tex
tweaks

@@
 foundations \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00}, synthesis of linear functional programs
-from linear types hasn't been previously explored in an extended context (e.g.
+from linear types hasn't been previously explored in an ``advanced'' context (e.g.
 with recursion, polymorphism, ...).
@@
 expressiveness. To enhance \synname\ and its language, synthetise more interesting
-programs -- and thus empirically proving the feasability of more relevant
+programs -- and thus empirically prove the feasability of more relevant
 synthesis in a linear context, we extend the syntax and type system with

### 2021-07-15 article/article.tex
retoques

@@
 program, i.e.~generating code, from a high-level specification. Synthesis can be
-seen as a means to improve a programmer's productivity and program correctness
+seen as a means to improve programmer productivity and program correctness
 (e.g. through suggestion and autocompletion).
 %
-Specifications can take many forms: natural language~\cite{chen2021evaluating},
-examples~\cite{DBLP:conf/popl/FrankleOWZ16}, polymorphic refinement
-types~\cite{DBLP:conf/pldi/PolikarpovaKS16}, graded linear
-types~\cite{DBLP:conf/lopstr/HughesO20}, \emph{etc.}, but, regardless the kind of
-specification, program synthetisers have to deal with two main inherent sources
-of complexity -- searching over a vast space of (potentially) valid programs,
-and interpreting user intent.
+Specifications can take many forms such as natural
+language~\cite{chen2021evaluating},
+examples~\cite{DBLP:conf/popl/FrankleOWZ16} or rich types such as
+polymorphic refinement types~\cite{DBLP:conf/pldi/PolikarpovaKS16} or
+graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
+of specification, program synthesizers have to deal with two main
+inherent sources of complexity -- searching over a vast space of
+(potentially) valid programs, and interpreting user intent.
 

### 2021-07-16 article/article.tex
WIP: Overview

--- 2021-07-15 article/article.tex
@@
 (i.e.~how fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
-
-% TODO: Why linear types?
-
-%%
-
-%%
+%TODO: And appendice with...
+
 %This work explores the synthesis of linear functional
@@
 
-The \synname\ synthetiser combines the features of linear types, recursion,
+% TODO: Why linear types?
+The \synname\ synthetiser combines the features of linear types with recursion,
 parametric polymorphism, recursive polymorphic? algebraic data types, and
-refinement types.
-
-The \synname\ language is an extension of the linear $\lambda$-calculus.
-
-... in proof search, our programs are correct by construction -- if
-we were able to synth it, then it typechecks.
+refinement types. The synthetiser is built on top of a system of proof-search in
+linear logic. Proof-search relates to program synthesis through the Curry-Howard
+correspondence~\cite{}, which states that \emph{propositions in a logic} map
+directly to \emph{types}, and \emph{proofs} of those propositions to well-typed
+\emph{programs} -- finding a proof of a proposition is finding a program with
+that type.
+
+The core of the synthesis is a \emph{sound} and \emph{complete} system
+consisting of \emph{bottom-up} proof search in linear logic augmented with
+resource management~\cite{} and focusing~\cite{} as techniques to, respectively,
+solve resource related non-determinism, and eliminate from the proof-search
+don't-care non-determinism. A more in depth explanation is in
+section~\ref{sec:formal_system}.
+
+With the core in place, we can synthetise programs from specifications that use
+canonical linear types ($\lolli, \with, \tensor, 1, \oplus, \bang$), such as
+fundamental linear logic equivalences. For example, given the specification $(A
+\lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$, the program
+\mintinline{haskell}{\a -> \b -> let c*d = b in a c d}
+% $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$ $c$ $d$
+(commonly known as \emph{uncurry}) is synthetised.
+
+However, a programming language only with the canonical linear types lacks
+expressiveness. To enhance \synname\ and its language, synthetise more
+interesting programs -- and thus empirically prove the feasability of more
+relevant synthesis in a linear context, we extend the syntax and type system
+with recursiveness and richer types: -- polimorfic, refinements, adts
+parametrizáveis...  todo
+
+The \synname\ programming language is naturally born alongside the synthetiser
+to create an environment for it. Synth goals are inserted in the program by use
+of a \emph{synth} keyword or syntatic \emph{mark}, indicating a program should
+be generated for a given type. Additionally, it opens the possibility of synthesis with
+context, that is, with knowledge of other functions and structures defined in
+the same program.
+
+All additions to the synthesis are extensions of the proof-search in linear
+logic system -- in the end, the process still consists of constructing a proof.
+Because we can only construct valid proofs (a wrong proof isn't a proof), we
+know that all produced synthesis results are correct: \textbf{all programs are
+correct by construction}, i.e. if we're able to synthetise it, then it typechecks,
+and therefore fullfills the specification.
 
@@
 assure, for example, the deconstruction of an ADT. Given the goal $\bang(a \lolli b)
-\lolli List\ a \lolli List\ b$, the program $\lambda x. \lambda y. Nil$ wouldn't
-be valid. To synthetise a valid program, $List\ a$
+\lolli List\ a \lolli List\ b$, the program \mintinline{haskell}{\x -> \y ->
+Nil} wouldn't be valid. To synthetise a valid program,
+\mintinline{haskell}{List a}
 would have to be deconstructed, and its constructor arguments used, resulting
@@
 \end{minted}
-
-At the core of the synthesis process, and of the developed synthetiser \synname,
-is a \emph{sound} and \emph{complete?} system of focusing for linear logic with
-resource management -- a type system consisting of inference rules in sequent
-calculus for linear logic connectives; with resource management, and focusing,
-as formalized techniques to, respectively, solve resource related
-non-determinism, and to eliminate from the proof search don't-care
-non-determinism.  Guided solely by this system's formal
-representation\ref{sec:formal_system}, we can (and did) create a very robust
-synthetiser for a select part of the linear types that's able to synthetise
-programs successfully from e.g. the major linear logic equivalences. For
-example, given the specification $(A \lolli (B \lolli C)) \lolli (A \tensor B)
-\lolli C$, the program $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$
-$c$ $d$ (commonly known as \emph{uncurry}) is synthetised.
-
-The \synname\ programming language is naturally born alongside the synthetiser
-to create an environment for it. Synth goals are inserted in the program by use
-of a \emph{synth} keyword or syntatic \emph{mark}, indicating a program should
-be generated for a given, or inferred, type. Synth results are functions in this
-language's syntax. Additionally, it opens the possibility of synthetisis with
-context, that is, with knowledge of other functions and eventual structures
-defined in the same program.
-
-However, a programming language only with the canonical linear types lacks
-expressiveness. To enhance \synname\ and its language, synthetise more interesting
-programs -- and thus empirically prove the feasability of more relevant
-synthesis in a linear context, we extend the syntax and type system with
-recursiveness and richer
-types: -- polimorfic, refinements, adts parametrizáveis...  todo
 
@@
 % é agora inutil. Talvez seja demais adicionar à regra
+%
+Finally, these rules end up ensuring that a recursive ADT will deconstruct
+itself once, and move the recursive argument from its constructor to the linear
+context -- essentially forcing it to be used as a function argument, and most
+usually that function will be the synthesis recursive call.
 

### 2021-07-16 STLLC/Array.hs

@@ -1,0 +1,13 @@
+
+data List a = Nil | Cons (a * List a);
+
+
+newMArray :: (Int -o (MArray a -o !b) -o b);
+
+write :: (MArray a -o (Int * a) -o MArray a);
+
+freeze :: (MArray a -o !(Array a));
+
+foldl :: ((a -o b -o a) -> a -o (List b) -o a)
+
+synth array :: (Int -o List (Int * a) -o Array a);

### 2021-07-16 STLLC/a.hs

@@ -1,0 +1,5 @@
+
+
+convert :: (A -o B);
+
+synth main :: (A -o A -o (B * B))

### 2021-07-16 STLLC/app/Syntax.hs

@@ -115,9 +115,9 @@ instance (Show Expr) where
             showexpr' d (Fst e) = "(fst " ++ showexpr' d e ++ ")"
             showexpr' d (Snd e) = "(snd " ++ showexpr' d e ++ ")"
             -- showexpr' d (InjL (Just t) e) = "inl " ++ showexpr' d e ++ " : " ++ show t
-            showexpr' d (InjL _ e) = "inl " ++ showexpr' d e 
+            showexpr' d (InjL _ e) = "inl (" ++ showexpr' d e ++ ")"
             -- showexpr' d (InjR (Just t) e) = "inr " ++ show t ++ " : " ++ showexpr' d e
-            showexpr' d (InjR _ e) = "inr " ++ showexpr' d e
+            showexpr' d (InjR _ e) = "inr (" ++ showexpr' d e ++ ")"
             showexpr' d (CaseOfPlus e1 x e2 y e3) = indent d ++ "case " ++ showexpr' d e1 ++ " of " ++
                                                         indent (d+1) ++ "  inl " ++ x ++ " -> " ++ showexpr' (d+2) e2 ++
                                                         indent (d+1) ++ "| inr " ++ y ++ " -> " ++ showexpr' (d+2) e3

### 2021-07-16 article/article.tex
minor intro tweaks

@@
 
+\paragraph{Contributions}
 In this paper we present a framework for synthesis of linear types extended with
@@
 
+\paragraph{Outline}
 First, we discuss the use of linear types as specifications, the outline of the

### 2021-07-16 article/article.tex

@@
 The \synname\ synthetiser combines the features of linear types with recursion,
-parametric polymorphism, recursive polymorphic? algebraic data types, and
+parametric polymorphism, recursive algebraic data types, and
 refinement types. The synthetiser is built on top of a system of proof-search in
@@
 The core of the synthesis is a \emph{sound} and \emph{complete} system
-consisting of \emph{bottom-up} proof search in linear logic augmented with
-resource management~\cite{} and focusing~\cite{} as techniques to, respectively,
-solve resource related non-determinism, and eliminate from the proof-search
-don't-care non-determinism. A more in depth explanation is in
-section~\ref{sec:formal_system}.
-
-With the core in place, we can synthetise programs from specifications that use
+consisting of \emph{bottom-up} proof search in propositional linear
+logic based on \emph{focusing}~\cite{}. Our approach, being grounded
+by propositions-as-types, ensures that all synthesized programs
+(i.e.~proofs) are well-typed \emph{by construction} (i.e.~if the
+synthesis procedure terminates, then the program satisfies its
+specification). Moreover, we can leverage the modularity of the
+proof-search based approach in two key axes: first, since proof search
+need not construct only closed proofs, we can effectively synthesize
+program sub-expressions (i.e.~synthesis based on typed holes);
+secondly, the framework is amenable to extension beyond the core
+propositional language, employing a richer type structure that
+includes polymorphism, algebraic data types and type refinements.
+
+\paragraph{A core linear calculus.}
+we can synthetise programs from specifications that use
 canonical linear types ($\lolli, \with, \tensor, 1, \oplus, \bang$), such as
@@
 
+\paragraph{Beyond propositional logic.}
 However, a programming language only with the canonical linear types lacks
@@
 the same program.
-
-All additions to the synthesis are extensions of the proof-search in linear
-logic system -- in the end, the process still consists of constructing a proof.
-Because we can only construct valid proofs (a wrong proof isn't a proof), we
-know that all produced synthesis results are correct: \textbf{all programs are
-correct by construction}, i.e. if we're able to synthetise it, then it typechecks,
-and therefore fullfills the specification.
 
@@
 
+\bibliographystyle{splncs04}
+\bibliography{references}
+
+\appendix
+
 \section{Background}\label{sec:background}
@@
 
-\bibliographystyle{splncs04}
-\bibliography{references}
 \end{document}

### 2021-07-17 article/article.tex
Preemptive instance of bangs and existentials, bug fix, trivial array example

@@
 
+\paragraph{Contributions.}
 Despite their long-known potential (\emph{linear types can change the
@@
 foundations \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00}, synthesis of linear functional programs
-from linear types hasn't been previously explored in an ``advanced'' context (e.g.
-with recursion, polymorphism, ...).
-%TODO: Sinto que faltam algumas palavras aqui..
-
-\paragraph{Contributions}
+from linear types in an advanced context has been greatly overlooked.
 In this paper we present a framework for synthesis of linear types extended with
@@
 
-\paragraph{Outline}
+\paragraph{Outline.}
 First, we discuss the use of linear types as specifications, the outline of the
@@
 variables. Additionally, it's required that all existential type
-variables were assigned a type, and provided they were, we can forget
-the constraints generated from the usage of the scheme. These concepts
+variables were assigned a type. These concepts
 are formalized with the rules:
@@
     }
-    {\Theta/\Theta; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
+    {\Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
     \and

### 2021-07-17 STLLC/Array.hs

@@ -1,13 +1,11 @@
-
 data List a = Nil | Cons (a * List a);
 
+newMArray :: (Int -> (MArray a -o !b) -o b);
 
-newMArray :: (Int -o (MArray a -o !b) -o b);
-
-write :: (MArray a -o (Int * a) -o MArray a);
+write :: (MArray a -o (Int * a) -> MArray a);
 
 freeze :: (MArray a -o !(Array a));
 
 foldl :: ((a -o b -o a) -> a -o (List b) -o a)
 
-synth array :: (Int -o List (Int * a) -o Array a);
+synth array :: (Int -> List (Int * a) -> Array a);

### 2021-07-17 STLLC/app/Constraints.hs

@@ -173,7 +173,7 @@ unifyExistential (ADT x ts) (ADT y ts') =
            foldM (\p n -> compose p <$> n) Map.empty maybesubs
        else Nothing
 unifyExistential (TypeVar x) (TypeVar y) = if x == y then Just Map.empty else Nothing
-unifyExistential (ExistentialTypeVar x) (ExistentialTypeVar y) = if x == y then Just Map.empty else Just $ Map.singleton x (ExistentialTypeVar y)
+unifyExistential (ExistentialTypeVar x) (ExistentialTypeVar y) = if x == y then Just Map.empty else Just $ Map.singleton (-x) (ExistentialTypeVar y)
 unifyExistential (ExistentialTypeVar x) y =
     if (-x) `notElem` ftv y then Just $ Map.singleton (-x) y else Nothing
 unifyExistential x (ExistentialTypeVar y) =

### 2021-07-17 STLLC/app/Program.hs

@@ -35,12 +35,13 @@ data ADTD = ADTD Name [Int] [(Name, Type)]    -- Algebraic Data Type Definition
 -------------------------------------------------------------------------------
 
 instance Show Program where
-    show (Program adts bs ts cs) = unlines (map show adts) ++ (if not $ null adts then "\n" else "") ++ displayBindingsWithTypes bs ts -- unlines (map show bs) ++ unlines (map show ts)
+    show (Program adts bs ts cs) = unlines (map show adts) ++ (if not $ null adts then "\n" else "") ++ showSoloAnnotations ++ displayBindingsWithTypes bs ts -- unlines (map show bs) ++ unlines (map show ts)
         where
             displayBindingsWithTypes bs ts = unlines $ map (\b@(Binding n e) -> showbity n ts ++ show b) bs
             showbity n ts = case lookup n $ map (\(TypeBinding n t) -> (n, t)) ts of
                               Nothing -> ""
                               Just ty -> show (TypeBinding n ty)
+            showSoloAnnotations = unlines $ map show $ filter (\(TypeBinding n _) -> (n `notElem` map (\(Binding n' _) -> n') bs) && (n `notElem` concatMap (\(ADTD _ _ ns) -> map fst ns) adts)) ts
 
 
 instance Show ADTD where

### 2021-07-17 STLLC/app/Synth.hs

@@ -53,7 +53,7 @@ data RestrictTag = ConstructADT | DeconstructADT | DecideLeftBangR deriving (Sho
 instance Hashable RestrictTag where
     hashWithSalt a r = hashWithSalt a (show r)
 
-data TraceTag = RightFun Ctxt Type | RightWith Ctxt Type | LeftTensor Ctxt Type Type | LeftUnit Ctxt Type | LeftPlus Ctxt Type Type | LeftSum Type Type | LeftADT Restrictions Ctxt Type Type | MoveLeftADT Restrictions Ctxt Type Type | LeftBang Ctxt Type Type | MoveToDelta (Name, Type) Type | Focus Restrictions Ctxt Type | DecideLeft Type Type | DecideRight Type | DecideLeftBang Restrictions Type Type | DecideLeftBangSch Restrictions Scheme Type | FocusLeftScheme (String, Type) FocusCtxt Type | RightTensor FocusCtxt Type | RightUnit FocusCtxt | RightPlus FocusCtxt Type | RightSum Type | RightADT Name Restrictions FocusCtxt Type | RightBang FocusCtxt Type | LeftFun FocusCtxt Type Type | LeftWith FocusCtxt Type Type | LeftExistentialTV | FocusLeftADT FocusCtxt Type Type | DefaultFocusLeft FocusCtxt (Name, Type) Type | DefaultFocusRight FocusCtxt Type | LeftRefinement Ctxt (Name, Type) Type | FocusLeftRefinement FocusCtxt (Name, Type) Type | RightRefinement FocusCtxt Type deriving (Show)
[…]

### 2021-07-17 STLLC/app/Util.hs

@@ -32,3 +32,6 @@ getNumCode n = fromJust $ lookup n [("a", 0), ("b", 1), ("c", 2), ("d", 3), ("e"
 
 mparens :: String -> String
 mparens s = "(" ++ s ++ ")"
+
+count :: Eq a => a -> [a] -> Int
+count x = length . filter (== x)

### 2021-07-17 STLLC/tests/support/List.hs

@@ -22,4 +22,4 @@ synth foldr :: (!(a -o b -o b) -o b -o List a -o b);
 
 synth insert :: (a -o List a -o List a)
 
-#synth concat :: (List (List a) -o List a);
+synth concat :: (List (List a) -o List a);

### 2021-07-18 article/article.tex
Overview first iteration

@@
 graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
-of specification, program synthesizers have to deal with two main
+of specification, program synthesisers have to deal with two main
 inherent sources of complexity -- searching over a vast space of
@@
 
-\paragraph{Contributions.}
-Despite their long-known potential (\emph{linear types can change the
-world!}\cite{Wadler90lineartypes}, \cite{DBLP:journals/mscs/CairesPT16,Bernardy_2018}) and,
-primarily, strong proof-theoretic
-foundations \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00}, synthesis of linear functional programs
-from linear types in an advanced context has been greatly overlooked.
-In this paper we present a framework for synthesis of linear types extended with
-recursive algebraic data types, parametric polymorphism and refinements.
+\paragraph{Contributions.} Despite their long-known potential (\emph{linear
+types can change the world!}\cite{Wadler90lineartypes},
+\cite{DBLP:journals/mscs/CairesPT16,Bernardy_2018}) and strong proof-theoretic
+foundations
+\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
+synthesis with linear types in a more advanced context
+has generally been overlooked in the literature.  In this paper we present a framework for synthesis
+with linear types extended with recursive algebraic data types, parametric
+polymorphism and refinements.
 % TODO: extensible? proving feasability of... ? showing potential for ... ?
@@
 \paragraph{Outline.}
-First, we discuss the use of linear types as specifications, the outline of the
+First, we discuss linear types as specifications, the outline of the
 synthesis process, and what that entails for the synthetiser -- ilustrated with
@@
 
+Linear types make for more precise specifications than simple traditional types
+because information on which resources must necessarily be used is encoded in
+the type. Their preciseness also influences the search space: all programs where
+a linear resource is used non-linearly (i.e. not \emph{exactly once}) are
+ill-typed. With linearity built-in to the synthesis process, usage of a
+linear proposition more than once is not even considered, and unused ones are caught
+mid-process, making proof-search in linear logic potentially faster than
+in classical logic.  % TODO: Reword?
+
 The core of the synthesis is a \emph{sound} and \emph{complete} system
-consisting of \emph{bottom-up} proof search in propositional linear
-logic based on \emph{focusing}~\cite{}. Our approach, being grounded
-by propositions-as-types, ensures that all synthesized programs
-(i.e.~proofs) are well-typed \emph{by construction} (i.e.~if the
-synthesis procedure terminates, then the program satisfies its
-specification). Moreover, we can leverage the modularity of the
-proof-search based approach in two key axes: first, since proof search
-need not construct only closed proofs, we can effectively synthesize
-program sub-expressions (i.e.~synthesis based on typed holes);
-secondly, the framework is amenable to extension beyond the core
-propositional language, employing a richer type structure that
-includes polymorphism, algebraic data types and type refinements.
-
-\paragraph{A core linear calculus.}
-we can synthetise programs from specifications that use
-canonical linear types ($\lolli, \with, \tensor, 1, \oplus, \bang$), such as
-fundamental linear logic equivalences. For example, given the specification $(A
-\lolli (B \lolli C)) \lolli (A \tensor B) \lolli C$, the program
-\mintinline{haskell}{\a -> \b -> let c*d = b in a c d}
-% $\lambda a .\ \lambda b .$ let $c \tensor d$ = $b$ in $a$ $c$ $d$
-(commonly known as \emph{uncurry}) is synthetised.
-
-\paragraph{Beyond propositional logic.}
-However, a programming language only with the canonical linear types lacks
-expressiveness. To enhance \synname\ and its language, synthetise more
-interesting programs -- and thus empirically prove the feasability of more
-relevant synthesis in a linear context, we extend the syntax and type system
-with recursiveness and richer types: -- polimorfic, refinements, adts
-parametrizáveis...  todo
+consisting of \emph{bottom-up} proof-search in propositional linear logic based
+on \emph{focusing}~\cite{}. Our approach, being grounded by
+propositions-as-types, ensures that all synthetised programs (i.e.~proofs) are
+well-typed \emph{by construction} (i.e.~if the synthesis procedure terminates,
+then the program satisfies its specification). Moreover, we can leverage the
+modularity of the proof-search based approach in two key axes: first, since
+proof search need not construct only closed proofs, we can effectively
+synthetise program sub-expressions (i.e.~synthesis based on typed holes);
+secondly, the framework is amenable to extensions to the core propositional
+language, allowing for the introduction of a richer type structure while
+preserving the correctness of programs by construction.
 
@@
 to create an environment for it. Synth goals are inserted in the program by use
-of a \emph{synth} keyword or syntatic \emph{mark}, indicating a program should
-be generated for a given type. Additionally, it opens the possibility of synthesis with
-context, that is, with knowledge of other functions and structures defined in
-the same program.
-
-Linear types provide a more certain specification, when compared to simple
-traditional types. By requiring some types to be used linearly (consumed) we can
-assure, for example, the deconstruction of an ADT. Given the goal $\bang(a \lolli b)
-\lolli List\ a \lolli List\ b$, the program \mintinline{haskell}{\x -> \y ->
-Nil} wouldn't be valid. To synthetise a valid program,
-\mintinline{haskell}{List a}
-would have to be deconstructed, and its constructor arguments used, resulting
-in the map function without relying on any additional information:
-
+of a \emph{synth} keyword or mark, indicating a program should be generated for a given
+type. Additionally, it opens the possibility of synthesis with context, that is,
+with knowledge of other functions and structures defined in the same program.
+
+\paragraph{Core linear calculus.} Initially, we synthetise from
+specifications that only use literals and canonical linear types ($\lolli, \with, \tensor, 1,
+\oplus, \bang$), such as linear logic theorems. Namely,
+the specification $(A \lolli B \lolli C) \lolli (A \tensor B) \lolli C$ produces
+the program \mintinline{haskell}{(\a -> \b -> let c*d = b in a c d)}
+(known as \emph{uncurry}).
+
+\paragraph{Beyond propositional logic.} To enhance \synname\ and its language,
+counter the core's lack of expressiveness, synthetise more interesting programs
+-- and thus, empirically prove the feasability of more relevant synthesis in a
+linear context, we extend the syntax and type system with recursiveness,
+parametric polymorphism, algebraic data types (ADTs) and type refinements.
+
+%TODO: Polymorphism ? read todo below
+
+\paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring some types to be
+used linearly (consumed), we can assure, e.g., the deconstruction of an
+ADT. As such, the specification $\bang(a \lolli b) \lolli List\ a \lolli
+List\ b$ isn't satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}.
+To synthetise a valid program from it, \mintinline{haskell}{(List a)} has to be
+deconstructed, and its constructor arguments used. \synname\ would output the following
+function (\emph{map}) without relying on any additional information:
+% TODO: Aqui não estou bem a falar de polymorphism mas parece estranho não por
+% visto que falo dos outros temas? se calhar faço um à parte?
+%
 \begin{minted}{haskell}
@@
           Nil  -> Nil
-        | Cons f ->
-            let g*h = f in Cons (e g, map (!e) h);
+        | Cons f -> let g*h = f in Cons (e g, map (!e) h);
+\end{minted}
+
+\paragraph{Refinements.} Refinements are predicates added to a type which must hold for any
+element that inhabits said type. The \synname\ language supports simple
+refinements on integers with arithmetic expressions in the predicates. They
+assert the robustness of our main framework, i.e., how it is amenable to
+significant additions without necessarily interfering with the algorithm.
+%
+As an example, from the specification $x \{Int\} \lolli y \{Int\} \lolli z
+\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$, the
+program\ \mintinline{haskell}{(fn a b c = add (mult (sub 0 1) a) (mult c b))} is
+synthetised.
+
+\paragraph{Configuration.} Specifications can be additionally augmented with three keywords: ``using'', ``depth'' and
+``assert''. Through them, we can fine-tune the synthesis process and, respectively,
+force certain functions to be present in the synthesis outcome, allow a
+\emph{deeper} search in the valid programs space, and require given predicates
+to hold true in the program after the synthesis is complete.
+
+Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
+complex example of synthesis: given the linear type signatures for array
+primitives (\emph{newMArray} passes a new mutable array to a function that uses
+it linearly, \emph{write} takes a mutable array and writes a pair to it,
+\emph{freeze} consumes a mutable array and produces an imutable array,
+\emph{foldl} has the default definition) we try to synthetise a function
+``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
+follows:
+%
+\begin{minted}{haskell}
+foldl :: ((a -o b -o a) -> a -o (List b) -o a);
+newMArray :: (Int -> (MArray a -o !b) -o b);
+write :: (MArray a -o (Int * a) -> MArray a);
+freeze :: (MArray a -o !(Array a));
+synth array :: (Int -> List (!(Int * a)) -> Array a)
+    | using (foldl) | depth 3;
+\end{minted}
+%
+And, in a small fraction of a second, matching exactly the linear definition for
+\emph{array} as seen in Linear Haskell~\cite{Bernardy_2018}, \synname\ outputs:
+%
+\begin{minted}{haskell}
+array b c = let !d = b in let !e = c in
+    newMArray (!d) (\j -> freeze (foldl (!write) j e));
 \end{minted}
@@
 % with Linear Types, Polymorphism and Refinements
-
-% \todo[inline]{Encurtar titulo, mas algo nesta onda que descreve
-%   ``tudo''}
 
@@
 \mypara{Resourceful Program Synthesis from Graded Linear Types} The
-work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
+work~\cite{DBLP:conf/lopstr/HughesO20} synthetises programs using an
 approach very similar to our own.  The work employs so-called graded
@@
 calculus system, for it simplifies the work to be done in the
-synthesizer.
+synthetiser.
 

### 2021-07-18 STLLC/Array.hs

@@ -1,10 +1,13 @@
 
+foldl :: ((a -o b -o a) -> a -o (List b) -o a);
+
 newMArray :: (Int -> (MArray a -o !b) -o b);
 
 write :: (MArray a -o (Int * a) -> MArray a);
 
+#read :: (MArray a -o Int -> (MArray a * !a));
+
 freeze :: (MArray a -o !(Array a));
 
-foldl :: ((a -o b -o a) -> a -o (List b) -o a)
-
-synth array :: (Int -> List (!(Int * a)) -> Array a) | using (foldl) | depth 3;
+synth array :: (Int -> List (!(Int * a)) -> Array a)
+  | using (foldl) | depth 3;

### 2021-07-18 STLLC/Maybe.hs

@@ -1,14 +1,0 @@
-data Maybe a = Nothing | Just a;
-
-data List a = Nil | Cons (a * List a);
-
-
-synth return :: (a -o Maybe a);
-
-synth empty :: Maybe a;
-
-synth bind :: (Maybe a -o (a -o Maybe b) -> Maybe b);
-
-synth maybe :: (b -> (a -o b) -> Maybe a -o b);
-
-#synth listToMaybe :: (List a -o Maybe a);

### 2021-07-18 STLLC/Replicate.hs

@@ -1,0 +1,10 @@
+data Bool = False | True;
+data List a = Nil | Cons (a * List a);
+
+
+deq :: (Int -o Int );
+
+leqzero :: (Int -o Bool);
+
+
+synth replicate :: (Int -> a -> List a);

### 2021-07-18 STLLC/Simple.hs

@@ -1,2 +1,0 @@
-
-synth x :: (a -o a);

### 2021-07-18 STLLC/SynthRef.hs

@@ -13,6 +13,8 @@ synth mm :: (x {Int} -o y {Int} -o z {Int} -o k { Int | k == x - y + z });
 
 synth plus20 :: (x { Int | x > 0 } -o y { Int | y == x + 20 });
 
+synth fn :: (x {Int} -o y {Int} -o z {Int} -o k {Int | x + k == y * z});
+
 synth number :: (x { Int | x > 345 && x != 346 });
 
 # correct me

### 2021-07-18 STLLC/Tree.hs

@@ -13,4 +13,4 @@ synth map :: ((a -o b) -> BST a -o BST b);
 synth uncons :: (BST a -o Maybe (a * (BST a) * (BST a)));
 
 # fails after a long time
-# synth foldl :: ((b -o a -o b) -> b -o BST a -o b);
+synth foldl :: ((b -o a -o b) -> b -o BST a -o b);

### 2021-07-18 STLLC/a.hs

@@ -1,5 +1,0 @@
-
-
-convert :: (A -o B);
-
-synth main :: (A -o A -o (B * B))

### 2021-07-18 STLLC/app/Parser.hs

@@ -409,7 +409,7 @@ letsynth :: Parser (Either TypeBinding Binding)
 letsynth = do
     reserved "synth"
     (Left (TypeBinding name (Forall _ t))) <- typeannot -- TODO: right now marks ignore the schemes, but we could make them such that marks have schemes and the synth function instantiates them
-    usenames <- option [] (do
+    usenames <- option [] (try $ do
         reservedOp "|"
         reserved "using"
         parens $ many1 identifier

### 2021-07-18 STLLC/tests/support/SynthRef.hs

@@ -18,6 +18,8 @@ synth number :: (x { Int | x > 345 && x != 346 });
 # correct me
 synth add3 :: (x {Int} -o y {Int} -o z {Int} -o k {Int | k == x + y + z});
 
+synth fn :: (x {Int} -o y {Int} -o z {Int} -o k {Int | x + k == y * z});
+
 # comment me
 #synth impossible :: (x {Int} -o y {Int} -o z {Int} -o k { Int | k == x * y + z });
 

### 2021-07-18 article/cr.hs

@@ -1,7 +1,0 @@
-
-data List a = Nil | Cons (a * List a);
-
-data Tree a = Leaf | Node (a * List (Tree a));
-
-
-synth insert :: (a -o Tree a -o Tree a);

### 2021-07-18 article/article.tex
WIP: Introduction to focusing, grammars...

@@
 \usepackage{hyperref}
-
 \usepackage{minted}
-
 \usepackage{todonotes}
-
 \usepackage{multirow}
@@
 
-\newcommand{\llet}[2]{\mathsf{let}\,#1\,\mathsf{in}\,#2}
+\newcommand{\llet}[2]{\mathsf{let}\ #1\ \mathsf{in}\ #2}
 
@@
 
-\paragraph{Outline.}
-First, we discuss linear types as specifications, the outline of the
-synthesis process, and what that entails for the synthetiser -- ilustrated with
-concrete results (Section~\ref{sec:overview}). Then, we present the
-algorithm and formal system driving the synthesis
-(Section~\ref{sec:formal_system}). Aftwerwards, we describe the architecture 
-of our synthetiser named \synname, and examine technical details and challenges of the
-implementation (Section~\ref{sec:architecture}). Finally, we evaluate our work
-through expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks
-(i.e.~how fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
+\paragraph{Outline.} First, we discuss linear types as specifications, the
+outline of the synthesis process, and what that entails for the synthetiser --
+ilustrated with concrete results (Section~\ref{sec:overview}). Then, we present
+the formal system driving the synthesis (Section~\ref{sec:formal_system}).
+Aftwerwards, we describe the architecture of our synthetiser named \synname, and
+examine technical details and challenges of the implementation
+(Section~\ref{sec:architecture}). Finally, we evaluate our work through
+expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks (i.e.~how
+fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
 %TODO: And appendice with...
@@
 
-\todo[inline]{Aqui parece-me que deves comecar por explicar o que vais
-  fazer nesta seccao. Sinto falta da sintaxe da linguagem,
-  possivelmente incrementalmente. Comecava por listar os tipos/termos
-  do lambda linear ``simples'', depois em cada subseccao acrescentava
-  as construcoes novas e como a sintese funciona para esse
-  contexto. Enfatizar que a novelty é a combinacao destas features todas.}
-
-The development of \synname\ was based on an almost direct reading of
-concrete inference rules and theoretical principles (~\cite{frank
-  pfenning notes});
-\todo{aqui queres dizer que começaste por te basear
-  num sistema de focusing para a logica linear (era sequer linear a
-  formulacao que comecaste a seguir?) O que são theoretical
-  principles?}
-and work on its environment was also guided by formal
-systems and methods (e.g.~Hindley-Milner type
-inference~\cite{}).\todo{esta frase parece-me desnecessaria. O que
-  queres dizer neste paragrafo, parece-me, é que partes de um sistema
-  focused da literatura que depois vais extender, inspirando te em
-  tecnicas standard de proof search / inferencia.}
-
-However, with the addition of recursion and
-richer types to the synthesis and language, formal guidelines
-\todo{nao sao formal guidelines, nao existem trabalhos a combinar
-  estas features todas, essa parte é importante. Além de que 
-ao incluir estas coisas deixamos de estar completamente em proof
-search, pois muitas destas coisas (e.g. recursao) nao sao
-necessariamente sound logicamente.
-  Tenta usar termos 
-  especificos. Richer types? Richer como?} to support
-the implementation become scarcer, or more complex \todo{nao existem
-  ou sao complicados? ambos? citacao?}. Accordingly,
-rather than derived from explicit rules, solutions appear first to
-tame the extensions' inherent complex problems (such as infinite
-recursion), and implicit rules come only as a byproduct. \todo{ Nao
-  percebo o que queres dizer aqui. O que sao explicit rules e implicit
-  rules? Ou há regras ou nao há. O que estas a dizer é que inventaste
-  tecnicas de sintese para estes casos, que correspondem as regras que
-  vais apresentar. E tao explicito ou implicito como o que esta na literatura.}
-
-With this section we aim to formalize the rules that guide
-the synthesis process for more expressive specifications, alongside the already
-well defined rules that model the core of the synthetiser, resulting in
-a \emph{sound} set of rules that characterizes our framework for synthesis of
-recursive functions from polymorphic and refinement types in a linear context
+%\todo[inline]{Aqui parece-me que deves comecar por explicar o que vais fazer
+%nesta seccao. Sinto falta da sintaxe da linguagem, possivelmente
+%incrementalmente. Comecava por listar os tipos/termos do lambda linear
+%``simples'', depois em cada subseccao acrescentava as construcoes novas e como a
+%sintese funciona para esse contexto. Enfatizar que a novelty é a combinacao
+%destas features todas.}
+
+%The development of \synname\ was based on an almost direct reading of concrete
+%inference rules and theoretical principles (~\cite{frank pfenning notes});
+%%
+%\todo{aqui queres dizer que começaste por te basear num sistema de focusing para
+%a logica linear (era sequer linear a formulacao que comecaste a seguir?) O que
+%são theoretical principles?}
+%%
+%and work on its environment was also guided by formal systems and methods
+%(e.g.~Hindley-Milner type inference~\cite{}).
+%%
+%\todo{esta frase parece-me desnecessaria. O que queres dizer neste paragrafo,
+%parece-me, é que partes de um sistema focused da literatura que depois vais
+%extender, inspirando te em tecnicas standard de proof search / inferencia.}
+
+%However, with the addition of recursion and richer types to the synthesis and
+%language, formal guidelines
+%%
+%\todo{nao sao formal guidelines, nao existem trabalhos a combinar estas features
+%todas, essa parte é importante. Além de que ao incluir estas coisas deixamos de
+%estar completamente em proof search, pois muitas destas coisas (e.g. recursao)
+%nao sao necessariamente sound logicamente.  Tenta usar termos especificos.
+%Richer types? Richer como?}
+%%
+%to support the implementation become scarcer, or more complex.
+%%
+%\todo{nao existem ou sao complicados? ambos? citacao?}
+%%
+%Accordingly, rather than derived
+%from rules, solutions appeared first as techniques to tame the extensions'
+%inherent complex problems (such as infinite recursion), and we now present
+%those techniques as rules.
+%%
+% \todo{ Nao percebo o que queres dizer aqui. O que sao explicit rules e implicit rules?
+% Ou há regras ou nao há. O que estas a
+% dizer é que inventaste tecnicas de sintese para estes casos, que correspondem
+% as regras que vais apresentar. E tao explicito ou implicito como o que esta na
+% literatura.}
+
+Program synthesis from linear types in combination with polymorphism, recursive
+algebraic data types, and refinements, is new in the synthesis 
+literature. Despite the substantial amount of research on linear logic and
+proof-search upon which we found our core synthetiser, formal guidelines for the
+richer types and their inherent complex problems (such as infinite recursion)
+don't exist. % TODO: Developed an algorithm...?
+
+With this section we aim to formalize as rules the novel techniques that guide
+synthesis from our more expressive specifications, alongside the already well
+defined rules that model the core of the synthetiser, putting together a
+\emph{sound} set of inference rules that characterizes our framework for synthesis of
+recursive programs from specifications with the select richer types in a linear context,
 % TODO: Better?....
-, and describes the system in enough detail for
-the synthetiser to be reproducible by a theory-driven implementation.
+and describes the system in enough detail for the synthetiser to be reproducible
+by a theory-driven implementation.
 
@@
 
-\mypara{Core Rules} The core system comprises of the sequent calculus rules for
-linear logic connectives with resource-management and focusing.
-\todo{Introdução para focusing...}
-Focusing separates the proof search into two
-phases: inversion ($\Uparrow$) in which we apply invertible rules eagerly, and
-focusing ($\Downarrow$) in which we decide a proposition to focus on, and then
+\mypara{Core Rules} The core system comprises of proof-search in sequent
+calculus rules for linear logic with resource-management and focusing.
+%
+Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
+focusing} is a technique to remove \emph{inessencial} non-determinism from
+proof-search by structuring the application of invertible and non-invertible
+inference rules. Rules are classified as \emph{left} or \emph{right},
+\emph{invertible} or \emph{non-invertible} as follows:...\todo{não estou a
+conseguir formular bem esta introdução}
+The proof search is separated into two phases:
+%
+inversion ($\Uparrow$), in which we apply all invertible rules eagerly, and
+focusing ($\Downarrow$), in which we decide a proposition to focus on, and then
 apply non-invertible rules, staying focused until either the focus is again an
@@
 invertible rules, a third, ordered, context ($\Omega$) is needed to hold
-propositions resulting from decomposing connectives; propositions we'll later try to
-invert, moving them to the linear context ($\Delta$) when we fail to.
-
-\todo{Primeiro explica os varios judgments (as ``formas'' das
-  conclusoes das regras) e o que significam as varias componentes,
-  nomeadamente a parte da gestao de recursos.  Depois podes explicar
-  regras concretas como fazes a seguir. Tambem seria util dizer quais
-  os tipos que sao invertiveis a esquerda/direita, pois o leitor nao
-  faz ideia do que estas a dizer.}
-
-Putting together linear logic, the Curry-Howard correspondence, resource
+propositions resulting from decomposing connectives; propositions we'll later
+try to invert, moving them to the linear context ($\Delta$) when we fail to.
+
+% TODO: Explicar vários judgments?
+% \todo{Primeiro explica os varios judgments (as ``formas'' das
+%   conclusoes das regras) e o que significam as varias componentes,
+%   nomeadamente a parte da gestao de recursos.  Depois podes explicar
+%   regras concretas como fazes a seguir. Tambem seria util dizer quais
+%   os tipos que sao invertiveis a esquerda/direita, pois o leitor nao
+%   faz ideia do que estas a dizer.}
+%
+% Esta parte em geral está a ficar pouco coerente, com a introdução do focusing,
+  % a suposta introdução dos judgments, e a introdução da gramática... deixa
+
+The core language is a simply typed linear lambda calculus for which the terms
+($M,N$)
+and types' ($\tau$) syntax is constructed with the following grammar:
+\[
+\begin{tabular}{lclc}
+$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
+    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
+    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
+    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
+    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
+    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
+    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
+    N_2)$ & $\oplus$ \\
+    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
+    \\
+$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
+    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
+\end{tabular}
+\]
+%
+Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
 management, and focusing, we get the following core formal system (heavily inspired
@@
     { }
-    {\Gamma; \Delta/\Delta \vdash \star : 1 \Downarrow}
+    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
 \and
@@
 propositions of type $A$, using its syntax: \mintinline{haskell}{data Container
-= None 1 | One A | Two (A * A)}
+= None 1 | One A | Two (A * A)}. The language grammar is extended as:
+\[
+\begin{tabular}{lclc}
+    $M,N$ & $\ ::=\ $ & $\dots\ \vert\ \emph{Cons}_n\ M\ \vert\ (\mathsf{case}\
+    M\ \mathsf{of}\ \dots\ \vert\ \emph{Cons}_n\ u \Rightarrow N)$ \\
+    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ \langle\emph{name}\rangle$ \\
+    $ADT$ & $\ ::=\ $ & (\textsf{data} $\langle\emph{name}\rangle$\ $=$\ $\emph{Cons}_1\ \tau$\
+$\vert$\ $\dots$\ $\vert$\ $\emph{Cons}_n\ \tau$) \\ \end{tabular}
+\]
+%TODO: Como definir gramática dos ADTs?
 
@@
 the inference rules for a simple ADT mimic the $\oplus$ rules, for an
-arbitrary ($> 1$) number of types, rather than exactly two. Therefore, there's one
+arbitrary ($>1$) number of types, rather than exactly two. Therefore, there's one
 left rule for ADTs, and an arbitrary number of right rules, one for each
@@
                 \infer*[right=(adtL)]
-                {
-                    \infer*[right=(adtL)]
                 {\dots}
-                    {\Gamma; \Delta/\Delta'; \Omega, z{:}T \Uparrow\ \vdash \textrm{case}\ z\ 
-                    \textrm{of}\ C_1\ k \rightarrow \dots : C}
-                }
                 {\Gamma; \Delta/\Delta'; \Omega, y{:}T \Uparrow\ \vdash \textrm{case}\ y\ 
@@
             \infer*[left=(adtR)]
-            {
-                \infer*[left=(adtR)]
             {\dots}
-                {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
-            }
             {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
@@
 context, or more complex specifications:
+%TODO: ignore constraints if type isnnt existential
 %
@@
 verification--and hence synthesis--of nontrivial programs''.  Their
-approach to refinement types consists a new algorithm that supports
+approach to refinement types consists of a new algorithm that supports
 decomposition of the refinement specification, allowing for separation

### 2021-07-19 article/article.tex
Fair conjunction ADT destruction + binary restrictions tweaks

@@
 Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
-focusing} is a technique to remove \emph{inessencial} non-determinism from
+focusing} is a technique to remove inessencial non-determinism from
 proof-search by structuring the application of invertible and non-invertible
@@
 
-The core language is a simply typed linear lambda calculus for which the terms
-($M,N$)
-and types' ($\tau$) syntax is constructed with the following grammar:
+The core language is a simply-typed linear lambda calculus with linear
+functions, additive and multiplicative pairs, multiplicative unit, additive sums
+and the exponential modality (to internalize unrestricted use of variables), for
+which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
+following grammar:
+%
+% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
+% usá-los como tipos gerais?
 \[

### 2021-07-19 STLLC/Array.hs

@@ -1,13 +1,0 @@
-
-foldl :: ((a -o b -o a) -> a -o (List b) -o a);
-
-newMArray :: (Int -> (MArray a -o !b) -o b);
-
-write :: (MArray a -o (Int * a) -> MArray a);
-
-#read :: (MArray a -o Int -> (MArray a * !a));
-
-freeze :: (MArray a -o !(Array a));
-
-synth array :: (Int -> List (!(Int * a)) -> Array a)
-  | using (foldl) | depth 3;

### 2021-07-19 STLLC/List.hs

@@ -28,4 +28,4 @@ synth foldr :: (!(a -o b -o b) -o b -o List a -o b);
 
 synth insert :: (a -o List a -o List a)
 
-#synth concat :: (List (List a) -o List a);
+synth concat :: (List (List a) -o List a);

### 2021-07-19 STLLC/ListT.hs

@@ -10,5 +10,6 @@ synth concat :: (List (List Int) -o List Int);
 
 synth concat2 :: (List (List A) -o List A);
 
+# Result w wrong infinite recursion
 synth zip :: ((A -o B -o C) -> List A -o List B -o List C);
 

### 2021-07-19 STLLC/SynthRef.hs

@@ -1,39 +1,0 @@
-
-
-# Right Focus Synth
-
-synth plus :: (x {Int} -o y {Int} -o z { Int | z == x + y });
-
-synth inc :: (x {Int} -o y { Int | y == x + 1 });
-
-synth deq :: (x {Int} -o y { Int | y == x - 1 });
-
-# is this correct?
-synth mm :: (x {Int} -o y {Int} -o z {Int} -o k { Int | k == x - y + z });
-
-synth plus20 :: (x { Int | x > 0 } -o y { Int | y == x + 20 });
-
-synth fn :: (x {Int} -o y {Int} -o z {Int} -o k {Int | x + k == y * z});
-
-synth number :: (x { Int | x > 345 && x != 346 });
-
-# correct me
-synth add3 :: (x {Int} -o y {Int} -o z {Int} -o k {Int | k == x + y + z});
-
-# comment me
-#synth impossible :: (x {Int} -o y {Int} -o z {Int} -o k { Int | k == x * y + z });
-
-
-# Left Focus Synth
-
-# how to avoid stupid functions so that add is attempted
-#synth s :: (x {Int | x > 5} -o y {Int | y > x} -o z {Int});
-
-
-synth inc :: (x {Int} -o y {Int | y == x + 1});
-
-synth shouldntfail :: (x {Int} -o y {Int | y > x});
-
-synth addplus1 :: (x {Int} -o y {Int} -o z {Int | z == x + y + 1});
-
-synth shouldntfail2 :: (x {Int | x > 0} -o y {Int | y > 0} -o z {Int | z > x + y});

### 2021-07-19 STLLC/Tree.hs

@@ -13,4 +13,4 @@ synth map :: ((a -o b) -> BST a -o BST b);
 synth uncons :: (BST a -o Maybe (a * (BST a) * (BST a)));
 
 # fails after a long time
-synth foldl :: ((b -o a -o b) -> b -o BST a -o b);
+#synth foldl :: ((b -o a -o b) -> b -o BST a -o b);

### 2021-07-19 STLLC/app/Synth.hs

@@ -143,7 +143,7 @@ addrestriction tag ty rs = (tag, Right ty):rs
 
 
 addbinaryrestriction :: RestrictTag -> Type -> Type -> Restrictions -> Restrictions
-addbinaryrestriction tag ty ty' rs = (tag, Left (ty, ty')):rs
+addbinaryrestriction tag ty ty' rs = if isFunction ty then (tag, Left (ty, ty')):rs else rs
 
 
 addtrace :: TraceTag -> Synth a -> Synth a
@@ -172,7 +172,7 @@ getExistentialDepth = lift (lift ask) >>= \(b,c,d,ra,ed) -> return ed
 checkrestrictions :: RestrictTag -> Type -> Restrictions -> Bool
 checkrestrictions tag ty@(ADT tyn tpl) rs = -- Restrictions only on ADT Construction and Destruction
     let reslist = rights $ map snd $ filter (\(n,x) -> n == tag) rs in
-    ty `notElem` reslist && (not (isExistentialType ty) || not (any isExistentialType $ filter (\(ADT tyn' _) -> tyn == tyn') reslist))
+    ty `notElem` reslist -- && (not (isExistentialType ty) || not (any isExistentialType $ filter (\(ADT tyn' _) -> tyn == tyn') reslist))
 
 checkbinaryrestrictions :: RestrictTag -> (Type, Type) -> Restrictions -> Synth Bool
 checkbinaryrestrictions tag typair rs = do
@@ -336,7 +336,7 @@ synth c@(cs, rs, g, d, p@(n, ADT tyn tps):o) t = addtrace (LeftADT rs c (ADT tyn
        then let rs' = addrestriction DeconstructADT (ADT tyn tps) rs in memoizesynth synth (cs, rs', g, p:d, o) t    -- An ADT that has no constructors might still be used to instantiate a proposition, but shouldn't leave synchronous mode (hence the restriction)
        else do
             isrectype <- isRecursiveType (ADT tyn tps)
-            (ls, cs'') <- runStateT (mapM (\(name, vtype) ->
+            (ls, cs'') <- runStateT (mapMFairConj (\(name, vtype) ->
                 let rs' = if isrectype then addrestriction DeconstructADT (ADT tyn tps) rs else rs in   -- For recursive types, restrict deconstruction of this type in further computations
                         case vtype of
                           Unit -> do
[…]

### 2021-07-19 STLLC/t2.hs

@@ -1,2 +1,2 @@
-main :: forall a b c d . ((! (a -o b)) -o (((! (a -o b)) -o (d -o c)) -o (d -o c)));
-main e f g = let !h = e in f ! (λk -o h k) g;
+#-- main :: forall a b c d . ((! (a -o b)) -o (((! (a -o b)) -o (d -o c)) -o (d -o c)));
+main e f g = let !h = e in f (!(λk -> h k)) g;

### 2021-07-20 article/article.tex
More restrictions paragraph

@@
 formulated, only rules for left focus on a scheme are needed. Our
-algorithm instances bound type vars of the focused scheme with
+algorithm instances bound type variables of the focused scheme with
 existential type variables, and the instanced type becomes the left
@@
 constraints -- a branch of the proof search is desired to fail as soon
-as possible; and more, two different existential vars do not unify
-against one another ($?\alpha \mapsto ?\beta$ fails) \todo{O que
-  significa $?\alpha$ ? Não foi explicado -- no Hindley-Milner nao ha
-  propriamente o problema de misturar variaveis existenciais com universais}. Note that we
-instance the scheme with \emph{existential} type variables ($\alpha$)
-rather than just type variables since the latter represent universal
+as possible.
+%TODO: Como fazer? \todo{O que
+% significa $?\alpha$ ? Não foi explicado -- no Hindley-Milner nao ha
+% propriamente o problema de misturar variaveis existenciais com universais}
+Note that we
+instance the scheme with \emph{existential} type variables ($?\alpha$)
+rather than just type variables ($\alpha$) since the latter represent universal
 types in the synth, and the first represent an existential instance of
-a scheme, that might inclusively constrain normal type
+a scheme, that might inclusively constrain traditional type
 variables. Additionally, it's required that all existential type
-variables were assigned a type. These concepts
+variables are assigned a type. These concepts
 are formalized with the rules:
@@
     x{:}A \Downarrow\ \vdash x : ?\alpha}
-\end{mathpar}
-where
-\begin{mathpar}
+    \\
     \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'\ \textrm{means type}\
@@
     \textrm{can be unified with those in}\ \Theta
+\end{mathpar}
+
+\mypara{More Restrictions} We can now look at two more sources of infinite
+recursion in the synthesis process. The first is the use of an unrestricted
+function to synthetise a term of type $\tau$ that in turn will require a term of
+the same type $\tau$. A good example is the sub-goal judgment $\Gamma,(a \lolli
+b \lolli b); (a \lolli b \lolli b) \Downarrow\ \vdash b$ that
+appears while synthetising \emph{foldr} -- we apply ($\lolli$L)
+until we can use \textsc{init} ($b \Downarrow\ \vdash b$), and then we must
+synthetise an argument of type $b$. Without any additional restrictions, we
+can end up left focused on $(a \lolli b \lolli b)$ again, and again require $b$,
+and on and on. The solution will be to disallow the usage of the same function
+to synthetise the same goal a second time further down in the derivation.
+
+% TODO TODO: Big rewrite
+The other situation occurs when using an unrestricted polymorphic function that
+requires synthesis of a term with an existential type when the goal is an
+existential type. In contrast to the previous problem, the type of the goal and
+of the argument that will cause the loop won't be the same, since instantiated
+bound variables are always fresh. For example, for $\Gamma,\forall \alpha,\beta
+.\ \alpha\lolli\beta\lolli\beta;?\alpha\lolli?\beta\lolli?\beta \Downarrow\
+\vdash\ ?\sigma$, we'll unify $?\beta$ with $?\sigma$, and then require a term
+of type $?\beta$ (not $?\sigma$). We want to forbid the usage of the \emph{same}
+function to attain \emph{any} existential goal, provided that function might
+create existential sub-goals (i.e. it's polymorphic). However, we noticed that,
+even though for most tried problems this ``same function'' approach worked,
+context-heavy problems such as \emph{array} (seen in Section~\ref{sec:overview})
+wouldn't terminate in a reasonable amount of time.  In fact, with $n$
+polymorphic functions in the unrestricted context, the complexity of searching for a
+program of any existential type $?\alpha$, while restricting solely the used
+function, is $O(n!)$~\ref{exemplo_apendice?}. As such, we'll instead forbid the
+usage of \emph{all} unrestricted polymorphic functions to synthetise any
+existential type if, altogether, we've already used $d_e$ times a polymorphic
+function for \emph{any} existential sub-goal, where $d_e$ is a \emph{constant
+existential depth} that controls the maximum depth of ``existential synthetis''.
+This approach reduces the complexity of our proof-search algorithm for
+existential types to $O(\tfrac{n!}{d_e!}) = O(n^{d_e})$.
+
+Extending the restrictions context ($\Rho$) with restrictions on using the
+unrestricted context ($\Rho_{L!}$), we modify \textsc{decideLeft!} to formalize
+the previous paragraph:
+\begin{mathpar}
+    \mprset{flushleft}
+    \infer*[right=(decideL!)]
+    {
+    (A, C) \notin \Rho_{L!}
+    \\
+    \textsc{isExist}(C) \Rightarrow |\{u\ |\
+    (f,u)\in\Rho_{L!},\textsc{isExist}(u),\textsc{isPoly}(f)\}| < d_e
+    \\
+    \Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!}');\Gamma, A; \Delta/\Delta' ; A
+    \Downarrow\ \vdash C
+    }
+    {\Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!});\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
+    \\
+    \Rho_{L!}' =\ \textbf{if}\ \textsc{isFunction}(A)\ \textbf{then}\ \Rho_{L!},(A, C)\ \textbf{else}\ \Rho_{L!}
 \end{mathpar}

### 2021-07-20 article/article.tex
ADT poly + memoization tweaks

@@
 
-\mypara{Polymorphic ADTs}
+\mypara{Polymorphic ADTs} To allow type parameters and the use of universally quantified type
+variables in ADT constructors, we must guarantee that the
+\textsc{adt-init} rule can unify the type parameters and that when constructing or
+destructing an ADT, type variables in constructor parameters are substituted by
+the actual type (i.e. to construct \mintinline{haskell}{List Int} with
+\mintinline{haskell}{data List a = Cons (a *
+List a)}, we wouldn't try to synthetise \mintinline{haskell}{(a * List a)},
+but rather \mintinline{haskell}{(Int * List Int)}). To unify $T_{\overline\alpha}$ with $T_{\overline\beta}$,
+the sets of type parameters $\overline\alpha$ and $\overline\beta$ must satisfy $|\overline\alpha| = |\overline\beta|$
+together with $\forall i\ 0 \leq i \land i < |\overline\alpha| \land
+\textsc{unify}(\overline\alpha_i \mapsto \overline\beta_i)$. The constructor
+type substitution needn't be explicit in the rule:
+\begin{mathpar}
+    \infer*[right=(adt-init)]
+    {\textsc{unify}(T_{\overline\alpha} \mapsto T_{\overline\beta}, \Theta)}
+    {\Theta/\Theta,T_{\overline\alpha} \mapsto T_{\overline\beta},\Rho; \Gamma; \Delta/\Delta'; x{:}T_{\overline\alpha} \Downarrow\ \vdash x :
+    T_{\overline\beta}}
+\end{mathpar}
 
@@
 ran benchmarks to measure the impact of memoization on the performance. It's
-especially noticeable in synthesis with multiple functions in the unrestricted
-context, or more complex specifications:
-%TODO: ignore constraints if type isnnt existential
-%
-\begin{center}
-    \begin{tabular}{ |c|c|c|c|c| }
-        \hline
-        Goal & Time + $\sigma$ & TimeM + $\sigma$ \\
-        \hline
-        insertWithKey & 13.31 s $\pm$ 593.4 ms & 2.667 s $\pm$ 284.3 ms \\
-        \hline
-    \end{tabular}
-\end{center}
+quite noticeable in synthesis with multiple functions in the unrestricted
+context, or more complex specifications. For example, if we add a new function
+\emph{read} to the unrestricted context in the \emph{array} problem seen in
+Section~\ref{sec:overview}, synthesis with memoization will take around
+$5s$, while synthesis without memoization doesn't terminate even after
+$15m$ running.
 %
@@
 memoization with the LogicT monad transformer (remembering decisions across
-backtracking branches) and by finetuning key-equality for polymorphic types.
+backtracking branches) and by finetuning key-equality by ignoring context
+components that aren't relevant to the synthesis call, to name some examples.
 
@@
 first couple of "infinite examples", the following problems were much harder,
-taking sometimes full days to solve (and as of now, some still remain unsolved,
-e.g. List (List a) $\lolli$ List a). To make the debugging experience better, we
+taking sometimes full days to solve. To make the debugging experience better, we
 developed a tracing system that prints information whenever a rule is applied,
@@
 what was supposed to happen.
-
-Unfortunately, it seems that sometimes in the process of backtracking, stack
-information is "lost", and the trace logs lose useability.
 

### 2021-07-20 article/article.tex
tweaks and further todos

--- 2021-07-20 article/article.tex
@@
 graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
-of specification, program synthesisers have to deal with two main
-inherent sources of complexity -- searching over a vast space of
-(potentially) valid programs, and interpreting user intent.
+of specification, program synthesis must deal with two main
+inherent sources of complexity -- search over the space of
+valid programs, and interpreting user intent.
 
@@
 %
-Linear types differ from more traditional types in that they constrain resource
+Linear types constrain resource
 usage in programs by \emph{statically} limiting the number of times certain
@@
 
-\paragraph{Contributions.} Despite their long-known potential (\emph{linear
-types can change the world!}\cite{Wadler90lineartypes},
-\cite{DBLP:journals/mscs/CairesPT16,Bernardy_2018}) and strong proof-theoretic
+\paragraph{Contributions and Outline.} Despite their long-known potential
+\cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018}
+and strong proof-theoretic
 foundations
 \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
-synthesis with linear types in a more advanced context
-has generally been overlooked in the literature.  In this paper we present a framework for synthesis
+synthesis with linear types combined with other advanced typing features
+has generally been overlooked in the literature.  In this work we present a framework for synthesis
 with linear types extended with recursive algebraic data types, parametric
@@
 % TODO: extensible? proving feasability of... ? showing potential for ... ?
-
-\paragraph{Outline.} First, we discuss linear types as specifications, the
-outline of the synthesis process, and what that entails for the synthetiser --
-ilustrated with concrete results (Section~\ref{sec:overview}). Then, we present
-the formal system driving the synthesis (Section~\ref{sec:formal_system}).
-Aftwerwards, we describe the architecture of our synthetiser named \synname, and
-examine technical details and challenges of the implementation
-(Section~\ref{sec:architecture}). Finally, we evaluate our work through
-expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks (i.e.~how
-fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
+We first introduce linear types as specifications
+and outline the synthesis process, leveraging linearity, by example
+(\S~\ref{sec:overview}). We then present the formal system driving the
+synthesis (\S~\ref{sec:formal_system}) and describe the architecture
+of our framework named \synname, examining technical details and key
+implementation challenges (\S~\ref{sec:architecture}). Finally,
+we evaluate our work through expressiveness and performance benchmarks
+(\S~\ref{sec:evaluation}) and discuss related work (\S~\ref{sec:related}).
 %TODO: And appendice with...
@@
 
-% TODO: Why linear types?
-The \synname\ synthetiser combines the features of linear types with recursion,
+The \synname\ synthetiser combines linear types with recursion,
 parametric polymorphism, recursive algebraic data types, and
-refinement types. The synthetiser is built on top of a system of proof-search in
-linear logic. Proof-search relates to program synthesis through the Curry-Howard
-correspondence~\cite{}, which states that \emph{propositions in a logic} map
-directly to \emph{types}, and \emph{proofs} of those propositions to well-typed
+refinement types. The synthetiser is built on top of a system of proof-search for
+linear logic. Proof search relates to program synthesis via the Curry-Howard
+correspondence~\cite{}, which states that \emph{propositions in a logic} are
+\emph{types}, and \emph{proofs} of those propositions are well-typed
 \emph{programs} -- finding a proof of a proposition is finding a program with
@@
 
-Linear types make for more precise specifications than simple traditional types
-because information on which resources must necessarily be used is encoded in
-the type. Their preciseness also influences the search space: all programs where
-a linear resource is used non-linearly (i.e. not \emph{exactly once}) are
-ill-typed. With linearity built-in to the synthesis process, usage of a
-linear proposition more than once is not even considered, and unused ones are caught
-mid-process, making proof-search in linear logic potentially faster than
-in classical logic.  % TODO: Reword?
+Linear types make for more precise specifications than simple types
+because information on which resources must necessarily be used is
+encoded in the type. For instance the type $\mathsf{Int} \lolli
+\mathsf{Int} \lolli \mathsf{Int}$ denotes a function that \emph{must}
+use its two integer arguments to produce an integer.
+Their preciseness also affects the search space:
+all programs where a linear resource is used non-linearly (i.e. not
+\emph{exactly once}) are ill-typed. With linearity built into the
+synthesis process, usage of a linear proposition more than once is not
+considered, and unused propositions are identified during synthesis,
+constraining the space of valid programs.  % TODO: Reword?
 
 The core of the synthesis is a \emph{sound} and \emph{complete} system
-consisting of \emph{bottom-up} proof-search in propositional linear logic based
-on \emph{focusing}~\cite{}. Our approach, being grounded by
-propositions-as-types, ensures that all synthetised programs (i.e.~proofs) are
-well-typed \emph{by construction} (i.e.~if the synthesis procedure terminates,
-then the program satisfies its specification). Moreover, we can leverage the
-modularity of the proof-search based approach in two key axes: first, since
-proof search need not construct only closed proofs, we can effectively
-synthetise program sub-expressions (i.e.~synthesis based on typed holes);
-secondly, the framework is amenable to extensions to the core propositional
-language, allowing for the introduction of a richer type structure while
-preserving the correctness of programs by construction.
-
-The \synname\ programming language is naturally born alongside the synthetiser
-to create an environment for it. Synth goals are inserted in the program by use
-of a \emph{synth} keyword or mark, indicating a program should be generated for a given
-type. Additionally, it opens the possibility of synthesis with context, that is,
-with knowledge of other functions and structures defined in the same program.
-
-\paragraph{Core linear calculus.} Initially, we synthetise from
+consisting of \emph{bottom-up} proof-search in propositional linear
+logic based on \emph{focusing}~\cite{}. Our approach, being grounded
+by propositions-as-types, ensures that all synthetised programs
+(i.e.~proofs) are well-typed \emph{by construction} (i.e.~if the
+synthesis procedure produces a program, then the program intrinsically
+satisfies its specification). Moreover, we can leverage the modularity
+of the proof-search based approach along two axes: first, since proof
+search need not construct only closed proofs, we can effectively
+synthesize program sub-expressions (i.e.~synthesis based on typed
+holes); secondly, the framework is amenable to extensions to the core
+propositional language, allowing for the introduction of a richer type
+structure while preserving the correctness of programs by
+construction.
+
+The \synname\ programming language is naturally developed alongside
+the synthetiser. Synthesis goals are inserted in the program by use of
+a \emph{synth} keyword or mark, indicating a program should be
+generated for a given type. Additionally, it opens the possibility of
+synthesis within a context, that is, with knowledge of other functions and
+structures defined in the same program.
+
+\paragraph{Core linear calculus.} Initially, we synthesize from
 specifications that only use literals and canonical linear types ($\lolli, \with, \tensor, 1,
@@
 the specification $(A \lolli B \lolli C) \lolli (A \tensor B) \lolli C$ produces
-the program \mintinline{haskell}{(\a -> \b -> let c*d = b in a c d)}
-(known as \emph{uncurry}).
-
-\paragraph{Beyond propositional logic.} To enhance \synname\ and its language,
-counter the core's lack of expressiveness, synthetise more interesting programs
--- and thus, empirically prove the feasability of more relevant synthesis in a
-linear context, we extend the syntax and type system with recursiveness,
-parametric polymorphism, algebraic data types (ADTs) and type refinements.
+the function \mintinline{haskell}{\a -> \b -> let c*d = b in a c d},
+which takes an argument \mintinline{haskell}{a}, a
+\emph{curried} linear function of type $A \lolli B \lolli C$,  and an
+argument \mintinline{haskell}{b}, a \emph{linear
+  pair} of an $A$ and a $B$ and produces a value of type $C$ by
+deconstructing the pair and applying \mintinline{haskell}{a} to the
+corresponding elements of the pair (i.e.,~\emph{uncurries} the
+function \mintinline{haskell}{a}).
+
+\paragraph{Beyond propositional logic.} To be able to synthesize more
+interesting programs -- and thus, empirically prove the feasability of
+more relevant synthesis in a linear context, we extend the syntax and
+type system with recursiveness, parametric polymorphism, algebraic
+data types (ADTs) and type refinements.
 
@@
 
+\todo[inline]{Tens de explicar o que significa o $\bang$, see above.}
 \paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring some types to be
@@
 
-\paragraph{Refinements.} Refinements are predicates added to a type which must hold for any
-element that inhabits said type. The \synname\ language supports simple
-refinements on integers with arithmetic expressions in the predicates. They
-assert the robustness of our main framework, i.e., how it is amenable to
-significant additions without necessarily interfering with the algorithm.
+\todo[inline]{Tens de dizer o que significa o refinamento concreto e o
+que e a funcao mult e add}
+\paragraph{Refinements.} Refinements are logical predicates
+which must hold for any element that inhabits a given type. The \synname\
+language supports simple refinements on integers with arithmetic
+expressions in the predicates. They assert the robustness of our main
+framework, i.e., how it is amenable to significant additions without
+necessarily interfering with the overall approach.
 %
@@
 
-\paragraph{Configuration.} Specifications can be additionally augmented with three keywords: ``using'', ``depth'' and
-``assert''. Through them, we can fine-tune the synthesis process and, respectively,
-force certain functions to be present in the synthesis outcome, allow a
-\emph{deeper} search in the valid programs space, and require given predicates
-to hold true in the program after the synthesis is complete.
+\paragraph{Guiding the Search.} Specifications can be additionally
+augmented with three keywords: ``using'', ``depth'' and
+``assert''. Through them, we can fine-tune the synthesis process and,
+respectively, force certain functions to be present in the synthesis
+outcome, allow a \emph{deeper} search in the valid programs space, and
+require given predicates to hold true in the program after the
+synthesis is complete.
 
 Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
-complex example of synthesis: given the linear type signatures for array
+more intricate example of synthesis: given the linear type signatures for array
 primitives (\emph{newMArray} passes a new mutable array to a function that uses
@@
 \emph{freeze} consumes a mutable array and produces an imutable array,
-\emph{foldl} has the default definition) we try to synthetise a function
+\emph{foldl} has the default definition) we synthesize a function
 ``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
@@
 %
-And, in a small fraction of a second, matching exactly the linear definition for
-\emph{array} as seen in Linear Haskell~\cite{Bernardy_2018}, \synname\ outputs:
+Matching exactly the linear definition for
+\emph{array} from~\cite{Bernardy_2018}, \synname\
+outputs (in a fraction of a second):
 %
@@
 
-\section{Synthesis Framework}\label{sec:formal_system}
+\section{The \synname\ Synthesis Framework}\label{sec:formal_system}
 % with Linear Types, Polymorphism and Refinements
@@
 
-Program synthesis from linear types in combination with polymorphism, recursive
-algebraic data types, and refinements, is new in the synthesis 
+Program synthesis from linear types with polymorphism, recursive
+algebraic data types, and refinements, is essentially new in the synthesis 
 literature. Despite the substantial amount of research on linear logic and
-proof-search upon which we found our core synthetiser, formal guidelines for the
-richer types and their inherent complex problems (such as infinite recursion)
-don't exist. % TODO: Developed an algorithm...?
-
-With this section we aim to formalize as rules the novel techniques that guide
+proof-search upon which we base our core synthesizer, formal guidelines for
+richer types and their intrinsic challenges (such as infinite
+recursion) must be developed.
+
+In this section we formalize  the techniques that guide
 synthesis from our more expressive specifications, alongside the already well
 defined rules that model the core of the synthetiser, putting together a
-\emph{sound} set of inference rules that characterizes our framework for synthesis of
-recursive programs from specifications with the select richer types in a linear context,
+\emph{sound} set of inference rules that characterizes our framework
+for linear synthesis of
+recursive programs from specifications with the select richer types,
 % TODO: Better?....
@@
 
-Even so, note that \emph{soundness} and \emph{completeness} have two very
-different meanings, and our \emph{sound} set of rules only guarantees we won't
-synthetise incorrect programs, but still hides a possibly infinite number of correct
-proofs; and that the valid programs derivable through it reflect the somewhat
-subjective trade-offs we decided on. Different choices and approaches outside
-the core might lead to completely distinct synths and spaces of valid programs.
-
-\mypara{Core Rules} The core system comprises of proof-search in sequent
-calculus rules for linear logic with resource-management and focusing.
-%
-Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
-focusing} is a technique to remove inessencial non-determinism from
-proof-search by structuring the application of invertible and non-invertible
-inference rules. Rules are classified as \emph{left} or \emph{right},
-\emph{invertible} or \emph{non-invertible} as follows:...\todo{não estou a
-conseguir formular bem esta introdução}
-The proof search is separated into two phases:
-%
-inversion ($\Uparrow$), in which we apply all invertible rules eagerly, and
-focusing ($\Downarrow$), in which we decide a proposition to focus on, and then
-apply non-invertible rules, staying focused until either the focus is again an
-asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
-invertible rules, a third, ordered, context ($\Omega$) is needed to hold
-propositions resulting from decomposing connectives; propositions we'll later
-try to invert, moving them to the linear context ($\Delta$) when we fail to.
+We note that a \emph{sound} set of rules guarantees we cannot synthesize
+incorrect programs; and that the valid programs derivable through them
+reflect the somewhat subjective trade-offs we committed to. Different
+choices and approaches outside the core might lead to completely
+distinct synths and spaces of valid programs.
+
+\todo[inline]{Indicar o judgment ``declarativo'' e depois os
+  judgments em focusing}
+
+\mypara{Core Rules} The core system comprises of proof search for
+(intuitionistic) linear logic sequent calculus proofs, based on a
+system of resource management and focusing.
+%
+In intuitionistic sequent calculi, each connective has a so-called
+\emph{left} and a \emph{right} rule, which effectively define how to
+decompose an ambient assumption of a given proposition and how to
+prove a certain proposition is true, respectively. 
+Andreoli's \emph{focusing} for linear logic~\cite{Andreolli focusing}
+is a technique to remove non-essential nondeterminism from
+proof-search by structuring the application of so-called invertible
+and non-invertible inference rules. Andreoli observed that the
+connectives of linear logic can be divided into two categories, dubbed
+synchronous and asynchronous. Synchronous connectives are those whose
+right rules are \emph{invertible}, i.e. they can be applied eagerly
+during proof search without losing provability and so the order in
+which these rules are applied is irrelevant, and whose left rules
+are not invertible. Asynchronous connectives are dual. \todo{Say which
+  are which}
+%
+Given this separation, focusing divides proof search into two
+phases: %
+the inversion phase ($\Uparrow$), in which we apply \emph{all} invertible rules
+eagerly, and the focusing phase  ($\Downarrow$), in which we decide a proposition
+to focus on, and then apply non-invertible rules, staying \emph{in focus}
+until we reach an asynchronous, i.e. invertible,
+proposition or the proof is complete.
+
+\todo[inline]{Isto so faz sentido olhando para os judgments}
+For the invertible rules, a
+third, ordered, context ($\Omega$) is needed to hold propositions
+resulting from decomposing connectives; propositions we'll later try
+to invert, moving them to the linear context ($\Delta$) when we fail
+to.
 
@@
 
+\todo[inline]{A frente de cada ``descricao'', escreve o tipo pois
+  ajuda o leitor}
 The core language is a simply-typed linear lambda calculus with linear
@@
 % TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
-% usá-los como tipos gerais?
+% usá-los como tipos gerais? NAO :)
 \[
@@
 by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
-first rule is read: to synthetise a program of type $A \lolli B$ while inverting
+first rule \todo{qual ``first''? referencia por nome} is read: to synthesize a program of type $A \lolli B$ while inverting
 right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
@@
 
+\todo[inline]{Aqui tens de explicar uma regra pelo menos}
 The decision rules are followed by either left or right focus rules:
@@
 \mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
-is a named sum type of any type, i.e. a named type that can be instantiated by one of many
-tags (also known as constructors) that take exactly one fixed type, which might
+is a tagged sum of any type, i.e. a named type that can be instantiated by one of many
+tags (or constructors) that take some value of  a fixed type, which might
 be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
@@
 language, the programmer can define custom ADTs; as an example, we show the
-definition of an ADT which might be constructed with zero, one, or two
-propositions of type $A$, using its syntax: \mintinline{haskell}{data Container
-= None 1 | One A | Two (A * A)}. The language grammar is extended as:
+definition of an ADT which holds zero, one, or two
+values of type $A$, using the syntax: \mintinline{haskell}{data Container
+= None 1 | One A | Two (A * A)}. The grammar is extended as:
 \[
@@
 
-The semantics of the ADTs relate to those of the plus ($\oplus$) type --
-both are additive disjunctions.  A proof of an ADT requires only a proof of one
-of the types that a constructor takes, similar to the way $\oplus$ requires only
-proof of either the left or right type it consists of. Analogously, a proof of
-$C$ by deconstructing an ADT stipulates that all types that can construct the
-ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
-the inference rules for a simple ADT mimic the $\oplus$ rules, for an
-arbitrary ($>1$) number of types, rather than exactly two. Therefore, there's one
-left rule for ADTs, and an arbitrary number of right rules, one for each
+The semantics of ADTs relate to those of the plus ($\oplus$) type
+-- both are additive disjunctions.  To construct a value of an ADT we
+must use one of its constructors, similar to the way $\oplus$ requires
+only proof of either the left or right type it consists
+of. Analogously, we can deconstruct a value of an ADT via pattern
+matching on its constructors, where all branches of the pattern match
+must have the same type -- akin to the left rule for the $\oplus$
+connective. In effect, the inference rules for a simple ADT are a
+generalized form of the $\oplus$ rules.  Therefore, there's one left
+rule for ADTs, and an arbitrary number of right rules, one for each
 constructor:
@@
 
-A more general (and interesting) formulation of ADTs says an ADT can be
-recursive (or "indutively defined"), meaning constructors can take as arguments
-values of the type they are defining. Seemingly small, this change has a big
-impact in the synthesis process. Take, for instance, the ADT defined as
-\mintinline{haskell}{data T = C1 T}, the synth goal $T \lolli C$, and its
-derivation:
+A more general formulation of ADTs says an ADT can be recursive (or
+"inductively defined"), meaning constructors can take as arguments
+values of the type they are defining. This change has
+a significant impact in the synthesis process. Take, for instance, the ADT
+defined as \mintinline{haskell}{data T = C1 T}, the synthesis goal
+$T \lolli C$, and its derivation:
 %
@@
 %
-Using our current system, we are to apply an infinite number of times the left
-ADT rule, never reaching a final state for the proof. Symmetrically, the
-derivation for the synth goal $T$ is also infinite.
+Using our current system, we are to apply an infinite number of times
+the left ADT rule, never closing the proof. Symmetrically, the
+derivation for the goal $T$ is also infinite.
 %
@@
 To account for this situation, we impede the decomposition of an ADT in subsequent
-proofs of its branches, and, symetrically, don't allow construction of an
+proofs of its branches, and, symmetrically, don't allow construction of an
 ADT when trying to synthetise an argument for its constructor.
@@
 for constraints on construction and $\Rho_D$ for constraints on
-deconstruction. Together, they hold a list of ADTs that can not be constructed or
-deconstructed at some point in the proof. For convenience, they are represented by a single $\Rho$ if
-unused. All rules must trivially propagate these. The ADT rules are explicitly
-modified to take this into consideration:
+deconstruction. Together, they hold a list of ADTs that cannot be constructed or
+deconstructed at a given point in the proof. For convenience, they are represented by a single $\Rho$ if
+unused. All non-ADT rules trivially propagate these. The ADT rules
+are then extended as follows, where $\Rho'_C = \Rho_C,T$ if $T$ is
+recursive and $\Rho'_C = \Rho_C$ otherwise (similarly for $\Rho'_D$):
 %
@@
 \end{mathpar}
-where
-\begin{mathpar}
-    \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
-    T\ \textrm{\textbf{else}}\ \Rho_C
-
-    \Rho'_D = \textrm{likewise}
-\end{mathpar}
-%
-These modifications prevent the infinite derivations in the scenarios described
-above, however, they also greatly limit the space of derivable programs, leaving
-the synthetiser effectively unable to synthetise from specifications with recursive
-types, which is undesirable, so we add three rules to complement the
-restrictions on construction and destruction of recursive types.
-
-First, after decomposing a recursive ADT, we must consume all constructor
-parameters, and since we can't decompose the self type any further because of a
-restriction, we should be able to use it from the linear context; even though,
-non-restricted types are not to take that course since they weren't unnaturally
-restricted, and because it would result in additional non-determinism and
-infinite recursion. Second, without any additional rules, an ADT in the linear
-context will loop back to the inversion context, and then loop again, going back and
-forth between the two contexts for infinity; instead, when focused on an ADT, we should
-either instance the goal (provided they're the same type), or switch to
-inversion if and only if its decomposition isn't restricted. The three
-following rules ensure this:
+% where
+% \begin{mathpar}
+%     \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
+%     T\ \textrm{\textbf{else}}\ \Rho_C
+
+%     \Rho'_D = \textrm{likewise}
+% \end{mathpar}
+
+These modifications prevent the infinite derivations in the scenarios
+described above. However, they also greatly limit the space of
+derivable programs, leaving the synthesizer effectively unable to
+synthesize from specifications with recursive types. To prevent this,
+so we add three rules to complement the restrictions on construction
+and destruction of recursive types.
+%
+First, after decomposing a recursive ADT, we must consume
+\todo{consume? deconstruct?} all constructor parameters, and since we
+can't decompose the self type any further because of a restriction, we
+should be able to use it from the linear context. The situation is
+slightly more involved when considering unrestricted types (i.e.~types
+of the form $\bang A$), which cannot be treated in this way since they
+weren't restricted, and because it would result in additional
+non-determinism and infinite recursion.\todo{check} 
+%
+Second, without any additional rules, an ADT in the linear context
+will loop back to the inversion context, looping back and forth
+between the two contexts; instead, when focusing on an ADT, we should
+either instantiate the goal (provided they're the same type), or switch
+to inversion if and only if its decomposition isn't restricted. The
+three following rules ensure this:
 \begin{mathpar}
@@
 %
-Finally, these rules end up ensuring that a recursive ADT will deconstruct
-itself once, and move the recursive argument from its constructor to the linear
-context -- essentially forcing it to be used as a function argument, and most
-usually that function will be the synthesis recursive call.
-
-\mypara{Recursion} The main idea behind synthesis of recursive programs is the
-labeling of the main goal and the addition of its type, under that name, to the
-unrestricted context. That is, to synthetise a function of type $A \lolli B$ named
-\emph{f}, the initial judgment can be written as
+The rules above ensure that a recursive ADT will be deconstructed
+once, and move the recursive argument from its constructor to the
+linear context -- essentially forcing it to be used as a function
+argument, which will typically be the recursive call.
+
+\mypara{Recursion} The main idea behind synthesis of recursive
+programs is the labeling of the main goal and the addition of its
+type, under that name, to the unrestricted context. That is, to
+synthesize a recursive function of type $A \lolli B$ named \emph{f},
+the initial judgment can be written as
 \begin{mathpar}
@@
 unrestricted context, but to the linear one as well.
-
-Having said that, we won't allow the use of the recursive call in $\Gamma$ right
-away. If we did, every synth goal would have a trivial proof (a non-terminating
-function that just calls itself) that could even shadow other valid, relevant,
-solutions.  Instead, our framework allows the use of recursion only after having
-deconstructed a recursive ADT. To avoid complicating the rules, we'll add the
-invariant: the recursive hypothesis can only be \textbf{used in recursive branches of
-ADT deconstruction}, i.e. the recursive call should only take "smaller", recursive,
-hypothesis as arguments. To illustrate, having \mintinline{haskell}{data List =
-Nil | Cons (A * List)}, recursion is only
-"available" when deconstructing ($\textrm {List} \vdash C$), and only in
-the \emph{Cons} branch. % (TODO: rewrite? O bold é meio estranho não?)
-
-\mypara{Polymorphic Types} A polymorphic specification is a type of form
-$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of bound
-variables that quantify universally the type variables in $\tau$, which is also
-called a \emph{scheme}.  Synthesis of a scheme comprises of turning it into a
-non-quantified type, and then treating its type variables agnostically.  First,
-\textbf{type variables are considered atomic types}, then, instantiate the bound
-variables of the scheme as described by the Hindley-Milner inference
-method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names
-for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
-\alpha$ could be instantiated to $\alpha0 \lolli \alpha0$. We add a rule for
-this:
+%
+However, we must restrict immediate uses of the recursive call since
+otherwise every goal would have a trivial proof (a non-terminating
+function that just calls itself), shadowing relevant solutions.
+Instead, our framework allows the use of recursion only after having
+deconstructed a recursive ADT via the following invariant: the
+recursive hypothesis can only be used in \emph{recursive branches of
+  ADT deconstruction}, i.e. the recursive call should only take
+``smaller'', recursive, hypothesis as arguments. To illustrate, in any
+recursive function with a list argument (whose type is defined as
+\mintinline{haskell}{data List = Nil | Cons (A * List)}), recursive
+calls are only allowed when considering a judgment of the form
+$\textrm {List} \vdash C$, i.e.~when a list value is available to
+produce the goal $C$, and only in the \emph{Cons}
+branch. % (TODO: rewrite? O bold é meio estranho não?)
+
+\mypara{Polymorphic Types} A polymorphic specification is a type of
+form $\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a
+set of variables that stand for any (non-polymorphic) type in
+$\tau$. Such a type is also called a \emph{scheme}.
+Synthesis of a scheme comprises of turning it into a non-quantified
+type, and then treating its type variables uniformly.  First,
+type variables are considered \emph{atomic types}, then, we instantiate
+the bound variables of the scheme as described by the Hindley-Milner
+inference method's~\cite{HM-infer} instantiation rule (put simply,
+generate fresh names for each bound type variable); e.g. the scheme
+$\forall \alpha.\ \alpha \lolli \alpha$ could be instantiated to
+$\alpha0 \lolli \alpha0$. We add a rule for this:
 \begin{mathpar}
@@
 %
-As such, the construction of a derivation in which, to some types, no specific
-rules can be applied (except for the \textsc{init} rule, a type variable can
-instance a goal that's a type variable of the same name), corresponds to the
-synthesis of a program where some expressions are treated agnostically --
-nothing constrains their type -- meaning the program is polymorphic. The
-simplest example is the polymorphic function \emph{id} of type $\forall \alpha
-.\ \alpha \lolli \alpha$. The
-program synthetised from that specification is $\lambda x . x$, a lambda
-abstraction that doesn't constrain its parameter $x$ in any way.
+As such, the construction of a derivation in which the only rule that
+can derive an atom is the \textsc{init} rule corresponds to the
+synthesis of a program where some expressions are treated agnostically
+-- nothing constrains their type -- i.e.~the program is
+polymorphic. The simplest example is the polymorphic function
+\emph{id} of type $\forall \alpha .\ \alpha \lolli \alpha$. The
+program synthesized from that specification is $\lambda x . x$, a
+lambda abstraction that does not constrain the type of its parameter
+$x$ in any way.
 % TODO: \todo{R : rewrite this? não estranho?}
 
-The interesting challenge of polymorphism in synthesis is the usage of
+The main challenge of polymorphism in synthesis is the usage of
 schemes from the unrestricted context.  To begin with, $\Gamma$ now
-holds both types and schemes. Consequently, after the rule
-\textsc{decideLeft!} is applied, we become left focused on either a
-type or a scheme. Since left focus on a type is already well
-formulated, only rules for left focus on a scheme are needed. Our
-algorithm instances bound type variables of the focused scheme with
-existential type variables, and the instanced type becomes the left
-focus. Inspired by the Hindley-Milner system, we also generate
+holds both (monomorphic) types and schemes. Consequently, after the
+rule \textsc{decideLeft!} is applied, we are left-focused on either a
+type or a scheme. Since left focus on a type is already well defined,
+we need only specify how to focus on a scheme.
+%
+Our algorithm instantiates bound type variables of the focused scheme
+with fresh \emph{existential} type variables, and the instantiated type becomes the
+left focus. Inspired by the Hindley-Milner system, we also generate
 inference constraints on the existential type variables (postponing
@@
 across derivation branches the same way the linear context is (by
-havin an input and output context ($\Theta/\Theta'$)).  In contrast to
-Hindley-Milner's inference, everytime a constraint is added it is solved against all other
-constraints -- a branch of the proof search is desired to fail as soon
-as possible.
+having an input and output context ($\Theta/\Theta'$)).  In contrast to
+Hindley-Milner's inference, everytime a constraint is added it is
+solved against all other constraints -- a branch of the proof search
+is desired to fail as soon as possible.
 %TODO: Como fazer? \todo{O que
@@
 % propriamente o problema de misturar variaveis existenciais com universais}
-Note that we
-instance the scheme with \emph{existential} type variables ($?\alpha$)
-rather than just type variables ($\alpha$) since the latter represent universal
-types in the synth, and the first represent an existential instance of
-a scheme, that might inclusively constrain traditional type
-variables. Additionally, it's required that all existential type
-variables are assigned a type. These concepts
-are formalized with the rules:
+Note that we instantiate the scheme with \emph{existential} type
+variables ($?\alpha$) rather than just type variables ($\alpha$) since
+the latter represent universal types during synthesis, and the former
+represent a concrete instance of a scheme, that might induce
+constraints on other type variables. Additionally, we require that
+all existential type variables are assigned a type. These concepts are
+formalized with the rules:
 \begin{mathpar}
@@
 
-\mypara{More Restrictions} We can now look at two more sources of infinite
+\mypara{Further Challenges} We now consider two more sources of infinite
 recursion in the synthesis process. The first is the use of an unrestricted
 function to synthetise a term of type $\tau$ that in turn will require a term of
-the same type $\tau$. A good example is the sub-goal judgment $\Gamma,(a \lolli
+the same type $\tau$. An example is the sub-goal judgment $\Gamma,(a \lolli
 b \lolli b); (a \lolli b \lolli b) \Downarrow\ \vdash b$ that
-appears while synthetising \emph{foldr} -- we apply ($\lolli$L)
+appears while synthesizing \emph{foldr} -- we apply ($\lolli$L)
 until we can use \textsc{init} ($b \Downarrow\ \vdash b$), and then we must
-synthetise an argument of type $b$. Without any additional restrictions, we
-can end up left focused on $(a \lolli b \lolli b)$ again, and again require $b$,
+synthesize an argument of type $b$. Without any additional restrictions, we
+may become again left focused on $(a \lolli b \lolli b)$, and again require $b$,
 and on and on. The solution will be to disallow the usage of the same function

### 2021-07-20 article/article.tex
tweaks

@@
 to focus on, and then apply non-invertible rules, staying \emph{in focus}
-until we reach an asynchronous, i.e. invertible,
-proposition or the proof is complete.
+until we reach an asynchronous (i.e. invertible proposition), the
+proof is complete or no rules are applicable, in which the proof must
+\emph{backtrack} to the state at which the focusing phase began.
 
@@
 The \synname\ synthetiser operates as part of the pipeline that processes a
-full program in the \synname\ language, and is called to generate parts marked
+full program in the \synname\ language, and is called to generate terms marked
 (by the syntax \emph{\{\{ ... \}\}}) for
-synthesis of said program. The main pipeline consists of:
+synthesis. The main pipeline consists of:
 \[
@@
 
-The desugaring module converts the frontend AST to a core AST without syntatic
-sugar and using the locally-nameless representation~\cite{locally nameless}, in
-which all bound variables have no name, and instead are addressed using the
-number of binders traversed since the binding.
+The desugaring module converts the frontend AST to a core AST without
+syntatic sugar and using the locally-nameless
+representation~\cite{locally nameless}, in which all bound variables
+are represented with Debruijn indices rather than names (i.e.~using
+the distance to the binding site in terms of the number of traversed
+binders),
 % TODO: não ficou muito mais claro :\todo{Explica o que e isto}
@@
 
-The synthesis module iterates over the synth marks and runs the synthetiser with
-its context and goal for each, minimizing the resulting expression, and
-replacing the mark in the frontend AST with the synthed program.
+The synthesis module iterates over the synthesis marks and runs the
+synthetiser with its context and goal for each mark, minimizing the
+resulting expression (i.e.~$\eta$-reducing it), and replacing the mark
+in the frontend AST with the synthed program.
 
@@
 
-\mypara{General} The complete implementation of the synthetiser and environment
-pipeline was done in Haskell...
-
-\mypara{Backtracking} Even though the synthetiser formal system helps, there's
-still non-determinism in the synthesis process. At diverse times in the proof
-construction we must choose one of multiple rules to apply, and in the event of
-not being able to construct the proof following that choice, we want to
-backtrack to the decision point and attempt to create a proof taking a different
-route. To this effect, our prototype synthetiser depends on \emph{logict}, "a backtracking logic-programming
-monad"~\cite{logict}.
-
-\mypara{SMT Solving} To typecheck and synthetise refinement types and dependent
-functions, we need to contact an SMT solver to check satisfiability. For this,
-we use the library \emph{sbv}~\cite{sbv}. Unfortunately, this library does not
-support the usage of the SMT solvers to satisfy and output more complicated
-first-order logic uninterpreted functions -- since this is needed to synthetise refinement types
-when right focused, we instead use unsafe library primitives to communicate with
-the solver directly, constructing the logical formulas ourselves. For 
-satisfying formulas, for instance, when typechecking, or left focused on a
-refinement type, we use the library api successfully to get a veridct on
-satisfiability.
-
-\mypara{Memoization} Some proofs attempt to synthetise the same sub-goal with
-the same premises more than once. Since these proof attempts might be expensive
-performance-wise, and we can rely on the program's determinism (equal calls to
-synthetise result in equal outcomes), we add to our synthetiser memoization, an
-optimization technique that stores function calls and returns cached values for
-equal inputs. For our synthetiser, two calls are equivalent when contexts
-$\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively the left focus), and
-the goal, are equivalent. Because comparing all of these is slow, our
-implementation hashes the relevant values and uses the result as the key.
-Furthermore, a branch we want to memoize might fail, not providing any result --
-we also want to record synthesis failure, as it happens more often than
-successful results. We capture failure of a logict computation using the $ifte$
-function and represent failure or success information with the Maybe type. We
-ran benchmarks to measure the impact of memoization on the performance. It's
-quite noticeable in synthesis with multiple functions in the unrestricted
-context, or more complex specifications. For example, if we add a new function
-\emph{read} to the unrestricted context in the \emph{array} problem seen in
-Section~\ref{sec:overview}, synthesis with memoization will take around
-$5s$, while synthesis without memoization doesn't terminate even after
-$15m$ running.
+\mypara{General} The implementation was carried out in Haskell. We
+highlight some key points.
+
+\mypara{Backtracking} Even though focusing reduces non-determinism
+during proof search, it cannot fully eliminate it. At several points
+during proof construction we must choose one of multiple rules to
+apply, and in the event of not being able to construct the proof
+following that choice, we want to backtrack to the decision point and
+attempt to derive a proof taking a different route. To this effect,
+our prototype synthetiser depends on \emph{LogicT}, "a backtracking
+logic-programming monad"~\cite{logict}.
+
+\mypara{SMT Solving} To typecheck and synthesize refinement types and
+dependent functions, we interface with an SMT solver to check
+satisfiability. For this, we use the library
+\emph{SBV}~\cite{sbv}. Unfortunately, this library does not support
+the usage of SMT solvers to check first-order logic with uninterpreted
+functions. Since this is needed for synthesis with refinement types
+when right focused, we instead use unsafe library primitives to
+communicate with the solver directly, constructing the logical
+formulas directly. For satisfying formulas, for instance, when
+typechecking, or left focused on a refinement type, we use the library
+API as an oracle.
+
+\mypara{Memoization} Some proofs attempt to synthetise the same
+sub-goal with the same premises more than once. Since these proof
+attempts might be expensive performance-wise, and we can rely on the
+program's determinism (equal calls to synthetise result in equal
+outcomes), we added memoization to our synthesizer -- an optimization
+technique that stores function calls and returns cached values for
+equal inputs.
+%
+For our synthesizer, two calls are equivalent when
+contexts $\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively
+the left focus), and the goal, are equivalent. Because comparing all
+of these is slow, our implementation hashes the relevant values and
+uses the result as the key\todo{via a hashconsing library?}.
+%
+Furthermore, a branch we want to memoize
+might fail, not providing any result -- we also want to record
+synthesis failure, as it happens more often than successful
+results. We capture failure of a LogicT computation using the $\mathit{ifte}$
+function and represent failure or success information with the \texttt{Maybe}
+type. We ran benchmarks to measure the impact of memoization on 
+performance. It's quite noticeable in synthesis with multiple
+functions in the unrestricted context, or more complex
+specifications. For example, if we add a new function \emph{read} to
+the unrestricted context in the \emph{array} problem seen in
+Section~\ref{sec:overview}, synthesis with memoization will take
+around $5s$, while synthesis without memoization doesn't terminate
+even after $15m$ running.
 %
@@
 backtracking branches) and by finetuning key-equality by ignoring context
-components that aren't relevant to the synthesis call, to name some examples.
+components that aren't relevant to the synthesis call, to name some
+examples. We leave such optimizations to future work.
 

### 2021-07-20 article/article.tex
Backtracking writing fair conjunction + disjunction +- & tweaks

--- 2021-07-20 article/article.tex
@@
 graded types~\cite{DBLP:conf/lopstr/HughesO20}. Regardless of the kind
-of specification, program synthesis must deal with two main
-inherent sources of complexity -- search over the space of
-valid programs, and interpreting user intent.
+of specification, program synthesisers have to deal with two main
+inherent sources of complexity -- searching over a vast space of
+(potentially) valid programs, and interpreting user intent.
 
@@
 %
-Linear types constrain resource
+Linear types differ from more traditional types in that they constrain resource
 usage in programs by \emph{statically} limiting the number of times certain
@@
 
-\paragraph{Contributions and Outline.} Despite their long-known potential
-\cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018}
-and strong proof-theoretic
+\paragraph{Contributions.} Despite their long-known potential (\emph{linear
+types can change the world!}\cite{Wadler90lineartypes},
+\cite{DBLP:journals/mscs/CairesPT16,Bernardy_2018}) and strong proof-theoretic
 foundations
 \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
-synthesis with linear types combined with other advanced typing features
-has generally been overlooked in the literature.  In this work we present a framework for synthesis
+synthesis with linear types in a more advanced context
+has generally been overlooked in the literature.  In this paper we present a framework for synthesis
 with linear types extended with recursive algebraic data types, parametric
@@
 % TODO: extensible? proving feasability of... ? showing potential for ... ?
-We first introduce linear types as specifications
-and outline the synthesis process, leveraging linearity, by example
-(\S~\ref{sec:overview}). We then present the formal system driving the
-synthesis (\S~\ref{sec:formal_system}) and describe the architecture
-of our framework named \synname, examining technical details and key
-implementation challenges (\S~\ref{sec:architecture}). Finally,
-we evaluate our work through expressiveness and performance benchmarks
-(\S~\ref{sec:evaluation}) and discuss related work (\S~\ref{sec:related}).
+
+\paragraph{Outline.} First, we discuss linear types as specifications, the
+outline of the synthesis process, and what that entails for the synthetiser --
+ilustrated with concrete results (Section~\ref{sec:overview}). Then, we present
+the formal system driving the synthesis (Section~\ref{sec:formal_system}).
+Aftwerwards, we describe the architecture of our synthetiser named \synname, and
+examine technical details and challenges of the implementation
+(Section~\ref{sec:architecture}). Finally, we evaluate our work through
+expressiveness benchmarks (i.e.~can we synth $X$?) and time benchmarks (i.e.~how
+fast can we synth a term of type $T$?) (Section~\ref{sec:evaluation}).
 %TODO: And appendice with...
@@
 
-The \synname\ synthetiser combines linear types with recursion,
+% TODO: Why linear types?
+The \synname\ synthetiser combines the features of linear types with recursion,
 parametric polymorphism, recursive algebraic data types, and
-refinement types. The synthetiser is built on top of a system of proof-search for
-linear logic. Proof search relates to program synthesis via the Curry-Howard
-correspondence~\cite{}, which states that \emph{propositions in a logic} are
-\emph{types}, and \emph{proofs} of those propositions are well-typed
+refinement types. The synthetiser is built on top of a system of proof-search in
+linear logic. Proof-search relates to program synthesis through the Curry-Howard
+correspondence~\cite{}, which states that \emph{propositions in a logic} map
+directly to \emph{types}, and \emph{proofs} of those propositions to well-typed
 \emph{programs} -- finding a proof of a proposition is finding a program with
@@
 
-Linear types make for more precise specifications than simple types
-because information on which resources must necessarily be used is
-encoded in the type. For instance the type $\mathsf{Int} \lolli
-\mathsf{Int} \lolli \mathsf{Int}$ denotes a function that \emph{must}
-use its two integer arguments to produce an integer.
-Their preciseness also affects the search space:
-all programs where a linear resource is used non-linearly (i.e. not
-\emph{exactly once}) are ill-typed. With linearity built into the
-synthesis process, usage of a linear proposition more than once is not
-considered, and unused propositions are identified during synthesis,
-constraining the space of valid programs.  % TODO: Reword?
+Linear types make for more precise specifications than simple traditional types
+because information on which resources must necessarily be used is encoded in
+the type. Their preciseness also influences the search space: all programs where
+a linear resource is used non-linearly (i.e. not \emph{exactly once}) are
+ill-typed. With linearity built-in to the synthesis process, usage of a
+linear proposition more than once is not even considered, and unused ones are caught
+mid-process, making proof-search in linear logic potentially faster than
+in classical logic.  % TODO: Reword?
 
 The core of the synthesis is a \emph{sound} and \emph{complete} system
-consisting of \emph{bottom-up} proof-search in propositional linear
-logic based on \emph{focusing}~\cite{}. Our approach, being grounded
-by propositions-as-types, ensures that all synthetised programs
-(i.e.~proofs) are well-typed \emph{by construction} (i.e.~if the
-synthesis procedure produces a program, then the program intrinsically
-satisfies its specification). Moreover, we can leverage the modularity
-of the proof-search based approach along two axes: first, since proof
-search need not construct only closed proofs, we can effectively
-synthesize program sub-expressions (i.e.~synthesis based on typed
-holes); secondly, the framework is amenable to extensions to the core
-propositional language, allowing for the introduction of a richer type
-structure while preserving the correctness of programs by
-construction.
-
-The \synname\ programming language is naturally developed alongside
-the synthetiser. Synthesis goals are inserted in the program by use of
-a \emph{synth} keyword or mark, indicating a program should be
-generated for a given type. Additionally, it opens the possibility of
-synthesis within a context, that is, with knowledge of other functions and
-structures defined in the same program.
-
-\paragraph{Core linear calculus.} Initially, we synthesize from
+consisting of \emph{bottom-up} proof-search in propositional linear logic based
+on \emph{focusing}~\cite{}. Our approach, being grounded by
+propositions-as-types, ensures that all synthetised programs (i.e.~proofs) are
+well-typed \emph{by construction} (i.e.~if the synthesis procedure terminates,
+then the program satisfies its specification). Moreover, we can leverage the
+modularity of the proof-search based approach in two key axes: first, since
+proof search need not construct only closed proofs, we can effectively
+synthetise program sub-expressions (i.e.~synthesis based on typed holes);
+secondly, the framework is amenable to extensions to the core propositional
+language, allowing for the introduction of a richer type structure while
+preserving the correctness of programs by construction.
+
+The \synname\ programming language is naturally born alongside the synthetiser
+to create an environment for it. Synth goals are inserted in the program by use
+of a \emph{synth} keyword or mark, indicating a program should be generated for a given
+type. Additionally, it opens the possibility of synthesis with context, that is,
+with knowledge of other functions and structures defined in the same program.
+
+\paragraph{Core linear calculus.} Initially, we synthetise from
 specifications that only use literals and canonical linear types ($\lolli, \with, \tensor, 1,
@@
 the specification $(A \lolli B \lolli C) \lolli (A \tensor B) \lolli C$ produces
-the function \mintinline{haskell}{\a -> \b -> let c*d = b in a c d},
-which takes an argument \mintinline{haskell}{a}, a
-\emph{curried} linear function of type $A \lolli B \lolli C$,  and an
-argument \mintinline{haskell}{b}, a \emph{linear
-  pair} of an $A$ and a $B$ and produces a value of type $C$ by
-deconstructing the pair and applying \mintinline{haskell}{a} to the
-corresponding elements of the pair (i.e.,~\emph{uncurries} the
-function \mintinline{haskell}{a}).
-
-\paragraph{Beyond propositional logic.} To be able to synthesize more
-interesting programs -- and thus, empirically prove the feasability of
-more relevant synthesis in a linear context, we extend the syntax and
-type system with recursiveness, parametric polymorphism, algebraic
-data types (ADTs) and type refinements.
+the program \mintinline{haskell}{(\a -> \b -> let c*d = b in a c d)}
+(known as \emph{uncurry}).
+
+\paragraph{Beyond propositional logic.} To enhance \synname\ and its language,
+counter the core's lack of expressiveness, synthetise more interesting programs
+-- and thus, empirically prove the feasability of more relevant synthesis in a
+linear context, we extend the syntax and type system with recursiveness,
+parametric polymorphism, algebraic data types (ADTs) and type refinements.
 
@@
 
-\todo[inline]{Tens de explicar o que significa o $\bang$, see above.}
 \paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring some types to be
@@
 
-\todo[inline]{Tens de dizer o que significa o refinamento concreto e o
-que e a funcao mult e add}
-\paragraph{Refinements.} Refinements are logical predicates
-which must hold for any element that inhabits a given type. The \synname\
-language supports simple refinements on integers with arithmetic
-expressions in the predicates. They assert the robustness of our main
-framework, i.e., how it is amenable to significant additions without
-necessarily interfering with the overall approach.
+\paragraph{Refinements.} Refinements are predicates added to a type which must hold for any
+element that inhabits said type. The \synname\ language supports simple
+refinements on integers with arithmetic expressions in the predicates. They
+assert the robustness of our main framework, i.e., how it is amenable to
+significant additions without necessarily interfering with the algorithm.
 %
@@
 
-\paragraph{Guiding the Search.} Specifications can be additionally
-augmented with three keywords: ``using'', ``depth'' and
-``assert''. Through them, we can fine-tune the synthesis process and,
-respectively, force certain functions to be present in the synthesis
-outcome, allow a \emph{deeper} search in the valid programs space, and
-require given predicates to hold true in the program after the
-synthesis is complete.
+\paragraph{Configuration.} Specifications can be additionally augmented with three keywords: ``using'', ``depth'' and
+``assert''. Through them, we can fine-tune the synthesis process and, respectively,
+force certain functions to be present in the synthesis outcome, allow a
+\emph{deeper} search in the valid programs space, and require given predicates
+to hold true in the program after the synthesis is complete.
 
 Taken directly from the Linear Haskell paper~\cite{Bernardy_2018}, we present a
-more intricate example of synthesis: given the linear type signatures for array
+complex example of synthesis: given the linear type signatures for array
 primitives (\emph{newMArray} passes a new mutable array to a function that uses
@@
 \emph{freeze} consumes a mutable array and produces an imutable array,
-\emph{foldl} has the default definition) we synthesize a function
+\emph{foldl} has the default definition) we try to synthetise a function
 ``array'' that provides an immutable array given a list of pairs to write. The input program is formulated as
@@
 %
-Matching exactly the linear definition for
-\emph{array} from~\cite{Bernardy_2018}, \synname\
-outputs (in a fraction of a second):
+And, in a small fraction of a second, matching exactly the linear definition for
+\emph{array} as seen in Linear Haskell~\cite{Bernardy_2018}, \synname\ outputs:
 %
@@
 
-\section{The \synname\ Synthesis Framework}\label{sec:formal_system}
+\section{Synthesis Framework}\label{sec:formal_system}
 % with Linear Types, Polymorphism and Refinements
@@
 
-Program synthesis from linear types with polymorphism, recursive
-algebraic data types, and refinements, is essentially new in the synthesis 
+Program synthesis from linear types in combination with polymorphism, recursive
+algebraic data types, and refinements, is new in the synthesis 
 literature. Despite the substantial amount of research on linear logic and
-proof-search upon which we base our core synthesizer, formal guidelines for
-richer types and their intrinsic challenges (such as infinite
-recursion) must be developed.
-
-In this section we formalize  the techniques that guide
+proof-search upon which we found our core synthetiser, formal guidelines for the
+richer types and their inherent complex problems (such as infinite recursion)
+don't exist. % TODO: Developed an algorithm...?
+
+With this section we aim to formalize as rules the novel techniques that guide
 synthesis from our more expressive specifications, alongside the already well
 defined rules that model the core of the synthetiser, putting together a
-\emph{sound} set of inference rules that characterizes our framework
-for linear synthesis of
-recursive programs from specifications with the select richer types,
+\emph{sound} set of inference rules that characterizes our framework for synthesis of
+recursive programs from specifications with the select richer types in a linear context,
 % TODO: Better?....
@@
 
-We note that a \emph{sound} set of rules guarantees we cannot synthesize
-incorrect programs; and that the valid programs derivable through them
-reflect the somewhat subjective trade-offs we committed to. Different
-choices and approaches outside the core might lead to completely
-distinct synths and spaces of valid programs.
-
-\todo[inline]{Indicar o judgment ``declarativo'' e depois os
-  judgments em focusing}
-
-\mypara{Core Rules} The core system comprises of proof search for
-(intuitionistic) linear logic sequent calculus proofs, based on a
-system of resource management and focusing.
-%
-In intuitionistic sequent calculi, each connective has a so-called
-\emph{left} and a \emph{right} rule, which effectively define how to
-decompose an ambient assumption of a given proposition and how to
-prove a certain proposition is true, respectively. 
-Andreoli's \emph{focusing} for linear logic~\cite{Andreolli focusing}
-is a technique to remove non-essential nondeterminism from
-proof-search by structuring the application of so-called invertible
-and non-invertible inference rules. Andreoli observed that the
-connectives of linear logic can be divided into two categories, dubbed
-synchronous and asynchronous. Synchronous connectives are those whose
-right rules are \emph{invertible}, i.e. they can be applied eagerly
-during proof search without losing provability and so the order in
-which these rules are applied is irrelevant, and whose left rules
-are not invertible. Asynchronous connectives are dual. \todo{Say which
-  are which}
-%
-Given this separation, focusing divides proof search into two
-phases: %
-the inversion phase ($\Uparrow$), in which we apply \emph{all} invertible rules
-eagerly, and the focusing phase  ($\Downarrow$), in which we decide a proposition
-to focus on, and then apply non-invertible rules, staying \emph{in focus}
-until we reach an asynchronous, i.e. invertible,
-proposition or the proof is complete.
-
-\todo[inline]{Isto so faz sentido olhando para os judgments}
-For the invertible rules, a
-third, ordered, context ($\Omega$) is needed to hold propositions
-resulting from decomposing connectives; propositions we'll later try
-to invert, moving them to the linear context ($\Delta$) when we fail
-to.
+Even so, note that \emph{soundness} and \emph{completeness} have two very
+different meanings, and our \emph{sound} set of rules only guarantees we won't
+synthetise incorrect programs, but still hides a possibly infinite number of correct
+proofs; and that the valid programs derivable through it reflect the somewhat
+subjective trade-offs we decided on. Different choices and approaches outside
+the core might lead to completely distinct synths and spaces of valid programs.
+
+\mypara{Core Rules} The core system comprises of proof-search in sequent
+calculus rules for linear logic with resource-management and focusing.
+%
+Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
+focusing} is a technique to remove inessencial non-determinism from
+proof-search by structuring the application of invertible and non-invertible
+inference rules. Rules are classified as \emph{left} or \emph{right},
+\emph{invertible} or \emph{non-invertible} as follows:...\todo{não estou a
+conseguir formular bem esta introdução}
+The proof search is separated into two phases:
+%
+inversion ($\Uparrow$), in which we apply all invertible rules eagerly, and
+focusing ($\Downarrow$), in which we decide a proposition to focus on, and then
+apply non-invertible rules, staying focused until either the focus is again an
+asynchronous, i.e. invertible, proposition, or the proof is complete.  For the
+invertible rules, a third, ordered, context ($\Omega$) is needed to hold
+propositions resulting from decomposing connectives; propositions we'll later
+try to invert, moving them to the linear context ($\Delta$) when we fail to.
 
@@
 
-\todo[inline]{A frente de cada ``descricao'', escreve o tipo pois
-  ajuda o leitor}
 The core language is a simply-typed linear lambda calculus with linear
@@
 % TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
-% usá-los como tipos gerais? NAO :)
+% usá-los como tipos gerais?
 \[
@@
 by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
-first rule \todo{qual ``first''? referencia por nome} is read: to synthesize a program of type $A \lolli B$ while inverting
+first rule is read: to synthetise a program of type $A \lolli B$ while inverting
 right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
@@
 
-\todo[inline]{Aqui tens de explicar uma regra pelo menos}
 The decision rules are followed by either left or right focus rules:
@@
 \mypara{Algebraic Data Types} In its simplest form, an algebraic data type (ADT)
-is a tagged sum of any type, i.e. a named type that can be instantiated by one of many
-tags (or constructors) that take some value of  a fixed type, which might
+is a named sum type of any type, i.e. a named type that can be instantiated by one of many
+tags (also known as constructors) that take exactly one fixed type, which might
 be, e.g., a product type ($A \tensor B$), or unit ($1$), in practice allowing
@@
 language, the programmer can define custom ADTs; as an example, we show the
-definition of an ADT which holds zero, one, or two
-values of type $A$, using the syntax: \mintinline{haskell}{data Container
-= None 1 | One A | Two (A * A)}. The grammar is extended as:
+definition of an ADT which might be constructed with zero, one, or two
+propositions of type $A$, using its syntax: \mintinline{haskell}{data Container
+= None 1 | One A | Two (A * A)}. The language grammar is extended as:
 \[
@@
 
-The semantics of ADTs relate to those of the plus ($\oplus$) type
--- both are additive disjunctions.  To construct a value of an ADT we
-must use one of its constructors, similar to the way $\oplus$ requires
-only proof of either the left or right type it consists
-of. Analogously, we can deconstruct a value of an ADT via pattern
-matching on its constructors, where all branches of the pattern match
-must have the same type -- akin to the left rule for the $\oplus$
-connective. In effect, the inference rules for a simple ADT are a
-generalized form of the $\oplus$ rules.  Therefore, there's one left
-rule for ADTs, and an arbitrary number of right rules, one for each
+The semantics of the ADTs relate to those of the plus ($\oplus$) type --
+both are additive disjunctions.  A proof of an ADT requires only a proof of one
+of the types that a constructor takes, similar to the way $\oplus$ requires only
+proof of either the left or right type it consists of. Analogously, a proof of
+$C$ by deconstructing an ADT stipulates that all types that can construct the
+ADT can prove $C$, akin to the left rule for the $\oplus$ connective. In effect,
+the inference rules for a simple ADT mimic the $\oplus$ rules, for an
+arbitrary ($>1$) number of types, rather than exactly two. Therefore, there's one
+left rule for ADTs, and an arbitrary number of right rules, one for each
 constructor:
@@
 
-A more general formulation of ADTs says an ADT can be recursive (or
-"inductively defined"), meaning constructors can take as arguments
-values of the type they are defining. This change has
-a significant impact in the synthesis process. Take, for instance, the ADT
-defined as \mintinline{haskell}{data T = C1 T}, the synthesis goal
-$T \lolli C$, and its derivation:
+A more general (and interesting) formulation of ADTs says an ADT can be
+recursive (or "indutively defined"), meaning constructors can take as arguments
+values of the type they are defining. Seemingly small, this change has a big
+impact in the synthesis process. Take, for instance, the ADT defined as
+\mintinline{haskell}{data T = C1 T}, the synth goal $T \lolli C$, and its
+derivation:
 %
@@
 %
-Using our current system, we are to apply an infinite number of times
-the left ADT rule, never closing the proof. Symmetrically, the
-derivation for the goal $T$ is also infinite.
+Using our current system, we are to apply an infinite number of times the left
+ADT rule, never reaching a final state for the proof. Symmetrically, the
+derivation for the synth goal $T$ is also infinite.
 %
@@
 To account for this situation, we impede the decomposition of an ADT in subsequent
-proofs of its branches, and, symmetrically, don't allow construction of an
+proofs of its branches, and, symetrically, don't allow construction of an
 ADT when trying to synthetise an argument for its constructor.
@@
 for constraints on construction and $\Rho_D$ for constraints on
-deconstruction. Together, they hold a list of ADTs that cannot be constructed or
-deconstructed at a given point in the proof. For convenience, they are represented by a single $\Rho$ if
-unused. All non-ADT rules trivially propagate these. The ADT rules
-are then extended as follows, where $\Rho'_C = \Rho_C,T$ if $T$ is
-recursive and $\Rho'_C = \Rho_C$ otherwise (similarly for $\Rho'_D$):
+deconstruction. Together, they hold a list of ADTs that can not be constructed or
+deconstructed at some point in the proof. For convenience, they are represented by a single $\Rho$ if
+unused. All rules must trivially propagate these. The ADT rules are explicitly
+modified to take this into consideration:
 %
@@
 \end{mathpar}
-% where
-% \begin{mathpar}
-%     \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
-%     T\ \textrm{\textbf{else}}\ \Rho_C
-
-%     \Rho'_D = \textrm{likewise}
-% \end{mathpar}
-
-These modifications prevent the infinite derivations in the scenarios
-described above. However, they also greatly limit the space of
-derivable programs, leaving the synthesizer effectively unable to
-synthesize from specifications with recursive types. To prevent this,
-so we add three rules to complement the restrictions on construction
-and destruction of recursive types.
-%
-First, after decomposing a recursive ADT, we must consume
-\todo{consume? deconstruct?} all constructor parameters, and since we
-can't decompose the self type any further because of a restriction, we
-should be able to use it from the linear context. The situation is
-slightly more involved when considering unrestricted types (i.e.~types
-of the form $\bang A$), which cannot be treated in this way since they
-weren't restricted, and because it would result in additional
-non-determinism and infinite recursion.\todo{check} 
-%
-Second, without any additional rules, an ADT in the linear context
-will loop back to the inversion context, looping back and forth
-between the two contexts; instead, when focusing on an ADT, we should
-either instantiate the goal (provided they're the same type), or switch
-to inversion if and only if its decomposition isn't restricted. The
-three following rules ensure this:
+where
+\begin{mathpar}
+    \Rho'_C = \textrm{\textbf{if}}\ T\ \textrm{is recursive \textbf{then}}\ \Rho_C,
+    T\ \textrm{\textbf{else}}\ \Rho_C
+
+    \Rho'_D = \textrm{likewise}
+\end{mathpar}
+%
+These modifications prevent the infinite derivations in the scenarios described
+above, however, they also greatly limit the space of derivable programs, leaving
+the synthetiser effectively unable to synthetise from specifications with recursive
+types, which is undesirable, so we add three rules to complement the
+restrictions on construction and destruction of recursive types.
+
+First, after decomposing a recursive ADT, we must consume all constructor
+parameters, and since we can't decompose the self type any further because of a
+restriction, we should be able to use it from the linear context; even though,
+non-restricted types are not to take that course since they weren't unnaturally
+restricted, and because it would result in additional non-determinism and
+infinite recursion. Second, without any additional rules, an ADT in the linear
+context will loop back to the inversion context, and then loop again, going back and
+forth between the two contexts for infinity; instead, when focused on an ADT, we should
+either instance the goal (provided they're the same type), or switch to
+inversion if and only if its decomposition isn't restricted. The three
+following rules ensure this:
 \begin{mathpar}
@@
 %
-The rules above ensure that a recursive ADT will be deconstructed
-once, and move the recursive argument from its constructor to the
-linear context -- essentially forcing it to be used as a function
-argument, which will typically be the recursive call.
-
-\mypara{Recursion} The main idea behind synthesis of recursive
-programs is the labeling of the main goal and the addition of its
-type, under that name, to the unrestricted context. That is, to
-synthesize a recursive function of type $A \lolli B$ named \emph{f},
-the initial judgment can be written as
+Finally, these rules end up ensuring that a recursive ADT will deconstruct
+itself once, and move the recursive argument from its constructor to the linear
+context -- essentially forcing it to be used as a function argument, and most
+usually that function will be the synthesis recursive call.
+
+\mypara{Recursion} The main idea behind synthesis of recursive programs is the
+labeling of the main goal and the addition of its type, under that name, to the
+unrestricted context. That is, to synthetise a function of type $A \lolli B$ named
+\emph{f}, the initial judgment can be written as
 \begin{mathpar}
@@
 unrestricted context, but to the linear one as well.
-%
-However, we must restrict immediate uses of the recursive call since
-otherwise every goal would have a trivial proof (a non-terminating
-function that just calls itself), shadowing relevant solutions.
-Instead, our framework allows the use of recursion only after having
-deconstructed a recursive ADT via the following invariant: the
-recursive hypothesis can only be used in \emph{recursive branches of
-  ADT deconstruction}, i.e. the recursive call should only take
-``smaller'', recursive, hypothesis as arguments. To illustrate, in any
-recursive function with a list argument (whose type is defined as
-\mintinline{haskell}{data List = Nil | Cons (A * List)}), recursive
-calls are only allowed when considering a judgment of the form
-$\textrm {List} \vdash C$, i.e.~when a list value is available to
-produce the goal $C$, and only in the \emph{Cons}
-branch. % (TODO: rewrite? O bold é meio estranho não?)
-
-\mypara{Polymorphic Types} A polymorphic specification is a type of
-form $\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a
-set of variables that stand for any (non-polymorphic) type in
-$\tau$. Such a type is also called a \emph{scheme}.
-Synthesis of a scheme comprises of turning it into a non-quantified
-type, and then treating its type variables uniformly.  First,
-type variables are considered \emph{atomic types}, then, we instantiate
-the bound variables of the scheme as described by the Hindley-Milner
-inference method's~\cite{HM-infer} instantiation rule (put simply,
-generate fresh names for each bound type variable); e.g. the scheme
-$\forall \alpha.\ \alpha \lolli \alpha$ could be instantiated to
-$\alpha0 \lolli \alpha0$. We add a rule for this:
+
+Having said that, we won't allow the use of the recursive call in $\Gamma$ right
+away. If we did, every synth goal would have a trivial proof (a non-terminating
+function that just calls itself) that could even shadow other valid, relevant,
+solutions.  Instead, our framework allows the use of recursion only after having
+deconstructed a recursive ADT. To avoid complicating the rules, we'll add the
+invariant: the recursive hypothesis can only be \textbf{used in recursive branches of
+ADT deconstruction}, i.e. the recursive call should only take "smaller", recursive,
+hypothesis as arguments. To illustrate, having \mintinline{haskell}{data List =
+Nil | Cons (A * List)}, recursion is only
+"available" when deconstructing ($\textrm {List} \vdash C$), and only in
+the \emph{Cons} branch. % (TODO: rewrite? O bold é meio estranho não?)
+
+\mypara{Polymorphic Types} A polymorphic specification is a type of form
+$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of bound
+variables that quantify universally the type variables in $\tau$, which is also
+called a \emph{scheme}.  Synthesis of a scheme comprises of turning it into a
+non-quantified type, and then treating its type variables agnostically.  First,
+\textbf{type variables are considered atomic types}, then, instantiate the bound
+variables of the scheme as described by the Hindley-Milner inference
+method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names
+for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
+\alpha$ could be instantiated to $\alpha0 \lolli \alpha0$. We add a rule for
+this:
 \begin{mathpar}
@@
 %
-As such, the construction of a derivation in which the only rule that
-can derive an atom is the \textsc{init} rule corresponds to the
-synthesis of a program where some expressions are treated agnostically
--- nothing constrains their type -- i.e.~the program is
-polymorphic. The simplest example is the polymorphic function
-\emph{id} of type $\forall \alpha .\ \alpha \lolli \alpha$. The
-program synthesized from that specification is $\lambda x . x$, a
-lambda abstraction that does not constrain the type of its parameter
-$x$ in any way.
+As such, the construction of a derivation in which, to some types, no specific
+rules can be applied (except for the \textsc{init} rule, a type variable can
+instance a goal that's a type variable of the same name), corresponds to the
+synthesis of a program where some expressions are treated agnostically --
+nothing constrains their type -- meaning the program is polymorphic. The
+simplest example is the polymorphic function \emph{id} of type $\forall \alpha
+.\ \alpha \lolli \alpha$. The
+program synthetised from that specification is $\lambda x . x$, a lambda
+abstraction that doesn't constrain its parameter $x$ in any way.
 % TODO: \todo{R : rewrite this? não estranho?}
 
-The main challenge of polymorphism in synthesis is the usage of
+The interesting challenge of polymorphism in synthesis is the usage of
 schemes from the unrestricted context.  To begin with, $\Gamma$ now
-holds both (monomorphic) types and schemes. Consequently, after the
-rule \textsc{decideLeft!} is applied, we are left-focused on either a
-type or a scheme. Since left focus on a type is already well defined,
-we need only specify how to focus on a scheme.
-%
-Our algorithm instantiates bound type variables of the focused scheme
-with fresh \emph{existential} type variables, and the instantiated type becomes the
-left focus. Inspired by the Hindley-Milner system, we also generate
+holds both types and schemes. Consequently, after the rule
+\textsc{decideLeft!} is applied, we become left focused on either a
+type or a scheme. Since left focus on a type is already well
+formulated, only rules for left focus on a scheme are needed. Our
+algorithm instances bound type variables of the focused scheme with
+existential type variables, and the instanced type becomes the left
+focus. Inspired by the Hindley-Milner system, we also generate
 inference constraints on the existential type variables (postponing
@@
 across derivation branches the same way the linear context is (by
-having an input and output context ($\Theta/\Theta'$)).  In contrast to
-Hindley-Milner's inference, everytime a constraint is added it is
-solved against all other constraints -- a branch of the proof search
-is desired to fail as soon as possible.
+havin an input and output context ($\Theta/\Theta'$)).  In contrast to
+Hindley-Milner's inference, everytime a constraint is added it is solved against all other
+constraints -- a branch of the proof search is desired to fail as soon
+as possible.
 %TODO: Como fazer? \todo{O que
@@
 % propriamente o problema de misturar variaveis existenciais com universais}
-Note that we instantiate the scheme with \emph{existential} type
-variables ($?\alpha$) rather than just type variables ($\alpha$) since
-the latter represent universal types during synthesis, and the former
-represent a concrete instance of a scheme, that might induce
-constraints on other type variables. Additionally, we require that
-all existential type variables are assigned a type. These concepts are
-formalized with the rules:
+Note that we
+instance the scheme with \emph{existential} type variables ($?\alpha$)
+rather than just type variables ($\alpha$) since the latter represent universal
+types in the synth, and the first represent an existential instance of
+a scheme, that might inclusively constrain traditional type
+variables. Additionally, it's required that all existential type
+variables are assigned a type. These concepts
+are formalized with the rules:
 \begin{mathpar}
@@
 
-\mypara{Further Challenges} We now consider two more sources of infinite
+\mypara{More Restrictions} We can now look at two more sources of infinite
 recursion in the synthesis process. The first is the use of an unrestricted
 function to synthetise a term of type $\tau$ that in turn will require a term of
-the same type $\tau$. An example is the sub-goal judgment $\Gamma,(a \lolli
+the same type $\tau$. A good example is the sub-goal judgment $\Gamma,(a \lolli
 b \lolli b); (a \lolli b \lolli b) \Downarrow\ \vdash b$ that
-appears while synthesizing \emph{foldr} -- we apply ($\lolli$L)
+appears while synthetising \emph{foldr} -- we apply ($\lolli$L)
 until we can use \textsc{init} ($b \Downarrow\ \vdash b$), and then we must
-synthesize an argument of type $b$. Without any additional restrictions, we
-may become again left focused on $(a \lolli b \lolli b)$, and again require $b$,
+synthetise an argument of type $b$. Without any additional restrictions, we
+can end up left focused on $(a \lolli b \lolli b)$ again, and again require $b$,
 and on and on. The solution will be to disallow the usage of the same function
@@
 
-\mypara{Refinement Types}
+\mypara{Refinement Types} Their syntax is... Their rules are...
+
+\begin{mathpar}
+    \infer*[right=(refR)]
+    {  }
+    {  }
+\end{mathpar}
 
@@
 
-\mypara{General} The complete implementation of the synthetiser and environment
+The complete implementation of the synthetiser and environment
 pipeline was done in Haskell...
@@
 backtrack to the decision point and attempt to create a proof taking a different
-route. To this effect, our prototype synthetiser depends on \emph{logict}, "a backtracking logic-programming
-monad"~\cite{logict}.
+route. To this effect, our prototype synthetiser depends on \emph{logict}, "a
+backtracking logic-programming monad"~\cite{logict}. Special attention is
+required for ADTs. When deconstructing an ADT, since all branches must be
+synthetisable and result in the same ouput linear context $\Delta'$, alternative
+constructor deconstructions should be considered fairly in conjunction, so the
+\textbf{fair conjunction} operator should be used. When constructing an ADT, the
+disjunctive choice of constructor should be fair, and so the \textbf{fair
+disjunction} operator used. Fairness at these two points usually increases
+the synthesis speed slightly, but can be a relevant factor in synthesis in which a
+combination of choices following the ADT deconstruction or construction can
+backtrack many times or even arbitrarily.
 

### 2021-07-20 STLLC/Simple.hs

@@ -1,0 +1,4 @@
+
+data List a = Nil | Cons (a * List a);
+
+synth foldr :: ((a -o b -o b) -> b -o List a -o b);

### 2021-07-20 STLLC/TreeG.hs

@@ -1,0 +1,15 @@
+data List a = Nil | Cons (a * List a);
+
+data Tree a = Node (a * List (Tree a));
+
+#synth map :: ((a -o b) -> List a -o List b);
+
+#unfoldTree f b = let !fx = f in
+#    let a*bs = fx b in Node (a, (map (unfoldTree fx) bs));
+
+#synth unfoldTree :: ((b -o (a * List b)) -> b -o Tree a) | using (map unfoldTree) | depth 5;
+
+synth insert :: (a -o Tree a -o Tree a);
+
+synth return :: (a -o Tree a);
+

### 2021-07-20 STLLC/app/Synth.hs

@@ -452,7 +452,6 @@ focus c goal =
                 -- (et, etvars) <- existencialInstantiate sch                                       -- tipo com existenciais
                 (se, d', cs') <- memoizefocus' focus' (Just (n, et)) ctxt goal                      -- fail ou success c restrições -- sempre que é adicionada uma constraint é feita a unificação
                                                                                                     -- resolve ou falha -- por conflito ou falta informação
-                -- trace ("DID I INSTANCE ALL VARIABLES " ++ show (apply cs' et) ++ "\nftv:" ++ show (ftv $ apply cs' et) ++ "\nSet etvars:" ++ show (Set.fromList etvars) ++ "? " ++ show (Set.disjoint (Set.fromList etvars) (ftv $ apply cs' et))) $ do
                                                                                                     -- por conflicto durante o processo
                 guard (Set.disjoint (Set.fromList etvars) (ftv $ apply cs' et))                     -- por falta de informação (não pode haver variáveis existenciais bound que fiquem por instanciar, i.e. não pode haver bound vars nas ftvs do tipo substituido)
                                                                                                     -- after making sure no instantiated variables escaped, the constraints added to the substitution can be forgotten so that it doesn't complicate further scheme computations

### 2021-07-20 STLLC/bench/list7.hs

@@ -1,4 +1,4 @@
 data List a = Nil | Cons (a * List a);
 
 # infinite loop
-#synth concat :: (List (List a) -o List a);
+synth concat :: (List (List a) -o List a);

### 2021-07-20 STLLC/tests/support/Array.hs

@@ -5,6 +5,8 @@ write :: (MArray a -o (Int * a) -> MArray a);
 
 freeze :: (MArray a -o !(Array a));
 
-foldl :: ((a -o b -o a) -> a -o (List b) -o a)
+foldl :: ((a -o b -o a) -> a -o (List b) -o a);
+
+#read :: (MArray a -o Int -> (MArray a * !a));
 
 synth array :: (Int -> List (!(Int * a)) -> Array a) | using (foldl) | depth 3;

### 2021-07-20 STLLC/tests/support/State.hs

@@ -1,0 +1,16 @@
+data State b a = State (!b -o (a * !b));
+
+synth runState :: (State b a -o (!b -o (a * !b)));
+
+synth bind :: (State c a -o (a -o State c b) -o State c b);
+
+synth return :: (a -o State b a);
+
+synth get :: (State a a);
+
+synth put :: (!a -o (State a 1));
+
+synth modify :: ((!a -o !a) -o State a 1);
+
+synth evalState :: (State b a -o !b -o a);
+

### 2021-07-20 article/article.tex

--- 2021-07-20 article/article.tex
@@
 
-\mypara{General} The implementation was carried out in Haskell. We
+The implementation was carried out in Haskell. We
 highlight some key points.
@@
 our prototype synthetiser depends on \emph{LogicT}, "a backtracking
-logic-programming monad"~\cite{logict}.
+logic-programming monad"~\cite{logict}. Special attention is
+required for ADTs. When deconstructing an ADT, since all branches must be
+synthetisable and result in the same ouput linear context $\Delta'$, alternative
+constructor deconstructions should be considered fairly in conjunction, so the
+\textbf{fair conjunction} operator should be used. When constructing an ADT, the
+disjunctive choice of constructor should be fair, and so the \textbf{fair
+disjunction} operator used. Fairness at these two points usually increases
+the synthesis speed slightly, but can be a relevant factor in synthesis in which a
+combination of choices following the ADT deconstruction or construction can
+backtrack many times or even arbitrarily.
 
@@
 even after $15m$ running.
+>>>>>>> 4290cfe31b8de10ca70f772fbc7d5fdc9542273c
 %

### 2021-07-22 article/article.tex
Add focusing judgments, explanations, do todos

@@
 
-% General motivation?
 Program synthesis is an automated or semi-automated process of deriving a
@@
 %
-% However, they can also be a burden -- the whole point of these type systems is
-% to ensure that \emph{fewer} programs are deemed well-typed, which can pose
-% additional challenges to the development process.
-%
 For instance, the type $\mathsf{Int} \rightarrow \mathsf{Int} \rightarrow
@@
 form of linear types.
-%
-% TODO: Deixar isto para mais à frente?  More concretely, we will explore the
-% proof-theoretic foundations of linear types via the Curry-Howard
-% correspondence with linear logic propositions, viewing program synthesis as
-% proof search in linear logic.  (Said correspondence maps \emph{types} to
-% \emph{propositions in a logic}, and \emph{programs} to \emph{proofs}).
-%
 Linear types constrain resource
@@
 implementation challenges (\S~\ref{sec:architecture}). Finally, 
+\todo{como formular a frase sobre o apêndice?}
 we evaluate our work through expressiveness and performance benchmarks
-(\S~\ref{sec:evaluation}) and discuss related work (\S~\ref{sec:related}).
-%TODO: And appendice with...
+(\S~\ref{sec:evaluation}) and discuss related work
+(\S~\ref{sec:related}).
 
@@
 \paragraph{Core linear calculus.} Initially, we synthesize from
-specifications that only use literals and canonical linear types ($\lolli, \with, \tensor, 1,
-\oplus, \bang$), such as linear logic theorems. Namely,
+specifications that only use literals and propositional linear types, such as linear logic theorems. Namely,
 the specification $(A \lolli B \lolli C) \lolli (A \tensor B) \lolli C$ produces
@@
 
-%TODO: Polymorphism ? read todo below
-
-\todo[inline]{Tens de explicar o que significa o $\bang$, see above.}
-\paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring some types to be
-used linearly (consumed), we can assure, e.g., the deconstruction of an
-ADT. As such, the specification $\bang(a \lolli b) \lolli List\ a \lolli
-List\ b$ isn't satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}.
-To synthetise a valid program from it, \mintinline{haskell}{(List a)} has to be
-deconstructed, and its constructor arguments used. \synname\ would output the following
+\paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring
+some types to be used linearly (consumed), we can assure, e.g., the
+deconstruction of an ADT. As such, the specification that identifies a function
+that takes as arguments an unrestricted ($\bang$) linear function\footnote{might be used non-linearly, but its parameter is used
+linearly in its body} 
+($\lolli$) that converts $a$s in $b$s, a list of $a$s, and produces a
+list of $b$s, written $\bang(a \lolli b) \lolli List\ a \lolli List\ b$, isn't
+satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}. To synthetise
+a valid program from it, \mintinline{haskell}{(List a)} has to be deconstructed,
+and its constructor arguments used. \synname\ would output the following
 function (\emph{map}) without relying on any additional information:
+%
 % TODO: Aqui não estou bem a falar de polymorphism mas parece estranho não por
@@
 
-\todo[inline]{Tens de dizer o que significa o refinamento concreto e o
-que e a funcao mult e add}
 \paragraph{Refinements.} Refinements are logical predicates
@@
 As an example, from the specification $x \{Int\} \lolli y \{Int\} \lolli z
-\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$, the
-program\ \mintinline{haskell}{(fn a b c = add (mult (sub 0 1) a) (mult c b))} is
+\{Int\} \lolli k \{Int\ \vert\ x + k = y * z\}$ (a function that takes three
+integers and produces an integer that satisfies the given predicate), the
+program\ \mintinline{haskell}{(\a -> \b -> \c -> ((0 - 1) * a) + (c * b))} is
 synthetised.
@@
 \paragraph{Guiding the Search.} Specifications can be additionally
-augmented with three keywords: ``using'', ``depth'' and
-``assert''. Through them, we can fine-tune the synthesis process and,
+augmented with four keywords: ``using'', ``depth'', ``assert'' and ``choose''.
+Through them, we can fine-tune the synthesis process and,
 respectively, force certain functions to be present in the synthesis
-outcome, allow a \emph{deeper} search in the valid programs space, and
+outcome, allow a \emph{deeper} search in the valid programs space,
 require given predicates to hold true in the program after the
-synthesis is complete.
+synthesis is complete, and select between alternative results.
 
@@
 \begin{minted}{haskell}
-foldl :: ((a -o b -o a) -> a -o (List b) -o a);
-newMArray :: (Int -> (MArray a -o !b) -o b);
-write :: (MArray a -o (Int * a) -> MArray a);
-freeze :: (MArray a -o !(Array a));
-synth array :: (Int -> List (!(Int * a)) -> Array a)
+foldl :: (a -o b -o a) -> a -o (List b) -o a;
+newMArray :: Int -> (MArray a -o !b) -o b;
+write :: MArray a -o (Int * a) -> MArray a;
+freeze :: MArray a -o !(Array a);
+synth array :: Int -> List (!(Int * a)) -> Array a
     | using (foldl) | depth 3;
@@
 
-\todo[inline]{Indicar o judgment ``declarativo'' e depois os
-  judgments em focusing}
-
 \mypara{Core Rules} The core system comprises of proof search for
-(intuitionistic) linear logic sequent calculus proofs, based on a
-system of resource management and focusing.
-%
-In intuitionistic sequent calculi, each connective has a so-called
-\emph{left} and a \emph{right} rule, which effectively define how to
-decompose an ambient assumption of a given proposition and how to
-prove a certain proposition is true, respectively. 
-Andreoli's \emph{focusing} for linear logic~\cite{Andreolli focusing}
-is a technique to remove non-essential nondeterminism from
-proof-search by structuring the application of so-called invertible
-and non-invertible inference rules. Andreoli observed that the
+(intuitionistic) linear logic sequent calculus proofs, based on a system of
+resource management~\cite{resman} and focusing.
+%
+% TODO: Onde falamos de resource management? ( 
+%
+In intuitionistic sequent calculi, each connective has a so-called \emph{left}
+and a \emph{right} rule, which effectively define how to decompose an ambient
+assumption of a given proposition and how to prove a certain proposition is
+true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
+focusing} is a technique to remove non-essential nondeterminism from
+proof-search by structuring the application of so-called invertible and
+non-invertible inference rules. Andreoli observed that the
 connectives of linear logic can be divided into two categories, dubbed
-synchronous and asynchronous. Synchronous connectives are those whose
+synchronous and asynchronous. Asynchronous connectives are those whose
 right rules are \emph{invertible}, i.e. they can be applied eagerly
@@
 which these rules are applied is irrelevant, and whose left rules
-are not invertible. Asynchronous connectives are dual. \todo{Say which
+are not invertible. Synchronous connectives are dual. \todo{Say which
   are which}
-%
+
 Given this separation, focusing divides proof search into two
@@
 until we reach an asynchronous (i.e. invertible proposition), the
-proof is complete or no rules are applicable, in which the proof must
+proof is complete, or no rules are applicable, in which case the proof must
 \emph{backtrack} to the state at which the focusing phase began.
-
-\todo[inline]{Isto so faz sentido olhando para os judgments}
-For the invertible rules, a
-third, ordered, context ($\Omega$) is needed to hold propositions
-resulting from decomposing connectives; propositions we'll later try
-to invert, moving them to the linear context ($\Delta$) when we fail
-to.
+As such, with focusing, a linear logic judgment in sequent calculus ($\Gamma; \Delta, B
+\vdash A$) is broken down into four different judgments, grouped into the two
+phases ($\Uparrow, \Downarrow$). For the invertible phase, a third % TODO: ordered??,
+context ($\Omega$) is needed to hold propositions that result from
+decomposing connectives (propositions we'll later try to invert, moving them to
+the linear context ($\Delta$) when we fail to). The judgments of right
+invertible rules are written $\Gamma; \Delta; \Omega,B \vdash A \Uparrow$, and
+those of left invertible rules are $\Gamma; \Delta; \Omega,B \Uparrow\ \vdash A$.
+In the focusing phase, the focus proposition will be either the goal or come
+from $\Gamma$ or $\Delta$, and appliable rules will be classified as right
+non-invertible, or left non-invertible, respectively. Right non-invertible
+rules have the form $\Gamma;\Delta,B \vdash A \Downarrow$, and left
+non-invertible rules are written $\Gamma;\Delta; B \Downarrow\ \vdash A$.
 
@@
 
-\todo[inline]{A frente de cada ``descricao'', escreve o tipo pois
-  ajuda o leitor}
 The core language is a simply-typed linear lambda calculus with linear
-functions, additive and multiplicative pairs, multiplicative unit, additive sums
-and the exponential modality (to internalize unrestricted use of variables), for
+functions ($\lolli$), additive ($\with$) and multiplicative ($\tensor$) pairs,
+multiplicative unit (\textbf{1}), additive sums ($\oplus$)
+and the exponential modality ($\bang$) (to internalize unrestricted use of variables), for
 which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
@@
 Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
-management, and focusing, we get the following core formal system (heavily inspired
+management, and focusing, we get the following core formal system (%heavily
+inspired
 by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
-first rule \todo{qual ``first''? referencia por nome} is read: to synthesize a program of type $A \lolli B$ while inverting
+rule ($\lolli R$) is read: to synthesize a program of type $A \lolli B$ while inverting
 right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
@@
 are then extended as follows, where $\Rho'_C = \Rho_C,T$ if $T$ is
-recursive and $\Rho'_C = \Rho_C$ otherwise (similarly for $\Rho'_D$):
-%
+recursive and $\Rho'_C = \Rho_C$ otherwise ($\Rho'_D$ is dual):
+%
+%TODO: ... we can actually instance ADTs that take no arguments even if they are restricted
 \begin{mathpar}
@@
 synthesize from specifications with recursive types. To prevent this,
-so we add three rules to complement the restrictions on construction
+we add three rules to complement the restrictions on construction
 and destruction of recursive types.
 %
-First, after decomposing a recursive ADT, we must consume
-\todo{consume? deconstruct?} all constructor parameters, and since we
-can't decompose the self type any further because of a restriction, we
-should be able to use it from the linear context. The situation is
-slightly more involved when considering unrestricted types (i.e.~types
-of the form $\bang A$), which cannot be treated in this way since they
-weren't restricted, and because it would result in additional
-non-determinism and infinite recursion.\todo{check} 
+First, since we can't deconstruct some ADTs any further because of a restriction,
+but must utilize all propositions linearly in some way, all propositions in $\Omega$ whose
+deconstruction is restricted are to be moved to the linear context $\Delta$.
 %
 Second, without any additional rules, an ADT in the linear context
-will loop back to the inversion context, looping back and forth
+will loop back to the inversion context, jumping back and forth
 between the two contexts; instead, when focusing on an ADT, we should
@@
 three following rules ensure this:
+
 \begin{mathpar}
@@
 The rules above ensure that a recursive ADT will be deconstructed
-once, and move the recursive argument from its constructor to the
-linear context -- essentially forcing it to be used as a function
-argument, which will typically be the recursive call.
+once, and that subsequent equal ADTs will only be useable from the linear
+context -- essentially forcing it to be used to instantiate another proposition,
+which will typically be an argument for the recursive call.
 
@@
 
-\mypara{Polymorphic Types} A polymorphic specification is a type of
-form $\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a
-set of variables that stand for any (non-polymorphic) type in
-$\tau$. Such a type is also called a \emph{scheme}.
-Synthesis of a scheme comprises of turning it into a non-quantified
-type, and then treating its type variables uniformly.  First,
-type variables are considered \emph{atomic types}, then, we instantiate
-the bound variables of the scheme as described by the Hindley-Milner
-inference method's~\cite{HM-infer} instantiation rule (put simply,
-generate fresh names for each bound type variable); e.g. the scheme
-$\forall \alpha.\ \alpha \lolli \alpha$ could be instantiated to
-$\alpha0 \lolli \alpha0$. We add a rule for this:
+\mypara{Polymorphic Types} A polymorphic specification is a type of form
+$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of
+variables that stand for any (non-polymorphic) type in $\tau$. Such a type is
+also called a \emph{scheme}.  Synthesis of a scheme comprises of turning it into
+a non-quantified type, and then treating its type variables uniformly.  First,
+type variables are considered \emph{atomic types}, then, we instantiate the
+bound variables of the scheme as described by the Hindley-Milner inference
+method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names
+for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
+\alpha$ could be instantiated to $\alpha0 \lolli \alpha0$. We add a rule for
+this, where $\forall \overline{\alpha}.\ \tau \sqsubseteq \tau'$ indicates type
+$\tau'$ is an \emph{instantiation} of type scheme $\forall \overline{\alpha}.\
+\tau$:
+%
 \begin{mathpar}
@@
 \end{mathpar}
-where
-\[
-    \forall \overline{\alpha}.\ \tau \sqsubseteq \tau'\ \textrm{indicates type}\
-    \tau'\ \textrm{is an \emph{instantiation} of type scheme}\ \forall
-    \overline{\alpha}.\ \tau
-\]
 %
@@
 
-The main challenge of polymorphism in synthesis is the usage of
-schemes from the unrestricted context.  To begin with, $\Gamma$ now
-holds both (monomorphic) types and schemes. Consequently, after the
-rule \textsc{decideLeft!} is applied, we are left-focused on either a
-type or a scheme. Since left focus on a type is already well defined,
-we need only specify how to focus on a scheme.
-%
-Our algorithm instantiates bound type variables of the focused scheme
-with fresh \emph{existential} type variables, and the instantiated type becomes the
-left focus. Inspired by the Hindley-Milner system, we also generate
-inference constraints on the existential type variables (postponing
-the decision of what type it should be to be used in the proof), and
-collect them in a new constraints context $\Theta$ that is propagated
-across derivation branches the same way the linear context is (by
-having an input and output context ($\Theta/\Theta'$)).  In contrast to
-Hindley-Milner's inference, everytime a constraint is added it is
-solved against all other constraints -- a branch of the proof search
-is desired to fail as soon as possible.
-%TODO: Como fazer? \todo{O que
-% significa $?\alpha$ ? Não foi explicado -- no Hindley-Milner nao ha
-% propriamente o problema de misturar variaveis existenciais com universais}
-Note that we instantiate the scheme with \emph{existential} type
-variables ($?\alpha$) rather than just type variables ($\alpha$) since
-the latter represent universal types during synthesis, and the former
-represent a concrete instance of a scheme, that might induce
-constraints on other type variables. Additionally, we require that
-all existential type variables are assigned a type. These concepts are
-formalized with the rules:
+The main challenge of polymorphism in synthesis is the usage of schemes from the
+unrestricted context.  To begin with, $\Gamma$ now holds both (monomorphic)
+types and schemes. Consequently, after the rule \textsc{decideLeft!} is applied,
+we are left-focused on either a type or a scheme. Since left focus on a type is
+already well defined, we need only specify how to focus on a scheme.
+%
+Our algorithm instantiates bound type variables of the focused scheme with fresh
+\emph{existential} type variables, and the instantiated type becomes the left
+focus. Inspired by the Hindley-Milner system, we also generate inference
+constraints on the existential type variables (postponing the decision of what
+type it should be to be used in the proof), and collect them in a new
+constraints context $\Theta$ that is propagated across derivation branches the
+same way the linear context is (by having an input and output context
+($\Theta/\Theta'$)).  In contrast to Hindley-Milner's inference, everytime a
+constraint is added it is solved against all other constraints -- a branch of
+the proof search is desired to fail as soon as possible.
+%TODO: Como fazer? \todo{O que significa $?\alpha$ ? Não foi explicado -- no
+%Hindley-Milner nao ha propriamente o problema de misturar variaveis
+%existenciais com universais}
+Note that we instantiate the scheme with \emph{existential} type variables
+($?\alpha$) rather than just type variables ($\alpha$) since the latter
+represent universal types during synthesis, and the former represent a concrete
+instance of a scheme, that might induce constraints on other type variables.
+Additionally, we require that all existential type variables are assigned a
+type. These concepts are formalized with the following rules, where $\forall
+\overline{\alpha}.\ \tau \sqsubseteq_E \tau'$ means type $\tau'$
+is an \emph{existential instantiation} of scheme $\forall \overline{\alpha}.\ \tau$,
+$\textrm{ftv}_E(\tau')$ is the set of free \emph{existential} type variables in
+type $\tau'$, $?\alpha \mapsto \tau_x$ is a mapping from \emph{existential} type
+$?\alpha$ to type $\tau_x$, and $\textsc{unify}(c, \Theta)$ indicates wether
+constrain $c$ can be unified with those in $\Theta$:
+
 \begin{mathpar}
@@
     x{:}A \Downarrow\ \vdash x : ?\alpha}
-    \\
-    \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'\ \textrm{means type}\
-    \tau'\ \textrm{is an \emph{existential instantiation} of scheme}\ \forall
-    \overline{\alpha}.\ \tau
-    \and
-    \textrm{ftv}_E(\tau')\ \textrm{is the set of free \emph{existential} type variables in type}\ \tau'
-    \and
-    ?\alpha \mapsto \tau_x\ \textrm{is a mapping from \emph{existential} type}\ ?\alpha\
-    \textrm{to type}\ \tau_x
-    \\
-    \textsc{unify}(c, \Theta)\ \textrm{indicates wether constraint}\ c\
-    \textrm{can be unified with those in}\ \Theta
 \end{mathpar}
@@
 function to synthetise a term of type $\tau$ that in turn will require a term of
-the same type $\tau$. An example is the sub-goal judgment $\Gamma,(a \lolli
+the same type $\tau$. An example is the sub-goal judgment $(a \lolli
 b \lolli b); (a \lolli b \lolli b) \Downarrow\ \vdash b$ that
@@
 existential type. In contrast to the previous problem, the type of the goal and
-of the argument that will cause the loop won't be the same, since instantiated
-bound variables are always fresh. For example, for $\Gamma,\forall \alpha,\beta
-.\ \alpha\lolli\beta\lolli\beta;?\alpha\lolli?\beta\lolli?\beta \Downarrow\
-\vdash\ ?\sigma$, we'll unify $?\beta$ with $?\sigma$, and then require a term
-of type $?\beta$ (not $?\sigma$). We want to forbid the usage of the \emph{same}
+of the argument that will cause the loop won't match exactly, since instantiated
+bound variables are always fresh. For example, for $\forall \alpha,\beta .\
+\alpha\lolli\beta\lolli\beta;?\alpha\lolli?\beta\lolli?\beta \Downarrow\ \vdash\
+?\sigma$, we'll unify $?\beta$ with $?\sigma$, and then require a term of type
+$?\beta$ (not $?\sigma$). We want to forbid the usage of the \emph{same}
 function to attain \emph{any} existential goal, provided that function might
@@
 context-heavy problems such as \emph{array} (seen in Section~\ref{sec:overview})
-wouldn't terminate in a reasonable amount of time.  In fact, with $n$
-polymorphic functions in the unrestricted context, the complexity of searching for a
-program of any existential type $?\alpha$, while restricting solely the used
-function, is $O(n!)$~\ref{exemplo_apendice?}. As such, we'll instead forbid the
-usage of \emph{all} unrestricted polymorphic functions to synthetise any
-existential type if, altogether, we've already used $d_e$ times a polymorphic
-function for \emph{any} existential sub-goal, where $d_e$ is a \emph{constant
-existential depth} that controls the maximum depth of ``existential synthetis''.
-This approach reduces the complexity of our proof-search algorithm for
-existential types to $O(\tfrac{n!}{d_e!}) = O(n^{d_e})$.
+wouldn't terminate in a reasonable amount of time.
+%
+% In fact, with $n$ polymorphic functions in the unrestricted context, the
+% complexity of searching for a program of any existential type $?\alpha$, while
+% restricting solely the used function, is $O(n!)$~\ref{exemplo_apendice?}.
+%
+As such, we'll instead forbid the usage of \emph{all} unrestricted polymorphic
+functions to synthetise any existential type if, altogether, we've already used
+$d_e$ times a polymorphic function for \emph{any} existential sub-goal, where
+$d_e$ is a \emph{constant existential depth} that controls the maximum depth of
+``existential synthetis''.
+%
+% This approach reduces the complexity of our proof-search algorithm for
+% existential types to $O(\tfrac{n!}{d_e!}) = O(n^{d_e})$.
 
@@
 even after $15m$ running.
->>>>>>> 4290cfe31b8de10ca70f772fbc7d5fdc9542273c
 %
@@
 
-\mypara{Type-and-Example-Directed Program Synthesis} This
-work~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
-explores a purely functional, recursive program synthesis technique
-based on types/proof search (as we intend to do), with examples as an
-auxiliar technique to trim down the program synthesis vast search
-space. A good use is
-made of the general strategy of using a type system to generate
-programs (by ``inverting'' the type system), rather than using it to
-type-check one.  Moreover, the paper uses a data structure called
-``refinement tree'' in conjunction with the extra information provided
-by the examples to allow for efficient synthesis of non-trivial
-programs. Overall it employs good engineering to achieve type-based
-synthesis of functional typed programs, with at least two additions
-that we might want to follow too (type refinements and recursive
-functions).
+% \mypara{Type-and-Example-Directed Program Synthesis} This
+% work~\cite{DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
+% explores a purely functional, recursive program synthesis technique
+% based on types/proof search (as we intend to do), with examples as an
+% auxiliar technique to trim down the program synthesis vast search
+% space. A good use is
+% made of the general strategy of using a type system to generate
+% programs (by ``inverting'' the type system), rather than using it to
+% type-check one.  Moreover, the paper uses a data structure called
+% ``refinement tree'' in conjunction with the extra information provided
+% by the examples to allow for efficient synthesis of non-trivial
+% programs. Overall it employs good engineering to achieve type-based
+% synthesis of functional typed programs, with at least two additions
+% that we might want to follow too (type refinements and recursive
+% functions).
 

### 2021-07-22 article/article.tex
explain -oL, do todos, tweaks

@@
 which these rules are applied is irrelevant, and whose left rules
-are not invertible. Synchronous connectives are dual. \todo{Say which
-  are which}
+are not invertible. Synchronous connectives are dual. The asynchronous \todo{Escrever isto depois
+de apresentar o linear $\lambda$-calculus?} 
+connectives are $\lolli$ and $\with$ and the synchronous are
+$\tensor,\textbf{1},\oplus,\bang$.
 
@@
 
-The core language is a simply-typed linear lambda calculus with linear
+The core language is a simply-typed linear $\lambda$-calculus with linear
 functions ($\lolli$), additive ($\with$) and multiplicative ($\tensor$) pairs,
@@
 by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
-rule ($\lolli R$) is read: to synthesize a program of type $A \lolli B$ while inverting
+rule $\lolli$R is read: to synthesize a program of type $A \lolli B$ while inverting
 right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
@@
 \begin{mathpar}
-    \infer*[right=($\Uparrow R$)]
+    \infer*[right=($\Uparrow$R)]
     {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
@@
 \begin{mathpar}
-    \infer*[right=($\Uparrow L$)]
+    \infer*[right=($\Uparrow$L)]
     {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
@@
 
-\todo[inline]{Aqui tens de explicar uma regra pelo menos}
-The decision rules are followed by either left or right focus rules:
+The decision rules are followed by either left or right focus rules. The left
+focus rule $\lolli$L is read ``to synthesize a program of type $C$ with input
+linear context $\Delta$ and while left focused on a proposition $x$ of type $A
+\lolli B$, synthesize a program $M$ of type $C$ with input $\Delta$ and 
+left focused ($\Downarrow$) on $y$ of type
+$B$, and output the context $\Delta'$ (with the unused
+linear propositions); then, switching to the right inversion judgment ($\Uparrow$), synthesize a program $N$ of type $A$ with input
+$\Delta'$ and output context $\Delta''$; finally, replace in $M$ all occurrences
+of variable $y$ with program $x\ N$ (apply $x$ to $N$)''. It can be understood
+as the synthesis of a program of type $C$ using the return value (of type $B$) from the function $A \lolli B$, and then the
+synthesis of an argument of type $A$ to apply the function to.
 \begin{mathpar}
@@
 \end{mathpar}
-
-where the ADT $T$ and its constructors stand for any ADT with form
-\begin{minted}{haskell}
+where the ADT $T$ and its constructors stand for any ADT with form \begin{minted}{haskell}
     data T = C1 X1 | C2 X2 | ... | Cn Xn
@@
 %
-The rules above ensure that a recursive ADT will be deconstructed
+Altogether, the rules above ensure that a recursive ADT will be deconstructed
 once, and that subsequent equal ADTs will only be useable from the linear
-context -- essentially forcing it to be used to instantiate another proposition,
+context -- essentially forcing them to be used to instantiate another proposition,
 which will typically be an argument for the recursive call.
@@
 %
-As such, the construction of a derivation in which the only rule that
-can derive an atom is the \textsc{init} rule corresponds to the
-synthesis of a program where some expressions are treated agnostically
--- nothing constrains their type -- i.e.~the program is
-polymorphic. The simplest example is the polymorphic function
-\emph{id} of type $\forall \alpha .\ \alpha \lolli \alpha$. The
-program synthesized from that specification is $\lambda x . x$, a
-lambda abstraction that does not constrain the type of its parameter
-$x$ in any way.
-% TODO: \todo{R : rewrite this? não estranho?}
+As such, the construction of a derivation in which the only rule that can derive
+an atom is the \textsc{init} rule corresponds to the synthesis of a program
+where some expressions are treated agnostically (nothing constrains their type),
+i.e.~a polymorphic program. The simplest example is the polymorphic function
+\emph{id} of type $\forall \alpha .\ \alpha \lolli \alpha$. The program
+synthesized from that specification is $\lambda x . x$, a lambda abstraction
+that does not constrain the type of its parameter $x$ in any way.
 
@@
 \[
-    \textrm{Parsing} \rightarrow \textrm{Desugaring} \rightarrow \textrm{Inference} \rightarrow \textrm{Synthesis}
+    \textrm{Parsing} \rightarrow \textrm{Desugaring} \rightarrow
+    \textrm{Inference} \rightarrow \textrm{Synthesis} \rightarrow
+    \textrm{Evaluation}
 \]
@@
 
-The synthesis module iterates over the synthesis marks and runs the
-synthetiser with its context and goal for each mark, minimizing the
-resulting expression (i.e.~$\eta$-reducing it), and replacing the mark
-in the frontend AST with the synthed program.
+The synthesis module iterates over the synthesis marks and runs the synthetiser
+with the respective context and goal for each mark. The resulting expression is checked
+against the synthesis constraints defined through some keywords, continuing
+synthesis if they are violated: ``using'' guarantees that certain functions are
+used in the program body, and ``assert'' evaluates the given assertion within the
+synthesized program to a validating boolean.  Additionally, the keyword
+``depth'' controls the \emph{existential depth} (seen
+in~\ref{sec:formal_system}), and ``choose'' selects a different result as long
+as one is available. Finally, the expression is minimized (i.e.~$\eta$-reduced),
+and the mark in the frontend AST is replaced with the synthesized program.
+
+The evaluation module evaluates the function named ``main'' within the
+complete synthesized program, returning a final value.
 
@@
 the left focus), and the goal, are equivalent. Because comparing all
-of these is slow, our implementation hashes the relevant values and
-uses the result as the key\todo{via a hashconsing library?}.
+of these is slow, our implementation hashes the relevant values via a hashing
+library (\emph{Hashable}~\cite{https://hackage.haskell.org/package/hashable}) and
+uses the result as the key.
 %

### 2021-07-22 article/article.tex
related work

@@
 
-Type-based program synthesis is a vast field of study and so it
-follows that a lot of literature is available to inspire and
-complement our work. Most works~\cite{DBLP:conf/lopstr/HughesO20,DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16} follow some variation of the
-synthesis-as-proof-search approach, however, the process is novel for
-each, due to a variety of different rich types explored, and their
-respective corresponding logics and programming languages; or nuances
-of the synthesis process itself, such as complementing types with
-program examples; or even the programming paradigm of the output
-produced (e.g. generating heap manipulating
-programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).  Some other
-common patterns we might want to follow are the use of type
-refinements~\cite{DBLP:conf/pldi/PolikarpovaKS16} and the support for synthesis of recursive
-functions~\cite{DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15}.  These works share some common ground, and base
-themselves of the same works we do -- most famously the Curry-Howard
-correspondence, but also, for example, works on focusing-based
-proof-search~\cite{10.1093/logcom/2.3.297}.
+Type-based program synthesis is a vast field of study and so it follows that a
+lot of literature is available to inspire and complement our work. Most
+works~\cite{DBLP:conf/lopstr/HughesO20,DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15,DBLP:conf/popl/FrankleOWZ16}
+follow some variation of the synthesis-as-proof-search approach. However, the
+process is novel for each due to a variety of different rich types explored and
+their corresponding logics and languages; or nuances of the synthesis process
+itself, such as complementing types with program examples; or even the
+programming paradigm of the output produced (e.g. generating heap manipulating
+programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).  Some other common
+patterns we follow are the use of type
+refinements~\cite{DBLP:conf/pldi/PolikarpovaKS16}, polymorphic types, and the
+support for synthesis of recursive
+functions~\cite{DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15}. These
+works share some common ground with each other and with us -- most
+famously the Curry-Howard correspondence, but also, for example, 
+focusing-based proof-search~\cite{10.1093/logcom/2.3.297}.
+% TODO: sinto que precisava de um toque...
 
@@
 \mypara{Program Synthesis from Polymorphic Refinement Types} The
-work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis
-recursive functional programs but in a more ``advanced'' setting. The
-specification is a combination of two rich forms of types: polymorphic and
-refinement types (which correspond to a first-order logic through the
-Curry-Howard isomorphism) -- this is an interesting combination that offers
-``expressive power and decidability, which enables automatic
-verification--and hence synthesis--of nontrivial programs''.  Their
-approach to refinement types consists of a new algorithm that supports
-decomposition of the refinement specification, allowing for separation
-between the language of specification and programs and making the
-approach amenable to compositional synthesis techniques.
+work~\cite{DBLP:conf/pldi/PolikarpovaKS16} also studies synthesis of recursive
+functional programs in an ``advanced'' context. Their specifications combine two
+rich forms of types: polymorphic and refinement types (which correspond to a
+first-order logic through the Curry-Howard isomorphism). Their approach to
+refinement types consists of a new algorithm that supports decomposition of the
+refinement specification, allowing for separation between the language of
+specification and programs and making the approach amenable to compositional
+synthesis techniques. We also support polymorphism and refinements, however, our
+refinements are somewhat underdeveloped and don't have an important role in our
+process, in contrast to their strong use of refinements to guide their
+synthesizer. Instead, our synthesizer leverages the expressiveness of linear
+types and techniques for proof-search in linear logic to guide its process.
 
@@
 approach very similar to our own.  The work employs so-called graded
-modal types, which is an refinement of pure linear types -- a more
+modal types, which is a refinement of pure linear types -- a more
 \emph{fine-grained} version, since it allows for quantitative
@@
 approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
--- the model we will be using.
+-- which, in contrast, is the model use in our work.
 
 They also use focusing as a solution to trim down search space and to
-ensure that synthesis only produces well-typed programs, and our
-solution will use the same literature as a base.  However, since their
+ensure that synthesis only produces well-typed programs. However, since their
 underlying logic is \emph{modal} rather than purely \emph{linear}, it
@@
 resulting program fullfills its intended purpose
-
-% TODO: Utilizamos qual argumento primeiro ??? para \omega com ++ ou : ?
 

### 2021-07-22 article/article.tex
Refinement types writing

@@
 
-\mypara{Refinement Types} Their syntax is... Their rules are...
+\mypara{Refinement Types} Refinement types are defined as a types
+with a name and with a predicate (a non-existing predicate is the same as it
+being always \emph{true}); dependent types are functions with
+refinement types in which the argument type name can be used in the predicates
+of the return type (e.g. $x\{\mathsf{Int}\} \lolli y\{\mathsf{Int}\ \vert\ y = x\}$ specifies a function that
+takes an Int named $x$ and produces an Int whose value is equal to the input $x$). We extend
+the types syntax with our refinement types syntax:
+\[
+\begin{tabular}{lclc}
+    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ \langle\emph{name}\rangle \{\tau\ \vert\ P\}$ \\
+    $P$ & $\ ::=\ $ & $P = P\ \vert\ P \neq P\ \vert\ P \vee P\ \vert\ P \wedge
+    P\ \vert\ P \Rightarrow P\ \vert\ n = n\ \vert\ n \neq n$ \\
+    & & $\vert\ n \leq n\ \vert\ n \geq n\ \vert\ n < n\ \vert\ n > n\ \vert\
+    true\ \vert\ false\ \vert\ \langle\emph{name}\rangle$ \\
+    $n$ & $\ ::=\ $ & $n * n\ \vert\ n + n\ \vert\ n - n\ \vert\
+    \langle\emph{natural}\rangle\ \vert\ \langle\emph{name}\rangle$
+    \\
+\end{tabular}
+\]
+The addition of refinement types to the synthesizer doesn't
+interfere with the rest of the process. We define the following right and
+left rule, to synthesize or consume in synthesis a refinement type, where
+$\textsc{getModel}(p)$ is a call to an SMT solver that returns a model of an
+uninterpreted function that satisfies
+$\forall_{\tilde{1},\tilde{2},\dots,\tilde{n}}\ h_{\tilde{1}} \Rightarrow h_{\tilde{2}} \Rightarrow \dots
+\Rightarrow h_{\tilde{n}} \Rightarrow p$, where $\tilde{i}$ is the refinement type
+name and $h_{\tilde{i}}$ its predicate for every refinement type in the
+propositional contexts; and $\textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}})$ is a call to an SMT solver that
+determines universal satisfiability of the implication between predicates (the
+left focused proposition subtypes the goal).
 
@@
     \infer*[right=(refR)]
-    {  }
-    {  }
+    { \textsc{getModel}(p) = M }
+    {\Theta/\Theta';\Rho;\Gamma;\Delta \vdash M : \tilde{x}\{A\ \vert\
+    p\}\Uparrow }
+    \and
+    \infer*[right=(refL)]
+    { \textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}}) }
+    {\Theta/\Theta';\Rho;\Gamma;\Delta; x{:}\tilde{y}\{A\ \vert\ p_{\tilde{y}}\}\Downarrow\ \vdash x : \tilde{x}\{A\ \vert\ p_{\tilde{x}}\} }
 \end{mathpar}

### 2021-07-22 article/article.tex
optimization, recursion extra

@@
 produce the goal $C$, and only in the \emph{Cons}
-branch. % (TODO: rewrite? O bold é meio estranho não?)
+branch. Furthermore, we also forbid the usage of the recursive function when
+synthesizing arguments to use it.
+% TODO : não sei se precisa de melhor explicação mas foi uma coisa que fiz para
+% não gerar um programa recursivo.
+% (TODO: rewrite? O bold é meio estranho não?)
 
@@
 left focused proposition subtypes the goal).
-
 \begin{mathpar}
@@
     {\Theta/\Theta';\Rho;\Gamma;\Delta; x{:}\tilde{y}\{A\ \vert\ p_{\tilde{y}}\}\Downarrow\ \vdash x : \tilde{x}\{A\ \vert\ p_{\tilde{x}}\} }
+\end{mathpar}
+
+\mypara{Optimizations} To speed up the process and get a cleaner and sometimes more correct
+output, we add a rule that lets us ``skip some rules'' if left focused on a
+$\bang$-ed proposition, and the goal is $\bang$-ed:
+\begin{mathpar}
+    \infer*[right=($\bang\Downarrow$L)]
+    { \Theta/\Theta';\Rho;\Gamma;\Delta; A\Downarrow\ \vdash M : C}
+    {\Theta/\Theta';\Rho;\Gamma;\Delta; \bang A\Downarrow\ \vdash M : \bang C}
 \end{mathpar}

### 2021-07-23 article/article.tex
Timeout on infinite evaluation, 5 hours lazy bug, $!, and reverse

@@
     \infer*[right=($\bang\Downarrow$L)]
-    { \Theta/\Theta';\Rho;\Gamma;\Delta; A\Downarrow\ \vdash M : C}
-    {\Theta/\Theta';\Rho;\Gamma;\Delta; \bang A\Downarrow\ \vdash M : \bang C}
+    { \Theta/\Theta';\Rho;\Gamma;\Delta; x{:}A\Downarrow\ \vdash M : C}
+    {\Theta/\Theta';\Rho;\Gamma;\Delta; x{:}\bang A\Downarrow\ \vdash M : \bang C}
 \end{mathpar}
@@
 Section~\ref{sec:overview}, synthesis with memoization will take
-around $5s$, while synthesis without memoization doesn't terminate
-even after $15m$ running.
+around $9s$, while synthesis without memoization doesn't terminate in less than
+$15m$ running.
 %

### 2021-07-23 STLLC/Reverse.hs

@@ -1,0 +1,27 @@
+
+
+
+
+
+
+
+
+
+
+
+
+
+
+data List a = Nil | Cons (a * List a);
+
+synth reverse :: List a -o List a -o List a
+  | assert
+  (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil)));
+
+#reverse l acc =
+#    case l of
+#        Nil -> acc
+#      | Cons x*xs -> reverse xs (Cons (x, acc));
+
+
+#main = reverse (Cons (1, Cons (2, Cons (3, Cons (4, Nil))))) Nil;

### 2021-07-23 STLLC/app/Synth.hs

@@ -1,6 +1,10 @@
 {-# LANGUAGE LambdaCase #-}
 module Synth (synthAllType, synthType, synthMarks, synthMarksModule, synth) where
 
+import System.Timeout
+import Control.Monad
+import Control.Concurrent.Async
+import Control.Concurrent
 import Data.List
 import Text.Read (readMaybe)
 import qualified Data.Set as Set
@@ -68,7 +72,74 @@ runSynth prog n (g, d) t st adtds ed touse assertion = do
             (e, d', _) <-  memoizesynth synth (Map.empty, [], recf, d, o) t -- synth only with the recursive function
                        <|> memoizesynth synth (Map.empty, [], g, d, o) t    -- synth with the whole context
             guard $ null d'
-            when (isJust assertion) $ guard $ fromADTBool $ evalModule $ desugarModule $ progAddBinds [Binding (fst $ head recf) e, Binding "main" $ fromJust assertion] prog
+            when (isJust assertion) (guard =<< liftIO (do
+                -- Attempt to evaluate the program, cancelling the thread running it after x time (deeming the program to be infinitely recursive (since no complex assertions are expected))
+
+                x <- timeout 5000 $ do
+                    return $! fromADTBool $ -- Important: The use of if is crucial to force strict evaluation, without it the timeout won't work and eval might run forever
+                                 evalModule $
+                                   desugarModule $
+                                     progAddBinds [Binding (fst $ head recf) e, Binding "main" $ fromJust assertion] prog
+                case x of
+                  Nothing -> return False
+                  Just y -> return y
+
+                -- x <- timeout 5000 $ do
+                --         if fromADTBool $ -- Important: The use of if is crucial to force strict evaluation, without it the timeout won't work and eval might run forever
+                --              evalModule $
+                --                desugarModule $
+                --                  progAddBinds [Binding (fst $ head recf) e, Binding "main" $ fromJust assertion] prog
+                --          then return True
+                --          else return False
+                -- case x of
+                --   Nothing -> return False
+                --   Just y -> return y
+
+                -- -- Loops forever, despite the timeout
+                -- x <- timeout 10000 $ do
+                --         return $ fromADTBool $
+                --              evalModule $
+                --                desugarModule $
+                --                  progAddBinds [Binding (fst $ head recf) e, Binding "main" $ fromJust assertion] prog
+                -- case x of
+                --   Nothing -> return False
+                --   Just y -> return y
+
+                -- resVar <- newEmptyMVar
+
[…]

### 2021-07-23 article/article.tex
tweaks

@@
 
-\paragraph{Contributions and Outline.} Despite their long-known potential
+\mypara{Contributions and Outline} Despite their long-known potential
 \cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018}
-and strong proof-theoretic
-foundations
+and strong proof-theoretic foundations
 \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
-synthesis with linear types combined with other advanced typing features
-has generally been overlooked in the literature.  In this work we present a framework for synthesis
-with linear types extended with recursive algebraic data types, parametric
-polymorphism and refinements.
+synthesis with linear types combined with other advanced typing
+features has generally been overlooked in the literature.  In this
+work we present a framework for synthesis with linear types extended
+with recursive algebraic data types, parametric polymorphism and
+refinements.
 % TODO: extensible? proving feasability of... ? showing potential for ... ?
@@
 implementation challenges (\S~\ref{sec:architecture}). Finally, 
-\todo{como formular a frase sobre o apêndice?}
 we evaluate our work through expressiveness and performance benchmarks
 (\S~\ref{sec:evaluation}) and discuss related work
-(\S~\ref{sec:related}).
+(\S~\ref{sec:related}). Appendix~\ref{sec:background} lists
+foo. Appendix~\ref{} details blah.
 
@@
 
-\paragraph{Core linear calculus.} Initially, we synthesize from
+\mypara{Core linear calculus} Initially, we synthesize from
 specifications that only use literals and propositional linear types, such as linear logic theorems. Namely,
@@
 
-\paragraph{Beyond propositional logic.} To be able to synthesize more
+\mypara{Beyond propositional logic} To be able to synthesize more
 interesting programs -- and thus, empirically prove the feasability of
@@
 
-\paragraph{ADTs.} We reiterate the expressiveness of linear types: by requiring
+\mypara{ADTs} We reiterate the expressiveness of linear types: by requiring
 some types to be used linearly (consumed), we can assure, e.g., the
@@
 linearly in its body} 
-($\lolli$) that converts $a$s in $b$s, a list of $a$s, and produces a
+that converts $a$s in $b$s, a list of $a$s, and produces a
 list of $b$s, written $\bang(a \lolli b) \lolli List\ a \lolli List\ b$, isn't
@@
 
-\paragraph{Refinements.} Refinements are logical predicates
+\mypara{Refinements} Refinements are logical predicates
 which must hold for any element that inhabits a given type. The \synname\
@@
 
-\paragraph{Guiding the Search.} Specifications can be additionally
+\mypara{Guiding the Search} Specifications can be additionally
 augmented with four keywords: ``using'', ``depth'', ``assert'' and ``choose''.
@@
 incorrect programs; and that the valid programs derivable through them
-reflect the somewhat subjective trade-offs we committed to. Different
+reflect the subjective trade-offs we committed to. Different
 choices and approaches outside the core might lead to completely
@@
 
-\mypara{Core Rules} The core system comprises of proof search for
-(intuitionistic) linear logic sequent calculus proofs, based on a system of
+\mypara{Core Rules} The system comprises of proof search in
+(intuitionistic) linear logic sequent calculus, based on a system of
 resource management~\cite{resman} and focusing.
+
+The core language is a simply-typed linear $\lambda$-calculus with linear
+functions ($\lolli$), additive ($\with$) and multiplicative
+($\tensor$) pairs, \todo{qual a diferenca}
+multiplicative unit (\textbf{1}), additive sums ($\oplus$)
+and the exponential modality ($\bang$) (to internalize unrestricted
+use of variables). The syntax of terms $(M,N)$ and types $\tau, \sigma$ is given below:
+
+%
+% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
+% usá-los como tipos gerais? NAO :)
+
+\[
+  \begin{array}{lclcc}
+    M,N & ::= & u, v & \quad & \\
+        & \mid & \lambda x.M \mid M\, N & & (\lolli) \\
+        & \mid & M \with N\ \vert\ \mathsf{fst}\ M\ \vert\
+                 \mathsf{snd}\ M && (\with) \\
+        & \mid & M \tensor N\ \vert\ \llet{u\tensor v = M}{N} && (\tensor)\\
+        & \mid & \star\ \vert\ \llet{\star = M}{N} && (\one) \\
+        & \mid & \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
+    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
+                 N_2) && (\oplus)\\
+        & \mid & \bang M\ \vert\ \llet{\bang u = M}{N} && (\bang)\\[0,5em]
+    \tau, \sigma & ::= & a\ \vert\ \tau \lolli \sigma\ \vert\ \tau \with \sigma\
+    \vert\ \tau \tensor \sigma\ \vert\ \mathbf{1}\ \vert\ \tau \oplus \sigma\ \vert\ \bang \tau
+  \end{array}
+  \]
+
+%\vspace{-0.5cm}
+% \[
+% \begin{tabular}{lclc}
+% $M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
+%     & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
+%     & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
+%     & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
+%     & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
+%     & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
+%     M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
+%     N_2)$ & $\oplus$ \\
+%     & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
+%     \\
+% $\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
+%     \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
+% \end{tabular}
+% \]
+
 %
@@
 phases: %
-the inversion phase ($\Uparrow$), in which we apply \emph{all} invertible rules
-eagerly, and the focusing phase  ($\Downarrow$), in which we decide a proposition
-to focus on, and then apply non-invertible rules, staying \emph{in focus}
-until we reach an asynchronous (i.e. invertible proposition), the
-proof is complete, or no rules are applicable, in which case the proof must
-\emph{backtrack} to the state at which the focusing phase began.
-As such, with focusing, a linear logic judgment in sequent calculus ($\Gamma; \Delta, B
-\vdash A$) is broken down into four different judgments, grouped into the two
-phases ($\Uparrow, \Downarrow$). For the invertible phase, a third % TODO: ordered??,
-context ($\Omega$) is needed to hold propositions that result from
-decomposing connectives (propositions we'll later try to invert, moving them to
-the linear context ($\Delta$) when we fail to). The judgments of right
-invertible rules are written $\Gamma; \Delta; \Omega,B \vdash A \Uparrow$, and
-those of left invertible rules are $\Gamma; \Delta; \Omega,B \Uparrow\ \vdash A$.
-In the focusing phase, the focus proposition will be either the goal or come
-from $\Gamma$ or $\Delta$, and appliable rules will be classified as right
-non-invertible, or left non-invertible, respectively. Right non-invertible
-rules have the form $\Gamma;\Delta,B \vdash A \Downarrow$, and left
-non-invertible rules are written $\Gamma;\Delta; B \Downarrow\ \vdash A$.
+the inversion phase ($\Uparrow$), in which we apply \emph{all}
+invertible rules eagerly, and the focusing phase ($\Downarrow$), in
+which we decide a proposition to focus on, and then apply
+non-invertible rules, staying \emph{in focus} until we reach an
+asynchronous (i.e. invertible proposition), the proof is complete, or
+no rules are applicable, in which case the proof must \emph{backtrack}
+to the state at which the focusing phase began.  As such, with
+focusing, the linear sequent calculus judgment 
+$\Gamma; \Delta \vdash A$, meaning that $A$ is derivable from the linear
+assumptions in $\Delta$ and unrestricted assumptions in $\Gamma$,
+is split into four judgments, grouped into the two phases ($\Uparrow,
+\Downarrow$).
+%
+For the invertible phase, an \emph{inversion} 
+context $\Omega$ holds propositions that result from
+decomposing connectives. % (propositions we'll later try to invert,
+% moving them to the linear context ($\Delta$) when we fail to).
+The right inversion and left inversion judgments are written
+$\Gamma; \Delta; \Omega \vdash A \Uparrow$ and $\Gamma; \Delta;
+\Omega \Uparrow\ \vdash A$, respectively, where the $\Uparrow$
+indicates the connective or context being inverted.
+For the focusing phase (i.e.~all non-invertible rules can apply), 
+the proposition under focus can be the goal or 
+one in $\Gamma$ or $\Delta$. The right focus judgment is written 
+$\Gamma;\Delta \vdash A \Downarrow$ and the left focus judgment is
+written $\Gamma;\Delta; B\Downarrow\ \vdash A$, where $\Downarrow$
+indicates the proposition under focus.
 
@@
 
-The core language is a simply-typed linear $\lambda$-calculus with linear
-functions ($\lolli$), additive ($\with$) and multiplicative ($\tensor$) pairs,
-multiplicative unit (\textbf{1}), additive sums ($\oplus$)
-and the exponential modality ($\bang$) (to internalize unrestricted use of variables), for
-which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
-following grammar:
-%
-% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
-% usá-los como tipos gerais? NAO :)
-\[
-\begin{tabular}{lclc}
-$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
-    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
-    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
-    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
-    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
-    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
-    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
-    N_2)$ & $\oplus$ \\
-    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
-    \\
-$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
-    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
-\end{tabular}
-\]
-%
+  %
+\todo[inline]{Context management -- To handle the context splitting
+  required ... we augment the judgments above with .... where ....}
+
 Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
@@
 inspired
-by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
+by~\cite{FPnotes...}) -- in which the
 rule $\lolli$R is read: to synthesize a program of type $A \lolli B$ while inverting
@@
 
-We begin with the right invertible rules:
+We begin with the right invertible rules:\todo{que decompoem a direita}
 \begin{mathpar}
@@
 
-Followed by the left invertible rules:
-\begin{mathpar}
+Followed by the left invertible rules, which ...\todo{fazem o que?}:
+\[
+  \begin{array}{c}
 
@@
     {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \tensor B \Uparrow\ \vdash\
-    \textrm{let}\ y \tensor z = x\ \textrm{in}\ M : C}
-
-\and
-
+    \textrm{let}\ y \tensor z = x\ \textrm{in}\ M : C}\\[0.5em]
     \infer*[right=($1 L$)]
@@
     \star =
-    x\ \textrm{in}\ M : C}
-
-\and
-
+    x\ \textrm{in}\ M : C}\\[0.5em]
     \mprset{flushleft}
@@
     \textrm{inr}\ z \rightarrow N : C}
-
-\and
-
+    \\[0.5em]
     \infer*[right=($\bang L$)]
@@
     \textrm{let}\ \bang y = x\ \textrm{in}\ M : C}
-
-\end{mathpar}
+\end{array}
+    \]
 
@@
 \Uparrow\ \vdash C$). At this point, we want to \emph{focus} on a proposition.
-The focus object can be any of the three: the proposition on the right (the
+The focus object will be: the proposition on the right (the
 goal), a proposition from the linear $\Delta$ context, or a proposition from the
@@
 rules:
-\begin{mathpar}
+\[
+  \begin{array}{c}
     \infer*[right=(decideR)]
     {\Gamma; \Delta/\Delta' \vdash C \Downarrow \and C\ \textrm{not atomic}}
-    {\Gamma; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
-\and
+    {\Gamma; \Delta/\Delta';\cdot \Uparrow\ \vdash C}\\[0.5em]
     \infer*[right=(decideL)]
@@
     {\Gamma; \Delta, A/\Delta';\cdot \Uparrow\ \vdash C}
-\and
+\qquad
     \infer*[right=(decideL!)]
@@
     {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
-\end{mathpar}
-
-The decision rules are followed by either left or right focus rules. The left
-focus rule $\lolli$L is read ``to synthesize a program of type $C$ with input
-linear context $\Delta$ and while left focused on a proposition $x$ of type $A
-\lolli B$, synthesize a program $M$ of type $C$ with input $\Delta$ and 
-left focused ($\Downarrow$) on $y$ of type
-$B$, and output the context $\Delta'$ (with the unused
-linear propositions); then, switching to the right inversion judgment ($\Uparrow$), synthesize a program $N$ of type $A$ with input
-$\Delta'$ and output context $\Delta''$; finally, replace in $M$ all occurrences
-of variable $y$ with program $x\ N$ (apply $x$ to $N$)''. It can be understood
-as the synthesis of a program of type $C$ using the return value (of type $B$) from the function $A \lolli B$, and then the
-synthesis of an argument of type $A$ to apply the function to.
+  \end{array}
+  \]
+
+  The decision rules are followed by either left or right focus rules.
+  To illustrate, consider the $\lolli$L left focus rule. The rule
+  states that to produce a program of type $C$ while left focused on
+  the function $x$ of type $A\lolli B$, we first check that we can produce a
+  program of type $C$ by using $B$. If this succeeds in producing some
+  program $M$, this means that we can apply $x$ to solve our
+  goal. We now synthesize a program $N$ of type $A$, switching to the
+  right inversion judgment ($\Uparrow$). To construct the overall
+  program, we replace in $M$ all occurrences of variable $y$ with the
+  application $x\,N$. The remaining left rules follow a similar
+  pattern. \todo{The focus right rules \dots }
+
   \begin{mathpar}
+      \infer*[right=($\lolli L$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
+    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
+    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
+\and
+    \infer*[right=($\with L_1$)]
+    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
+\and
+    \infer*[right=($\with L_2$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash
+      M\{(\textrm{snd}\ x)/y\} : C}
+    \and
     \infer*[right=($\tensor R$)]
@@
     {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
-\and
-    \infer*[right=($\lolli L$)]
-    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
-    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
-    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
-\and
-    \infer*[right=($\with L_1$)]
-    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
-    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
-\and
-    \infer*[right=($\with L_2$)]
-    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
-    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
 \end{mathpar}
@@
 %
-\begin{mathpar}
+\[
+  \begin{array}{c}
     \infer*[right=(init)]
@@
     {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
-\and
+    \qquad
+     \infer*[right=($\Downarrow R$)]
+    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
+    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}\\[0.5em]
     \infer*[right=($\Downarrow L$)]
@@
     {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
-\and
-    \infer*[right=($\Downarrow R$)]
-    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
-    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
-\end{mathpar}
+
+\end{array}
+  \]
 The rules written above together make the core of our synthetiser. Next, we'll
@@
 values of type $A$, using the syntax: \mintinline{haskell}{data Container
-= None 1 | One A | Two (A * A)}. The grammar is extended as:
+= None 1 | One A | Two (A * A)}. The grammar is extended as (where $C$
+is and $T$ is \dots):\todo{look here}
 \[
 \begin{tabular}{lclc}
-    $M,N$ & $\ ::=\ $ & $\dots\ \vert\ \emph{Cons}_n\ M\ \vert\ (\mathsf{case}\
-    M\ \mathsf{of}\ \dots\ \vert\ \emph{Cons}_n\ u \Rightarrow N)$ \\
-    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ \langle\emph{name}\rangle$ \\
-    $ADT$ & $\ ::=\ $ & (\textsf{data} $\langle\emph{name}\rangle$\ $=$\ $\emph{Cons}_1\ \tau$\
-$\vert$\ $\dots$\ $\vert$\ $\emph{Cons}_n\ \tau$) \\ \end{tabular}
+    $M,N$ & $\ ::=\ $ & $\dots\ \vert\ \emph{C}_n\ M\ \vert\ (\mathsf{case}\
+    M\ \mathsf{of}\ \dots\ \vert\ \emph{C}_n\ u \Rightarrow N)$ \\
+    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ T$ \\
+ %    $ADT$ & $\ ::=\ $ & (\textsf{data} $\langle\emph{name}\rangle$\ $=$\ $\emph{Cons}_1\ \tau$\
+% $\vert$\ $\dots$\ $\vert$\ $\emph{Cons}_n\ \tau$) \\
+\end{tabular}
 \]
@@
 $d_e$ is a \emph{constant existential depth} that controls the maximum depth of
-``existential synthetis''.
+``existential synthetis''. \todo{alinhar mais com a premissa
+  respectiva da regra}
 %
@@
 \begin{tabular}{lclc}
-    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ \langle\emph{name}\rangle \{\tau\ \vert\ P\}$ \\
+    $\tau$ & $\ ::=\ $ & $ \dots \vert\  (x:\tau) \lolli \sigma\ \vert\  \{x:\tau\ \vert\ P\}$ \\
     $P$ & $\ ::=\ $ & $P = P\ \vert\ P \neq P\ \vert\ P \vee P\ \vert\ P \wedge
@@
 
-\mypara{Optimizations} To speed up the process and get a cleaner and sometimes more correct
+\mypara{Optimizations} To speed up the process and get a cleaner %and sometimes more correct
 output, we add a rule that lets us ``skip some rules'' if left focused on a
@@
 
-The implementation was carried out in Haskell. We
-highlight some key points.
-
-\mypara{Backtracking} Even though focusing reduces non-determinism
+\mypara{Backtracking} Although focusing reduces non-determinism
 during proof search, it cannot fully eliminate it. At several points
@@
 attempt to derive a proof taking a different route. To this effect,
-our prototype synthetiser depends on \emph{LogicT}, "a backtracking
-logic-programming monad"~\cite{logict}. Special attention is
-required for ADTs. When deconstructing an ADT, since all branches must be
-synthetisable and result in the same ouput linear context $\Delta'$, alternative
-constructor deconstructions should be considered fairly in conjunction, so the
-\textbf{fair conjunction} operator should be used. When constructing an ADT, the
-disjunctive choice of constructor should be fair, and so the \textbf{fair
-disjunction} operator used. Fairness at these two points usually increases
-the synthesis speed slightly, but can be a relevant factor in synthesis in which a
-combination of choices following the ADT deconstruction or construction can
-backtrack many times or even arbitrarily.
+our prototype synthetiser depends on the Haskell library \emph{LogicT}, "a backtracking
+logic-programming monad"~\cite{logict}.
+% Special attention is
+% required for ADTs. When deconstructing an ADT, since all branches must be
+% synthetisable and result in the same ouput linear context $\Delta'$, alternative
+% constructor deconstructions should be considered fairly in conjunction, so the
+% \textbf{fair conjunction} operator should be used. When constructing an ADT, the
+% disjunctive choice of constructor should be fair, and so the \textbf{fair
+% disjunction} operator used. Fairness at these two points usually increases
+% the synthesis speed slightly, but can be a relevant factor in synthesis in which a
+% combination of choices following the ADT deconstruction or construction can
+% backtrack many times or even arbitrarily.
 
@@
 \emph{SBV}~\cite{sbv}. Unfortunately, this library does not support
-the usage of SMT solvers to check first-order logic with uninterpreted
+first-order logic with uninterpreted
 functions. Since this is needed for synthesis with refinement types
@@
 the unrestricted context in the \emph{array} problem seen in
-Section~\ref{sec:overview}, synthesis with memoization will take
-around $9s$, while synthesis without memoization doesn't terminate in less than
-$15m$ running.
+\S~\ref{sec:overview}, synthesis with memoization will take
+around $9s$, while synthesis without memoization doesn't terminate
+after $15m$.
 %
@@
 \mypara{Debugging} The rules described in the formal
-system\ref{sec:formal_system} to handle infinite recursion might seem obvious in
+system (\S~\ref{sec:formal_system}) to handle infinite recursion might seem obvious in
 retrospective, however, when faced with a program that doesn't terminate, what
@@
 
+\begin{figure}[t]
+{\scriptsize
 \begin{center}
@@
   \end{center}
+}
+\caption{Table with \dots}
+\end{figure}
 

### 2021-07-23 article/article.tex
tweaks

@@
 %TODO: ... we can actually instance ADTs that take no arguments even if they are restricted
+
 \begin{mathpar}
@@
 specification and programs and making the approach amenable to compositional
-synthesis techniques. We also support polymorphism and refinements, however, our
-refinements are somewhat underdeveloped and don't have an important role in our
-process, in contrast to their strong use of refinements to guide their
-synthesizer. Instead, our synthesizer leverages the expressiveness of linear
+synthesis techniques. We also support refinements (and polymorphism),
+but they are not as integrated in the synthesis process as in \cite{DBLP:conf/pldi/PolikarpovaKS16}.
+Instead, our synthesizer leverages the expressiveness of linear
 types and techniques for proof-search in linear logic to guide its process.

### 2021-07-23 article/article.tex
list appendices + synthet_ -> synthes_

@@
 (\S~\ref{sec:evaluation}) and discuss related work
-(\S~\ref{sec:related}). Appendix~\ref{sec:background} lists
-foo. Appendix~\ref{} details blah.
+(\S~\ref{sec:related}). Appendix~\ref{sec:background} covers background
+concepts such as \emph{linear logic} and \emph{sequent calculus}.
+Appendix~\ref{sec:final_system} presents the final set of inference
+rules, putting together everything seen in \S~\ref{sec:formal_system}.
+Appendix~\ref{sec:examples} lists concrete examples of synthesis with \synname.
 
@@
 
-The \synname\ synthetiser combines linear types with recursion,
+The \synname\ synthesizer combines linear types with recursion,
 parametric polymorphism, recursive algebraic data types, and
-refinement types. The synthetiser is built on top of a system of proof-search for
+refinement types. The synthesizer is built on top of a system of proof-search for
 linear logic. Proof search relates to program synthesis via the Curry-Howard
@@
 logic based on \emph{focusing}~\cite{}. Our approach, being grounded
-by propositions-as-types, ensures that all synthetised programs
+by propositions-as-types, ensures that all synthesized programs
 (i.e.~proofs) are well-typed \emph{by construction} (i.e.~if the
@@
 The \synname\ programming language is naturally developed alongside
-the synthetiser. Synthesis goals are inserted in the program by use of
+the synthesizer. Synthesis goals are inserted in the program by use of
 a \emph{synth} keyword or mark, indicating a program should be
@@
 list of $b$s, written $\bang(a \lolli b) \lolli List\ a \lolli List\ b$, isn't
-satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}. To synthetise
+satisfied by the program \mintinline{haskell}{(\x -> \y -> Nil)}. To synthesize
 a valid program from it, \mintinline{haskell}{(List a)} has to be deconstructed,
@@
 program\ \mintinline{haskell}{(\a -> \b -> \c -> ((0 - 1) * a) + (c * b))} is
-synthetised.
+synthesized.
 
@@
 synthesis from our more expressive specifications, alongside the already well
-defined rules that model the core of the synthetiser, putting together a
+defined rules that model the core of the synthesizer, putting together a
 \emph{sound} set of inference rules that characterizes our framework
@@
 % TODO: Better?....
-and describes the system in enough detail for the synthetiser to be reproducible
+and describes the system in enough detail for the synthesizer to be reproducible
 by a theory-driven implementation.
@@
 right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
-context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
+context $\Delta$, and inversion context $\Omega$, synthesize a program of type $B$ with
 an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
@@
   \]
-The rules written above together make the core of our synthetiser. Next, we'll
-present new rules that align and build on top of these to synthetise recursive
+The rules written above together make the core of our synthesizer. Next, we'll
+present new rules that align and build on top of these to synthesize recursive
 programs from more expressive (richer) types.
@@
 proofs of its branches, and, symmetrically, don't allow construction of an
-ADT when trying to synthetise an argument for its constructor.
+ADT when trying to synthesize an argument for its constructor.
 %
@@
 recursion in the synthesis process. The first is the use of an unrestricted
-function to synthetise a term of type $\tau$ that in turn will require a term of
+function to synthesize a term of type $\tau$ that in turn will require a term of
 the same type $\tau$. An example is the sub-goal judgment $(a \lolli
@@
 and on and on. The solution will be to disallow the usage of the same function
-to synthetise the same goal a second time further down in the derivation.
+to synthesize the same goal a second time further down in the derivation.
 
@@
 As such, we'll instead forbid the usage of \emph{all} unrestricted polymorphic
-functions to synthetise any existential type if, altogether, we've already used
+functions to synthesize any existential type if, altogether, we've already used
 $d_e$ times a polymorphic function for \emph{any} existential sub-goal, where
 $d_e$ is a \emph{constant existential depth} that controls the maximum depth of
-``existential synthetis''. \todo{alinhar mais com a premissa
+``existential synthesis''. \todo{alinhar mais com a premissa
   respectiva da regra}
@@
 \mintinline{haskell}{data List a = Cons (a *
-List a)}, we wouldn't try to synthetise \mintinline{haskell}{(a * List a)},
+List a)}, we wouldn't try to synthesize \mintinline{haskell}{(a * List a)},
 but rather \mintinline{haskell}{(Int * List Int)}). To unify $T_{\overline\alpha}$ with $T_{\overline\beta}$,
@@
 
-The \synname\ synthetiser operates as part of the pipeline that processes a
+The \synname\ synthesizer operates as part of the pipeline that processes a
 full program in the \synname\ language, and is called to generate terms marked
@@
 
-The synthesis module iterates over the synthesis marks and runs the synthetiser
+The synthesis module iterates over the synthesis marks and runs the synthesizer
 with the respective context and goal for each mark. The resulting expression is checked
@@
 attempt to derive a proof taking a different route. To this effect,
-our prototype synthetiser depends on the Haskell library \emph{LogicT}, "a backtracking
+our prototype synthesizer depends on the Haskell library \emph{LogicT}, "a backtracking
 logic-programming monad"~\cite{logict}.
@@
 
-\mypara{Memoization} Some proofs attempt to synthetise the same
+\mypara{Memoization} Some proofs attempt to synthesize the same
 sub-goal with the same premises more than once. Since these proof
 attempts might be expensive performance-wise, and we can rely on the
-program's determinism (equal calls to synthetise result in equal
+program's determinism (equal calls to synthesize result in equal
 outcomes), we added memoization to our synthesizer -- an optimization
@@
 
-\section{Related Work}
+\section{Related Work}\label{sec:related}
 
@@
 \mypara{Resourceful Program Synthesis from Graded Linear Types} The
-work~\cite{DBLP:conf/lopstr/HughesO20} synthetises programs using an
+work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
 approach very similar to our own.  The work employs so-called graded
@@
 calculus system, for it simplifies the work to be done in the
-synthetiser.
+synthesizer.
 
@@
 
+\section{Formal System}%
+\label{sec:final_system}
+
+\section{Examples}%
+\label{sec:examples}
+
 \end{document}

### 2021-07-23 STLLC/BST.hs

@@ -13,11 +13,4 @@ size b = case b of
 
 synth singleton :: (a -o BST a);
 
-
-synth insert :: (!a -o BST a -o BST a)
-  | assert
-  let s*t = size (insert (!1) (singleton 1)) in
-      let s1 * t1 = size (Node ((1, Empty), singleton 1)) in
-          ((s == (s1 + 2)) && (t != t1));
-
-
+synth insert :: (!a -o BST a -o BST a);

### 2021-07-23 STLLC/app/Synth.hs

@@ -74,71 +74,14 @@ runSynth prog n (g, d) t st adtds ed touse assertion = do
             guard $ null d'
             when (isJust assertion) (guard =<< liftIO (do
                 -- Attempt to evaluate the program, cancelling the thread running it after x time (deeming the program to be infinitely recursive (since no complex assertions are expected))
-
                 x <- timeout 5000 $ do
-                    return $! fromADTBool $ -- Important: The use of if is crucial to force strict evaluation, without it the timeout won't work and eval might run forever
+                    return $! fromADTBool $ -- Important: The use of strict evaluation is crucial, without it the timeout won't work and eval might run forever
                                  evalModule $
                                    desugarModule $
                                      progAddBinds [Binding (fst $ head recf) e, Binding "main" $ fromJust assertion] prog
                 case x of
                   Nothing -> return False
                   Just y -> return y
-
-                -- x <- timeout 5000 $ do
-                --         if fromADTBool $ -- Important: The use of if is crucial to force strict evaluation, without it the timeout won't work and eval might run forever
-                --              evalModule $
-                --                desugarModule $
-                --                  progAddBinds [Binding (fst $ head recf) e, Binding "main" $ fromJust assertion] prog
-                --          then return True
-                --          else return False
-                -- case x of
-                --   Nothing -> return False
-                --   Just y -> return y
-
-                -- -- Loops forever, despite the timeout
-                -- x <- timeout 10000 $ do
-                --         return $ fromADTBool $
-                --              evalModule $
-                --                desugarModule $
-                --                  progAddBinds [Binding (fst $ head recf) e, Binding "main" $ fromJust assertion] prog
-                -- case x of
-                --   Nothing -> return False
-                --   Just y -> return y
-
-                -- resVar <- newEmptyMVar
-
-                -- evalt  <- forkIO (do            -- If the program doesn't terminate fast enough, the thread above will have already placed "False" in the MVar, 
-                --         let evalres = 1 == length [0..]
-                --         -- let evalres = fromADTBool $
-                --         --                   evalModule $
-                --         --                       desugarModule $
-                --         --                           progAddBinds [Binding (fst $ head recf) e, Binding "main" $ fromJust assertion] prog
-                --         putMVar resVar evalres -- Program successfully terminated, and the assertion result is returned
-                --         return ()
-                --     )
-
-                -- delayt <- forkIO (do
-                --     threadDelay 200000
-                --     putMVar resVar False
[…]

### 2021-07-23 article/article.tex
Todos, tweaks, page composition :)

@@
 \mypara{Contributions and Outline} Despite their long-known potential
-\cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018}
-and strong proof-theoretic foundations
+\cite{Wadler90lineartypes,DBLP:journals/mscs/CairesPT16,Bernardy_2018} and
+strong proof-theoretic foundations
 \cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05,DBLP:journals/tcs/CervesatoHP00},
-synthesis with linear types combined with other advanced typing
-features has generally been overlooked in the literature.  In this
-work we present a framework for synthesis with linear types extended
-with recursive algebraic data types, parametric polymorphism and
-refinements.
+synthesis with linear types combined with other advanced typing features has
+generally been overlooked in the literature.  In this work we present a
+framework for synthesis with linear types extended with recursive algebraic data
+types, parametric polymorphism and refinements.
 % TODO: extensible? proving feasability of... ? showing potential for ... ?
-We first introduce linear types as specifications
-and outline the synthesis process, leveraging linearity, by example
-(\S~\ref{sec:overview}). We then present the formal system driving the
-synthesis (\S~\ref{sec:formal_system}) and describe the architecture
-of our framework named \synname, examining technical details and key
-implementation challenges (\S~\ref{sec:architecture}). Finally, 
+We first introduce linear types as specifications and outline the synthesis
+process, leveraging linearity, by example (\S~\ref{sec:overview}). We then
+discuss the formal system driving the synthesis (\S~\ref{sec:formal_system}) and
+describe the architecture of our framework named \synname, examining technical
+details and key implementation challenges (\S~\ref{sec:architecture}). Finally,
 we evaluate our work through expressiveness and performance benchmarks
-(\S~\ref{sec:evaluation}) and discuss related work
-(\S~\ref{sec:related}). Appendix~\ref{sec:background} covers background
-concepts such as \emph{linear logic} and \emph{sequent calculus}.
-Appendix~\ref{sec:final_system} presents the final set of inference
-rules, putting together everything seen in \S~\ref{sec:formal_system}.
-Appendix~\ref{sec:examples} lists concrete examples of synthesis with \synname.
+(\S~\ref{sec:evaluation}) and discuss related work (\S~\ref{sec:related}).
+Appendix~\ref{sec:background} covers background concepts such as \emph{linear
+logic} and \emph{sequent calculus}. Appendix~\ref{sec:final_system} presents
+the final set of inference rules. Appendix~\ref{sec:examples} lists concrete examples
+of synthesis with \synname.
 
@@
 considered, and unused propositions are identified during synthesis,
-constraining the space of valid programs.  % TODO: Reword?
+constraining the space of valid programs.
 
@@
 \begin{minted}{haskell}
-map c d = let !e = c in
-  case d of
+map !e d = case d of
       Nil -> Nil
-    | Cons f -> let g*h = f in Cons (e g, map (!e) h);
+    | Cons g*h -> Cons (e g, map (!e) h);
 \end{minted}
@@
 \mypara{Refinements} Refinements are logical predicates
-which must hold for any element that inhabits a given type. The \synname\
+which must hold for any term that inhabits a given type. The \synname\
 language supports simple refinements on integers with arithmetic
@@
 %
-Matching exactly the linear definition for
-\emph{array} from~\cite{Bernardy_2018}, \synname\
-outputs (in a fraction of a second):
-%
+Matching the linear definition for \emph{array} from~\cite{Bernardy_2018}, \synname\
+outputs (\emph{in 0.1s}):
+%
+% TODO: Make it actually output this... need pattern matching on lambdas + minimization...
 \begin{minted}{haskell}
-array b c = let !d = b in let !e = c in
-    newMArray (!d) (\j -> freeze (foldl (!write) j e));
+array !d !e = newMArray (!d) (\j -> freeze (foldl (!write) j e));
 \end{minted}
@@
 % with Linear Types, Polymorphism and Refinements
-
-%\todo[inline]{Aqui parece-me que deves comecar por explicar o que vais fazer
-%nesta seccao. Sinto falta da sintaxe da linguagem, possivelmente
-%incrementalmente. Comecava por listar os tipos/termos do lambda linear
-%``simples'', depois em cada subseccao acrescentava as construcoes novas e como a
-%sintese funciona para esse contexto. Enfatizar que a novelty é a combinacao
-%destas features todas.}
-
-%The development of \synname\ was based on an almost direct reading of concrete
-%inference rules and theoretical principles (~\cite{frank pfenning notes});
-%%
-%\todo{aqui queres dizer que começaste por te basear num sistema de focusing para
-%a logica linear (era sequer linear a formulacao que comecaste a seguir?) O que
-%são theoretical principles?}
-%%
-%and work on its environment was also guided by formal systems and methods
-%(e.g.~Hindley-Milner type inference~\cite{}).
-%%
-%\todo{esta frase parece-me desnecessaria. O que queres dizer neste paragrafo,
-%parece-me, é que partes de um sistema focused da literatura que depois vais
-%extender, inspirando te em tecnicas standard de proof search / inferencia.}
-
-%However, with the addition of recursion and richer types to the synthesis and
-%language, formal guidelines
-%%
-%\todo{nao sao formal guidelines, nao existem trabalhos a combinar estas features
-%todas, essa parte é importante. Além de que ao incluir estas coisas deixamos de
-%estar completamente em proof search, pois muitas destas coisas (e.g. recursao)
-%nao sao necessariamente sound logicamente.  Tenta usar termos especificos.
-%Richer types? Richer como?}
-%%
-%to support the implementation become scarcer, or more complex.
-%%
-%\todo{nao existem ou sao complicados? ambos? citacao?}
-%%
-%Accordingly, rather than derived
-%from rules, solutions appeared first as techniques to tame the extensions'
-%inherent complex problems (such as infinite recursion), and we now present
-%those techniques as rules.
-%%
-% \todo{ Nao percebo o que queres dizer aqui. O que sao explicit rules e implicit rules?
-% Ou há regras ou nao há. O que estas a
-% dizer é que inventaste tecnicas de sintese para estes casos, que correspondem
-% as regras que vais apresentar. E tao explicito ou implicito como o que esta na
-% literatura.}
 
@@
 recursive programs from specifications with the select richer types,
-% TODO: Better?....
 and describes the system in enough detail for the synthesizer to be reproducible
 by a theory-driven implementation.
-
+%
 We note that a \emph{sound} set of rules guarantees we cannot synthesize
@@
 functions ($\lolli$), additive ($\with$) and multiplicative
-($\tensor$) pairs, \todo{qual a diferenca}
+($\tensor$) pairs
+(denoting alternative \emph{vs} simultaneous occurrence of resources),
 multiplicative unit (\textbf{1}), additive sums ($\oplus$)
 and the exponential modality ($\bang$) (to internalize unrestricted
-use of variables). The syntax of terms $(M,N)$ and types $\tau, \sigma$ is given below:
-
-%
-% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
-% usá-los como tipos gerais? NAO :)
-
+use of variables). The syntax of terms $(M,N)$ and types $(\tau, \sigma)$ is given below:
 \[
@@
 
-%\vspace{-0.5cm}
-% \[
-% \begin{tabular}{lclc}
-% $M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
-%     & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
-%     & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
-%     & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
-%     & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
-%     & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
-%     M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
-%     N_2)$ & $\oplus$ \\
-%     & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
-%     \\
-% $\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
-%     \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
-% \end{tabular}
-% \]
-
-%
-% TODO: Onde falamos de resource management? ( 
-%
 In intuitionistic sequent calculi, each connective has a so-called \emph{left}
@@
 proof-search by structuring the application of so-called invertible and
-non-invertible inference rules. Andreoli observed that the
-connectives of linear logic can be divided into two categories, dubbed
-synchronous and asynchronous. Asynchronous connectives are those whose
-right rules are \emph{invertible}, i.e. they can be applied eagerly
-during proof search without losing provability and so the order in
-which these rules are applied is irrelevant, and whose left rules
-are not invertible. Synchronous connectives are dual. The asynchronous \todo{Escrever isto depois
-de apresentar o linear $\lambda$-calculus?} 
+non-invertible inference rules. Andreoli observed that the connectives of linear
+logic can be divided into two categories, dubbed synchronous and asynchronous.
+Asynchronous connectives are those whose right rules are \emph{invertible}, i.e.
+they can be applied eagerly during proof search without losing provability and
+so the order in which these rules are applied is irrelevant, and whose left
+rules are not invertible. Synchronous connectives are dual. The asynchronous
 connectives are $\lolli$ and $\with$ and the synchronous are
@@
 context $\Omega$ holds propositions that result from
-decomposing connectives. % (propositions we'll later try to invert,
+decomposing connectives.
+% (propositions we'll later try to invert,
 % moving them to the linear context ($\Delta$) when we fail to).
@@
 
-% TODO: Explicar vários judgments?
-% \todo{Primeiro explica os varios judgments (as ``formas'' das
-%   conclusoes das regras) e o que significam as varias componentes,
-%   nomeadamente a parte da gestao de recursos.  Depois podes explicar
-%   regras concretas como fazes a seguir. Tambem seria util dizer quais
-%   os tipos que sao invertiveis a esquerda/direita, pois o leitor nao
-%   faz ideia do que estas a dizer.}
-%
-% Esta parte em geral está a ficar pouco coerente, com a introdução do focusing,
-  % a suposta introdução dos judgments, e a introdução da gramática... deixa
-
-  %
-\todo[inline]{Context management -- To handle the context splitting
-  required ... we augment the judgments above with .... where ....}
+To handle the context splitting required to prove subgoals, we augment the
+judgments above using Hodas and Miller's resource management technique where a pair of
+input/output linear contexts is used to propagate the yet unused linear
+resources across subgoals; e.g. the left inversion judment is written
+$\Gamma; \Delta/\Delta'; \Omega \Uparrow\ \nolinebreak\vdash A$ where $\Delta$ is the input
+linear context and $\Delta'$ is the output one.
 
 Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
-management, and focusing, we get the following core formal system (%heavily
-inspired
-by~\cite{FPnotes...}) -- in which the
+management, and focusing, we get the following core formal system
+(inspired by~\cite{FPnotes...}) -- in which the
 rule $\lolli$R is read: to synthesize a program of type $A \lolli B$ while inverting
@@
 
-We begin with the right invertible rules:\todo{que decompoem a direita}
+We start with right invertible rules, which decompose the goal proposition until it's synchronous.
+%
 \begin{mathpar}
@@
 $\Omega$ context. The rule to transition to inversion on the left is:
+%
 \begin{mathpar}
@@
 
-Followed by the left invertible rules, which ...\todo{fazem o que?}:
+We follow with left invertible rules for asynchronous connectives, which
+decompose asynchronous propositions in $\Omega$.
 \[
@@
 application $x\,N$. The remaining left rules follow a similar
-  pattern. \todo{The focus right rules \dots }
-
+pattern. The right focus rules are read similarly to right inversion ones,
+albeit the goal and sub-goals are under focus (except for $\bang R$).
+%
 \begin{mathpar}
@@
 values of type $A$, using the syntax: \mintinline{haskell}{data Container
-= None 1 | One A | Two (A * A)}. The grammar is extended as (where $C$
-is and $T$ is \dots):\todo{look here}
+= None 1 | One A | Two (A * A)}. The grammar is extended as (where $C$ is an ADT
+constructor and $T$ is an ADT):
 \[
-\begin{tabular}{lclc}
-    $M,N$ & $\ ::=\ $ & $\dots\ \vert\ \emph{C}_n\ M\ \vert\ (\mathsf{case}\
-    M\ \mathsf{of}\ \dots\ \vert\ \emph{C}_n\ u \Rightarrow N)$ \\
-    $\tau$ & $\ ::=\ $ & $\dots\ \vert\ T$ \\
+\begin{array}{lclc}
+    M,N & ::= & \dots\ \vert\ \emph{C}_n\ M\ \vert\ (\mathsf{case}\ M\ \mathsf{of}\ \dots\ \vert\ \emph{C}_n\ u \Rightarrow N) \\
+    \tau, \sigma & ::= & \dots\ \vert\ T \\
  %    $ADT$ & $\ ::=\ $ & (\textsf{data} $\langle\emph{name}\rangle$\ $=$\ $\emph{Cons}_1\ \tau$\
 % $\vert$\ $\dots$\ $\vert$\ $\emph{Cons}_n\ \tau$) \\
-\end{tabular}
+\end{array}
 \]
-%TODO: Como definir gramática dos ADTs?
-
-The semantics of ADTs relate to those of the plus ($\oplus$) type
--- both are additive disjunctions.  To construct a value of an ADT we
-must use one of its constructors, similar to the way $\oplus$ requires
-only proof of either the left or right type it consists
-of. Analogously, we can deconstruct a value of an ADT via pattern
-matching on its constructors, where all branches of the pattern match
-must have the same type -- akin to the left rule for the $\oplus$
-connective. In effect, the inference rules for a simple ADT are a
-generalized form of the $\oplus$ rules.  Therefore, there's one left
-rule for ADTs, and an arbitrary number of right rules, one for each
-constructor:
-
+
+The semantics of ADTs relate to those of the plus ($\oplus$) type -- both are
+additive disjunctions.  To construct a value of an ADT we must use one of its
+constructors, similar to the way $\oplus$ requires only proof of either the left
+or right type it consists of. Analogously, we can deconstruct a value of an ADT
+via pattern matching on its constructors, where all branches of the pattern
+match must have the same type -- akin to the left rule for the $\oplus$
+connective. In effect, the inference rules for a simple ADT are a generalized
+form of the $\oplus$ rules.  Therefore, there's one left rule for ADTs, and an
+arbitrary number of right rules, one for each constructor, where ADT $T$ and
+its constructors stand for any ADT defined as \mintinline{haskell}{data T = C1
+X1 | C2 X2 | ... | Cn Xn}
+%
 \begin{mathpar}
@@
 \end{mathpar}
-where the ADT $T$ and its constructors stand for any ADT with form \begin{minted}{haskell}
-    data T = C1 X1 | C2 X2 | ... | Cn Xn
-\end{minted}
 
@@
 defined as \mintinline{haskell}{data T = C1 T}, the synthesis goal
-$T \lolli C$, and its derivation:
-%
+$T \lolli C$, and part of its derivation:
+%
+{\small
 \begin{mathpar}
-    \infer*[right=($\lolli R$)]
-    {
     \infer*[right=($\Uparrow R$)]
@@
     \Uparrow}
-    }
-    {\Gamma; \Delta/\Delta'; \Omega \vdash \lambda x . \dots :
-    T \lolli C \Uparrow}
-\end{mathpar}
+\end{mathpar}}
 %
 Using our current system, we are to apply an infinite number of times
-the left ADT rule, never closing the proof. Symmetrically, the
-derivation for the goal $T$ is also infinite.
-%
+(\textsc{adtL}), never closing the proof. Symmetrically, the
+derivation for goal $T$ is also infinite.
+%
+{\small
 \begin{mathpar}
@@
     {\Gamma; \Delta/\Delta'; \Omega \vdash C_1 \dots : T \Downarrow}
-\end{mathpar}
+\end{mathpar}}
 %
@@
     & & $\vert\ n \leq n\ \vert\ n \geq n\ \vert\ n < n\ \vert\ n > n\ \vert\
-    true\ \vert\ false\ \vert\ \langle\emph{name}\rangle$ \\
+    true\ \vert\ false\ \vert\ x$ \\
     $n$ & $\ ::=\ $ & $n * n\ \vert\ n + n\ \vert\ n - n\ \vert\
-    \langle\emph{natural}\rangle\ \vert\ \langle\emph{name}\rangle$
+    \langle\emph{natural}\rangle\ \vert\ x$
     \\
@@
 approach as one scales to more sophisticated settings (e.g.~refinement
-types).
+types). \todo{but we have polymorphism and recursion}
 

### 2021-07-23 article/article.tex
Table and almost done

@@
 (intuitionistic) linear logic sequent calculus, based on a system of
-resource management~\cite{resman} and focusing.
+resource management~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
+and focusing.
 
@@
 its constructors stand for any ADT defined as \mintinline{haskell}{data T = C1
-X1 | C2 X2 | ... | Cn Xn}
+X1 | C2 X2 | ... | Cn Xn}:
 %
@@
 
-% TODO TODO: Big rewrite
 The other situation occurs when using an unrestricted polymorphic function that
@@
 even though for most tried problems this ``same function'' approach worked,
-context-heavy problems such as \emph{array} (seen in Section~\ref{sec:overview})
+context-heavy problems such as \emph{array} (seen in \S~\ref{sec:overview})
 wouldn't terminate in a reasonable amount of time.
@@
 %
-As such, we'll instead forbid the usage of \emph{all} unrestricted polymorphic
-functions to synthesize any existential type if, altogether, we've already used
-$d_e$ times a polymorphic function for \emph{any} existential sub-goal, where
-$d_e$ is a \emph{constant existential depth} that controls the maximum depth of
-``existential synthesis''. \todo{alinhar mais com a premissa
-  respectiva da regra}
+As such, we'll instead define that, given an existential\footnote{a type 
+is existential when any of its components is an existential type variable} goal $C$,
+we can only ``decide left!'' on a proposition $A$ if, altogether, 
+the amount of times we've ``decided left!'' on an polymorphic function to produce
+an existential goal is less than a \emph{constant ``existential depth''} $d_e$
+(which controls a \emph{depth} aspect of the synthesis process).
 %
@@
 unrestricted context ($\Rho_{L!}$), we modify \textsc{decideLeft!} to formalize
-the previous paragraph:
+the two previous paragraphs, where $\textsc{isExist}(C)$ is true if $C$ is an
+existential type, $\textsc{isPoly}(f)$ is true if $f$ is universally
+quantified (i.e. $f$ has form $\forall\overline{\alpha} f'$), and
+$\Rho_{L!}' = \Rho_{L!},(A, C)$ if $A$ is a function and $\Rho_{L!}' =
+\Rho_{L!}$ otherwise:
+%
 \begin{mathpar}
@@
     \textsc{isExist}(C) \Rightarrow |\{u\ |\
-    (f,u)\in\Rho_{L!},\textsc{isExist}(u),\textsc{isPoly}(f)\}| < d_e
+    (f,u)\in\Rho_{L!},\textsc{isPoly}(f),\textsc{isExist}(u)\}| < d_e
     \\
@@
     \\
-    \Rho_{L!}' =\ \textbf{if}\ \textsc{isFunction}(A)\ \textbf{then}\ \Rho_{L!},(A, C)\ \textbf{else}\ \Rho_{L!}
 \end{mathpar}
@@
 
-\mypara{Refinement Types} Refinement types are defined as a types
-with a name and with a predicate (a non-existing predicate is the same as it
-being always \emph{true}); dependent types are functions with
-refinement types in which the argument type name can be used in the predicates
-of the return type (e.g. $x\{\mathsf{Int}\} \lolli y\{\mathsf{Int}\ \vert\ y = x\}$ specifies a function that
-takes an Int named $x$ and produces an Int whose value is equal to the input $x$). We extend
-the types syntax with our refinement types syntax:
+\mypara{Refinement Types} Refinement types are types with a predicate (a non-existing predicate is the same as it
+being \emph{true}); dependent types are functions with
+refinement types in which the argument type is labeled and said label can be used in the predicates
+of the return type (e.g. $(x : \mathsf{Int}) \lolli \{y : \mathsf{Int}\ \vert\ y = x\}$ specifies a function that
+takes an Int and returns an Int of equal value). We extend
+the types syntax with our refinement types:
 \[
@@
 uninterpreted function that satisfies
-$\forall_{\tilde{1},\tilde{2},\dots,\tilde{n}}\ h_{\tilde{1}} \Rightarrow h_{\tilde{2}} \Rightarrow \dots
-\Rightarrow h_{\tilde{n}} \Rightarrow p$, where $\tilde{i}$ is the refinement type
-name and $h_{\tilde{i}}$ its predicate for every refinement type in the
-propositional contexts; and $\textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}})$ is a call to an SMT solver that
+$\forall_{a,b,\dots,n}\ h_{a} \Rightarrow h_{b} \Rightarrow \dots
+\Rightarrow h_{n} \Rightarrow p$, where $n$ is the refinement type label
+and $h_{n}$ its predicate, with $a,\dots,n$ standing for the label of every refinement type
+in the propositional contexts;
+and $\textsc{sat}(p_{a} \Rightarrow p_{b})$ is a call to an SMT solver that
 determines universal satisfiability of the implication between predicates (the
@@
     { \textsc{getModel}(p) = M }
-    {\Theta/\Theta';\Rho;\Gamma;\Delta \vdash M : \tilde{x}\{A\ \vert\
+    {\Theta/\Theta';\Rho;\Gamma;\Delta/\Delta' \vdash M : \{a : A\ \vert\
     p\}\Uparrow }
@@
     \infer*[right=(refL)]
-    { \textsc{sat}(p_{\tilde{y}} \Rightarrow p_{\tilde{x}}) }
-    {\Theta/\Theta';\Rho;\Gamma;\Delta; x{:}\tilde{y}\{A\ \vert\ p_{\tilde{y}}\}\Downarrow\ \vdash x : \tilde{x}\{A\ \vert\ p_{\tilde{x}}\} }
+    { \textsc{sat}(p_{a} \Rightarrow p_{b}) }
+    {\Theta/\Theta';\Rho;\Gamma;\Delta/\Delta'; x{:}(\{a : A\ \vert\
+    p_{a}\})\Downarrow\ \vdash x : \{b : A\ \vert\ p_{b}\} }
 \end{mathpar}
@@
 binders),
-% TODO: não ficou muito mais claro :\todo{Explica o que e isto}
 This way inference can be done without worrying about name conflicts.
@@
 infer the type of all functions defined by the user, checking the inferred
-type against a possible user given type annotation. Whenever a synth mark is
+type against a possible user given type annotation. Whenever a synthesis mark is
 found in the AST, all the inferrence context up to that point is added to the
-mark, and if not explicit, its type is inferred, essentially defining the synth
-context and synth goal.
-
-The synthesis module iterates over the synthesis marks and runs the synthesizer
+mark, and, if not explicit, its type is inferred, essentially defining the
+synthesis context and synthesis goal.
+
+The synthesis module iterates over the marks and runs the synthesizer
 with the respective context and goal for each mark. The resulting expression is checked
@@
 synthesis if they are violated: ``using'' guarantees that certain functions are
-used in the program body, and ``assert'' evaluates the given assertion within the
+used in the program body (to invalidate valid program without them), and ``assert'' evaluates the given assertion within the
 synthesized program to a validating boolean.  Additionally, the keyword
 ``depth'' controls the \emph{existential depth} (seen
-in~\ref{sec:formal_system}), and ``choose'' selects a different result as long
+in \S~\ref{sec:formal_system}), and ``choose'' selects a different result as long
 as one is available. Finally, the expression is minimized (i.e.~$\eta$-reduced),
@@
 The evaluation module evaluates the function named ``main'' within the
-complete synthesized program, returning a final value.
+complete synthesized program, returning a final value. When evaluating
+synthesis assertions, a timeout is used to interrupt possibly non-terminating
+programs.
 
 \subsection{Implementation}
+
+The implementation was carried out in Haskell. We highlight some key points. The
+project and source code are available from the repository~\cite{sili_repo}.
 
@@
 language without having to download and compile the complete toolchain. The web
-interface is available from the repository~\cite{github repo}.
+interface is available from the repository~\cite{sili_repo}.
 
@@
 programming paradigm of the output produced (e.g. generating heap manipulating
-programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).  Some other common
-patterns we follow are the use of type
-refinements~\cite{DBLP:conf/pldi/PolikarpovaKS16}, polymorphic types, and the
-support for synthesis of recursive
-functions~\cite{DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15}. These
-works share some common ground with each other and with us -- most
-famously the Curry-Howard correspondence, but also, for example, 
-focusing-based proof-search~\cite{10.1093/logcom/2.3.297}.
-% TODO: sinto que precisava de um toque...
+programs~\cite{DBLP:journals/pacmpl/PolikarpovaS19}).
+% Some other common
+% patterns we follow are the use of type
+% refinements~\cite{DBLP:conf/pldi/PolikarpovaKS16}, polymorphic types, and the
+% support for synthesis of recursive
+% functions~\cite{DBLP:conf/pldi/PolikarpovaKS16,DBLP:conf/pldi/OseraZ15}. These
+% works share some common ground with each other and with us -- most
+% famously the Curry-Howard correspondence, but also, for example, 
+% focusing-based proof-search~\cite{10.1093/logcom/2.3.297}.
 
@@
 functional programs in an ``advanced'' context. Their specifications combine two
-rich forms of types: polymorphic and refinement types (which correspond to a
-first-order logic through the Curry-Howard isomorphism). Their approach to
+rich forms of types: polymorphic and refinement types.
+% (which correspond to a
+% first-order logic through the Curry-Howard isomorphism).
+Their approach to
 refinement types consists of a new algorithm that supports decomposition of the
-refinement specification, allowing for separation between the language of
-specification and programs and making the approach amenable to compositional
-synthesis techniques. We also support refinements (and polymorphism),
+refinement specification.
+%,allowing for separation between the language of
+%specification and programs and making the approach amenable to compositional
+%synthesis techniques.
+We also support refinements (and polymorphism),
 but they are not as integrated in the synthesis process as in \cite{DBLP:conf/pldi/PolikarpovaKS16}.
@@
 work~\cite{DBLP:conf/lopstr/HughesO20} synthesizes programs using an
-approach very similar to our own.  The work employs so-called graded
-modal types, which is a refinement of pure linear types -- a more
-\emph{fine-grained} version, since it allows for quantitative
+approach similar ours. It employs so-called graded
+modal types, which are a refinement of pure linear types
+% -- a more
+% \emph{fine-grained} version 
+that allows for quantitative
 specification of resource usage, in contrast to ours either
@@
 complex, and so they provide solutions which adapt Hodas and Miller's
-resource management
-approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}
--- which, in contrast, is the model use in our work.
-
+approach~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:journals/tcs/LiangM09}.
+% -- which, in contrast, is the one used in our work.
+%
 They also use focusing as a solution to trim down search space and to
@@
 programs~\cite{DBLP:journals/mscs/CairesPT16,DBLP:conf/concur/CairesP10},
-which is a crucial avenue of future work. Moreover, their use grading
+which is a crucial avenue of future work. Moreover, their use of grading
 effectively requires constraint solving to be integrated with the
 synthesis procedure, which can limit the effectiveness of the overall
-approach as one scales to more sophisticated settings (e.g.~refinement
-types). \todo{but we have polymorphism and recursion}
+approach.
+% as one scales to more sophisticated settings (e.g.~refinement
+% types).
+Additionally, Our system extends the focusing-based system with
+recursion, ADTs, polymorphism and refinements to synthesize
+more expressive programs.
 
@@
 
-The benchmarks are separated into groups: theorems of linear logic, recursive
-ADTs, refinements. For each goal we analyze if the synthesis process terminated
-according to what was expected, the mean time and standard deviation, if the
-resulting program fullfills its intended purpose
-
-\begin{figure}[t]
+The benchmarks\footnote{Considering the full pipeline pass; 2017 MBP; 8GB RAM;
+2.3Ghz Intel Core i5; with benchmarking library Criterion} are separated into groups: theorems of linear logic, recursive
+ADTs, refinements. The first table displays successful synthesis -- for each
+goal we analyze the mean time and standard deviation, the conditions used to
+guide the output, and the components (other functions) used in the resulting
+program.
+
+% TODO Verificar nomes dos linear logic theoresm...
 {\scriptsize
@@
         \hline
-        Group & Goal & Avg. time + $\sigma$ & Found & Is expected \\
+        Group & Goal & Avg. time $\pm\,\sigma$ & Keywords & Components \\
         \hline
-        \multirow{4}{4em}{Linear Logic Theor.} & uncurry & 1s & yes & yes \\ 
-        & curry & 2s & yes & yes \\ 
-        & call by name?? & 2s & yes & yes \\ 
-        & 0/1?? & 2s & yes & yes \\ 
+        \multirow{4}{4em}{Linear Logic Theor.} & uncurry & $133\mu s\pm 4.9\mu s$ &&\\ 
+        & distributivity & $179\mu s\pm 5.0\mu s$ && \\ 
+        & call by name & $196\mu s\pm 4.6\mu s$ && \\ 
+        & 0/1 & $294\mu s\pm 5.3\mu s$ && \\ 
         \hline
-        \multirow{8}{4em}{List} & singleton & 3s & yes & yes \\ 
-        & insert & 2s & yes & unordered \\ 
-        & map & 1s & yes & yes \\ 
-        & append & 2s & yes & yes \\ 
-        & foldl & 2s & yes & got (foldl . reverse) \\ 
-        & foldr & 2s & yes & yes \\ 
-        & concat & $\infty$ & no & no \\ 
-        & uncons & 1s & yes & yes \\
+        \multirow{7}{4em}{List} & map & $288\mu s\pm 7.2\mu s$  && \\ 
+        & append & $292\mu s\pm 7.0\mu s$ & & \\ 
+        & foldl & $1.69ms\pm 5.3\mu s$ & \emph{choose 1} & \\ 
+        & foldr & $704\mu s\pm 10\mu s$ && \\ 
+        & concat & $505\mu s\pm 18\mu s$ && \\
+        & uncons & $215\mu s\pm 15\mu s$ && \\
+        & reverse & $17.4ms\pm 515\mu s$ & \emph{reverse [1,2] == [2,1]} & \\ 
         \hline
-        \multirow{3}{4em}{Maybe} & return & 1s & yes & yes \\ 
-        & $>>=$ & 1s & yes & yes \\
-        & maybe & 1s & yes & yes \\
-        & list to maybe & 12s & no & yes \\
+        \multirow{2}{4em}{Maybe} & $>>=$ & $194\mu s\pm 5.3\mu s$ && \\ 
+        & maybe & $161\mu s\pm 4.8\mu s$&& \\
         \hline
-        \multirow{5}{4em}{Binary Tree} & singleton & 1s & yes & yes \\ 
-        & insert & 1s & yes & unordered \\
-        & map & 1s & yes & yes \\
-        & append & 1s & yes & appends to rightmost leaf \\
-        & join node & 1s & yes & joins at rightmost leaf \\
+        \multirow{7}{4em}{State} & runState & $190\mu s\pm 6.8\mu s$ && \\ 
+        & $>>=$ & $979\mu s\pm 23\mu s$ && \\
+        & $>>=$ & $\infty$ & \emph{using (runState)} & \\
+        & get & $133\mu s\pm 3.8\mu s$ && \\
+        & put & $146\mu s\pm 3.4\mu s$ && \\
+        & modify & $219\mu s\pm 4.9\mu s$ && \\
+        & evalState & $156\mu s\pm 4.0\mu s$ && \\
+        % \multirow{5}{4em}{Binary Tree} & singleton & 1s & yes & yes \\ 
+        % & insert & 1s & yes & unordered \\
+        % & map & 1s & yes & yes \\
+        % & append & 1s & yes & appends to rightmost leaf \\
+        % & join node & 1s & yes & joins at rightmost leaf \\
+        % \hline
+        % \multirow{7}{4em}{Map} & singleton & 1s & yes & yes \\ 
+        % & insert & 1s & yes & unordered \\
+        % & insertWithKey & 1s & yes & ignores function \\
+        % & union & 1s & yes & ignores keys \\
+        % & mapAccum & 1s & no & ?? \\
+        % & map & 1s & yes & yes \\
+        % & mapKeys & 1s & yes & yes \\
         \hline
-        \multirow{7}{4em}{Map} & singleton & 1s & yes & yes \\ 
-        & insert & 1s & yes & unordered \\
-        & insertWithKey & 1s & yes & ignores function \\
-        & union & 1s & yes & ignores keys \\
-        & mapAccum & 1s & no & ?? \\
-        & map & 1s & yes & yes \\
-        & mapKeys & 1s & yes & yes \\
+        \multirow{1}{4em}{Misc} & either & $197\mu s\pm 5.3\mu s$ && \\ 
         \hline
-        \multirow{1}{4em}{Misc} & either & 1s & yes & yes \\ 
+        \multirow{2}{4em}{Array} & & & \emph{depth 3} & \emph{freeze, foldl} \\
+        & array & $80ms\pm 870\mu s$ & \emph{using (foldl),depth 3} & \emph{newMArray,write} \\
         \hline
@@
 }
-\caption{Table with \dots}
-\end{figure}
-
-Synthesis of
 

### 2021-07-23 STLLC/State.hs

@@ -1,0 +1,20 @@
+data State b a = State (!b -o (a * !b));
+
+synth runState :: State b a -o (!b -o (a * !b));
+
+synth bind :: State c a -o (a -o State c b) -o State c b;
+
+#synth bind :: (State c a -o (a -o State c b) -o State c b) | using (runState);
+
+synth return :: a -o State b a;
+
+synth get :: State a a;
+
+synth put :: !a -o (State a 1);
+
+synth modify :: (!a -o !a) -o State a 1;
+
+synth evalState :: State b a -o !b -o a;
+
+
+main = runState (bind (return 2) (\x -> return x)) (!0);

### 2021-07-23 STLLC/app/Evaluate.hs

@@ -243,7 +243,7 @@ evalExpr e = fromJust $ runReaderT (eval ([], []) e) []
 
 evalModule :: Program -> CoreExpr
 evalModule p@(Program adts _ _ cbs) =
-    traceShow p $ case find (\(CoreBinding n _) -> n == "main") cbs of
+    case find (\(CoreBinding n _) -> n == "main") cbs of
       Nothing -> errorWithoutStackTrace "[Eval] No main function defined."
       Just (CoreBinding _ exp) -> case runReaderT (eval ([], map (\(CoreBinding n e) -> (n, e)) cbs) exp) adts of
                                     Nothing -> error "[Eval] Evaluation failed, this shouldn't happen and might be because maybe an assertion typechecked incorrectly (haven't tried fixing this bug yet :))"

### 2021-07-23 STLLC/app/Main.hs

@@ -99,7 +99,7 @@ main = do
     (action:args) <- getArgs
     if action == "bench"
     then do
-        let benchmarks = map (\x -> "bench/" ++ x ++ ".hs") ["llt1", "llt2", "llt3", "llt4", "list1", "list2", "list3", "list4", "list5", "list6", "list7", "list8", "maybe1", "maybe2", "maybe3", "tree1", "tree2", "tree3", "tree4", "tree5", "map2", "map3", "map5", "map6", "map8", "map9"]
+        let benchmarks = map (\x -> "bench/final-" ++ x ++ ".hs") ["1-1", "1-2", "1-3", "1-4", "2-1", "2-2", "2-3", "2-4", "2-5", "2-6", "2-7", "3-1", "3-2", "4-1", "4-2", "4-4", "4-5", "4-6", "4-7", "5-1", "6-1"]
         defaultMain [bgroup "synth" (map (\n -> bench n $ nfIO $ (mainsynthMarksModule >=> \(Program _ bs _ _) -> return bs) n) benchmarks)]
     else
         let arg =

### 2021-07-23 STLLC/bench/final-1-1.hs

@@ -1,0 +1,1 @@
+synth goal :: (a -o (b -o c)) -o ((a * b) -o c)

### 2021-07-23 STLLC/bench/final-1-2.hs

@@ -1,0 +1,2 @@
+# distributivity
+synth goal :: ((a + b) * c) -o (( a * c ) + ( b * c ))

### 2021-07-23 STLLC/bench/final-1-3.hs

@@ -1,0 +1,2 @@
+# call by name
+synth goal :: !(!A -o B) -o !(!B -o C) -o (!A -o C);

### 2021-07-23 STLLC/bench/final-1-4.hs

@@ -1,0 +1,2 @@
+# 0 / 1
+synth goal :: !(!A -o !B) -o !(!B -o !C) -o !(!A -o C)

### 2021-07-23 STLLC/bench/final-2-1.hs

@@ -1,0 +1,5 @@
+#map
+
+data List a = Nil | Cons (a * List a);
+
+synth map :: (a -o b) -> List a -o List b;

### 2021-07-23 STLLC/bench/final-2-2.hs

@@ -1,0 +1,4 @@
+# append
+data List a = Nil | Cons (a * List a);
+
+synth append :: List a -o List a -o List a;

### 2021-07-23 STLLC/bench/final-2-3.hs

@@ -1,0 +1,4 @@
+# foldl
+data List a = Nil | Cons (a * List a);
+
+synth foldl :: (b -o a -o b) -> b -o List a -o b | choose 1;

### 2021-07-23 STLLC/bench/final-2-4.hs

@@ -1,0 +1,4 @@
+# foldr
+data List a = Nil | Cons (a * List a);
+
+synth foldr :: ((a -o b -o b) -> b -o List a -o b);

### 2021-07-23 STLLC/bench/final-2-5.hs

@@ -1,0 +1,4 @@
+# concat
+data List a = Nil | Cons (a * List a);
+
+synth concat :: List (List a) -o List a;

### 2021-07-23 STLLC/bench/final-2-6.hs

@@ -1,0 +1,6 @@
+# uncons
+data List a = Nil | Cons (a * List a);
+
+data Maybe a = Nothing | Just a;
+
+synth uncons :: List a -o Maybe (a * List a);

### 2021-07-23 STLLC/bench/final-2-7.hs

@@ -1,0 +1,5 @@
+data List a = Nil | Cons (a * List a);
+
+synth reverse :: List a -o List a -o List a
+  | assert
+  (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil)));

### 2021-07-23 STLLC/bench/final-3-1.hs

@@ -1,0 +1,4 @@
+# >>=
+data Maybe a = Nothing | Just a;
+
+synth bind :: Maybe a -o (a -o Maybe b) -> Maybe b;

### 2021-07-23 STLLC/bench/final-3-2.hs

@@ -1,0 +1,4 @@
+# maybe
+data Maybe a = Nothing | Just a;
+
+synth maybe :: b -> (a -o b) -> Maybe a -o b;

### 2021-07-23 STLLC/bench/final-4-1.hs

@@ -1,0 +1,4 @@
+# runState
+data State b a = State (!b -o (a * !b));
+
+synth runState :: State b a -o (!b -o (a * !b));

### 2021-07-23 STLLC/bench/final-4-2.hs

@@ -1,0 +1,4 @@
+# >>=
+data State b a = State (!b -o (a * !b));
+
+synth bind :: State c a -o (a -o State c b) -o State c b;

### 2021-07-23 STLLC/bench/final-4-3.hs

@@ -1,0 +1,6 @@
+# >>= w/ runState
+data State b a = State (!b -o (a * !b));
+
+synth runState :: State b a -o (!b -o (a * !b));
+
+synth bind :: (State c a -o (a -o State c b) -o State c b) | using (runState); # | optimize 1

### 2021-07-23 STLLC/bench/final-4-4.hs

@@ -1,0 +1,4 @@
+# get
+data State b a = State (!b -o (a * !b));
+
+synth get :: State a a;

### 2021-07-23 STLLC/bench/final-4-5.hs

@@ -1,0 +1,4 @@
+# put
+data State b a = State (!b -o (a * !b));
+
+synth put :: !a -o (State a 1);

### 2021-07-23 STLLC/bench/final-4-6.hs

@@ -1,0 +1,4 @@
+# modify
+data State b a = State (!b -o (a * !b));
+
+synth modify :: (!a -o !a) -o State a 1;

### 2021-07-23 STLLC/bench/final-4-7.hs

@@ -1,0 +1,4 @@
+# evalState
+data State b a = State (!b -o (a * !b));
+
+synth evalState :: State b a -o !b -o a;

### 2021-07-23 STLLC/bench/final-5-1.hs

@@ -1,0 +1,4 @@
+# either
+data Either a b = Left a | Right b;
+
+synth either :: ((a -o c) -> (b -o c) -> Either a b -o c);

### 2021-07-23 STLLC/bench/final-6-1.hs

@@ -1,0 +1,9 @@
+newMArray :: Int -> (MArray a -o !b) -o b;
+
+write :: MArray a -o (Int * a) -> MArray a;
+
+freeze :: MArray a -o !(Array a);
+
+foldl :: (a -o b -o a) -> a -o (List b) -o a;
+
+synth array :: Int -> List (!(Int * a)) -> Array a | using (foldl) | depth 3;

### 2021-07-23 STLLC/bench/list7.hs

@@ -1,4 +1,3 @@
 data List a = Nil | Cons (a * List a);
 
-# infinite loop
 synth concat :: (List (List a) -o List a);

### 2021-07-23 STLLC/bench/llt2.hs

@@ -1,1 +1,1 @@
-synth goal :: (( (a * b) -o c ) -o ( a -o ( b -o c ) ))
+synth goal :: (((a + b) * c) -o (( a * c ) + ( b * c)));

### 2021-07-23 article/article.tex
Add example to end of evaluation, cut some text

@@
 linear logic. Proof search relates to program synthesis via the Curry-Howard
-correspondence~\cite{}, which states that \emph{propositions in a logic} are
+correspondence~\cite{curry:34,howard:80,DBLP:journals/cacm/Wadler15}, which states that \emph{propositions in a logic} are
 \emph{types}, and \emph{proofs} of those propositions are well-typed
@@
 assumption of a given proposition and how to prove a certain proposition is
-true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
-focusing} is a technique to remove non-essential nondeterminism from
+true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{10.1093/logcom/2.3.297} 
+is a technique to remove non-essential nondeterminism from
 proof-search by structuring the application of so-called invertible and
@@
 management, and focusing, we get the following core formal system
-(inspired by~\cite{FPnotes...}) -- in which the
+(inspired by~\cite{DBLP:conf/cade/ChaudhuriP05,fpnotes}) -- in which the
 rule $\lolli$R is read: to synthesize a program of type $A \lolli B$ while inverting
@@
 bound variables of the scheme as described by the Hindley-Milner inference
-method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names
+method's~\cite{DBLP:journals/jcss/Milner78,10.2307/1995158} instantiation rule (put simply, generate fresh names
 for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
@@
 syntatic sugar and using the locally-nameless
-representation~\cite{locally nameless}, in which all bound variables
+representation~\cite{DBLP:journals/jar/Chargueraud12}, in which all bound variables
 are represented with Debruijn indices rather than names (i.e.~using
@@
 of these is slow, our implementation hashes the relevant values via a hashing
-library (\emph{Hashable}~\cite{https://hackage.haskell.org/package/hashable}) and
+library (\emph{Hashable}~\cite{hashable}) and
 uses the result as the key.
@@
 synthesis failure, as it happens more often than successful
-results. We capture failure of a LogicT computation using the $\mathit{ifte}$
-function and represent failure or success information with the \texttt{Maybe}
-type. We ran benchmarks to measure the impact of memoization on 
+results. We ran benchmarks to measure the impact of memoization on 
 performance. It's quite noticeable in synthesis with multiple
@@
 %
-The implementation could be improved by better combining the
-memoization with the LogicT monad transformer (remembering decisions across
-backtracking branches) and by finetuning key-equality by ignoring context
-components that aren't relevant to the synthesis call, to name some
-examples. We leave such optimizations to future work.
+The implementation could be greatly improved, but we
+% by better combining the
+% memoization with the LogicT monad transformer (remembering decisions across
+% backtracking branches) and by finetuning key-equality by ignoring context
+% components that aren't relevant to the synthesis call, to name some
+% examples. We 
+leave such optimizations to future work.
 
@@
 retrospective, however, when faced with a program that doesn't terminate, what
-went wrong is not so clear. While we were able to figure out a solution for the
-first couple of "infinite examples", the following problems were much harder,
-taking sometimes full days to solve. To make the debugging experience better, we
+went wrong is not so clear.
+% While we were able to figure out a solution for the
+% first couple of "infinite examples", the following problems were much harder,
+% taking sometimes full days to solve.
+To make the debugging experience better, we
 developed a tracing system that prints information whenever a rule is applied,
-alongside the stack of rules applied up until that one. This improved our
-debugging speed considerably, especially when put together with a correct
-derivation done by hand, with which we could match what was happening against
-what was supposed to happen.
+alongside the stack of rules applied so far.
+% This improved our
+% debugging speed considerably, especially when put together with a correct
+% derivation done by hand, with which we could match what was happening against
+% what was supposed to happen.
 
@@
 
-The benchmarks\footnote{Considering the full pipeline pass; 2017 MBP; 8GB RAM;
-2.3Ghz Intel Core i5; with benchmarking library Criterion} are separated into groups: theorems of linear logic, recursive
+The benchmarks\footnote{ran on an Intel Core i5 machine with 8GB of RAM using the benchmarking library
+Criterion, and considering the full pipeline pass, from parsing to synthesis} are separated into groups: theorems of linear logic, recursive
 ADTs, refinements. The first table displays successful synthesis -- for each
@@
         \multirow{2}{4em}{Array} & & & \emph{depth 3} & \emph{freeze, foldl} \\
-        & array & $80ms\pm 870\mu s$ & \emph{using (foldl),depth 3} & \emph{newMArray,write} \\
+        & array~\ref{sec:overview} & $80ms\pm 870\mu s$ & \emph{using (foldl),depth 3} & \emph{newMArray,write} \\
         \hline
@@
   \end{center}
+}
+
+To further showcase the expressiveness of our synthesizer, we present as
+examples the input and output program for \emph{reverse}.
+Note how we can guide the synthesis using the assertion mechanism to 
+supply an example:
+{\small
+\begin{minted}{haskell}
+synth reverse :: List a -o List a -o List a
+  | assert (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil))); 
+\end{minted}
+
+\begin{minted}{haskell}
+reverse b c = case b of
+    Nil -> c
+  | Cons e*f -> reverse f (Cons (e, c));
+\end{minted}
 }

### 2021-07-23 article/article.tex
break line in example
moved example

@@
 % live demo com exemplos pré-feitos em Y?
+
+To showcase the expressiveness of our synthesizer, we present as
+examples the input and output program for \emph{reverse}.
+Note how we can guide the synthesis using the assertion mechanism to 
+supply an example:
+
+{\small
+\begin{minted}{haskell}
+synth reverse :: List a -o List a -o List a
+  | assert (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil))); 
+\end{minted}
+
+\begin{minted}{haskell}
+reverse b c = case b of
+    Nil -> c
+  | Cons e*f -> reverse f (Cons (e, c));
+\end{minted}
+}
 
@@
 
-To further showcase the expressiveness of our synthesizer, we present as
-examples the input and output program for \emph{reverse}.
-Note how we can guide the synthesis using the assertion mechanism to 
-supply an example:
-{\small
-\begin{minted}{haskell}
-synth reverse :: List a -o List a -o List a
-  | assert (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil))); 
-\end{minted}
-
-\begin{minted}{haskell}
-reverse b c = case b of
-    Nil -> c
-  | Cons e*f -> reverse f (Cons (e, c));
-\end{minted}
-}
-
 \bibliographystyle{splncs04}

### 2021-07-23 article/article.tex
Formal system and tweaks

--- 2021-07-23 article/article.tex
@@
         \hline
-        \multirow{1}{4em}{Misc} & either & $197\mu s\pm 5.3\mu s$ && \\ 
+        Misc & either & $197\mu s\pm 5.3\mu s$ && \\ 
         \hline
@@
         & array~\ref{sec:overview} & $80ms\pm 870\mu s$ & \emph{using (foldl),depth 3} & \emph{newMArray,write} \\
+        \hline
+        Refinements & add3 & $39ms\pm 1.1ms$ & & + \\
         \hline
@@
 synth reverse :: List a -o List a -o List a
-  | assert (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil))); 
+  | assert 
+  (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil))); 
 \end{minted}
@@
 \label{sec:final_system}
+{\small
+\begin{mathparpagebreakable}
+    \infer*[right=($\lolli R$)]
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \vdash M : B \Uparrow \and x
+    \notin \Delta'}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash \lambda x . M : A
+    \lolli B \Uparrow}
+\and
+    \infer*[right=($\with R$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \vdash M : A \Uparrow \and \Gamma ;
+    \Delta/ \Delta'' ; \Omega \vdash N : B \Uparrow \and \Delta' = \Delta''}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash  (M \with N) : A
+    \with B \Uparrow}
+\and
+    \infer*[right=($\Uparrow$R)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
+    right asynchronous}}
+    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
+\and
+    \infer*[right=($\tensor L$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega, y{:}A, z{:}B \Uparrow\ \vdash M : C
+    \and y,z \notin \Delta'}
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \tensor B \Uparrow\ \vdash\
+    \textrm{let}\ y \tensor z = x\ \textrm{in}\ M : C}
+\and
+    \infer*[right=($1 L$)]
+    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}1 \Uparrow\ \vdash\ \textrm{let}\
+    \star =
+    x\ \textrm{in}\ M : C}
+\and
+    \mprset{flushleft}
+    \infer*[right=($\oplus L$)]
+    {
+    \Gamma ; \Delta/ \Delta' ; \Omega, y{:}A \Uparrow\ \vdash M : C \and
+    y \notin \Delta' \\
+    \Gamma ; \Delta/ \Delta'' ; \Omega, z{:}B \Uparrow\ \vdash N : C \\
+    z \notin \Delta'' \\
+    \Delta' = \Delta''
+    }
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}A \oplus B \Uparrow\ \vdash\
+    \textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y \rightarrow M\ |\
+    \textrm{inr}\ z \rightarrow N : C}
+\and
+    \infer*[right=($\bang L$)]
+    {\Gamma, y{:}A ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash M : C}
+    {\Gamma ; \Delta/\Delta' ; \Omega, x{:}\bang A \Uparrow\ \vdash\
+    \textrm{let}\ \bang y = x\ \textrm{in}\ M : C}
+\and
+    \infer*[right=($\Uparrow$L)]
+    {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
+    \textrm{not left asynchronous}}
+    {\Gamma; \Delta/\Delta'; \Omega, A \Uparrow\ \vdash C}
+\and
+    \infer*[right=(decideR)]
+    {\Gamma; \Delta/\Delta' \vdash C \Downarrow \and C\ \textrm{not atomic}}
+    {\Gamma; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
+\and
+    \infer*[right=(decideL)]
+    {\Gamma; \Delta/\Delta' ; A \Downarrow\ \vdash C}
+    {\Gamma; \Delta, A/\Delta';\cdot \Uparrow\ \vdash C}
+\and
+    \infer*[right=(decideL!)]
+    {\Gamma, A; \Delta/\Delta' ; A \Downarrow\ \vdash C}
+    {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
+\and
+  \infer*[right=($\lolli L$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
+    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
+    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
+\and
+    \infer*[right=($\with L_1$)]
+    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
+\and
+    \infer*[right=($\with L_2$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash
+      M\{(\textrm{snd}\ x)/y\} : C}
+\and
+    \infer*[right=($\tensor R$)]
+    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
+    : B \Downarrow}
+    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
+\and
+    \infer*[right=($1 R$)]
+    { }
+    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
+\and
+    \infer*[right=($\oplus R_1$)]
+    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
+\and
+    \infer*[right=($\oplus R_2$)]
+    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
+\and
+    \infer*[right=($\bang R$)]
+    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
+    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
+\and
+    \infer*[right=(init)]
+    {  }
+    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
+\and
+     \infer*[right=($\Downarrow R$)]
+    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
+    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
+\and
+    \infer*[right=($\Downarrow L$)]
+    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
+    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
+\and
+    \infer*[right=(adtR)]
+    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
+\and
+    \mprset{flushleft}
+    \infer*[right=(adtL)]
+    {
+        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
+        \and
+        y_1 \notin \Delta'_1
+        \\
+        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
+        \\
+        y_2 \notin \Delta'_2
+        \\\\
+        \\ \dots
+        \\
+        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
+        \\
+        y_n \notin \Delta'_n
+        \\
+        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
+    }
+    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
+    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
+    \rightarrow M_n : C}
+\and
+    \infer*[right=(adtR)]
+    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
+    T \notin \Rho_C}
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
+\and
+    \mprset{flushleft}
+    \infer*[right=(adtL)]
+    {
+        T \notin \Rho_D
+        \and
+        \Delta'_1 = \dots = \Delta'_n 
+        \\
+        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
+        \and
+        y_1 \notin \Delta'_1
+        \\\\
+        \\ \dots
+        \\\\
+        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
+        \and
+        y_n \notin \Delta'_n
+    }
+    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
+    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
+    \rightarrow M_n : C}
+\and
+    \infer*[right=(adt$\Uparrow$L)]
+    {
+        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
+        \and
+        T \in \Rho_D
+    }
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
+\and
+    \infer*[right=(adt-init)]
+    {  }
+    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}
+\and
+    \infer*[right=(adt$\Downarrow$L)]
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
+    T \and T \notin \Rho_D}
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
+\and
+    \infer*[right=($\forall R$)]
+    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
+    \overline{\alpha}.\ \tau
+    \sqsubseteq \tau'}
+    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
+    \tau \Uparrow}
+\and
+    \infer*[right=($\forall L$)]
+    {
+        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
+        \\
+        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
+        \\
+        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
+    }
+    {\Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
+\and
+    \infer*[right=($?L$)]
+    {\textsc{unify}(?\alpha
+    \mapsto C, \Theta)}
+    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
+    x{:}?\alpha \Downarrow\ \vdash x : C}
+\and
+    \infer*[right=($\Downarrow L?$)]
+    {\textsc{unify}(?\alpha
+    \mapsto A, \Theta)}
+    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
+    x{:}A \Downarrow\ \vdash x : ?\alpha}
+\and
+    \mprset{flushleft}
+    \infer*[right=(decideL!)]
+    {
+    (A, C) \notin \Rho_{L!}
+    \\
+    \textsc{isExist}(C) \Rightarrow |\{u\ |\
+    (f,u)\in\Rho_{L!},\textsc{isPoly}(f),\textsc{isExist}(u)\}| < d_e
+    \\
+    \Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!}');\Gamma, A; \Delta/\Delta' ; A
+    \Downarrow\ \vdash C
+    }
+    {\Theta/\Theta';(\Rho_C,\Rho_D,\Rho_{L!});\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
+\and
+    \infer*[right=(adt-init)]
+    {\textsc{unify}(T_{\overline\alpha} \mapsto T_{\overline\beta}, \Theta)}
+    {\Theta/\Theta,T_{\overline\alpha} \mapsto T_{\overline\beta},\Rho; \Gamma; \Delta/\Delta'; x{:}T_{\overline\alpha} \Downarrow\ \vdash x :
+    T_{\overline\beta}}
+\and
+    \infer*[right=(refR)]
+    { \textsc{getModel}(p) = M }
+    {\Theta/\Theta';\Rho;\Gamma;\Delta/\Delta' \vdash M : \{a : A\ \vert\
+    p\}\Uparrow }
+\and
+    \infer*[right=(refL)]
+    { \textsc{sat}(p_{a} \Rightarrow p_{b}) }
+    {\Theta/\Theta';\Rho;\Gamma;\Delta/\Delta'; x{:}(\{a : A\ \vert\
+    p_{a}\})\Downarrow\ \vdash x : \{b : A\ \vert\ p_{b}\} }
+\and
+    \infer*[right=($\bang\Downarrow$L)]
+    { \Theta/\Theta';\Rho;\Gamma;\Delta; x{:}A\Downarrow\ \vdash M : C}
+    {\Theta/\Theta';\Rho;\Gamma;\Delta; x{:}\bang A\Downarrow\ \vdash M : \bang C}
+\end{mathparpagebreakable}
+}
 
@@
 
+\paragraph{Maybe}
+input:
+\begin{minted}{haskell}
+data Maybe a = Nothing | Just a;
+data List a = Nil | Cons (a * List a);
+
+synth return :: a -o Maybe a;
+synth empty :: Maybe a;
+synth bind :: Maybe a -o (a -o Maybe b) -> Maybe b;
+synth maybe :: b -> (a -o b) -> Maybe a -o b;
+\end{minted}
+output:
+\begin{minted}{haskell}
+return :: forall a . (a -o Maybe a);
+return = Just;
+
+empty :: forall a . Maybe a;
+empty = Nothing;
+
+bind :: forall a b . (Maybe a -o (!(a -o Maybe b) -o Maybe b));
+bind c d = case c of
+    Nothing ->
+      let !e = d in Nothing
+  | Just f -> let !g = d in g f;
+
+maybe :: forall a b . (!b -o (!(a -o b) -o (Maybe a -o b)));
+maybe c d e = let !f = c in
+  let !g = d in
+    case e of
+        Nothing -> f
+      | Just h -> g h;
+\end{minted}
+
 \end{document}

### 2021-07-23 STLLC/app/Main.hs

@@ -99,7 +99,8 @@ main = do
     (action:args) <- getArgs
     if action == "bench"
     then do
-        let benchmarks = map (\x -> "bench/final-" ++ x ++ ".hs") ["1-1", "1-2", "1-3", "1-4", "2-1", "2-2", "2-3", "2-4", "2-5", "2-6", "2-7", "3-1", "3-2", "4-1", "4-2", "4-4", "4-5", "4-6", "4-7", "5-1", "6-1"]
+        -- let benchmarks = map (\x -> "bench/final-" ++ x ++ ".hs") ["1-1", "1-2", "1-3", "1-4", "2-1", "2-2", "2-3", "2-4", "2-5", "2-6", "2-7", "3-1", "3-2", "4-1", "4-2", "4-4", "4-5", "4-6", "4-7", "5-1", "6-1"]
+        let benchmarks = map (\x -> "bench/final-" ++ x ++ ".hs") ["7-1"]
         defaultMain [bgroup "synth" (map (\n -> bench n $ nfIO $ (mainsynthMarksModule >=> \(Program _ bs _ _) -> return bs) n) benchmarks)]
     else
         let arg =

### 2021-07-23 STLLC/bench/final-7-1.hs

@@ -1,0 +1,1 @@
+synth add3 :: x {Int} -o y {Int} -o z {Int} -o k {Int | k == x + y + z};

### 2021-07-23 article/article.tex

@@
 
-To further showcase the expressiveness of our synthesizer, we present as
-examples the input and output program for \emph{reverse}.
-Note how we can guide the synthesis using the assertion mechanism to 
-supply an example:
-{\small
-\begin{minted}{haskell}
-synth reverse :: List a -o List a -o List a
-  | assert 
-  (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil))); 
-\end{minted}
-
-\begin{minted}{haskell}
-reverse b c = case b of
-    Nil -> c
-  | Cons e*f -> reverse f (Cons (e, c));
-\end{minted}
-}
-
 \bibliographystyle{splncs04}

### 2021-07-23 article/article.tex
add examples

@@
 
-To showcase the expressiveness of our synthesizer, we present as
+\noindent \emph{Example.} To showcase the expressiveness of our synthesizer, we present as
 examples the input and output program for \emph{reverse}.
@@
 
-\paragraph{Maybe}
-input:
+\mypara{Maybe}\
+\\
+Input program:
 \begin{minted}{haskell}
@@
 \end{minted}
-output:
+Output program:
 \begin{minted}{haskell}
@@
 
+\mypara{List}\
+\\
+Input program:
+\begin{minted}{haskell}
+data List a = Nil | Cons (a * List a);
+data Maybe a = Nothing | Just a;
+
+synth singleton :: a -o List a;
+synth append :: List a -o List a -o List a;
+synth map :: (!(a -o b)) -o List a -o List b;
+synth foldl :: !(b -o a -o b) -o b -o List a -o b | choose 1;
+synth uncons :: List a -o Maybe (a * List a);
+synth foldr :: !(a -o b -o b) -o b -o List a -o b;
+synth insert :: a -o List a -o List a;
+synth concat :: List (List a) -o List a;
+\end{minted}
+Ouput program:
+\begin{minted}{haskell}
+singleton :: forall a . (a -o List a);
+singleton b = Cons (b, Nil);
+
+append :: forall a . (List a -o (List a -o List a));
+append b c = case b of
+    Nil -> c
+  | Cons d ->
+      let e*f = d in
+        Cons (e, append f c);
+
+map :: forall a b . (!(a -o b) -o (List a -o List b));
+map c d = let !e = c in
+  case d of
+      Nil -> Nil
+    | Cons f ->
+        let g*h = f in
+          Cons (e g, map (!e) h);
+
+foldl :: forall a b . (!(b -o (a -o b)) -o (b -o (List a -o b)));
+foldl c d e = let !f = c in
+  case e of
+      Nil -> d
+    | Cons g ->
+        let h*i = g in
+          foldl (!f) (f d h) i;
+
+uncons :: forall a . (List a -o Maybe (a * List a));
+uncons b = case b of
+    Nil -> Nothing
+  | Cons c ->
+      let d*e = c in Just (d, e);
+
+foldr :: forall a b . (!(a -o (b -o b)) -o (b -o (List a -o b)));
+foldr c d e = let !f = c in
+  case e of
+      Nil -> d
+    | Cons g ->
+        let h*i = g in
+          f h (foldr (!f) d i);
+
+insert :: forall a . (a -o (List a -o List a));
+insert b c = case c of
+    Nil -> Cons (b, Nil)
+  | Cons g ->
+      let h*i = g in
+        Cons (h, insert b i);
+
+concat :: forall a . (List (List a) -o List a);
+concat b = case b of
+    Nil -> Nil
+  | Cons c ->
+      let d*e = c in
+        case d of
+            Nil -> concat e
+          | Cons k ->
+              let l*m = k in
+                Cons (l, concat (Cons (m, e)));
+\end{minted}
+
+\mypara{State} (with a slight optimization that will be added as a control keyword
+futurely, that allows bind using runState to terminate in a reasonable time)
+\\
+Input program:
+\begin{minted}{haskell}
+data State b a = State (!b -o (a * !b));
+
+synth runState :: State b a -o (!b -o (a * !b));
+synth bind :: (State c a -o (a -o State c b) -o State c b) | using (runState);
+synth return :: a -o State b a;
+synth get :: State a a;
+synth put :: !a -o (State a 1);
+synth modify :: (!a -o !a) -o State a 1;
+synth evalState :: State b a -o !b -o a;
+\end{minted}
+Output program:
+\begin{minted}{haskell}
+data State b a = State (!b -o (a * !b));
+
+runState :: forall a b . (State b a -o (!b -o (a * !b)));
+runState c !f = case c of
+    State e ->
+        let h*i = e (!f) in
+          let !j = i in (h, (!j));
+
+bind :: forall a b c . (State c a -o ((a -o State c b) -o State c b));
+bind bs bt = case bs of
+    State bu ->
+      State (λbv -> let !bw = bv in
+                      let cu*cv = bu (!bw) in
+                        let !cw = cv in
+                          (let dr*ds = runState (bt cu) (!cw) in
+                             let !dt = ds in dr, (!cw)));
+
+return :: forall a b . (a -o State b a);
+return c = State (λd -> let !e = d in
+               (c, (!e)));
+
+get :: forall a . State a a;
+get = State (λb -> let !c = b in
+               (c, (!c)));
+
+put :: forall a . (!a -o State a 1);
+put b = let !c = b in
+  State (λd -> let !e = d in
+                 ((), (!e)));
+
+modify :: forall a . ((!a -o !a) -o State a 1);
+modify b = State (λc -> let !d = c in
+               let !f = b (!d) in ((), (!f)));
+
+evalState :: forall a b . (State b a -o (!b -o a));
+evalState c d = case c of
+    State e ->
+      let !f = d in
+        let h*i = e (!f) in
+          let !j = i in h;
+\end{minted}
+
 \end{document}

### 2021-07-23 article/article.tex
remove lambdas unicode

@@
     State bu ->
-      State (λbv -> let !bw = bv in
+      State (\bv -> let !bw = bv in
                       let cu*cv = bu (!bw) in
@@
 return :: forall a b . (a -o State b a);
-return c = State (λd -> let !e = d in
+return c = State (\d -> let !e = d in
                (c, (!e)));
@@
 get :: forall a . State a a;
-get = State (λb -> let !c = b in
+get = State (\b -> let !c = b in
                (c, (!c)));
@@
 put b = let !c = b in
-  State (λd -> let !e = d in
+  State (\d -> let !e = d in
                  ((), (!e)));
@@
 modify :: forall a . ((!a -o !a) -o State a 1);
-modify b = State (λc -> let !d = c in
+modify b = State (\c -> let !d = c in
                let !f = b (!d) in ((), (!f)));

### 2021-07-23 article/article.tex
final :)
Adjustments pos-delivery

@@
 \title{Synthesis of Linear Functional Programs}
-\author{Rodrigo Mesquita - Nº 55902 \\
-    Supervisor: Bernardo Toninho}
+\author{Rodrigo Mesquita \and Bernardo Toninho}
 \date{July 2021}
@@
 correspondence~\cite{curry:34,howard:80,DBLP:journals/cacm/Wadler15}, which states that \emph{propositions in a logic} are
-\emph{types}, and \emph{proofs} of those propositions are well-typed
+\emph{types}, and their \emph{proofs} are well-typed
 \emph{programs} -- finding a proof of a proposition is finding a program with
@@
     \and
-    \infer*[right=($\Downarrow L?$)]
+    \infer*[right=($\Downarrow ?L$)]
     {\textsc{unify}(?\alpha
@@
 
+\mypara{Example} To showcase the expressiveness of our synthesizer, we present as
+examples the input and output program for \emph{reverse}.
+Note how we can guide the synthesis using the assertion mechanism to 
+supply an example:
+
+{\small
+\begin{minted}{haskell}
+synth reverse :: List a -o List a -o List a
+  | assert (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil))); 
+\end{minted}
+
+\begin{minted}{haskell}
+reverse b c = case b of
+    Nil -> c
+  | Cons e*f -> reverse f (Cons (e, c));
+\end{minted}
+}
+
 \subsection{Implementation}
@@
 % live demo com exemplos pré-feitos em Y?
-
-\noindent \emph{Example.} To showcase the expressiveness of our synthesizer, we present as
-examples the input and output program for \emph{reverse}.
-Note how we can guide the synthesis using the assertion mechanism to 
-supply an example:
-
-{\small
-\begin{minted}{haskell}
-synth reverse :: List a -o List a -o List a
-  | assert (reverse (Cons (1, Cons (2, Nil))) Nil) == (Cons (2, Cons (1, Nil))); 
-\end{minted}
-
-\begin{minted}{haskell}
-reverse b c = case b of
-    Nil -> c
-  | Cons e*f -> reverse f (Cons (e, c));
-\end{minted}
-}
 
@@
 \and
-    \infer*[right=(decideL!)]
-    {\Gamma, A; \Delta/\Delta' ; A \Downarrow\ \vdash C}
-    {\Gamma, A; \Delta/\Delta';\cdot \Uparrow\ \vdash C}
-\and
-  \infer*[right=($\lolli L$)]
-    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
-    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
-    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
-\and
-    \infer*[right=($\with L_1$)]
-    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
-    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
-\and
-    \infer*[right=($\with L_2$)]
-    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
-    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash
-      M\{(\textrm{snd}\ x)/y\} : C}
-\and
-    \infer*[right=($\tensor R$)]
-    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
-    : B \Downarrow}
-    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
-\and
-    \infer*[right=($1 R$)]
-    { }
-    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
-\and
-    \infer*[right=($\oplus R_1$)]
-    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
-    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
-\and
-    \infer*[right=($\oplus R_2$)]
-    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
-    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
-\and
-    \infer*[right=($\bang R$)]
-    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
-    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
-\and
-    \infer*[right=(init)]
-    {  }
-    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
-\and
-     \infer*[right=($\Downarrow R$)]
-    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
-    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
-\and
-    \infer*[right=($\Downarrow L$)]
-    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
-    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
-\and
-    \infer*[right=(adtR)]
-    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
-    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
-\and
-    \mprset{flushleft}
-    \infer*[right=(adtL)]
-    {
-        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
-        \and
-        y_1 \notin \Delta'_1
-        \\
-        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
-        \\
-        y_2 \notin \Delta'_2
-        \\\\
-        \\ \dots
-        \\
-        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
-        \\
-        y_n \notin \Delta'_n
-        \\
-        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
-    }
-    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
-    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
-    \rightarrow M_n : C}
-\and
-    \infer*[right=(adtR)]
-    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
-    T \notin \Rho_C}
-    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
-\and
-    \mprset{flushleft}
-    \infer*[right=(adtL)]
-    {
-        T \notin \Rho_D
-        \and
-        \Delta'_1 = \dots = \Delta'_n 
-        \\
-        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
-        \and
-        y_1 \notin \Delta'_1
-        \\\\
-        \\ \dots
-        \\\\
-        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
-        \and
-        y_n \notin \Delta'_n
-    }
-    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
-    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
-    \rightarrow M_n : C}
-\and
-    \infer*[right=(adt$\Uparrow$L)]
-    {
-        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
-        \and
-        T \in \Rho_D
-    }
-    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
-\and
-    \infer*[right=(adt-init)]
-    {  }
-    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}
-\and
-    \infer*[right=(adt$\Downarrow$L)]
-    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
-    T \and T \notin \Rho_D}
-    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
-\and
-    \infer*[right=($\forall R$)]
-    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
-    \overline{\alpha}.\ \tau
-    \sqsubseteq \tau'}
-    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
-    \tau \Uparrow}
-\and
-    \infer*[right=($\forall L$)]
-    {
-        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
-        \\
-        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
-        \\
-        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
-    }
-    {\Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
-\and
-    \infer*[right=($?L$)]
-    {\textsc{unify}(?\alpha
-    \mapsto C, \Theta)}
-    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
-    x{:}?\alpha \Downarrow\ \vdash x : C}
-\and
-    \infer*[right=($\Downarrow L?$)]
-    {\textsc{unify}(?\alpha
-    \mapsto A, \Theta)}
-    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
-    x{:}A \Downarrow\ \vdash x : ?\alpha}
-\and
     \mprset{flushleft}
@@
 \and
+  \infer*[right=($\lolli L$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
+    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
+    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
+\and
+    \infer*[right=($\with L_1$)]
+    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
+\and
+    \infer*[right=($\with L_2$)]
+    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
+    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash
+      M\{(\textrm{snd}\ x)/y\} : C}
+\and
+    \infer*[right=($\tensor R$)]
+    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
+    : B \Downarrow}
+    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
+\and
+    \infer*[right=($1 R$)]
+    { }
+    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
+\and
+    \infer*[right=($\oplus R_1$)]
+    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
+\and
+    \infer*[right=($\oplus R_2$)]
+    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
+    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
+\and
+    \infer*[right=($\bang R$)]
+    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
+    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
+\and
+    \infer*[right=(init)]
+    {  }
+    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
+\and
+     \infer*[right=($\Downarrow R$)]
+    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
+    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
+\and
+    \infer*[right=($\Downarrow L$)]
+    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
+    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
+\and
+    \infer*[right=(adtR)]
+    {(\Rho_C'; \Rho_D) ; \Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow \and
+    T \notin \Rho_C}
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
+\and
+    \mprset{flushleft}
+    \infer*[right=(adtL)]
+    {
+        T \notin \Rho_D
+        \and
+        \Delta'_1 = \dots = \Delta'_n 
+        \\
+        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
+        \and
+        y_1 \notin \Delta'_1
+        \\\\
+        \\ \dots
+        \\\\
+        (\Rho_C; \Rho'_D);\Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
+        \and
+        y_n \notin \Delta'_n
+    }
+    {(\Rho_C; \Rho_D); \Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
+    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
+    \rightarrow M_n : C}
+\and
+    \infer*[right=(adt$\Uparrow$L)]
+    {
+        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
+        \and
+        T \in \Rho_D
+    }
+    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
+\and
+    \infer*[right=(adt$\Downarrow$L)]
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
+    T \and T \notin \Rho_D}
+    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
+\and
     \infer*[right=(adt-init)]
@@
     T_{\overline\beta}}
+\and
+    \infer*[right=($\forall R$)]
+    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
+    \overline{\alpha}.\ \tau
+    \sqsubseteq \tau'}
+    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
+    \tau \Uparrow}
+\and
+    \infer*[right=($\forall L$)]
+    {
+        \Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \tau' \Downarrow\ \vdash C
+        \\
+        \forall \overline{\alpha}.\ \tau \sqsubseteq_E \tau'
+        \\
+        \textrm{ftv}_E(\tau') \cap \{ ?\alpha\ \vert\ (?\alpha \mapsto \tau_x) \in \Theta'\} = \emptyset
+    }
+    {\Theta/\Theta'; \Rho; \Gamma; \Delta/\Delta'; \forall \overline{\alpha}.\ \tau \Downarrow\ \vdash C}
+\and
+    \infer*[right=($?L$)]
+    {\textsc{unify}(?\alpha
+    \mapsto C, \Theta)}
+    {\Theta/\Theta, ?\alpha \mapsto C ; \Rho; \Gamma; \Delta/\Delta';
+    x{:}?\alpha \Downarrow\ \vdash x : C}
+\and
+    \infer*[right=($\Downarrow ?L$)]
+    {\textsc{unify}(?\alpha
+    \mapsto A, \Theta)}
+    {\Theta/\Theta, ?\alpha \mapsto A ; \Rho; \Gamma; \Delta/\Delta';
+    x{:}A \Downarrow\ \vdash x : ?\alpha}
 \and
@@
     Nil -> Nothing
-  | Cons c ->
-      let d*e = c in Just (d, e);
+  | Cons c -> let d*e = c in Just (d, e);
 
@@
     Nil -> Cons (b, Nil)
-  | Cons g ->
-      let h*i = g in
-        Cons (h, insert b i);
+  | Cons h*i -> Cons (h, insert b i);
 
@@
     Nil -> Nil
-  | Cons c ->
-      let d*e = c in
+  | Cons d*e ->
         case d of
